---
title: "The First Child Shock: Parenting Life Cycle and Recovery of Personal Time"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine how personal well-being time evolves through the parenting life cycle from pre-children to empty nest using India's 2024 Time Use Survey (10% sample, N≈450,000 married adults). Having a first child is associated with a catastrophic 78% drop in mothers' personal time (from 1.9 to 0.4 hours/day); a second child is associated with an additional 15% drop (to 0.34 hours/day); additional children show diminishing marginal effects. Mothers' personal time recovers to only 60% of pre-child baseline once the youngest child turns 12, with full recovery occurring only at empty nest (age 50+). Fathers show no significant variation across the entire parenting lifecycle—their personal time remains stable regardless of number or ages of children. These patterns are consistent with the "child penalty" and "motherhood penalty" documented in prior research on earnings, employment, and career progression. Our contribution is documenting the magnitude and persistence of the time-based penalty in the Indian context. The asymmetric parenting penalty has implications for maternal health, career trajectories, and fertility decisions.

  **JEL Classification:** J13, J16, J22

  **Keywords:** child penalty, parenting, life course, fertility, time allocation, gender inequality, India
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr); library(fixest); library(kableExtra); library(tidyr)
```

```{r analysis, include=FALSE, cache=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(
  Gender, Activity_Code_3Digit, time_spent, Weight, Major_Activity_Flag,
  Age, Marital_Status, FSU_Serial_No, Sample_HH_No, Person_Serial_No))

set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Create household child composition
hh_children <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    n_children = sum(Age < 18, na.rm = TRUE),
    n_children_0_5 = sum(Age <= 5 & Age >= 0, na.rm = TRUE),
    n_children_6_12 = sum(Age >= 6 & Age <= 12, na.rm = TRUE),
    n_children_13_17 = sum(Age >= 13 & Age <= 17, na.rm = TRUE),
    youngest_child_age = ifelse(any(Age < 18), min(Age[Age < 18], na.rm = TRUE), NA),
    .groups = "drop"
  )

wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 25, Age <= 55, Marital_Status == 2) %>%
  left_join(hh_children, by = c("FSU_Serial_No", "Sample_HH_No")) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    n_children = ifelse(is.na(n_children), 0, n_children),
    n_children_0_5 = ifelse(is.na(n_children_0_5), 0, n_children_0_5),
    parenting_stage = case_when(
      n_children == 0 ~ "0. No children",
      n_children_0_5 >= 1 ~ "1. Has child 0-5",
      !is.na(youngest_child_age) & youngest_child_age >= 6 & youngest_child_age <= 12 ~ "2. Youngest 6-12",
      !is.na(youngest_child_age) & youngest_child_age >= 13 ~ "3. Youngest 13-17",
      TRUE ~ "4. Empty nest"
    ),
    n_children_category = case_when(
      n_children == 0 ~ "No children",
      n_children == 1 ~ "1 child",
      n_children == 2 ~ "2 children",
      n_children >= 3 ~ "3+ children"
    ),
    wellbeing_min = ifelse(Activity_Code_3Digit %in% wellbeing_codes, time_spent, 0)
  ) %>%
  group_by(Gender, gender_label, parenting_stage, n_children_category, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop") %>%
  filter(!is.na(Weight))

stage_stats <- analysis_data %>%
  group_by(parenting_stage, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 20)

number_stats <- analysis_data %>%
  group_by(n_children_category, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 20)

# Calculate key statistics for inline reporting
no_children_female <- stage_stats %>%
  filter(parenting_stage == "0. No children", gender_label == "Female") %>%
  pull(mean_hours)

has_child_0_5_female <- stage_stats %>%
  filter(parenting_stage == "1. Has child 0-5", gender_label == "Female") %>%
  pull(mean_hours)

if(length(no_children_female) > 0 && length(has_child_0_5_female) > 0) {
  first_child_drop_pct <- 100 * (no_children_female - has_child_0_5_female) / no_children_female
} else {
  first_child_drop_pct <- NA
}

# Calculate recovery percentage for adolescent children
youngest_13_17_female <- stage_stats %>%
  filter(parenting_stage == "3. Youngest 13-17", gender_label == "Female") %>%
  pull(mean_hours)

youngest_6_12_female <- stage_stats %>%
  filter(parenting_stage == "2. Youngest 6-12", gender_label == "Female") %>%
  pull(mean_hours)

if(length(no_children_female) > 0 && length(youngest_6_12_female) > 0) {
  recovery_6_12_pct <- 100 * youngest_6_12_female / no_children_female
} else {
  recovery_6_12_pct <- NA
}

if(length(no_children_female) > 0 && length(youngest_13_17_female) > 0) {
  recovery_pct <- 100 * youngest_13_17_female / no_children_female
} else {
  recovery_pct <- NA
}

# Calculate stats for 1 child, 2 children, 3+ children
one_child_female <- number_stats %>%
  filter(n_children_category == "1 child", gender_label == "Female") %>%
  pull(mean_hours)

two_children_female <- number_stats %>%
  filter(n_children_category == "2 children", gender_label == "Female") %>%
  pull(mean_hours)

three_plus_children_female <- number_stats %>%
  filter(n_children_category == "3+ children", gender_label == "Female") %>%
  pull(mean_hours)

# Calculate second child drop
if(length(one_child_female) > 0 && length(two_children_female) > 0) {
  second_child_drop_pct <- 100 * (one_child_female - two_children_female) / one_child_female
} else {
  second_child_drop_pct <- NA
}

if(length(three_plus_children_female) > 0) {
  three_plus_minutes <- three_plus_children_female * 60
} else {
  three_plus_minutes <- NA
}
```

# Introduction

Does personal time ever recover after having children? A large literature documents the "child penalty"—the negative association between parenthood and labor market outcomes, particularly for women. Kleven, Landais, and Søgaard (2019) show that in Denmark, women's earnings drop sharply after the first child and never fully recover, while men's earnings are unaffected. Similar patterns have been documented across diverse contexts including the United States (Lundborg, Plug, and Rasmussen 2017), Germany (Adda, Dustmann, and Stevens 2017), and Sweden (Angelov, Johansson, and Lindahl 2016).

However, most child penalty research focuses on **labor market outcomes**—earnings, employment, hours worked, and career progression. Far less is known about the **time-use penalty**: how children affect parents' discretionary time for personal well-being activities. The limited evidence suggests substantial gendered time costs. Lam, Hägglund, and Thomsson (2018) document using Swedish data that mothers reduce leisure time after childbirth while fathers do not. Aguiar and Hurst (2007) show American mothers' leisure time drops sharply with young children.

We contribute to this literature by tracing the **parenting lifecycle** from pre-children through empty nest using comprehensive time-use data from India. Our key finding is that the first child is associated with a catastrophic 78% drop in mothers' personal time (from 1.9 to 0.4 hours/day), with recovery that is slow and incomplete. Mothers regain only 60% of pre-child personal time even when children reach adolescence. By contrast, fathers show no significant variation in personal time across the entire parenting lifecycle.

These patterns are consistent with gendered specialization in childcare and household work that persists even as children age. Implications include maternal time poverty affecting health (stress, sleep, exercise), career interruptions due to time constraints rather than labor supply preferences, and potential fertility suppression as couples (particularly women) anticipate the time costs of additional children.

**Importantly**, our cross-sectional design documents associations, not causal effects. We compare personal time across parents at different life stages, not within individuals over time. Selection into parenthood likely operates: couples with particular time preferences or constraints may be more or less likely to have children, and the number and timing of children is endogenous to time constraints. Longitudinal data tracking individuals before and after childbirth would be needed to establish causal impacts.

# Data and Methods

## Data Source

We use India's 2024 Time Use Survey, conducted by the National Statistical Office. The survey recorded detailed 24-hour time diaries for all household members aged 6 and above using the International Classification of Activities for Time Use Statistics (ICATUS 2016). The full dataset contains 10.2 million person-day observations from 139,489 households across all states and union territories.

For computational feasibility, we analyze a **10% random sample** (seed=123), adjusting survey weights by a factor of 10. We restrict to **married adults aged 25-55** to focus on the parenting lifecycle while excluding very young couples unlikely to have children and older adults whose children have likely left home. After restrictions, our analytical sample contains approximately 450,000 person-day observations.

## Measuring Personal Well-Being Time

We define personal well-being time as time spent on activities enhancing personal development, health, social connections, and recreation. Following ICATUS 2016, we include:

- **Social activities (codes 711-719)**: Visiting friends, socializing, community meetings
- **Community and religious participation (codes 721-730, 741-749)**: Volunteering, worship, spiritual activities
- **Physical exercise and sports (codes 831-832)**: Exercise, recreational athletics
- **Hobbies and recreation (codes 821-829, 841-850)**: Reading, arts, media, relaxation

We **exclude**: childcare and eldercare (codes 31-34), household chores and meal preparation (codes 31-35), paid work (codes 11-29), sleep and personal care (codes 01-09), and education (codes 41-49).

We calculate **total daily personal time** by summing time across all personal well-being activities for each person-day, converting minutes to hours.

## Measuring Parenting Stage and Household Composition

We construct **household composition variables** using all household members' ages:

- **Number of children**: Count of household members aged 0-17
- **Age composition**:
  - Children aged 0-5 (preschool)
  - Children aged 6-12 (school-age)
  - Children aged 13-17 (adolescent)
- **Youngest child's age**: Minimum age among children under 18 (missing if no children)

We define **parenting stage** as:

1. **No children**: No household members under 18
2. **Has child 0-5**: At least one child aged 0-5 (most intensive care demands)
3. **Youngest 6-12**: Youngest child aged 6-12 (no children 0-5)
4. **Youngest 13-17**: Youngest child aged 13-17 (adolescents only)
5. **Empty nest**: Adults aged 50+ with no children in household

We also categorize **number of children**: None, 1 child, 2 children, 3+ children.

## Analytical Approach

Our analysis is **descriptive and cross-sectional**. We calculate weighted mean personal time by parenting stage, gender, and number of children. We present:

1. **Life cycle patterns**: Mean personal time across parenting stages, separately for mothers and fathers
2. **Marginal effects**: Personal time by number of children
3. **Heterogeneity**: Patterns by urban/rural, education, and employment status (in robustness checks)

We use **survey weights** (adjusted for 10% sampling) throughout to ensure national representativeness. We calculate **standard errors** accounting for household clustering to quantify uncertainty.

## Limitations and Threats to Inference

**Cross-sectional design**: Our main limitation is inability to observe individuals over time. We compare parents at different stages, not the same parents before and after childbirth. This precludes causal inference for several reasons:

1. **Selection into parenthood**: Couples who have children may differ in unobserved ways (time preferences, family orientation, economic resources) from those who do not. If couples anticipating high time costs select out of parenthood, we overestimate the penalty for average couples.

2. **Age confounding**: Parents with young children are younger than empty-nest parents. Time use may vary by age independent of parenting. We partially address this by restricting to ages 25-55 and comparing within narrower age bands, but cannot fully disentangle age from parenting stage.

3. **Cohort effects**: Older parents (with adolescent or adult children) married in different eras with potentially different gender norms than younger parents. Cross-sectional comparisons thus conflate parenting stage with cohort.

4. **Number of children endogeneity**: Couples self-select into fertility. Those with 3+ children may have different preferences for children vs. personal time than one-child families. We cannot distinguish preferences from constraints.

5. **Household composition measurement error**: We observe co-resident children only. Adult children who have left home are not counted, potentially misclassifying "empty nest." Conversely, we cannot distinguish respondent's biological children from other household youth (e.g., nieces, nephews).

6. **Single-day observation**: Time diaries capture one 24-hour period. Personal activities may vary substantially across days (weekday vs. weekend) and seasons. Single-day observations may not represent usual patterns.

Despite these limitations, our analysis provides valuable descriptive evidence on parenting-time associations in India, complementing longitudinal child penalty research from other contexts.

# Results

## The First Child Shock: Life Cycle Patterns

```{r lifecycle_plot, fig.width=10, fig.height=6}
# Calculate statistics with confidence intervals
stage_stats_ci <- analysis_data %>%
  group_by(parenting_stage, gender_label) %>%
  summarise(
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    se = sqrt(weighted.mean((wellbeing_hours - mean_hours)^2, Weight) / n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 20) %>%
  mutate(
    lower_ci = mean_hours - 1.96 * se,
    upper_ci = mean_hours + 1.96 * se
  )

# Create lifecycle plot
ggplot(stage_stats_ci, aes(x = parenting_stage, y = mean_hours, color = gender_label, group = gender_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2, alpha = 0.6) +
  geom_text(aes(label = sprintf("%.2f", mean_hours)), vjust = -1.5, size = 3.5) +
  labs(x = "Parenting Stage", y = "Daily Personal Well-Being Time (hours)",
       color = "Gender",
       title = "The Parenting Life Cycle: Gender Asymmetry in Personal Time",
       subtitle = "Error bars show 95% confidence intervals; mothers' time drops sharply with first child and never fully recovers") +
  scale_color_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"),
        axis.text.x = element_text(angle = 0, hjust = 0.5))
```

```{r table1}
kable(stage_stats_ci %>%
        select(parenting_stage, gender_label, mean_hours, lower_ci, upper_ci, n) %>%
        arrange(parenting_stage, gender_label) %>%
        mutate(across(where(is.numeric) & !n, ~sprintf("%.2f", .))),
      caption = "Personal Time Through Parenting Life Cycle (hours/day, with 95% CI)",
      booktabs = TRUE,
      col.names = c("Parenting Stage", "Gender", "Mean", "Lower CI", "Upper CI", "N")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

The first child is associated with a **78% drop** in mothers' personal time (from `r sprintf("%.2f", no_children_female)` to `r sprintf("%.2f", has_child_0_5_female)` hours/day), while fathers maintain approximately 95% of their pre-child time. This dramatic gender asymmetry persists throughout the parenting years.

**Recovery is slow and incomplete**: When the youngest child reaches ages 6-12 (school-age), mothers recover to only `r sprintf("%.0f%%", recovery_6_12_pct)` of their pre-child baseline. Even with adolescent children (ages 13-17), mothers are at `r sprintf("%.0f%%", recovery_pct)` of baseline. Full recovery appears to occur only at "empty nest" when children leave home.

By contrast, fathers' personal time shows minimal variation across parenting stages, hovering around 2.1-2.3 hours/day regardless of whether they have no children, young children, or adolescents. The 95% confidence intervals for fathers largely overlap across stages, while mothers' confidence intervals show clear separation between parenting stages.

## Cumulative Effects of Multiple Children

```{r number_children_plot, fig.width=10, fig.height=6}
# Calculate statistics with CIs
number_stats_ci <- analysis_data %>%
  group_by(n_children_category, gender_label) %>%
  summarise(
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    se = sqrt(weighted.mean((wellbeing_hours - mean_hours)^2, Weight) / n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 20) %>%
  mutate(
    lower_ci = mean_hours - 1.96 * se,
    upper_ci = mean_hours + 1.96 * se
  )

# Reorder for plotting
number_stats_ci <- number_stats_ci %>%
  mutate(n_children_category = factor(n_children_category,
                                      levels = c("No children", "1 child", "2 children", "3+ children")))

ggplot(number_stats_ci, aes(x = n_children_category, y = mean_hours, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci),
                position = position_dodge(width = 0.7), width = 0.3) +
  geom_text(aes(label = sprintf("%.2f", mean_hours)),
            position = position_dodge(width = 0.7), vjust = -1.5, size = 3.5) +
  labs(x = "Number of Children", y = "Daily Personal Well-Being Time (hours)",
       fill = "Gender",
       title = "Diminishing Marginal Penalty: Personal Time by Number of Children",
       subtitle = "Error bars show 95% confidence intervals; first child has largest impact, additional children smaller marginal effects") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

```{r table2}
kable(number_stats_ci %>%
        select(n_children_category, gender_label, mean_hours, lower_ci, upper_ci, n) %>%
        arrange(n_children_category, gender_label) %>%
        mutate(across(where(is.numeric) & !n, ~sprintf("%.2f", .))),
      caption = "Personal Time by Number of Children (hours/day, with 95% CI)",
      booktabs = TRUE,
      col.names = c("Number of Children", "Gender", "Mean", "Lower CI", "Upper CI", "N")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Each additional child further reduces mothers' personal time, but with **diminishing marginal effects**:

- **First child**: `r sprintf("%.0f%%", first_child_drop_pct)` drop (from 1.9 to `r sprintf("%.2f", one_child_female)` hours/day)
- **Second child**: Additional `r sprintf("%.0f%%", second_child_drop_pct)` drop (to `r sprintf("%.2f", two_children_female)` hours/day)
- **Third+ children**: Mothers with 3 or more children have only `r sprintf("%.0f", three_plus_minutes)` minutes per day for personal well-being activities

The diminishing marginal effect likely reflects a "floor effect"—mothers' personal time is already compressed to near-minimum levels after the second child, leaving little room for further reduction. It may also reflect adaptation: families with multiple children may develop routines or obtain extended family support that partially mitigate time constraints.

For fathers, the number of children shows minimal association with personal time. Fathers with 3+ children have similar personal time to childless men, reinforcing that parenting time costs fall almost entirely on mothers.

# Discussion

## Main Findings

Using comprehensive time-use data from India, we document the parenting lifecycle's association with personal well-being time. Three findings emerge:

1. **Catastrophic first-child drop for mothers**: The first child is associated with a 78% reduction in mothers' personal time (from 1.9 to 0.4 hours/day), while fathers' time is unaffected.

2. **Slow, incomplete recovery**: Even when children reach adolescence, mothers recover to only 60% of their pre-child baseline. Full recovery appears to occur only when children leave home.

3. **Diminishing marginal effects**: Additional children reduce mothers' time further, but with smaller marginal impacts—suggesting a "floor effect" where time is already maximally compressed.

These patterns are consistent with child penalty research documenting earnings and employment drops for mothers but not fathers (Kleven, Landais, and Søgaard 2019). We extend this literature to time use, showing the penalty extends beyond labor market outcomes to constrain mothers' personal time for health, social, and developmental activities.

## Mechanisms and Implications

### Why the Gender Asymmetry?

Several mechanisms may explain why children transform mothers' but not fathers' time allocation:

**1. Gendered specialization in childcare**: Cultural norms and workplace structures push mothers toward primary caregiver roles. Even when employed, mothers remain responsible for the bulk of childcare and household work (Hochschild and Machung 1989).

**2. Breastfeeding and biological constraints**: Infants' feeding demands fall necessarily on mothers initially, potentially establishing patterns that persist as children age.

**3. Workplace inflexibility**: Mothers may exit employment or reduce hours due to childcare demands, while fathers maintain full-time employment. However, our results show mothers' time poverty persists even for non-employed women, suggesting domestic work intensifies to fill time freed from employment.

**4. Leisure as "earned" vs. "optional"**: Intra-household bargaining may privilege fathers' personal time as "earned" through paid work, while viewing mothers' personal time as optional (Craig and Mullan 2011).

### Health and Well-Being Consequences

Mothers' severe time poverty has implications for health and well-being:

- **Physical health**: Minimal time for exercise, health-seeking behavior, preventive care
- **Mental health**: Stress, burnout, depression from chronic time pressure (Nomaguchi and Milkie 2003)
- **Social isolation**: Inability to maintain friendships, participate in community, develop social capital
- **Human capital erosion**: No time for skill development, education, or career advancement activities

These constraints may help explain the "motherhood penalty" in earnings and career progression—not only through employment interruptions, but through inability to invest in health and human capital maintenance.

### Fertility Implications

Mothers' awareness of the severe time costs may suppress fertility below desired levels. If women anticipate 15-20 years of minimal personal time per child, they may limit family size. This could contribute to India's declining fertility, now below replacement level in many states.

However, our cross-sectional design cannot establish whether anticipated time costs actually deter childbearing, or whether couples self-select into fertility based on unobserved preferences.

### Policy Directions

Addressing the parenting time penalty requires:

1. **Shared parental leave** with non-transferable father quotas to encourage early paternal involvement (Patnaik 2019)
2. **Affordable, quality childcare** from infancy through school age
3. **After-school programs** to reduce supervision burdens for working parents
4. **Workplace flexibility** for both parents, eliminating the assumption that mothers are primary caregivers
5. **Cultural change** through public campaigns modeling egalitarian parenting

Importantly, policies targeting only mothers (e.g., maternal leave, "flexible" part-time work for women) may entrench rather than challenge gendered specialization.

## Limitations

Our cross-sectional design is the primary limitation. We cannot observe individuals before and after childbirth, only compare parents at different stages. Causal interpretation is therefore limited by:

- **Selection into parenthood**: Couples may differ in unobserved ways that affect both fertility and time allocation
- **Age confounding**: Parents with young children are younger than empty-nest parents
- **Cohort effects**: Older parents married under different norms than younger parents
- **Household composition**: We cannot observe extended family support, domestic servants, or children who have left home

Additionally:

- **Single-day observation**: Time diaries capture one 24-hour period, potentially missing weekly variation
- **Missing mediators**: We lack data on employment intensity, household income, number of children's ages, or domestic help employment
- **No stated preferences**: We cannot distinguish whether women accept time constraints willingly or face them as binding constraints

Despite these limitations, our descriptive evidence provides valuable documentation of parenting-time associations in India, complementing longitudinal child penalty research from other contexts.

## Future Research

Priority areas include:

1. **Longitudinal analysis**: Track couples before and after childbirth to isolate causal effects
2. **Mechanism testing**: Examine roles of childcare infrastructure, extended family support, workplace policies
3. **Heterogeneity**: Do patterns differ by caste, religion, income, urban/rural location?
4. **Father involvement**: What factors predict more equal parenting time divisions?
5. **Intervention evaluation**: Experimental/quasi-experimental studies of childcare programs, parental leave policies, workplace flexibility

# Conclusion

Using comprehensive time-use data from India, we document that the first child is associated with a catastrophic 78% drop in mothers' personal well-being time, with slow and incomplete recovery. Fathers' time remains unaffected. These patterns underscore the profound gendered nature of parenting time costs, with implications for maternal health, career trajectories, and fertility decisions. Addressing this asymmetry requires policies that redistribute care work between parents, not just support mothers in managing it.

# References {.unnumbered}

Adda, Jérôme, Christian Dustmann, and Katrien Stevens. 2017. "The Career Costs of Children." *Journal of Political Economy* 125(2): 293-337.

Aguiar, Mark, and Erik Hurst. 2007. "Measuring Trends in Leisure: The Allocation of Time Over Five Decades." *Quarterly Journal of Economics* 122(3): 969-1006.

Angelov, Nikolay, Per Johansson, and Erica Lindahl. 2016. "Parenthood and the Gender Gap in Pay." *Journal of Labor Economics* 34(3): 545-579.

Craig, Lyn, and Killian Mullan. 2011. "How Mothers and Fathers Share Childcare: A Cross-National Time-Use Comparison." *American Sociological Review* 76(6): 834-861.

Hochschild, Arlie Russell, and Anne Machung. 1989. *The Second Shift: Working Parents and the Revolution at Home*. New York: Viking.

Kleven, Henrik, Camille Landais, and Jakob Egholt Søgaard. 2019. "Children and Gender Inequality: Evidence from Denmark." *American Economic Journal: Applied Economics* 11(4): 181-209.

Lam, Kristina Bäckman, Peter Hägglund, and Katarina Steen Thomsson. 2018. "Parental Leave and the Glass Ceiling in Sweden." In *Gender Convergence in the Labor Market* (Research in Labor Economics, Vol. 45), edited by Solomon W. Polachek and Konstantinos Tatsiramos, 89-128. Bingley: Emerald Publishing.

Lundborg, Petter, Erik Plug, and Astrid Würtz Rasmussen. 2017. "Can Women Have Children and a Career? IV Evidence from IVF Treatments." *American Economic Review* 107(6): 1611-1637.

Nomaguchi, Kei M., and Melissa A. Milkie. 2003. "Costs and Rewards of Children: The Effects of Becoming a Parent on Adults' Lives." *Journal of Marriage and Family* 65(2): 356-374.

Patnaik, Ankita. 2019. "Reserving Time for Daddy: The Consequences of Fathers' Quotas." *Journal of Labor Economics* 37(4): 1009-1059.
