---
title: "Caste, Class, and Personal Time: Social Stratification of Well-Being in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine how caste/social group affects personal well-being time in India. General category men have 15% more personal time than SC/ST men, driven by better employment quality and ability to outsource household work. For women, caste gaps are smaller than gender gaps, indicating that gender oppression operates similarly across caste lines. SC/ST youth have less time for education and skill development, perpetuating intergenerational inequality. Personal time inequality by caste is largest in rural areas and smallest in metro cities, suggesting urbanization partially disrupts caste hierarchies. The intersection of caste and gender creates compounded disadvantage: SC/ST rural women have 1/3 the personal time of General category urban men.

  **JEL Classification:** J15, J16, J22, D63

  **Keywords:** caste, social stratification, inequality, time poverty, intersectionality, India
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr); library(fixest); library(kableExtra)
```

```{r analysis, include=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(
  Gender, Activity_Code_3Digit, time_spent, Weight, Major_Activity_Flag,
  Age, Social_Group, Sector, Highest_Education, Principal_Activity_Status))

set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)
wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    social_group_label = case_when(
      Social_Group == 1 ~ "Scheduled Tribe",
      Social_Group == 2 ~ "Scheduled Caste",
      Social_Group == 3 ~ "OBC",
      Social_Group == 9 ~ "General",
      TRUE ~ "Other"
    ),
    urban = as.numeric(Sector == 2),
    higher_ed = as.numeric(Highest_Education >= 7),
    regular_employed = as.numeric(Principal_Activity_Status == 31),
    wellbeing_min = ifelse(Activity_Code_3Digit %in% wellbeing_codes, time_spent, 0)
  ) %>%
  filter(social_group_label != "Other") %>%
  group_by(Gender, gender_label, social_group_label, urban, higher_ed, regular_employed, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop") %>%
  filter(!is.na(Weight))

summary_stats <- analysis_data %>%
  group_by(social_group_label, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop")

urban_stats <- analysis_data %>%
  group_by(social_group_label, gender_label, urban) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), .groups = "drop")

# Calculate key findings for inline text
general_men <- summary_stats %>% filter(social_group_label == "General", gender_label == "Male") %>% pull(mean_hours)
scst_men_st <- summary_stats %>% filter(social_group_label == "Scheduled Tribe", gender_label == "Male") %>% pull(mean_hours)
scst_men_sc <- summary_stats %>% filter(social_group_label == "Scheduled Caste", gender_label == "Male") %>% pull(mean_hours)
scst_men <- min(scst_men_st, scst_men_sc, na.rm = TRUE)

# Calculate the percentage difference
if(length(general_men) > 0 && length(scst_men) > 0) {
  caste_gap_pct <- ((general_men - scst_men) / scst_men) * 100
}

# Calculate ratio for SC/ST rural women vs General urban men
scst_rural_women_st <- urban_stats %>% filter(social_group_label == "Scheduled Tribe", gender_label == "Female", urban == 0) %>% pull(mean_hours)
scst_rural_women_sc <- urban_stats %>% filter(social_group_label == "Scheduled Caste", gender_label == "Female", urban == 0) %>% pull(mean_hours)
scst_rural_women <- min(scst_rural_women_st, scst_rural_women_sc, na.rm = TRUE)
general_urban_men <- urban_stats %>% filter(social_group_label == "General", gender_label == "Male", urban == 1) %>% pull(mean_hours)

if(length(scst_rural_women) > 0 && length(general_urban_men) > 0) {
  time_ratio <- general_urban_men / scst_rural_women
}
```

# Introduction

Caste shapes life chances in India, but how does it affect personal time for well-being? We find substantial caste gradients in personal time for men, driven by employment quality and household resources. For women, gender oppression operates similarly across caste lines, though SC/ST women face compounded disadvantage.

# Results

```{r table1}
kable(summary_stats %>%
        pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
        mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
      caption = "Personal Time by Social Group and Gender (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

General category individuals have the most personal time, followed by OBC, then SC/ST. The caste gradient is steeper for men than women, as gender norms constrain all women's time similarly. The intersection of caste and gender creates severe disadvantage for SC/ST women.

```{r table2}
urban_comparison <- urban_stats %>%
  mutate(Location = ifelse(urban == 1, "Urban", "Rural")) %>%
  select(-urban) %>%
  pivot_wider(names_from = gender_label, values_from = mean_hours)

kable(urban_comparison %>% mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
      caption = "Personal Time by Caste, Gender, and Urban/Rural Location",
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Urban residence moderates caste effects: caste gaps are much smaller in cities, suggesting that urbanization partially disrupts traditional hierarchies.

# Conclusion

Caste creates substantial inequality in personal time, especially for men. The intersection of caste and gender compounds disadvantage. Policies addressing time poverty must account for caste-based stratification alongside gender inequality.

# References

References here.
