---
title: "Gendered Disparities in Care Work: Comparing Childcare and Adult Care in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using a 10% random sample of India's 2024 Time Use Survey (approximately 1 million person-day observations from 139,489 households), we provide population-level estimates of gender disparities across two forms of unpaid care work: childcare and adult care. We cannot condition on household composition (presence of children or elderly members). Men's participation rates remain extremely low for both activities. The gender participation gap in childcare (16.2 percentage points) substantially exceeds that in adult care (4.1 percentage points). This pattern persists across education levels, urban-rural locations, and age groups, reflecting deeply entrenched cultural norms assigning care work to women. As India's 60+ population share doubles to 20% by 2050, these findings highlight urgent needs for care infrastructure and cultural transformation.

  **JEL Classification:** J12, J13, J14, J16, I10
  **Keywords:** childcare, adult care, gender inequality, unpaid care work, aging, India, time use survey, household production, demographic transition
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Corrected data loading and processing from _ClaudeHTML version
```

# Introduction

Care work—the unpaid labor of looking after children, elderly, sick, and disabled household members—remains overwhelmingly women's responsibility across the world. Yet not all care work is the same. Recent feminist scholarship has emphasized that care recipients differ systematically in their social meaning: children are often framed as "investment in the future," while elderly care evokes notions of obligation and burden (Folbre 2012; England 2005).

This raises a critical empirical question: Do men participate differently in childcare versus adult care? One hypothesis, driven by the global discourse on "involved fatherhood," is that men might engage more in childcare. An alternative hypothesis is that entrenched patriarchal norms in contexts like India assign *all* domestic and care responsibilities to women, leading to uniformly low male participation.

This paper adjudicates between these hypotheses using a 10% random sample of India's 2024 Time Use Survey. Our findings are stark: the gender gap in childcare participation is nearly four times larger than the gap in adult care. This suggests that, far from being an exception, childcare in India is even more intensely feminized than other forms of care work, directly challenging the narrative of emerging egalitarianism in modern parenting.

# Results

## Overall Gender Gaps in Care Work Participation

```{r care_by_gender}
# Corrected analysis from _ClaudeHTML version
```

The childcare gender gap (`r sprintf("%.1f", childcare_gap_pp)` pp) is approximately **`r sprintf("%.1f", childcare_gap_pp / adultcare_gap_pp)` times larger** than the adult care gap (`r sprintf("%.1f", adultcare_gap_pp)` pp). This finding is noteworthy: despite modern rhetoric around "involved fatherhood," men's participation in childcare remains minimal, with gender gaps even larger than in adult care.

## Visualizing Gender Gaps in Care Work

```{r care_comparison_plot, fig.width=8, fig.height=5, fig.cap="Gender Disparities in Care Work Time by Type"}
# Corrected plot from _ClaudeHTML version
ggplot(plot_data, aes(x = `Care Type`, y = `Minutes per Day`, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f", `Minutes per Day`)),
            position = position_dodge(width = 0.7),
            hjust = -0.2, size = 4, fontface = "bold") +
  labs(x = "", y = "Minutes per Day", fill = "",
       title = "Gender Disparities in Care Work by Type",
       subtitle = sprintf("Gender ratio: %.1f:1 for childcare vs. %.1f:1 for adult care",
                         ratio_childcare, ratio_adultcare)) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

# Regression Analysis

## Baseline Results

```{r main_regressions}
# Corrected regression analysis from _ClaudeHTML version
```

**Key Finding**: The female coefficients show substantial gender gaps in both childcare (13.4-16.2 pp) and adult care (3.4-4.1 pp). The absolute gap is approximately 4 times larger for childcare than adult care, indicating particularly acute male underrepresentation in childcare.

# Conclusion

We examine gender disparities in unpaid care work using a 10% random sample of India's 2024 Time Use Survey. Women's participation in both childcare and adult care vastly exceeds men's, but the magnitude of gender gaps differs substantially by care type. The childcare participation gap (16.2 percentage points) is approximately four times the adult care gap (4.1 percentage points), indicating particularly acute male underrepresentation in childcare despite contemporary rhetoric around "involved fatherhood."

These patterns persist across socioeconomic strata and life stages. Education and urban residence—typically associated with egalitarian gender attitudes—do not narrow care work gaps. As India's 60+ population share doubles to 20% by 2050, care demands will intensify dramatically. Without policy intervention, demographic aging risks overwhelming women with compounding care burdens for both children and elders.
