---
title: "Men Do 22% of Childcare but Only 8% of Elder Care: The Selective Help Gap in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10.2 million observations, 139,489 households), we document stark gender asymmetries in care work that vary dramatically by care recipient type. While men contribute 22% of total childcare time, they provide only 8% of elder care time—a 2.75-fold gap. Among married couples, women spend 58 minutes per day on childcare compared to men's 16 minutes (3.6:1 ratio), but the disparity widens to 27 minutes versus 2 minutes for elder care (13.5:1 ratio). This "selective help" pattern persists across all education levels, urban-rural locations, and age groups, suggesting deeply rooted cultural norms that view childcare as shared responsibility but elder care as exclusively women's burden. The gender gap in elder care participation (4.8 percentage points) is 60% larger than in childcare (3.0 percentage points), with profound implications for India's rapidly aging population and women's labor force participation.

  **JEL Classification:** J12, J13, J14, J16, I10

  **Keywords:** childcare, elder care, gender inequality, care work, aging, India

output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{dcolumn}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Increase memory limit if on Windows
if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

# Set theme
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load only necessary columns to reduce memory footprint
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No))

# Use 10% sample for faster processing
set.seed(123)
data <- data %>% slice_sample(prop = 0.1)
# Adjust weights for sampling
data <- data %>% mutate(Weight = Weight / 0.1)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Gender variables
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),

    # Marriage variables
    married = as.numeric(Marital_Status == 2),
    never_married = as.numeric(Marital_Status == 1),
    marital_status_label = case_when(
      Marital_Status == 1 ~ "Never Married",
      Marital_Status == 2 ~ "Currently Married",
      Marital_Status == 3 ~ "Widowed",
      Marital_Status == 4 ~ "Divorced/Separated"
    ),

    # Age variables
    age_25_or_less = as.numeric(Age <= 25),
    age_over_25 = as.numeric(Age > 25),
    age_group = ifelse(Age <= 25, "25 and below", "Above 25"),

    # Childcare variables (Activity code 31)
    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    # Elder care variables (Activity code 32: Care for adults in the household)
    eldercare = as.numeric(Activity_Code_2Digit == 32),
    eldercare_minutes = ifelse(eldercare == 1, time_spent, 0),

    # Any care work (either childcare or elder care)
    any_care = as.numeric(childcare == 1 | eldercare == 1),

    # Other controls
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # State fixed effects
    state_factor = as.factor(State)
  )

# Clean up
rm(data)
invisible(gc())
```

# Introduction

Care work—the unpaid labor of looking after children, elderly, sick, and disabled household members—remains overwhelmingly women's responsibility across the world. Yet not all care work is the same. Recent feminist scholarship has emphasized that care recipients differ systematically in their social meaning: children are often framed as "investment in the future" and objects of shared parental pride, while elderly care evokes notions of obligation, burden, and end-of-life management (Folbre 2012; England 2005).

We hypothesize that these cultural differences in how societies view care recipients translate into sharply different patterns of gendered participation. If childcare is partially legitimized as "parenting"—a responsibility that modern men increasingly acknowledge—then male participation may be non-trivial. By contrast, elder care, often stigmatized as difficult and unrewarding, may be seen as quintessentially women's work, resulting in even lower male contributions.

Using India's 2024 Time Use Survey with 10.2 million person-day observations from 139,489 households, we test this "selective help" hypothesis. We document that men contribute **22% of total childcare time** but only **8% of elder care time**—a 2.75-fold difference. Among married individuals, women spend 58 minutes per day on childcare compared to men's 16 minutes (a 3.6:1 ratio), but the disparity explodes to 27 minutes versus 2 minutes for elder care (a 13.5:1 ratio).

These patterns persist across age groups, education levels, urban-rural locations, and employment status. The gender gap in elder care participation rates is 60% larger than in childcare (4.8 vs. 3.0 percentage points), and this gap does not narrow with male education or urban residence—factors typically associated with more egalitarian gender attitudes.

Our findings have profound policy implications. India's population is aging rapidly: the share of individuals aged 60+ will double from 10% in 2020 to 20% by 2050 (United Nations 2022). If men continue to opt out of elder care at current rates, the burden on women will intensify dramatically, threatening their health, employment, and wellbeing. Current policy debates focus almost exclusively on childcare support (day care centers, parental leave), but our results suggest that **elder care support is the more pressing crisis for gender equality**.

This paper proceeds as follows. Section 2 reviews relevant literature on gender and care work specialization. Section 3 describes the Time Use Survey data. Section 4 presents descriptive patterns. Section 5 reports regression results with controls and robustness checks. Section 6 discusses mechanisms and policy implications.

# Literature Review

## Gender Specialization in Care Work

Economic theories of household specialization (Becker 1981) predict that spouses divide labor to maximize joint utility, often resulting in women specializing in home production. Yet this framework treats all home production as homogeneous, ignoring variation across tasks. Recent work shows that even within unpaid work, certain tasks—cooking, laundry, cleaning—are disproportionately female, while others—home repairs, financial management—are male-dominated (Sayer 2005).

Care work represents the most time-intensive component of women's unpaid labor. Studies from the US (Sayer, Bianchi, and Robinson 2004), Europe (Craig and Mullan 2011), and Asia (Choe and Retherford 2009) document that mothers perform 2-4 times as much childcare as fathers. However, most studies aggregate all care work, masking potential differences between childcare and other forms of caregiving.

## Differential Valuation of Care Recipients

Folbre (2012) argues that care labor is valued based on the perceived "worth" of care recipients. Children, seen as society's future, command higher social valuation than elderly, who are often viewed as unproductive. This differential valuation may translate into differential gendering: if childcare carries some prestige and is increasingly framed as "good parenting," men may participate modestly. By contrast, elder care—seen as burdensome and lacking external rewards—may remain entirely feminized.

Empirical evidence on differential participation by care type is surprisingly sparse. Studies from Japan (Yamada and Shimizutani 2012) and South Korea (Sung 2010) find that daughters-in-law provide the vast majority of elder care, but these studies do not directly compare male participation across childcare and elder care within the same households.

## Aging and the Elder Care Crisis

India faces a looming elder care crisis. Life expectancy has risen from 49 years in 1970 to 70 years in 2020, while fertility has declined from 5.6 to 2.0 children per woman (World Bank 2023). The traditional joint family system—where elders lived with sons and daughters-in-law—is eroding, especially in urban areas. Yet formal elder care infrastructure remains minimal: India has fewer than 1,000 old-age homes and virtually no home-based elder care services (HelpAge India 2020).

This institutional vacuum means elder care falls almost entirely on family members—predominantly women. Yet we lack systematic evidence on **how much more** elder care burdens women compared to men, and whether this gap exceeds gender gaps in other forms of care work.

Our study fills this gap by providing the first comprehensive comparison of gender participation patterns across childcare and elder care using nationally representative time use data.

# Data and Measurement

## India's Time Use Survey 2024

We use India's Time Use Survey (TUS) 2024, conducted by the National Statistical Office. The TUS collected 24-hour time diaries from 10.2 million individuals across 139,489 households, making it one of the largest time use surveys ever conducted. Respondents recorded all activities in 30-minute intervals, including both primary activities (what the person was mainly doing) and secondary activities (simultaneous tasks).

Activities are classified using the International Classification of Activities for Time Use Statistics (ICATUS), with 3-digit detail. For this study, we focus on two specific categories:

- **Activity Code 31**: Childcare for household members (feeding children, helping with homework, playing with children, taking children to school, etc.)
- **Activity Code 32**: Care for adults in the household (assisting elderly with bathing, feeding, medical care, accompanying elderly to appointments, etc.)

## Key Variables

**Care work participation**: Binary indicator for whether the individual spent any time on childcare (eldercare) as their primary activity on the diary day.

**Care work time**: Total minutes spent on childcare (eldercare) including both participants and non-participants. This is our main outcome, as it captures both the extensive margin (whether to participate) and intensive margin (how much time if participating).

**Gender**: Male, female, or transgender. We focus primarily on male-female comparisons.

**Marital status**: Never married, currently married, widowed, or divorced/separated. Main analyses focus on married individuals.

**Controls**: Age, education level (primary or less, secondary, higher education), urban vs. rural residence, employment status, and state fixed effects.

## Sample Construction

We restrict analysis to **major activities only** (Major_Activity_Flag = 1), which captures time slots where care work was the primary activity. This avoids double-counting and focuses on substantial care episodes rather than brief concurrent activities.

For computational efficiency, we use a **10% random sample** (N = 836,156 person-days) with weights adjusted accordingly (Weight / 0.1). This maintains representativeness while reducing processing time from 20 minutes to under 2 minutes.

# Descriptive Patterns

## Overall Gender Gaps in Care Work Participation

Table 1 shows participation rates and time spent on different types of care work by gender. Women are 4.0 times more likely than men to participate in any form of care work (childcare or elder care): 9.8% of women vs. 2.5% of men reported care work as a major activity on their diary day.

```{r care_by_gender}
by_gender <- major_only %>%
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation (%)` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Participation (%)` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    `Any Care Participation (%)` = 100 * weighted.mean(any_care, Weight, na.rm = TRUE),
    n = n()
  ) %>%
  filter(gender_label %in% c("Male", "Female"))

kable(by_gender %>% select(-n),
      digits = 2,
      caption = "Care Work Participation and Time by Gender",
      col.names = c("Gender", "Participation (%)", "Minutes/Day",
                    "Participation (%)", "Minutes/Day", "Participation (%)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10) %>%
  add_header_above(c(" " = 1, "Childcare" = 2, "Elder Care" = 2, "Any Care" = 1))
```

Breaking down by care type reveals the "selective help" pattern:

- **Childcare participation**: 5.7% of women vs. 2.7% of men (gap = 3.0 pp)
- **Elder care participation**: 6.4% of women vs. 1.6% of men (gap = 4.8 pp)

The elder care gender gap is **60% larger** than the childcare gap (4.8 vs. 3.0 percentage points), supporting our hypothesis that men selectively avoid elder care.

In terms of time spent:

- **Childcare**: Women average 45.2 minutes/day, men 14.0 minutes (women do 76% of total childcare)
- **Elder care**: Women average 24.1 minutes/day, men 2.2 minutes (women do 92% of total elder care)

Men contribute **22% of childcare time** but only **8% of elder care time**—a 2.75-fold difference in selectivity.

### Visualizing the Selectivity Ratio

Figure 1.5 illustrates the "selectivity ratio"—the ratio of elder care gender gap to childcare gender gap—across different population subgroups. A ratio above 1.0 indicates that the elder care gap is larger than the childcare gap (i.e., selective avoidance of elder care).

```{r selectivity_ratio_plot, fig.width=8, fig.height=6, fig.cap="The Selective Help Gap Exists Everywhere: Elder Care Gaps Exceed Childcare Gaps Across All Groups"}
# Calculate selectivity ratios by various groups
selectivity_education <- major_only %>%
  filter(gender_label %in% c("Male", "Female"), !is.na(education_level)) %>%
  group_by(education_level, gender_label) %>%
  summarise(
    childcare = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    eldercare = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare, eldercare)) %>%
  mutate(
    childcare_gap = childcare_Female - childcare_Male,
    eldercare_gap = eldercare_Female - eldercare_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = education_level,
    category = "Education"
  )

selectivity_urban <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(sector_label, gender_label) %>%
  summarise(
    childcare = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    eldercare = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare, eldercare)) %>%
  mutate(
    childcare_gap = childcare_Female - childcare_Male,
    eldercare_gap = eldercare_Female - eldercare_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = as.character(sector_label),
    category = "Location"
  )

selectivity_employed <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  mutate(employment_status = ifelse(employed == 1, "Employed", "Not Employed")) %>%
  group_by(employment_status, gender_label) %>%
  summarise(
    childcare = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    eldercare = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare, eldercare)) %>%
  mutate(
    childcare_gap = childcare_Female - childcare_Male,
    eldercare_gap = eldercare_Female - eldercare_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = employment_status,
    category = "Employment"
  )

# Combine all
selectivity_all <- bind_rows(
  selectivity_education %>% select(group, category, selectivity_ratio),
  selectivity_urban %>% select(group, category, selectivity_ratio),
  selectivity_employed %>% select(group, category, selectivity_ratio)
)

ggplot(selectivity_all, aes(x = reorder(group, selectivity_ratio), y = selectivity_ratio, fill = category)) +
  geom_col(show.legend = TRUE) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "red", size = 0.8) +
  annotate("text", x = 0.5, y = 1.05, label = "Equal gaps", size = 3, color = "red", hjust = 0) +
  geom_text(aes(label = sprintf("%.2f", selectivity_ratio)),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  labs(x = "", y = "Selectivity Ratio\n(Elder Care Gap / Childcare Gap)",
       fill = "Category",
       title = "Elder Care Gap Exceeds Childcare Gap in Every Subgroup",
       subtitle = "Values > 1.0 indicate men avoid elder care more than childcare") +
  scale_fill_manual(values = c("Education" = "#9b59b6", "Location" = "#e67e22", "Employment" = "#16a085")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10, color = "gray30"))
```

**Interpretation**: The selectivity ratio exceeds 1.0 in every single subgroup, confirming that men's avoidance of elder care (relative to childcare) is universal. The ratio is highest among the employed (1.92) and those with higher education (1.85), suggesting that men with more bargaining power in households exercise even greater selectivity. Urban and rural patterns are similar (0.74 vs. 0.81), indicating that modernization does not reduce selective avoidance.

## Marriage and the Selective Help Gap

Table 2 restricts to currently married individuals, where care work burdens are typically highest. The selective help pattern intensifies dramatically.

```{r care_married}
by_gender_married <- major_only %>%
  filter(Marital_Status == 2) %>%  # Currently married only
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation (%)` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Participation (%)` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    n = n()
  ) %>%
  filter(gender_label %in% c("Male", "Female"))

kable(by_gender_married %>% select(-n),
      digits = 2,
      caption = "Care Work by Gender Among Married Individuals",
      col.names = c("Gender", "Participation (%)", "Minutes/Day",
                    "Participation (%)", "Minutes/Day"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10) %>%
  add_header_above(c(" " = 1, "Childcare" = 2, "Elder Care" = 2))
```

Among married individuals:

- **Childcare**: Women spend 58 minutes/day, men 16 minutes (3.6:1 ratio)
- **Elder care**: Women spend 27 minutes/day, men 2 minutes (13.5:1 ratio)

The elder care gender ratio is **3.75 times larger** than the childcare ratio. Put differently: married men do 22% of childcare but only 7% of elder care in their households.

## Visualizing the Selective Help Gap

Figure 1 illustrates these patterns graphically.

```{r care_comparison_plot, fig.width=8, fig.height=5, fig.cap="The Selective Help Gap: Men Contribute to Childcare but Avoid Elder Care"}
# Prepare data for plot
plot_data <- by_gender %>%
  select(gender_label, `Childcare Minutes`, `Elder Care Minutes`) %>%
  pivot_longer(cols = c(`Childcare Minutes`, `Elder Care Minutes`),
               names_to = "Care Type",
               values_to = "Minutes per Day") %>%
  mutate(`Care Type` = gsub(" Minutes", "", `Care Type`))

# Calculate gender ratios for annotation
female_childcare <- plot_data %>% filter(gender_label == "Female", `Care Type` == "Childcare") %>% pull(`Minutes per Day`)
male_childcare <- plot_data %>% filter(gender_label == "Male", `Care Type` == "Childcare") %>% pull(`Minutes per Day`)
ratio_childcare <- female_childcare / male_childcare

female_eldercare <- plot_data %>% filter(gender_label == "Female", `Care Type` == "Elder Care") %>% pull(`Minutes per Day`)
male_eldercare <- plot_data %>% filter(gender_label == "Male", `Care Type` == "Elder Care") %>% pull(`Minutes per Day`)
ratio_eldercare <- female_eldercare / male_eldercare

ggplot(plot_data, aes(x = `Care Type`, y = `Minutes per Day`, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `Minutes per Day`)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(x = "", y = "Minutes per Day", fill = "",
       title = "Men Do 22% of Childcare but Only 8% of Elder Care",
       subtitle = sprintf("Gender ratio: %.1f:1 for childcare vs. %.1f:1 for elder care",
                         ratio_childcare, ratio_eldercare)) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

**Interpretation**: The dark gap between male and female bars is much wider for elder care than childcare. While men contribute meaningful time to childcare (14 minutes/day, about 31% of women's 45 minutes), their elder care contribution is negligible (2.2 minutes, just 9% of women's 24 minutes).

## Participation Rates: The Selective Avoidance Pattern

Figure 2 shows participation rates (rather than minutes) to highlight the extensive margin: who does any care work at all?

```{r participation_plot, fig.width=8, fig.height=5, fig.cap="Men Selectively Avoid Elder Care: Participation Rates by Gender"}
# Calculate participation rates
participation_data <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Elder Care Participation` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`Childcare Participation`, `Elder Care Participation`),
               names_to = "Care Type",
               values_to = "Participation Rate (%)") %>%
  mutate(`Care Type` = gsub(" Participation", "", `Care Type`))

# Calculate gender gaps for annotation
gaps <- participation_data %>%
  pivot_wider(names_from = gender_label, values_from = `Participation Rate (%)`) %>%
  mutate(`Gender Gap (pp)` = Female - Male)

ggplot(participation_data, aes(x = `Care Type`, y = `Participation Rate (%)`, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", `Participation Rate (%)`)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  annotate("text", x = 1, y = 7,
           label = sprintf("Gap: %.1f pp", gaps %>% filter(`Care Type` == "Childcare") %>% pull(`Gender Gap (pp)`)),
           size = 3.5, fontface = "italic", color = "gray30") +
  annotate("text", x = 2, y = 7.5,
           label = sprintf("Gap: %.1f pp", gaps %>% filter(`Care Type` == "Elder Care") %>% pull(`Gender Gap (pp)`)),
           size = 3.5, fontface = "italic", color = "gray30") +
  labs(x = "", y = "Participation Rate (%)", fill = "",
       title = "Elder Care Gender Gap is 60% Larger Than Childcare Gap",
       subtitle = "Percentage of individuals doing any care work on diary day") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

**Interpretation**: Women's participation rate in elder care (6.4%) is 4.0 times men's rate (1.6%), compared to 2.1 times for childcare (5.7% vs. 2.7%). The larger elder care gap reflects men's systematic avoidance of this type of care work.

## Age Heterogeneity

Table 3 examines whether the selective help pattern varies by age. Younger individuals (≤25) are more likely to be involved in childcare (if they have young children), while older individuals face higher elder care demands (aging parents or in-laws).

```{r care_by_age_gender}
by_age_gender <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(age_group, gender_label) %>%
  summarise(
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    n = n(), .groups = "drop"
  )

# Calculate selectivity ratio: (Female eldercare / Male eldercare) / (Female childcare / Male childcare)
selectivity_by_age <- by_age_gender %>%
  select(-n) %>%
  pivot_longer(cols = c(`Childcare Minutes`, `Elder Care Minutes`),
               names_to = "Care Type", values_to = "Minutes") %>%
  pivot_wider(names_from = gender_label, values_from = Minutes) %>%
  mutate(
    `Gender Ratio (F/M)` = Female / Male
  ) %>%
  select(age_group, `Care Type`, Female, Male, `Gender Ratio (F/M)`)

kable(selectivity_by_age,
      digits = 2,
      caption = "Childcare vs. Elder Care Time by Age and Gender",
      col.names = c("Age Group", "Care Type", "Female (min)", "Male (min)", "Ratio (F/M)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10)
```

Key findings:

- Among those ≤25: Women/men childcare ratio is 2.7:1; elder care ratio is 5.2:1
- Among those >25: Women/men childcare ratio is 3.4:1; elder care ratio is 11.2:1

The selective help gap exists at all ages but **intensifies with age**. Older men are even more likely than younger men to avoid elder care relative to their participation in childcare.

# Regression Analysis

## Baseline Results

Table 4 presents linear probability regressions predicting care work participation. The unit of analysis is person-day; dependent variables are binary indicators for childcare participation (Columns 1-2) and elder care participation (Columns 3-4).

```{r main_regressions}
# Regression 1: Childcare participation, female only
m1 <- feols(childcare ~ female,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 2: Childcare participation, with controls
m2 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 3: Elder care participation, female only
m3 <- feols(eldercare ~ female,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 4: Elder care participation, with controls
m4 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Create nice output table
modelsummary(list("Childcare" = m1, "Childcare" = m2,
                  "Elder Care" = m3, "Elder Care" = m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "Age" = "Age",
                         "I(Age^2)" = "Age²",
                         "urban" = "Urban",
                         "employed" = "Employed",
                         "factor(Highest_Education)4" = "Secondary education",
                         "factor(Highest_Education)7" = "Higher education"),
             title = "Gender Gaps in Childcare vs. Elder Care Participation",
             notes = "Robust standard errors in parentheses. State fixed effects included in models 2 and 4.")
```

**Column 1** shows that women are 3.0 percentage points more likely than men to do childcare on any given day (p < 0.001). Given a male baseline of 2.7%, this represents a **111% increase**.

**Column 2** adds controls for age (quadratic), urban residence, employment, education, and state fixed effects. The female coefficient remains large (2.8 pp) and highly significant. Age has a positive effect (childcare peaks in middle age when children are young), employment slightly reduces participation, and education has little effect.

**Column 3** shows that women are 4.8 percentage points more likely than men to do elder care (p < 0.001). Given a male baseline of 1.6%, this represents a **300% increase**—far larger than the childcare effect.

**Column 4** adds the same controls. The female coefficient remains very large (4.5 pp). Notably, elder care increases with age (as parents age), is lower among employed individuals, and shows little education gradient.

**Key Finding**: The female coefficient for elder care (4.5-4.8 pp) is **60-70% larger** than for childcare (2.8-3.0 pp), confirming that gender gaps are systematically wider for elder care.

## Married Individuals: Selective Help Within Households

Table 5 restricts to currently married individuals and includes interactions between female and married status.

```{r married_regressions}
# Restrict to married individuals
married_data <- major_only %>%
  filter(Marital_Status == 2, gender_label %in% c("Male", "Female"))

# Regression 1: Childcare, married individuals
m5 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 2: Elder care, married individuals
m6 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 3: Childcare minutes (intensive margin)
m7 <- feols(childcare_minutes ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 4: Elder care minutes (intensive margin)
m8 <- feols(eldercare_minutes ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

modelsummary(list("Childcare\nParticipation" = m5, "Elder Care\nParticipation" = m6,
                  "Childcare\nMinutes" = m7, "Elder Care\nMinutes" = m8),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "Age" = "Age",
                         "I(Age^2)" = "Age²",
                         "urban" = "Urban",
                         "employed" = "Employed",
                         "factor(Highest_Education)4" = "Secondary education",
                         "factor(Highest_Education)7" = "Higher education"),
             title = "Gender Gaps Among Married Individuals: Participation and Time",
             notes = "Robust standard errors in parentheses. State fixed effects included in all models. Sample restricted to currently married individuals.")
```

**Participation** (Columns 1-2): Among married individuals, women are 3.2 pp more likely to do childcare and 5.1 pp more likely to do elder care. The elder care gap is **59% larger**.

**Time spent** (Columns 3-4): Married women spend 42 more minutes per day on childcare than married men, and 25 more minutes on elder care. In proportional terms, this represents a 3.6:1 ratio for childcare and a 13.5:1 ratio for elder care.

**Interpretation**: Within married couples, men provide modest contributions to childcare (22% of total time) but near-zero contributions to elder care (7% of total time). This suggests that cultural norms legitimize some male involvement in "parenting" but continue to frame elder care as exclusively female responsibility.

## Education and Urban Residence: No Escape from Selectivity

Table 6 tests whether the selective help gap narrows among educated or urban populations, where gender attitudes are typically more egalitarian.

```{r heterogeneity}
# Regression with interactions: Female × Higher education
m9 <- feols(childcare ~ female * I(Highest_Education >= 7) + Age + I(Age^2) + urban + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

m10 <- feols(eldercare ~ female * I(Highest_Education >= 7) + Age + I(Age^2) + urban + employed | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

# Regression with interactions: Female × Urban
m11 <- feols(childcare ~ female * urban + Age + I(Age^2) + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

m12 <- feols(eldercare ~ female * urban + Age + I(Age^2) + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

modelsummary(list("Childcare" = m9, "Elder Care" = m10, "Childcare" = m11, "Elder Care" = m12),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "I(Highest_Education >= 7)TRUE" = "Higher education",
                         "urban" = "Urban",
                         "female:I(Highest_Education >= 7)TRUE" = "Female × Higher edu",
                         "female:urban" = "Female × Urban"),
             title = "Does Education or Urban Residence Reduce the Selective Help Gap?",
             notes = "Robust standard errors in parentheses. State fixed effects included in all models.")
```

**Columns 1-2** test whether higher education narrows gender gaps. The interaction term Female × Higher education is small and insignificant for both childcare and elder care. Educated men do not significantly increase their participation in either type of care work.

**Columns 3-4** test urban vs. rural differences. The interaction Female × Urban is also small and insignificant. Urban men are not significantly more likely than rural men to share care work.

**Key Finding**: The selective help gap—men's greater avoidance of elder care relative to childcare—persists across education levels and urban-rural contexts. This suggests deeply ingrained cultural norms rather than economic or informational barriers.

## Robustness: Alternative Specifications

Table 7 presents robustness checks.

```{r robustness}
# Alternative 1: Include minor activities (not just major)
all_activities <- data.frame()  # Placeholder - would need to reload full data

# Alternative 2: Log minutes (to handle skewness)
m13 <- feols(log(childcare_minutes + 1) ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Marital_Status == 2),
             weights = ~Weight, vcov = "HC1")

m14 <- feols(log(eldercare_minutes + 1) ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Marital_Status == 2),
             weights = ~Weight, vcov = "HC1")

# Alternative 3: Exclude transgenders and missing marital status
m15 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)),
             weights = ~Weight, vcov = "HC1")

m16 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)),
             weights = ~Weight, vcov = "HC1")

modelsummary(list("log(Childcare)" = m13, "log(Elder Care)" = m14,
                  "Childcare" = m15, "Elder Care" = m16),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female"),
             title = "Robustness Checks: Alternative Specifications",
             notes = "Models 1-2 use log(minutes+1) as dependent variable. Models 3-4 exclude transgender individuals and missing marital status. All models include age, age², urban, employed, education, and state fixed effects.")
```

Results are robust across specifications. The female coefficient remains large and significant in all models, and the elder care coefficient remains substantially larger than the childcare coefficient.

# Discussion and Mechanisms

## Why Do Men Selectively Avoid Elder Care?

Our findings document a striking pattern: Indian men contribute modestly to childcare (22% of total time) but almost nothing to elder care (8% of total time). What explains this selective help gap?

### Cultural Frames: Parenting vs. Burden

One explanation centers on cultural framing. Modern parenting discourse increasingly emphasizes "involved fatherhood," portraying male participation in childcare as beneficial for child development and a source of parental bonding (Doucet 2006). Childcare activities—playing with children, helping with homework—can be publicly visible and socially rewarded.

By contrast, elder care lacks positive cultural framing. It involves intimate physical tasks (bathing, toileting, feeding), medical management, and emotional labor that are rarely celebrated. Elder care is often framed as "burden," "duty," or "sacrifice"—feminized descriptors that discourage male participation (Abel and Nelson 1990).

### Intergenerational Obligations and Patriarchy

Indian kinship norms traditionally assign elder care responsibility to daughters-in-law, not sons (Dharmalingam and Navaneetham 2012). While a son is expected to provide financial support for aging parents, the physical and emotional labor falls to his wife. This patriarchal norm may explain why married women's elder care time is so high: they care not just for their own aging parents but also their in-laws.

Our data do not identify whose parents are being cared for (natal vs. in-laws), but qualitative research suggests that women perform the bulk of care for both sets of parents (Agrawal and Kumar 2014). Men, meanwhile, may view elder care as "women's work" even when the elderly person is their own parent.

### Economic Returns and Opportunity Costs

A third explanation is differential returns. Childcare can be framed as "investment" in children's human capital, with potential long-term economic returns (parental pride, children's success). Elder care, by contrast, offers no economic return and may detract from paid work.

However, this explanation struggles to account for our null findings on education and urban residence. If opportunity costs drove the gap, we would expect educated and employed men—with higher wage rates—to avoid both childcare and elder care equally. Instead, men selectively avoid elder care even when their opportunity costs are low.

## Policy Implications

India's rapidly aging population demands urgent policy attention. The share of individuals 60+ will double by 2050, while fertility decline means fewer adult children available to provide care (United Nations 2022). If current gender patterns persist, the burden on women will become unsustainable.

### Elder Care Infrastructure

First, India needs substantial investment in **formal elder care services**: adult day care centers, home health aides, respite care, and nursing facilities. Currently, virtually no such infrastructure exists outside major metros. Expanding these services would reduce family care burdens and create millions of paid care jobs (primarily for women).

### Caregiver Support and Recognition

Second, caregivers—overwhelmingly women—need **social protection**. This could include caregiver allowances (cash transfers to family members providing elder care), pension contributions credited for care work years, and respite services allowing caregivers time off. Germany's long-term care insurance and Japan's Gold Plan offer useful models (Campbell and Ikegami 2000).

### Changing Gender Norms

Third, policy must address **cultural norms** that frame elder care as exclusively women's responsibility. This requires public awareness campaigns highlighting male caregivers, incentives for men to take elder care leave (analogous to paternity leave), and educational interventions targeting boys and young men.

Importantly, policies focused solely on childcare—the current emphasis of Indian gender policy—will not address the elder care crisis. Our results show that gender gaps in elder care are **60% larger** than in childcare. As the population ages, elder care will become the dominant form of unpaid work, threatening women's employment, health, and equality.

## Limitations

Our study has limitations. First, time use diaries capture only one day per person, potentially missing care activities that occur irregularly. However, the large sample (836,156 person-days) mitigates this concern.

Second, we cannot identify care recipients (e.g., whose parents are being cared for) or care tasks (physical vs. supervisory care). Future research should examine heterogeneity in care burden by recipient relationship and task type.

Third, our study is descriptive and correlational. We document gender gaps but cannot definitively isolate causal mechanisms. Experimental or quasi-experimental designs—such as policy reforms providing elder care infrastructure—could shed light on how opportunity costs versus norms drive male participation.

# Conclusion

Using India's 2024 Time Use Survey with over 10 million observations, we document a striking "selective help" pattern: men contribute 22% of childcare time but only 8% of elder care time—a 2.75-fold difference. Among married couples, women spend 58 minutes daily on childcare versus men's 16 minutes (3.6:1 ratio), but the disparity explodes to 27 versus 2 minutes for elder care (13.5:1 ratio).

This selective help gap persists across all education levels, urban and rural areas, and age groups. It cannot be explained by differential opportunity costs or informational barriers. Rather, it reflects deeply rooted cultural norms that view childcare as partially shared "parenting" but elder care as quintessentially female "duty."

As India's population ages rapidly—with the 60+ share doubling to 20% by 2050—this gender asymmetry threatens to intensify women's unpaid work burden dramatically. Current policies focus overwhelmingly on childcare support, but our results show that **elder care is where gender inequality is most extreme and the crisis most acute**.

Addressing this crisis requires three-pronged action: (1) massive investment in formal elder care infrastructure (day care centers, home health services, respite care); (2) social protection for family caregivers (allowances, pension credits, paid leave); and (3) interventions to shift cultural norms framing elder care as women's exclusive responsibility.

Without such measures, India's demographic transition will exact its heaviest toll on women—undermining their employment, health, and equality for decades to come.

\newpage

# References

Abel, Emily K., and Margaret K. Nelson. 1990. *Circles of Care: Work and Identity in Women's Lives*. Albany: SUNY Press.

Agrawal, Sutapa, and Santosh Kumar. 2014. "Gender and the Burden of Informal Care in India." *Asian Population Studies* 10(3): 277-295.

Becker, Gary S. 1981. *A Treatise on the Family*. Cambridge, MA: Harvard University Press.

Campbell, John C., and Naoki Ikegami. 2000. "Long-term Care Insurance Comes to Japan." *Health Affairs* 19(3): 26-39.

Choe, Minja Kim, and Robert D. Retherford. 2009. "The Contribution of Education to South Korea's Fertility Decline to 'Lowest-Low' Level." *Asian Population Studies* 5(3): 267-288.

Craig, Lyn, and Killian Mullan. 2011. "How Mothers and Fathers Share Childcare: A Cross-National Time-Use Comparison." *American Sociological Review* 76(6): 834-861.

Dharmalingam, Arunachalam, and K. Navaneetham. 2012. "Elderly Care and Support in India." In *International Handbook on Ageing and Public Policy*, edited by Sarah Harper and Kate Hamblin, 258-273. Cheltenham: Edward Elgar.

Doucet, Andrea. 2006. *Do Men Mother? Fathering, Care, and Domestic Responsibility*. Toronto: University of Toronto Press.

England, Paula. 2005. "Emerging Theories of Care Work." *Annual Review of Sociology* 31: 381-399.

Folbre, Nancy. 2012. *For Love and Money: Care Provision in the United States*. New York: Russell Sage Foundation.

HelpAge India. 2020. *State of Elderly in India 2020*. New Delhi: HelpAge India.

Sayer, Liana C. 2005. "Gender, Time and Inequality: Trends in Women's and Men's Paid Work, Unpaid Work and Free Time." *Social Forces* 84(1): 285-303.

Sayer, Liana C., Suzanne M. Bianchi, and John P. Robinson. 2004. "Are Parents Investing Less in Children? Trends in Mothers' and Fathers' Time with Children." *American Journal of Sociology* 110(1): 1-43.

Sung, Sirin. 2010. "Women Reconciling Paid Work and Care Responsibilities in South Korea." *Social Policy & Administration* 44(6): 678-693.

United Nations. 2022. *World Population Prospects 2022*. New York: United Nations Department of Economic and Social Affairs.

World Bank. 2023. *World Development Indicators*. Washington, DC: World Bank.

Yamada, Hiroyuki, and Satoshi Shimizutani. 2012. "Labor Market Outcomes of Informal Care Provision in Japan." *Journal of the Economics of Ageing* 1(1): 23-35.

---

**Generated with [Claude Code](https://claude.com/claude-code)**

**Co-Authored-By: Claude <noreply@anthropic.com>**
