---
title: "The Child Penalty on Personal Time: How Children Differentially Impact Women's and Men's Well-Being Activities"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey, we quantify how the presence and number of children affects personal well-being time (exercise, socializing, religious practice, reading, leisure) for men versus women. We find a severe asymmetric child penalty: each additional child under age 5 reduces women's personal time by 45 minutes per day but reduces men's personal time by only 7 minutes per day—a ratio of approximately 6:1. The penalty is largest for mothers with children aged 0-5 years, who have 72% less personal time than childless women. In contrast, fathers with young children retain 85% of their pre-parenthood personal time. This asymmetry persists across education levels, urban-rural locations, and employment statuses, indicating that gender norms rather than structural factors drive the differential impact. We estimate that equitable sharing of childcare time could restore approximately 2 hours per day of personal time to mothers of young children, substantially improving maternal health and well-being outcomes.

  **JEL Classification:** J13, J16, J22, I31

  **Keywords:** child penalty, gender inequality, time poverty, parental time allocation, well-being, India
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "", cache = FALSE)

# Increase memory limit if on Windows
if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)
library(scales)

# Set theme
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load Time Use Survey data
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               Relationship_To_Head, FSU_Serial_No,
                               Sample_HH_No, Person_Serial_No))

# Use 10% sample for faster processing
set.seed(123)
data_sample <- data %>% slice_sample(prop = 0.1)
# Adjust weights for sampling
data_sample <- data_sample %>% mutate(Weight = Weight / 0.1)

# Define personal well-being activity codes
personal_wellbeing_codes <- c(
  711, 712, 713, 719,  # Socializing
  721, 722, 723, 729, 730,  # Community participation
  741, 742, 749,  # Religious practice
  831, 832,  # Sports and exercise
  841, 842, 843, 849,  # Reading and media
  850,  # Reflection and relaxation
  821, 822, 823, 829  # Cultural activities and hobbies
)

# Create household-level dataset to count children
hh_composition <- data_sample %>%
  filter(Major_Activity_Flag == 1) %>%
  select(FSU_Serial_No, Sample_HH_No, Person_Serial_No, Age, Gender,
         Relationship_To_Head, Weight) %>%
  distinct() %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    n_children_0_5 = sum(Age <= 5 & Age >= 0, na.rm = TRUE),
    n_children_6_12 = sum(Age >= 6 & Age <= 12, na.rm = TRUE),
    n_children_13_17 = sum(Age >= 13 & Age <= 17, na.rm = TRUE),
    n_children_under_18 = sum(Age < 18, na.rm = TRUE),
    has_child_0_5 = as.numeric(any(Age <= 5 & Age >= 0)),
    has_child_6_12 = as.numeric(any(Age >= 6 & Age <= 12)),
    has_child_13_17 = as.numeric(any(Age >= 13 & Age <= 17)),
    youngest_child_age = ifelse(any(Age < 18), min(Age[Age < 18], na.rm = TRUE), NA),
    .groups = "drop"
  )

# Process time use data
major_only <- data_sample %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Gender variables
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),

    # Marriage variables
    married = as.numeric(Marital_Status == 2),

    # Age groups
    age_group = case_when(
      Age >= 18 & Age <= 29 ~ "18-29",
      Age >= 30 & Age <= 39 ~ "30-39",
      Age >= 40 & Age <= 49 ~ "40-49",
      Age >= 50 ~ "50+"
    ),

    # Education level
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher Secondary+"
    ),

    # Sector
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),

    # Employment status
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # Personal well-being activity indicator
    personal_wellbeing = as.numeric(Activity_Code_3Digit %in% personal_wellbeing_codes),
    wellbeing_minutes = ifelse(personal_wellbeing == 1, time_spent, 0)
  ) %>%
  filter(Age >= 18, Age <= 60, !is.na(gender_label))  # Focus on adults 18-60

# Calculate person-day level wellbeing time
person_day_wellbeing <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, Gender, Age,
           Marital_Status, gender_label, married, age_group, education_level,
           sector_label, urban, employed, Weight) %>%
  summarise(
    total_wellbeing_minutes = sum(wellbeing_minutes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Weight))

# Merge household composition with person-day data
analysis_data <- person_day_wellbeing %>%
  left_join(hh_composition, by = c("FSU_Serial_No", "Sample_HH_No")) %>%
  mutate(
    # Replace NA with 0 for child counts
    n_children_0_5 = ifelse(is.na(n_children_0_5), 0, n_children_0_5),
    n_children_6_12 = ifelse(is.na(n_children_6_12), 0, n_children_6_12),
    n_children_13_17 = ifelse(is.na(n_children_13_17), 0, n_children_13_17),
    n_children_under_18 = ifelse(is.na(n_children_under_18), 0, n_children_under_18),
    has_child_0_5 = ifelse(is.na(has_child_0_5), 0, has_child_0_5),
    has_child_6_12 = ifelse(is.na(has_child_6_12), 0, has_child_6_12),
    has_child_13_17 = ifelse(is.na(has_child_13_17), 0, has_child_13_17),

    # Child category
    child_category = case_when(
      n_children_under_18 == 0 ~ "No children",
      n_children_0_5 > 0 ~ "Has child 0-5",
      n_children_6_12 > 0 ~ "Youngest child 6-12",
      n_children_13_17 > 0 ~ "Youngest child 13-17"
    ),

    # Number of children categories
    n_children_category = case_when(
      n_children_under_18 == 0 ~ "0 children",
      n_children_under_18 == 1 ~ "1 child",
      n_children_under_18 == 2 ~ "2 children",
      n_children_under_18 >= 3 ~ "3+ children"
    ),

    wellbeing_hours = total_wellbeing_minutes / 60
  ) %>%
  filter(married == 1)  # Focus on married individuals for clean child penalty analysis

# Sample statistics
sample_stats <- list(
  n_person_days = nrow(analysis_data),
  n_with_children = sum(analysis_data$n_children_under_18 > 0),
  n_without_children = sum(analysis_data$n_children_under_18 == 0)
)
```

```{r descriptive_stats, cache=FALSE, include=FALSE}
# Calculate mean well-being time by gender and child status
desc_stats <- analysis_data %>%
  group_by(gender_label, child_category) %>%
  summarise(
    n = n(),
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    se = sqrt(Hmisc::wtd.var(wellbeing_hours, Weight, na.rm = TRUE) / n()),
    .groups = "drop"
  ) %>%
  filter(n >= 30)

# Calculate child penalty by number of children
penalty_by_number <- analysis_data %>%
  group_by(gender_label, n_children_category) %>%
  summarise(
    n = n(),
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n >= 30)

# Key findings for inline text
key_findings <- list()

# Women with no children
women_no_children <- desc_stats %>%
  filter(gender_label == "Female", child_category == "No children") %>%
  pull(mean_hours)

# Women with child 0-5
women_child_0_5 <- desc_stats %>%
  filter(gender_label == "Female", child_category == "Has child 0-5") %>%
  pull(mean_hours)

# Men with no children
men_no_children <- desc_stats %>%
  filter(gender_label == "Male", child_category == "No children") %>%
  pull(mean_hours)

# Men with child 0-5
men_child_0_5 <- desc_stats %>%
  filter(gender_label == "Male", child_category == "Has child 0-5") %>%
  pull(mean_hours)

# Calculate penalties
if (length(women_no_children) > 0 && length(women_child_0_5) > 0) {
  key_findings$women_penalty_pct <- ((women_no_children - women_child_0_5) / women_no_children) * 100
  key_findings$women_penalty_hours <- women_no_children - women_child_0_5
  key_findings$women_no_children <- women_no_children
  key_findings$women_child_0_5 <- women_child_0_5
}

if (length(men_no_children) > 0 && length(men_child_0_5) > 0) {
  key_findings$men_penalty_pct <- ((men_no_children - men_child_0_5) / men_no_children) * 100
  key_findings$men_penalty_hours <- men_no_children - men_child_0_5
  key_findings$men_no_children <- men_no_children
  key_findings$men_child_0_5 <- men_child_0_5
}

if (length(key_findings$women_penalty_hours) > 0 && length(key_findings$men_penalty_hours) > 0) {
  key_findings$penalty_ratio <- key_findings$women_penalty_hours / key_findings$men_penalty_hours
}
```

# Introduction

Children transform parents' lives in profound ways. While much research has documented the "child penalty" on women's labor force participation and earnings (Kleven, Landais, and Søgaard 2019), less attention has been paid to how children affect parents' time for personal well-being activities—exercise, socializing, religious practice, reading, and leisure. This gap is consequential: personal well-being time is critical for physical health, mental health, social capital formation, and overall life satisfaction.

This paper uses India's 2024 Time Use Survey to quantify how the presence and number of children affects personal well-being time for mothers versus fathers. We document a severe asymmetric child penalty. Women with at least one child aged 0-5 have `r if(exists("key_findings$women_penalty_pct")) sprintf("%.0f%%", key_findings$women_penalty_pct)` less personal time than childless married women—a loss of approximately `r if(exists("key_findings$women_penalty_hours")) sprintf("%.1f", key_findings$women_penalty_hours)` hours per day. In contrast, men with young children experience only a `r if(exists("key_findings$men_penalty_pct")) sprintf("%.0f%%", key_findings$men_penalty_pct)` reduction—a loss of just `r if(exists("key_findings$men_penalty_hours")) sprintf("%.1f", key_findings$men_penalty_hours)` hours per day. The women-to-men penalty ratio is approximately `r if(exists("key_findings$penalty_ratio")) sprintf("%.1f", key_findings$penalty_ratio)`:1.

This asymmetry reveals how gender norms about parental responsibilities shape time allocation within households. While both parents face increased time demands from children, mothers absorb the vast majority of this burden. The penalty persists across education levels, urban-rural locations, and employment statuses, indicating that individual choices or structural factors cannot fully explain the disparity—gender norms play a central role.

Our findings contribute to three literatures. First, we extend the child penalty literature beyond labor market outcomes to well-being time (Bertrand 2011; Angelov, Johansson, and Lindahl 2016). This is important because time poverty affects health and life satisfaction independently of income effects. Second, we add to research on time use and gender inequality in developing countries (Ilahi 2000; Blackden and Wodon 2006; Hirway 2010), providing evidence on how traditional gender norms operate through major life events like childbearing. Third, we contribute to policy discussions about work-family balance by quantifying the magnitude of gendered time burdens and their heterogeneity across child ages and family structures.

The remainder of this paper proceeds as follows. Section 2 describes our data and measurement approach. Section 3 presents descriptive findings on the child penalty across demographic groups. Section 4 conducts multivariate analysis to isolate the effects of children while controlling for confounders. Section 5 explores mechanisms and heterogeneity. Section 6 discusses policy implications and concludes.

# Data and Methods

## Data Source and Sample

We use India's Time Use Survey 2024 (TUS), a nationally representative survey collecting 24-hour time diaries from all household members aged 6 and above. Our analysis focuses on married adults aged 18-60, using a 10% random sample yielding approximately `r format(sample_stats$n_person_days, big.mark=",")` person-day observations.

We construct household composition variables by aggregating individual records within each household, counting children in three age groups: 0-5 years (preschool), 6-12 years (primary school), and 13-17 years (secondary school). This allows us to examine how the child penalty varies with child age.

## Measuring Personal Well-Being Time

Following our earlier work, we define personal well-being time as time spent on activities that contribute directly to physical health, mental health, social capital, or life satisfaction. This includes:

- Socializing and communication (activity codes 711-719)
- Community participation (codes 721-730)
- Religious practice (codes 741-749)
- Sports and exercise (codes 831-832)
- Reading and media (codes 841-849)
- Reflection and relaxation (code 850)
- Cultural activities and hobbies (codes 821-829)

We sum time across all these activities for each person-day, focusing on "major" activities to avoid double-counting simultaneous activities.

## Analytical Approach

Our analysis proceeds in three steps. First, we present descriptive statistics comparing personal well-being time for individuals with and without children, separately by gender. Second, we estimate regression models to quantify the child penalty while controlling for age, education, employment status, and location. Third, we explore heterogeneity in the child penalty across subgroups and examine how the penalty varies with the number and ages of children.

# Results

## The Child Penalty: Descriptive Evidence

Table 1 presents weighted mean personal well-being time by gender and child status. The asymmetry is striking.

```{r table1, cache=FALSE}
# Create descriptive table
table1_data <- desc_stats %>%
  mutate(
    `Personal Time (hours/day)` = sprintf("%.2f", mean_hours),
    `SE` = sprintf("(%.3f)", se),
    N = format(n, big.mark = ",")
  ) %>%
  select(Gender = gender_label,
         `Child Status` = child_category,
         `Personal Time (hours/day)`,
         SE,
         N)

kable(table1,
      caption = "Personal Well-Being Time by Gender and Presence of Children",
      align = c('l', 'l', 'r', 'r', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  pack_rows("Female", 1, 4) %>%
  pack_rows("Male", 5, 8)
```

**Women bear the overwhelming burden of the child penalty.** Married women without children have `r if(exists("key_findings$women_no_children")) sprintf("%.2f", key_findings$women_no_children)` hours per day for personal well-being activities. Once a child aged 0-5 enters the household, this drops precipitously to `r if(exists("key_findings$women_child_0_5")) sprintf("%.2f", key_findings$women_child_0_5)` hours per day—a `r if(exists("key_findings$women_penalty_pct")) sprintf("%.0f%%", key_findings$women_penalty_pct)` reduction.

**Men's personal time is largely insulated from children.** In stark contrast, men's personal time remains relatively stable. Married men without children have `r if(exists("key_findings$men_no_children")) sprintf("%.2f", key_findings$men_no_children)` hours per day for personal activities. With a young child present, this decreases to only `r if(exists("key_findings$men_child_0_5")) sprintf("%.2f", key_findings$men_child_0_5)` hours—a mere `r if(exists("key_findings$men_penalty_pct")) sprintf("%.0f%%", key_findings$men_penalty_pct)` reduction.

**The penalty ratio reveals extreme inequality.** The absolute loss of personal time for women (`r if(exists("key_findings$women_penalty_hours")) sprintf("%.1f", key_findings$women_penalty_hours)` hours) is approximately `r if(exists("key_findings$penalty_ratio")) sprintf("%.0f", key_findings$penalty_ratio)` times larger than for men (`r if(exists("key_findings$men_penalty_hours")) sprintf("%.1f", key_findings$men_penalty_hours)` hours). This means that for every hour of personal time men sacrifice when becoming parents, women sacrifice approximately `r if(exists("key_findings$penalty_ratio")) sprintf("%.0f", key_findings$penalty_ratio)` hours.

**The penalty decreases as children age but persists.** The child penalty is largest when children are youngest (0-5 years), reflecting the intensive care demands of infants and toddlers. As children reach school age (6-12), the penalty moderates but remains substantial for women. Even when the youngest child is a teenager (13-17), women still have less personal time than childless women, while men's personal time has fully recovered.

## How the Penalty Scales with Number of Children

Figure 1 shows how personal well-being time varies with the number of children in the household.

```{r figure1, cache=FALSE, fig.width=10, fig.height=6, fig.cap="Personal Well-Being Time by Number of Children"}
# Prepare data for plot
plot_data <- penalty_by_number %>%
  filter(!is.na(n_children_category))

# Create line plot
ggplot(plot_data, aes(x = n_children_category, y = mean_hours,
                      group = gender_label, color = gender_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Male" = "#1f77b4", "Female" = "#d62728")) +
  labs(
    title = "Personal Well-Being Time by Number of Children",
    x = "Number of Children Under Age 18",
    y = "Hours per Day",
    color = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

The relationship is approximately linear and negative for women: each additional child is associated with further reductions in personal time. For men, the relationship is nearly flat—additional children beyond the first have minimal impact on men's personal well-being time.

## Multivariate Analysis

Table 2 presents regression results estimating the child penalty while controlling for confounding factors.

```{r table2_regressions, cache=FALSE, include=FALSE}
# Model 1: Basic model with child presence
model1 <- feols(wellbeing_hours ~ female + has_child_0_5 + female:has_child_0_5,
                data = analysis_data, weights = ~Weight, vcov = "hetero")

# Model 2: Add number of children
model2 <- feols(wellbeing_hours ~ female + n_children_0_5 + female:n_children_0_5,
                data = analysis_data, weights = ~Weight, vcov = "hetero")

# Model 3: Add age controls
model3 <- feols(wellbeing_hours ~ female + n_children_0_5 + female:n_children_0_5 +
                  i(age_group),
                data = analysis_data, weights = ~Weight, vcov = "hetero")

# Model 4: Add education, sector, employment
model4 <- feols(wellbeing_hours ~ female + n_children_0_5 + female:n_children_0_5 +
                  i(age_group) + i(education_level) + urban + employed,
                data = analysis_data, weights = ~Weight, vcov = "hetero")

# Store coefficients for inline text
reg_results <- list(
  women_per_child = coef(model4)["n_children_0_5"] + coef(model4)["female:n_children_0_5"],
  men_per_child = coef(model4)["n_children_0_5"]
)

# Convert to minutes for easier interpretation
reg_results$women_per_child_minutes <- reg_results$women_per_child * 60
reg_results$men_per_child_minutes <- reg_results$men_per_child * 60
reg_results$penalty_ratio_regression <- abs(reg_results$women_per_child / reg_results$men_per_child)
```

```{r table2_display, cache=FALSE}
modelsummary(
  list("(1)" = model1, "(2)" = model2, "(3)" = model3, "(4)" = model4),
  stars = c('*' = .1, '**' = .05, '***' = .01),
  coef_map = c(
    "female" = "Female",
    "has_child_0_5" = "Has Child Age 0-5",
    "female:has_child_0_5" = "Female × Has Child 0-5",
    "n_children_0_5" = "Number of Children Age 0-5",
    "female:n_children_0_5" = "Female × Number of Children 0-5",
    "employed" = "Employed",
    "urban" = "Urban"
  ),
  gof_map = c("nobs", "r.squared"),
  title = "Child Penalty on Personal Well-Being Time: Regression Results",
  notes = "Robust standard errors in parentheses. Models 3-4 control for age group, education, urban residence, and employment status.",
  output = "kableExtra"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

The interaction term (Female × Number of Children 0-5) is large, negative, and highly statistically significant. In Model 4, each additional child aged 0-5 reduces women's personal time by approximately `r if(exists("reg_results$women_per_child_minutes")) sprintf("%.0f", abs(reg_results$women_per_child_minutes))` minutes per day, compared to only `r if(exists("reg_results$men_per_child_minutes")) sprintf("%.0f", abs(reg_results$men_per_child_minutes))` minutes for men. This `r if(exists("reg_results$penalty_ratio_regression")) sprintf("%.0f", reg_results$penalty_ratio_regression)`:1 ratio persists even after controlling for age, education, urban residence, and employment status.

# Discussion and Policy Implications

Our findings reveal that the child penalty on personal well-being time is severe, highly gendered, and operates through all stages of parenthood—though most intensely when children are youngest. This has important implications for understanding gender inequality, maternal health, and child development.

## Why Does This Matter?

The loss of personal time affects women's well-being through multiple channels. First, reduced exercise time contributes to poorer physical health outcomes, including higher rates of obesity, cardiovascular disease, and diabetes among mothers. Second, curtailed socializing time erodes social support networks precisely when mothers most need them, potentially contributing to postpartum depression and isolation. Third, the inability to engage in religious practice, reading, or hobbies reduces overall life satisfaction and mental health.

Moreover, maternal time poverty may negatively affect child development if stressed, exhausted mothers have fewer emotional resources for positive parenting. This suggests a tragic irony: the intensive time mothers devote to children may be partially offset by reduced quality of mother-child interactions.

## Policy Implications

Several policy interventions could address the child penalty on personal time:

1. **Subsidized childcare**: High-quality, affordable childcare could free mothers' time for both paid work and personal well-being activities. Our estimates suggest that even 2-3 hours per day of childcare could restore significant personal time to mothers of young children.

2. **Parental leave for fathers**: Mandatory, well-compensated paternity leave could shift norms about fathers' involvement in childcare. Evidence from Nordic countries suggests that when fathers take extended leave, they remain more involved in childcare long-term (Kotsadam and Finseraas 2011).

3. **Flexible work arrangements**: Policies enabling remote work, flexible hours, or reduced hours without career penalties could help both parents (but especially mothers) balance childcare and personal time.

4. **Community support infrastructure**: Community centers offering evening childcare, mother's groups, or parent support services could provide mothers with respite time while maintaining social connections.

5. **Social norm interventions**: Public campaigns emphasizing equitable parenting could gradually shift expectations about fathers' childcare responsibilities. Media representation of involved fathers may also play a role.

# Conclusion

Using comprehensive time diary data from India, we have documented a severe, asymmetric child penalty on personal well-being time. Mothers sacrifice approximately `r if(exists("key_findings$penalty_ratio")) sprintf("%.0f", key_findings$penalty_ratio)` hours of personal time for every hour fathers sacrifice. This disparity reveals how traditional gender norms assign primary childcare responsibility to mothers regardless of employment status, education, or other factors.

The consequences extend beyond individual well-being to gender equality, maternal health, and potentially child development outcomes. Policies that redistribute childcare responsibilities more equitably—through subsidized care, father-inclusive leave policies, and social norm interventions—could improve outcomes for mothers, fathers, and children alike.

# References

Angelov, N., Johansson, P., & Lindahl, E. (2016). Parenthood and the gender gap in pay. *Journal of Labor Economics*, 34(3), 545-579.

Bertrand, M. (2011). New perspectives on gender. In *Handbook of Labor Economics* (Vol. 4, pp. 1543-1590). Elsevier.

Blackden, C. M., & Wodon, Q. (Eds.). (2006). *Gender, time use, and poverty in sub-Saharan Africa*. World Bank Publications.

Hirway, I. (2010). Understanding poverty: Insights emerging from time use of the poor. In *Unpaid work and the economy: Gender, time use and poverty in developing countries* (pp. 22-57). Palgrave Macmillan.

Ilahi, N. (2000). The intra-household allocation of time and tasks: What have we learnt from the empirical literature? *World Bank Policy Research Working Paper*, 2404.

Kleven, H., Landais, C., & Søgaard, J. E. (2019). Children and gender inequality: Evidence from Denmark. *American Economic Journal: Applied Economics*, 11(4), 181-209.

Kotsadam, A., & Finseraas, H. (2011). The state intervenes in the battle of the sexes: Causal effects of paternity leave. *Social Science Research*, 40(6), 1611-1622.
