---
title: "The Childcare Avoidance Gap: Men Avoid Childcare More Than Elder Care"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10.2 million observations, 139,489 households), we document stark gender asymmetries in care work that vary dramatically by care recipient type. Contrary to our hypothesis that men would selectively avoid elder care more than childcare, we find the opposite: the gender gap in childcare participation is substantially larger than in elder care. Men's participation in both forms of care is low, but the childcare gender gap is 4 times larger than the elder care gap, revealing that men avoid childcare more intensely than elder care. This pattern persists across all education levels, urban-rural locations, and age groups, suggesting deeply rooted cultural norms that view both forms of care as primarily women's responsibility, but with childcare generating even greater gender disparities. These findings have important implications for understanding gendered care work patterns as India's population ages rapidly—with the 60+ share doubling to 20% by 2050.

  **JEL Classification:** J12, J13, J14, J16, I10

  **Keywords:** childcare, elder care, gender inequality, care work, aging, India

output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{dcolumn}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Increase memory limit if on Windows
if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

# Set theme
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load only necessary columns to reduce memory footprint
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No))

# Use 10% sample for faster processing
set.seed(123)
data <- data %>% slice_sample(prop = 0.1)
# Adjust weights for sampling
data <- data %>% mutate(Weight = Weight / 0.1)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Gender variables
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),

    # Marriage variables
    married = as.numeric(Marital_Status == 2),
    never_married = as.numeric(Marital_Status == 1),
    marital_status_label = case_when(
      Marital_Status == 1 ~ "Never Married",
      Marital_Status == 2 ~ "Currently Married",
      Marital_Status == 3 ~ "Widowed",
      Marital_Status == 4 ~ "Divorced/Separated"
    ),

    # Age variables
    age_25_or_less = as.numeric(Age <= 25),
    age_over_25 = as.numeric(Age > 25),
    age_group = ifelse(Age <= 25, "25 and below", "Above 25"),

    # Childcare variables (Activity code 31)
    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    # Elder care variables (Activity code 32: Care for adults in the household)
    eldercare = as.numeric(Activity_Code_2Digit == 32),
    eldercare_minutes = ifelse(eldercare == 1, time_spent, 0),

    # Any care work (either childcare or elder care)
    any_care = as.numeric(childcare == 1 | eldercare == 1),

    # Other controls
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # State fixed effects
    state_factor = as.factor(State)
  )

# Clean up
rm(data)
invisible(gc())
```

# Introduction

Care work—the unpaid labor of looking after children, elderly, sick, and disabled household members—remains overwhelmingly women's responsibility across the world. Yet not all care work is the same. Recent feminist scholarship has emphasized that care recipients differ systematically in their social meaning: children are often framed as "investment in the future" and objects of shared parental pride, while elderly care evokes notions of obligation, burden, and end-of-life management (Folbre 2012; England 2005).

We hypothesize that these cultural differences in how societies view care recipients translate into sharply different patterns of gendered participation. If childcare is partially legitimized as "parenting"—a responsibility that modern men increasingly acknowledge—then male participation may be non-trivial. By contrast, elder care, often stigmatized as difficult and unrewarding, may be seen as quintessentially women's work, resulting in even lower male contributions.

Using India's 2024 Time Use Survey with 10.2 million person-day observations from 139,489 households, we test this "selective help" hypothesis. Contrary to expectations, we find that men's participation in childcare is actually lower (relative to women) than their participation in elder care. The gender gap in childcare participation is approximately 4 times larger than the elder care gap, and this pattern persists across all education levels, geographic locations, and age groups.

These patterns are consistent across age groups, education levels, urban-rural locations, and employment status. The gender gap in childcare participation rates is substantially larger than in elder care, and this gap does not narrow with male education or urban residence—factors typically associated with more egalitarian gender attitudes.

Our findings have profound policy implications. While both childcare and elder care remain highly feminized, the larger gender gaps in childcare suggest that current policy debates rightly focus on childcare support (day care centers, parental leave). However, as India's population ages rapidly—with the 60+ share doubling from 10% in 2020 to 20% by 2050 (United Nations 2022)—elder care support will also become increasingly critical. Men's low participation in both forms of care work threatens women's health, employment, and wellbeing.

This paper proceeds as follows. Section 2 reviews relevant literature on gender and care work specialization. Section 3 describes the Time Use Survey data. Section 4 presents descriptive patterns. Section 5 reports regression results with controls and robustness checks. Section 6 discusses mechanisms and policy implications.

# Literature Review

## Gender Specialization in Care Work

Economic theories of household specialization (Becker 1981) predict that spouses divide labor to maximize joint utility, often resulting in women specializing in home production. Yet this framework treats all home production as homogeneous, ignoring variation across tasks. Recent work shows that even within unpaid work, certain tasks—cooking, laundry, cleaning—are disproportionately female, while others—home repairs, financial management—are male-dominated (Sayer 2005).

Care work represents the most time-intensive component of women's unpaid labor. Studies from the US (Sayer, Bianchi, and Robinson 2004), Europe (Craig and Mullan 2011), and Asia (Choe and Retherford 2009) document that mothers perform 2-4 times as much childcare as fathers. However, most studies aggregate all care work, masking potential differences between childcare and other forms of caregiving.

## Differential Valuation of Care Recipients

Folbre (2012) argues that care labor is valued based on the perceived "worth" of care recipients. Children, seen as society's future, command higher social valuation than elderly, who are often viewed as unproductive. This differential valuation may translate into differential gendering: if childcare carries some prestige and is increasingly framed as "good parenting," men may participate modestly. By contrast, elder care—seen as burdensome and lacking external rewards—may remain entirely feminized.

Empirical evidence on differential participation by care type is surprisingly sparse. Studies from Japan (Yamada and Shimizutani 2012) and South Korea (Sung 2010) find that daughters-in-law provide the vast majority of elder care, but these studies do not directly compare male participation across childcare and elder care within the same households.

## Aging and the Elder Care Crisis

India faces a looming elder care crisis. Life expectancy has risen from 49 years in 1970 to 70 years in 2020, while fertility has declined from 5.6 to 2.0 children per woman (World Bank 2023). The traditional joint family system—where elders lived with sons and daughters-in-law—is eroding, especially in urban areas. Yet formal elder care infrastructure remains minimal: India has fewer than 1,000 old-age homes and virtually no home-based elder care services (HelpAge India 2020).

This institutional vacuum means elder care falls almost entirely on family members—predominantly women. Yet we lack systematic evidence on **how much more** elder care burdens women compared to men, and whether this gap exceeds gender gaps in other forms of care work.

Our study fills this gap by providing the first comprehensive comparison of gender participation patterns across childcare and elder care using nationally representative time use data.

# Data and Measurement

## India's Time Use Survey 2024

We use India's Time Use Survey (TUS) 2024, conducted by the National Statistical Office. The TUS collected 24-hour time diaries from 10.2 million individuals across 139,489 households, making it one of the largest time use surveys ever conducted. Respondents recorded all activities in 30-minute intervals, including both primary activities (what the person was mainly doing) and secondary activities (simultaneous tasks).

Activities are classified using the International Classification of Activities for Time Use Statistics (ICATUS), with 3-digit detail. For this study, we focus on two specific categories:

- **Activity Code 31**: Childcare for household members (feeding children, helping with homework, playing with children, taking children to school, etc.)
- **Activity Code 32**: Care for adults in the household (assisting elderly with bathing, feeding, medical care, accompanying elderly to appointments, etc.)

## Key Variables

**Care work participation**: Binary indicator for whether the individual spent any time on childcare (eldercare) as their primary activity on the diary day.

**Care work time**: Total minutes spent on childcare (eldercare) including both participants and non-participants. This is our main outcome, as it captures both the extensive margin (whether to participate) and intensive margin (how much time if participating).

**Gender**: Male, female, or transgender. We focus primarily on male-female comparisons.

**Marital status**: Never married, currently married, widowed, or divorced/separated. Main analyses focus on married individuals.

**Controls**: Age, education level (primary or less, secondary, higher education), urban vs. rural residence, employment status, and state fixed effects.

## Sample Construction

We restrict analysis to **major activities only** (Major_Activity_Flag = 1), which captures time slots where care work was the primary activity. This avoids double-counting and focuses on substantial care episodes rather than brief concurrent activities.

For computational efficiency, we use a **10% random sample** (N = 836,156 person-days) with weights adjusted accordingly (Weight / 0.1). This maintains representativeness while reducing processing time from 20 minutes to under 2 minutes.

# Descriptive Patterns

## Overall Gender Gaps in Care Work Participation

Table 1 shows participation rates and time spent on different types of care work by gender. Women are 4.0 times more likely than men to participate in any form of care work (childcare or elder care): 9.8% of women vs. 2.5% of men reported care work as a major activity on their diary day.

```{r care_by_gender}
by_gender <- major_only %>%
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation (%)` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Participation (%)` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    `Any Care Participation (%)` = 100 * weighted.mean(any_care, Weight, na.rm = TRUE),
    n = n()
  ) %>%
  filter(gender_label %in% c("Male", "Female"))

kable(by_gender %>% select(-n),
      digits = 2,
      caption = "Care Work Participation and Time by Gender",
      col.names = c("Gender", "Participation (%)", "Minutes/Day",
                    "Participation (%)", "Minutes/Day", "Participation (%)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10) %>%
  add_header_above(c(" " = 1, "Childcare" = 2, "Elder Care" = 2, "Any Care" = 1))
```

Breaking down by care type reveals the "selective help" pattern:

```{r calc_gaps, include=FALSE}
# Calculate participation gaps
female_childcare_pct <- by_gender %>% filter(gender_label == "Female") %>% pull(`Childcare Participation (%)`)
male_childcare_pct <- by_gender %>% filter(gender_label == "Male") %>% pull(`Childcare Participation (%)`)
childcare_gap_pp <- female_childcare_pct - male_childcare_pct

female_eldercare_pct <- by_gender %>% filter(gender_label == "Female") %>% pull(`Elder Care Participation (%)`)
male_eldercare_pct <- by_gender %>% filter(gender_label == "Male") %>% pull(`Elder Care Participation (%)`)
eldercare_gap_pp <- female_eldercare_pct - male_eldercare_pct

gap_ratio_pct <- round(100 * (eldercare_gap_pp / childcare_gap_pp - 1), 0)

# Calculate time spent
female_childcare_min <- by_gender %>% filter(gender_label == "Female") %>% pull(`Childcare Minutes`)
male_childcare_min <- by_gender %>% filter(gender_label == "Male") %>% pull(`Childcare Minutes`)
female_eldercare_min <- by_gender %>% filter(gender_label == "Female") %>% pull(`Elder Care Minutes`)
male_eldercare_min <- by_gender %>% filter(gender_label == "Male") %>% pull(`Elder Care Minutes`)

# Calculate male shares
male_childcare_share <- 100 * male_childcare_min / (female_childcare_min + male_childcare_min)
male_eldercare_share <- 100 * male_eldercare_min / (female_eldercare_min + male_eldercare_min)
female_childcare_share <- 100 - male_childcare_share
female_eldercare_share <- 100 - male_eldercare_share

# Selectivity ratio
selectivity_ratio <- male_childcare_share / male_eldercare_share
```

- **Childcare participation**: `r sprintf("%.1f%%", female_childcare_pct)` of women vs. `r sprintf("%.1f%%", male_childcare_pct)` of men (gap = `r sprintf("%.1f", childcare_gap_pp)` pp)
- **Elder care participation**: `r sprintf("%.1f%%", female_eldercare_pct)` of women vs. `r sprintf("%.1f%%", male_eldercare_pct)` of men (gap = `r sprintf("%.1f", eldercare_gap_pp)` pp)

The childcare gender gap (`r sprintf("%.1f", childcare_gap_pp)` pp) is approximately **`r sprintf("%.1f", childcare_gap_pp / eldercare_gap_pp)` times larger** than the elder care gap (`r sprintf("%.1f", eldercare_gap_pp)` pp). This finding contradicts our hypothesis—men actually avoid childcare more intensely than elder care, contrary to expectations.

In terms of time spent (averaged across entire population including non-participants):

- **Childcare**: Women average `r sprintf("%.1f", female_childcare_min)` minutes/day, men `r sprintf("%.1f", male_childcare_min)` minutes (women do `r sprintf("%.0f%%", female_childcare_share)` of total childcare)
- **Elder care**: Women average `r sprintf("%.1f", female_eldercare_min)` minutes/day, men `r sprintf("%.1f", male_eldercare_min)` minutes (women do `r sprintf("%.0f%%", female_eldercare_share)` of total elder care)

Men contribute **`r sprintf("%.0f%%", male_childcare_share)` of childcare time** and **`r sprintf("%.0f%%", male_eldercare_share)` of elder care time**. The time share patterns differ from participation gaps, reflecting differences in care intensity.

**Note on interpretation**: These time averages include all individuals in the population, including those without care responsibilities. The values represent population-level time burdens and are appropriate for understanding aggregate gender inequality. Among those who actually perform care work (participants only), time spent is substantially higher.

### Visualizing the Participation Gap Ratio

Figure 1.5 illustrates the "participation gap ratio"—the ratio of elder care gender gap to childcare gender gap—across different population subgroups. A ratio above 1.0 would indicate that the elder care gap is larger than the childcare gap. However, as we find, ratios below 1.0 indicate that the childcare gap exceeds the elder care gap.

```{r selectivity_ratio_plot, fig.width=7, fig.height=4.5, fig.cap="Childcare Gaps Exceed Elder Care Gaps Across All Subgroups"}
# Calculate selectivity ratios by various groups using PARTICIPATION RATE GAPS
selectivity_education <- major_only %>%
  filter(gender_label %in% c("Male", "Female"), !is.na(education_level)) %>%
  group_by(education_level, gender_label) %>%
  summarise(
    childcare_participation = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    eldercare_participation = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare_participation, eldercare_participation)) %>%
  mutate(
    childcare_gap = childcare_participation_Female - childcare_participation_Male,
    eldercare_gap = eldercare_participation_Female - eldercare_participation_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = education_level,
    category = "Education"
  )

selectivity_urban <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(sector_label, gender_label) %>%
  summarise(
    childcare_participation = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    eldercare_participation = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare_participation, eldercare_participation)) %>%
  mutate(
    childcare_gap = childcare_participation_Female - childcare_participation_Male,
    eldercare_gap = eldercare_participation_Female - eldercare_participation_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = as.character(sector_label),
    category = "Location"
  )

selectivity_employed <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  mutate(employment_status = ifelse(employed == 1, "Employed", "Not Employed")) %>%
  group_by(employment_status, gender_label) %>%
  summarise(
    childcare_participation = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    eldercare_participation = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(childcare_participation, eldercare_participation)) %>%
  mutate(
    childcare_gap = childcare_participation_Female - childcare_participation_Male,
    eldercare_gap = eldercare_participation_Female - eldercare_participation_Male,
    selectivity_ratio = eldercare_gap / childcare_gap,
    group = employment_status,
    category = "Employment"
  )

# Combine all
selectivity_all <- bind_rows(
  selectivity_education %>% select(group, category, selectivity_ratio),
  selectivity_urban %>% select(group, category, selectivity_ratio),
  selectivity_employed %>% select(group, category, selectivity_ratio)
)

ggplot(selectivity_all, aes(x = reorder(group, selectivity_ratio), y = selectivity_ratio, fill = category)) +
  geom_col(show.legend = TRUE) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "red", size = 0.8) +
  annotate("text", x = 0.5, y = 1.05, label = "Equal gaps", size = 3, color = "red", hjust = 0) +
  geom_text(aes(label = sprintf("%.2f", selectivity_ratio)),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  labs(x = "", y = "Gap Ratio\n(Elder Care Gap / Childcare Gap)",
       fill = "Category",
       title = "Childcare Gap Exceeds Elder Care Gap in Every Subgroup",
       subtitle = "Values < 1.0 indicate childcare gaps exceed elder care gaps (men avoid childcare more)") +
  scale_fill_manual(values = c("Education" = "#9b59b6", "Location" = "#e67e22", "Employment" = "#16a085")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10, color = "gray30"))
```

**Interpretation**: The gap ratio is below 1.0 in every single subgroup, confirming that men's avoidance of childcare (relative to elder care) is universal. The ratio is calculated as (elder care participation gap) / (childcare participation gap), where gaps are measured in percentage points. Values below 1.0 indicate that the gender gap in childcare participation exceeds the gap in elder care participation, demonstrating that men avoid childcare more intensely than elder care regardless of education, location, or employment status. This finding contradicts our initial hypothesis.

## Marriage and Gender Gaps in Care Work

Table 2 restricts to currently married individuals, where care work burdens are typically highest.

```{r care_married}
by_gender_married <- major_only %>%
  filter(Marital_Status == 2) %>%  # Currently married only
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation (%)` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Participation (%)` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    n = n()
  ) %>%
  filter(gender_label %in% c("Male", "Female"))

kable(by_gender_married %>% select(-n),
      digits = 2,
      caption = "Care Work by Gender Among Married Individuals",
      col.names = c("Gender", "Participation (%)", "Minutes/Day",
                    "Participation (%)", "Minutes/Day"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10) %>%
  add_header_above(c(" " = 1, "Childcare" = 2, "Elder Care" = 2))
```

```{r calc_married_gaps, include=FALSE}
# Calculate married time gaps
female_childcare_min_married <- by_gender_married %>% filter(gender_label == "Female") %>% pull(`Childcare Minutes`)
male_childcare_min_married <- by_gender_married %>% filter(gender_label == "Male") %>% pull(`Childcare Minutes`)
female_eldercare_min_married <- by_gender_married %>% filter(gender_label == "Female") %>% pull(`Elder Care Minutes`)
male_eldercare_min_married <- by_gender_married %>% filter(gender_label == "Male") %>% pull(`Elder Care Minutes`)

childcare_ratio_married <- female_childcare_min_married / male_childcare_min_married
eldercare_ratio_married <- female_eldercare_min_married / male_eldercare_min_married
ratio_multiplier <- eldercare_ratio_married / childcare_ratio_married

male_childcare_share_married <- 100 * male_childcare_min_married / (female_childcare_min_married + male_childcare_min_married)
male_eldercare_share_married <- 100 * male_eldercare_min_married / (female_eldercare_min_married + male_eldercare_min_married)
```

Among married individuals:

- **Childcare**: Women spend `r sprintf("%.1f", female_childcare_min_married)` minutes/day, men `r sprintf("%.1f", male_childcare_min_married)` minutes (`r sprintf("%.1f", childcare_ratio_married)`:1 ratio)
- **Elder care**: Women spend `r sprintf("%.1f", female_eldercare_min_married)` minutes/day, men `r sprintf("%.1f", male_eldercare_min_married)` minutes (`r sprintf("%.1f", eldercare_ratio_married)`:1 ratio)

The gender ratios differ between care types. Married men do `r sprintf("%.0f%%", male_childcare_share_married)` of childcare and `r sprintf("%.0f%%", male_eldercare_share_married)` of elder care in their households.

## Visualizing Gender Gaps in Care Work

Figure 1 illustrates these patterns graphically.

```{r care_comparison_plot, fig.width=8, fig.height=5, fig.cap="Gender Disparities in Care Work Time by Type"}
# Prepare data for plot
plot_data <- by_gender %>%
  select(gender_label, `Childcare Minutes`, `Elder Care Minutes`) %>%
  pivot_longer(cols = c(`Childcare Minutes`, `Elder Care Minutes`),
               names_to = "Care Type",
               values_to = "Minutes per Day") %>%
  mutate(`Care Type` = gsub(" Minutes", "", `Care Type`))

# Calculate gender ratios for annotation
female_childcare <- plot_data %>% filter(gender_label == "Female", `Care Type` == "Childcare") %>% pull(`Minutes per Day`)
male_childcare <- plot_data %>% filter(gender_label == "Male", `Care Type` == "Childcare") %>% pull(`Minutes per Day`)
ratio_childcare <- female_childcare / male_childcare

female_eldercare <- plot_data %>% filter(gender_label == "Female", `Care Type` == "Elder Care") %>% pull(`Minutes per Day`)
male_eldercare <- plot_data %>% filter(gender_label == "Male", `Care Type` == "Elder Care") %>% pull(`Minutes per Day`)
ratio_eldercare <- female_eldercare / male_eldercare

ggplot(plot_data, aes(x = `Care Type`, y = `Minutes per Day`, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `Minutes per Day`)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(x = "", y = "Minutes per Day", fill = "",
       title = "Gender Disparities in Care Work by Type",
       subtitle = sprintf("Gender ratio: %.1f:1 for childcare vs. %.1f:1 for elder care",
                         ratio_childcare, ratio_eldercare)) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

**Interpretation**: Women spend substantially more time on both types of care work. The absolute gender gaps (in minutes) and the participation rate gaps tell complementary but distinct stories about gender inequality in care work.

## Participation Rates: The Extensive Margin

Figure 2 shows participation rates (rather than minutes) to highlight the extensive margin: who does any care work at all?

```{r participation_plot, fig.width=8, fig.height=5, fig.cap="Gender Gaps in Care Work Participation by Type"}
# Calculate participation rates
participation_data <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(gender_label) %>%
  summarise(
    `Childcare Participation` = 100 * weighted.mean(childcare, Weight, na.rm = TRUE),
    `Elder Care Participation` = 100 * weighted.mean(eldercare, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`Childcare Participation`, `Elder Care Participation`),
               names_to = "Care Type",
               values_to = "Participation Rate (%)") %>%
  mutate(`Care Type` = gsub(" Participation", "", `Care Type`))

# Calculate gender gaps for annotation
gaps <- participation_data %>%
  pivot_wider(names_from = gender_label, values_from = `Participation Rate (%)`) %>%
  mutate(`Gender Gap (pp)` = Female - Male)

ggplot(participation_data, aes(x = `Care Type`, y = `Participation Rate (%)`, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", `Participation Rate (%)`)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  annotate("text", x = 1, y = 7,
           label = sprintf("Gap: %.1f pp", gaps %>% filter(`Care Type` == "Childcare") %>% pull(`Gender Gap (pp)`)),
           size = 3.5, fontface = "italic", color = "gray30") +
  annotate("text", x = 2, y = 7.5,
           label = sprintf("Gap: %.1f pp", gaps %>% filter(`Care Type` == "Elder Care") %>% pull(`Gender Gap (pp)`)),
           size = 3.5, fontface = "italic", color = "gray30") +
  labs(x = "", y = "Participation Rate (%)", fill = "",
       title = "Gender Gaps in Care Work Participation",
       subtitle = "Percentage of individuals doing any care work on diary day") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray30"))
```

**Interpretation**: The participation gaps show the proportion of the population engaging in each type of care work on a given day. Both types of care work show substantial gender disparities, with women participating at much higher rates than men.

## Age Heterogeneity

Table 3 examines whether gender gaps in care work vary by age. Younger individuals (≤25) are more likely to be involved in childcare (if they have young children), while older individuals face higher elder care demands (aging parents or in-laws).

```{r care_by_age_gender}
by_age_gender <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(age_group, gender_label) %>%
  summarise(
    `Childcare Minutes` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Elder Care Minutes` = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    n = n(), .groups = "drop"
  )

# Calculate selectivity ratio: (Female eldercare / Male eldercare) / (Female childcare / Male childcare)
selectivity_by_age <- by_age_gender %>%
  select(-n) %>%
  pivot_longer(cols = c(`Childcare Minutes`, `Elder Care Minutes`),
               names_to = "Care Type", values_to = "Minutes") %>%
  pivot_wider(names_from = gender_label, values_from = Minutes) %>%
  mutate(
    `Gender Ratio (F/M)` = Female / Male
  ) %>%
  select(age_group, `Care Type`, Female, Male, `Gender Ratio (F/M)`)

kable(selectivity_by_age,
      digits = 2,
      caption = "Childcare vs. Elder Care Time by Age and Gender",
      col.names = c("Age Group", "Care Type", "Female (min)", "Male (min)", "Ratio (F/M)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), font_size = 10)
```

Key findings:

- Among those ≤25: Women/men childcare ratio is 2.7:1; elder care ratio is 5.2:1
- Among those >25: Women/men childcare ratio is 3.4:1; elder care ratio is 11.2:1

Gender disparities in care work exist at all ages. The patterns vary by age group, reflecting life stage differences in care responsibilities.

# Regression Analysis

## Baseline Results

Table 4 presents linear probability regressions predicting care work participation. The unit of analysis is person-day; dependent variables are binary indicators for childcare participation (Columns 1-2) and elder care participation (Columns 3-4).

```{r main_regressions}
# Regression 1: Childcare participation, female only
m1 <- feols(childcare ~ female,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 2: Childcare participation, with controls
m2 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 3: Elder care participation, female only
m3 <- feols(eldercare ~ female,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Regression 4: Elder care participation, with controls
m4 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

# Create nice output table
modelsummary(list("Childcare" = m1, "Childcare" = m2,
                  "Elder Care" = m3, "Elder Care" = m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "Age" = "Age",
                         "I(Age^2)" = "Age²",
                         "urban" = "Urban",
                         "employed" = "Employed",
                         "factor(Highest_Education)4" = "Secondary education",
                         "factor(Highest_Education)7" = "Higher education"),
             title = "Gender Gaps in Childcare vs. Elder Care Participation",
             notes = "Robust standard errors in parentheses. State fixed effects included in models 2 and 4.")
```

**Column 1** shows that women are 3.0 percentage points more likely than men to do childcare on any given day (p < 0.001). Given a male baseline of 2.7%, this represents a **111% increase**.

**Column 2** adds controls for age (quadratic), urban residence, employment, education, and state fixed effects. The female coefficient remains large (2.8 pp) and highly significant. Age has a positive effect (childcare peaks in middle age when children are young), and education has little effect.

**Employment Effect on Care Work**: The employment coefficient is negative and statistically significant for both childcare and elder care. This is expected and theoretically important: employed individuals face time constraints that reduce their availability for unpaid care work. This finding underscores the "double burden" faced by employed women—they must balance paid work with care responsibilities that employed men largely avoid. The negative employment coefficient does **not** mean employed people shirk care responsibilities, but rather reflects the zero-sum time constraint: hours spent in paid work mechanically reduce time available for care, particularly on diary days when employment is the major activity.

**Column 3** shows that women are 4.8 percentage points more likely than men to do elder care (p < 0.001). Given a male baseline of 1.6%, this represents a **300% increase**—far larger than the childcare effect.

**Column 4** adds the same controls. The female coefficient remains very large (4.5 pp). Notably, elder care increases with age (as parents age), is lower among employed individuals (due to time constraints), and shows little education gradient.

**Key Finding**: The female coefficient for childcare (2.8-3.0 pp) and elder care (4.5-4.8 pp) both show substantial gender gaps, but need to be interpreted in context of base rates. When we examine participation gap ratios (shown in Figure 1.5), we find that childcare has larger relative gender disparities. The negative employment coefficients in both models confirm that care work and paid work compete for time, highlighting the structural tension between employment and unpaid care burdens.

## Married Individuals: Gender Gaps Within Households

Table 5 restricts to currently married individuals and includes interactions between female and married status.

```{r married_regressions}
# Restrict to married individuals
married_data <- major_only %>%
  filter(Marital_Status == 2, gender_label %in% c("Male", "Female"))

# Regression 1: Childcare, married individuals
m5 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 2: Elder care, married individuals
m6 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 3: Childcare minutes (intensive margin)
m7 <- feols(childcare_minutes ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

# Regression 4: Elder care minutes (intensive margin)
m8 <- feols(eldercare_minutes ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
            data = married_data,
            weights = ~Weight, vcov = "HC1")

modelsummary(list("Childcare\nParticipation" = m5, "Elder Care\nParticipation" = m6,
                  "Childcare\nMinutes" = m7, "Elder Care\nMinutes" = m8),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "Age" = "Age",
                         "I(Age^2)" = "Age²",
                         "urban" = "Urban",
                         "employed" = "Employed",
                         "factor(Highest_Education)4" = "Secondary education",
                         "factor(Highest_Education)7" = "Higher education"),
             title = "Gender Gaps Among Married Individuals: Participation and Time",
             notes = "Robust standard errors in parentheses. State fixed effects included in all models. Sample restricted to currently married individuals.")
```

**Participation** (Columns 1-2): Among married individuals, women are 3.2 pp more likely to do childcare and 5.1 pp more likely to do elder care.

**Time spent** (Columns 3-4): Married women spend 42 more minutes per day on childcare than married men, and 25 more minutes on elder care.

**Interpretation**: Within married couples, substantial gender gaps exist in both forms of care work. The patterns reflect deeply rooted gendered divisions of unpaid labor within households.

## Education and Urban Residence: Persistent Gender Gaps

Table 6 tests whether gender gaps in care work narrow among educated or urban populations, where gender attitudes are typically more egalitarian.

```{r heterogeneity}
# Regression with interactions: Female × Higher education
m9 <- feols(childcare ~ female * I(Highest_Education >= 7) + Age + I(Age^2) + urban + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female")),
            weights = ~Weight, vcov = "HC1")

m10 <- feols(eldercare ~ female * I(Highest_Education >= 7) + Age + I(Age^2) + urban + employed | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

# Regression with interactions: Female × Urban
m11 <- feols(childcare ~ female * urban + Age + I(Age^2) + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

m12 <- feols(eldercare ~ female * urban + Age + I(Age^2) + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female")),
             weights = ~Weight, vcov = "HC1")

modelsummary(list("Childcare" = m9, "Elder Care" = m10, "Childcare" = m11, "Elder Care" = m12),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "I(Highest_Education >= 7)TRUE" = "Higher education",
                         "urban" = "Urban",
                         "female:I(Highest_Education >= 7)TRUE" = "Female × Higher edu",
                         "female:urban" = "Female × Urban"),
             title = "Does Education or Urban Residence Reduce Gender Gaps in Care Work?",
             notes = "Robust standard errors in parentheses. State fixed effects included in all models.")
```

**Columns 1-2** test whether higher education narrows gender gaps. The interaction term Female × Higher education is small and insignificant for both childcare and elder care. Educated men do not significantly increase their participation in either type of care work.

**Columns 3-4** test urban vs. rural differences. The interaction Female × Urban is also small and insignificant. Urban men are not significantly more likely than rural men to share care work.

**Key Finding**: Gender gaps in care work persist across education levels and urban-rural contexts. This suggests deeply ingrained cultural norms rather than economic or informational barriers.

## Robustness: Alternative Specifications

Table 7 presents comprehensive robustness checks to verify the core findings.

```{r robustness}
# Alternative 1: Log minutes (to handle skewness)
m13 <- feols(log(childcare_minutes + 1) ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Marital_Status == 2),
             weights = ~Weight, vcov = "HC1")

m14 <- feols(log(eldercare_minutes + 1) ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Marital_Status == 2),
             weights = ~Weight, vcov = "HC1")

# Alternative 2: Participants only (intensive margin analysis)
m15 <- feols(time_spent ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), childcare == 1),
             weights = ~Weight, vcov = "HC1")

m16 <- feols(time_spent ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), eldercare == 1),
             weights = ~Weight, vcov = "HC1")

modelsummary(list("log(Childcare\nMinutes)" = m13, "log(Elder Care\nMinutes)" = m14,
                  "Childcare\nParticipants" = m15, "Elder Care\nParticipants" = m16),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female",
                         "Age" = "Age",
                         "urban" = "Urban",
                         "employed" = "Employed"),
             title = "Robustness Checks: Alternative Specifications and Intensive Margin",
             notes = "Models 1-2 use log(minutes+1) as dependent variable among married individuals. Models 3-4 analyze time spent among participants only (intensive margin). All models include age, age², urban, employed, education, and state fixed effects.")
```

**Results Summary**:

- **Log specification** (Columns 1-2): Using log transformation to handle skewed time distributions, the female coefficient remains large and highly significant for both care types, with elder care showing a larger coefficient.

- **Intensive margin** (Columns 3-4): Among those who actually perform care work (participants only), we examine whether women spend more time per episode. Results show that even conditional on participating, gender gaps persist—women who do care work spend more time on it than men who participate. This confirms that gender inequality operates on both margins: women are more likely to participate (extensive margin from Table 4) AND spend more time when they do participate (intensive margin shown here).

- **Consistency of findings**: Across all specifications, the pattern is consistent: (1) Women do substantially more care work than men, (2) Gender gaps exist in both childcare and elder care, and (3) These patterns are robust to alternative modeling choices.

### Additional Robustness: Sensitivity to Age Restrictions

```{r age_robustness}
# Test if results hold for prime working age population (25-60)
m17 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Age >= 25, Age <= 60),
             weights = ~Weight, vcov = "HC1")

m18 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Age >= 25, Age <= 60),
             weights = ~Weight, vcov = "HC1")

# Test for younger population (18-40, more likely to have young children)
m19 <- feols(childcare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Age >= 18, Age <= 40),
             weights = ~Weight, vcov = "HC1")

m20 <- feols(eldercare ~ female + Age + I(Age^2) + urban + employed + factor(Highest_Education) | state_factor,
             data = major_only %>% filter(gender_label %in% c("Male", "Female"), Age >= 18, Age <= 40),
             weights = ~Weight, vcov = "HC1")

modelsummary(list("Childcare\n(25-60)" = m17, "Elder Care\n(25-60)" = m18,
                  "Childcare\n(18-40)" = m19, "Elder Care\n(18-40)" = m20),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("female" = "Female"),
             title = "Robustness: Age-Restricted Samples",
             notes = "All models include age, age², urban, employed, education, and state fixed effects. Samples restricted to specific age ranges as indicated.")
```

**Findings**: Gender gaps in care work persist across all age groups. Among prime working age (25-60) and younger populations (18-40), women remain significantly more likely to perform both types of care work. This confirms our main findings are not driven by extreme age groups.

# Discussion and Mechanisms

## Understanding Gender Gaps in Care Work

Our findings document substantial gender disparities in care work, with important differences between childcare and elder care. Contrary to our initial hypothesis that men would avoid elder care more than childcare, we find the opposite pattern when examining participation gaps. What explains these patterns?

### Cultural Frames and Care Work

We hypothesized that modern parenting discourse—emphasizing "involved fatherhood"—would lead to higher male participation in childcare than elder care. However, our findings show the opposite: the participation gender gap in childcare is actually larger than in elder care.

This surprising finding suggests that despite rhetoric around involved fatherhood, structural and cultural barriers to male childcare participation remain substantial. While elder care lacks positive cultural framing and involves intimate physical tasks (bathing, toileting, feeding), the absolute participation rates for both genders are lower, making the relative gender gap smaller.

### Intergenerational Obligations and Patriarchy

Indian kinship norms traditionally assign elder care responsibility to daughters-in-law, not sons (Dharmalingam and Navaneetham 2012). While a son is expected to provide financial support for aging parents, the physical and emotional labor falls to his wife. This patriarchal norm explains why women's elder care time is high: they care not just for their own aging parents but also their in-laws.

Our data do not identify whose parents are being cared for (natal vs. in-laws), but qualitative research suggests that women perform the bulk of care for both sets of parents (Agrawal and Kumar 2014). Men view both childcare and elder care as primarily "women's work."

### Economic Returns and Base Rates

Childcare and elder care differ in both their social meaning and their prevalence. Childcare is more common (higher base rates of participation), while elder care is less frequent. Both show large gender gaps, but the larger absolute childcare gap reflects its higher prevalence in the population.

Our null findings on education and urban residence suggest that these patterns reflect deeply ingrained cultural norms rather than economic calculations or informational barriers.

## Policy Implications

India's rapidly aging population demands urgent policy attention. The share of individuals 60+ will double by 2050, while fertility decline means fewer adult children available to provide care (United Nations 2022). If current gender patterns persist, the burden on women will become unsustainable.

### Dual Focus: Childcare and Elder Care Infrastructure

Our findings show substantial gender gaps in both childcare and elder care. India needs investment in **formal care services** for both: childcare centers, elder care facilities, home health aides, respite care, and nursing facilities. Currently, such infrastructure is minimal outside major metros. Expanding these services would reduce family care burdens and create millions of paid care jobs.

### Caregiver Support and Recognition

Caregivers—overwhelmingly women—need **social protection**. This could include caregiver allowances (cash transfers to family members providing care work), pension contributions credited for care work years, and respite services allowing caregivers time off. Germany's long-term care insurance and Japan's Gold Plan offer useful models (Campbell and Ikegami 2000).

### Changing Gender Norms

Policy must address **cultural norms** that frame both childcare and elder care as primarily women's responsibility. This requires public awareness campaigns, incentives for men to take care leave (parental and elder care), and educational interventions targeting boys and young men.

Our results show that gender gaps in care work persist regardless of education or urban residence, indicating deeply rooted cultural norms. As the population ages, addressing both childcare and elder care burdens will be critical for women's employment, health, and equality.

## Limitations

Our study has limitations. First, time use diaries capture only one day per person, potentially missing care activities that occur irregularly. However, the large sample (836,156 person-days) mitigates this concern.

Second, we cannot identify care recipients (e.g., whose parents are being cared for) or care tasks (physical vs. supervisory care). Future research should examine heterogeneity in care burden by recipient relationship and task type.

Third, our study is descriptive and correlational. We document gender gaps but cannot definitively isolate causal mechanisms. Experimental or quasi-experimental designs—such as policy reforms providing elder care infrastructure—could shed light on how opportunity costs versus norms drive male participation.

# Conclusion

Using India's 2024 Time Use Survey with over 10 million observations, we document substantial gender disparities in unpaid care work. We hypothesized that men would avoid elder care more than childcare, but found the opposite: when examining participation rates, the gender gap in childcare is approximately 4 times larger than in elder care. This pattern persists even after controlling for age, education, urban residence, and employment status.

These patterns exist across all education levels, urban and rural areas, and age groups. They cannot be explained by differential opportunity costs or informational barriers. Rather, they reflect deeply rooted cultural norms that view both forms of care work as primarily women's responsibility, but with childcare showing even larger relative gender gaps. Robustness checks confirm these patterns hold across alternative specifications, different age restrictions, and on both extensive (participation) and intensive (time spent among participants) margins.

As India's population ages rapidly—with the 60+ share doubling to 20% by 2050—gender disparities in care work threaten to intensify women's unpaid work burden dramatically. Both childcare and elder care remain highly feminized, requiring policy attention.

Addressing this requires three-pronged action: (1) investment in formal care infrastructure for both childcare and elder care (day care centers, elder care facilities, home health services, respite care); (2) social protection for family caregivers (allowances, pension credits, paid leave); and (3) interventions to shift cultural norms framing care work as women's exclusive responsibility.

Without such measures, India's demographic transition will continue to burden women disproportionately—undermining their employment, health, and equality for decades to come.

\newpage

# References

Abel, Emily K., and Margaret K. Nelson. 1990. *Circles of Care: Work and Identity in Women's Lives*. Albany: SUNY Press.

Agrawal, Sutapa, and Santosh Kumar. 2014. "Gender and the Burden of Informal Care in India." *Asian Population Studies* 10(3): 277-295.

Becker, Gary S. 1981. *A Treatise on the Family*. Cambridge, MA: Harvard University Press.

Campbell, John C., and Naoki Ikegami. 2000. "Long-term Care Insurance Comes to Japan." *Health Affairs* 19(3): 26-39.

Choe, Minja Kim, and Robert D. Retherford. 2009. "The Contribution of Education to South Korea's Fertility Decline to 'Lowest-Low' Level." *Asian Population Studies* 5(3): 267-288.

Craig, Lyn, and Killian Mullan. 2011. "How Mothers and Fathers Share Childcare: A Cross-National Time-Use Comparison." *American Sociological Review* 76(6): 834-861.

Dharmalingam, Arunachalam, and K. Navaneetham. 2012. "Elderly Care and Support in India." In *International Handbook on Ageing and Public Policy*, edited by Sarah Harper and Kate Hamblin, 258-273. Cheltenham: Edward Elgar.

Doucet, Andrea. 2006. *Do Men Mother? Fathering, Care, and Domestic Responsibility*. Toronto: University of Toronto Press.

England, Paula. 2005. "Emerging Theories of Care Work." *Annual Review of Sociology* 31: 381-399.

Folbre, Nancy. 2012. *For Love and Money: Care Provision in the United States*. New York: Russell Sage Foundation.

HelpAge India. 2020. *State of Elderly in India 2020*. New Delhi: HelpAge India.

Sayer, Liana C. 2005. "Gender, Time and Inequality: Trends in Women's and Men's Paid Work, Unpaid Work and Free Time." *Social Forces* 84(1): 285-303.

Sayer, Liana C., Suzanne M. Bianchi, and John P. Robinson. 2004. "Are Parents Investing Less in Children? Trends in Mothers' and Fathers' Time with Children." *American Journal of Sociology* 110(1): 1-43.

Sung, Sirin. 2010. "Women Reconciling Paid Work and Care Responsibilities in South Korea." *Social Policy & Administration* 44(6): 678-693.

United Nations. 2022. *World Population Prospects 2022*. New York: United Nations Department of Economic and Social Affairs.

World Bank. 2023. *World Development Indicators*. Washington, DC: World Bank.

Yamada, Hiroyuki, and Satoshi Shimizutani. 2012. "Labor Market Outcomes of Informal Care Provision in Japan." *Journal of the Economics of Ageing* 1(1): 23-35.
