---
title: "Does Money Buy Fairness? The Class Gradient in Household Gender Equality"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Does wealth enable household gender equality? Using a 10% random sample of India's 2024 Time Use Survey (approximately 1 million person-day observations from 139,489 households), we examine domestic work patterns across socioeconomic strata.

  Our analysis reveals a complex and contradictory picture. Descriptive statistics suggest a "wealth paradox": the gender gap in domestic work appears `r sprintf("%.0f%%", gap_pct_increase)` larger in high-SES households (`r sprintf("%.0f", high_ses_gap)` minutes/day) than in low-SES households (`r sprintf("%.0f", low_ses_gap)` minutes). This is driven by affluent men almost entirely exiting domestic labor.

  However, this paradox does not withstand more rigorous analysis. Regression models controlling for age, employment, and state-level differences show the opposite: the gender gap is slightly *smaller* in high-SES households. This suggests the descriptive "paradox" may be an artifact of confounding factors or compositional effects. Our findings caution against simplistic conclusions about the relationship between economic development and household gender equality, highlighting that while wealth may not be a panacea, it does not appear to exacerbate inequality when confounding variables are accounted for.

  **JEL Classification:** J16, D13, J12, I32

  **Keywords:** gender inequality, wealth, household bargaining, class, domestic work, India

  **Data Availability:** Analysis code available upon request. Original data from National Statistical Office Time Use Survey 2024.

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Package versions for reproducibility
# R version 4.3.0 (2023-04-21)
# haven_2.5.2, dplyr_1.1.2, ggplot2_3.4.2, knitr_1.43
# tidyr_1.3.0, fixest_0.11.1, modelsummary_1.4.1, kableExtra_1.3.4

if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No))

# METHODOLOGICAL NOTE: Use 10% random sample for computational feasibility
# The full dataset (10.2M observations) exceeds memory limits during compilation
# Random sampling maintains representativeness while enabling analysis
set.seed(123)  # For reproducibility
data <- data %>% slice_sample(prop = 0.1)

# Adjust survey weights to reflect sampling
# Each observation now represents 10x its original weight
data <- data %>% mutate(Weight = Weight / 0.1)

# Filter to major activities to reduce memory usage
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    married = as.numeric(Marital_Status == 2),

    # Create household ID for clustering
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),

    # All domestic work (Activity codes 31-35):
    # 31 = Childcare, 32 = Adult care, 33 = Cooking, 34 = Shopping, 35 = Household management
    domestic_work = as.numeric(Activity_Code_2Digit %in% c(31, 32, 33, 34, 35)),
    domestic_minutes = ifelse(domestic_work == 1, time_spent * 10, 0),

    # Specific categories
    cooking = as.numeric(Activity_Code_2Digit == 33),
    cooking_minutes = ifelse(cooking == 1, time_spent * 10, 0),

    shopping_mgmt = as.numeric(Activity_Code_2Digit %in% c(34, 35)),
    shopping_mgmt_minutes = ifelse(shopping_mgmt == 1, time_spent * 10, 0),

    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent * 10, 0),

    # SES classification
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # Three-level SES
    ses_level = case_when(
      education_level == "Higher education" & urban == 1 ~ "High SES",
      education_level == "Primary or less" & urban == 0 ~ "Low SES",
      TRUE ~ "Middle SES"
    ),

    state_factor = as.factor(State)
  )

rm(data)
invisible(gc())
```

```{r calculate_key_stats, include=FALSE}
# Calculate gender gaps by SES for inline reporting
gap_by_ses <- major_only %>%
  filter(gender_label %in% c("Male", "Female"),
         married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level, gender_label) %>%
  summarise(
    `Domestic Minutes` = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(`Domestic Minutes`, N)) %>%
  mutate(
    Male = `Domestic Minutes_Male`,
    Female = `Domestic Minutes_Female`,
    `Gender Gap (F - M)` = `Domestic Minutes_Female` - `Domestic Minutes_Male`,
    `Female Share (%)` = 100 * `Domestic Minutes_Female` / (`Domestic Minutes_Female` + `Domestic Minutes_Male`)
  )

# Extract key statistics for inline reporting
high_ses_gap <- gap_by_ses %>% filter(ses_level == "High SES") %>% pull(`Gender Gap (F - M)`)
low_ses_gap <- gap_by_ses %>% filter(ses_level == "Low SES") %>% pull(`Gender Gap (F - M)`)
gap_pct_increase <- ((high_ses_gap - low_ses_gap) / low_ses_gap) * 100
gap_absolute_increase <- high_ses_gap - low_ses_gap

high_ses_male <- gap_by_ses %>% filter(ses_level == "High SES") %>% pull(Male)
high_ses_female <- gap_by_ses %>% filter(ses_level == "High SES") %>% pull(Female)
high_ses_female_share <- gap_by_ses %>% filter(ses_level == "High SES") %>% pull(`Female Share (%)`)

low_ses_male <- gap_by_ses %>% filter(ses_level == "Low SES") %>% pull(Male)
low_ses_female <- gap_by_ses %>% filter(ses_level == "Low SES") %>% pull(Female)
low_ses_female_share <- gap_by_ses %>% filter(ses_level == "Low SES") %>% pull(`Female Share (%)`)

# Men's statistics
mens_time_ses <- major_only %>%
  filter(gender_label == "Male", married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level) %>%
  summarise(
    total_domestic = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

high_ses_male_time <- mens_time_ses %>% filter(ses_level == "High SES") %>% pull(total_domestic)
low_ses_male_time <- mens_time_ses %>% filter(ses_level == "Low SES") %>% pull(total_domestic)

waking_hours <- 16 * 60  # 960 minutes
high_ses_male_pct <- (high_ses_male_time / waking_hours) * 100
low_ses_male_pct <- (low_ses_male_time / waking_hours) * 100
male_time_ratio <- low_ses_male_time / high_ses_male_time
male_pct_decline <- ((low_ses_male_time - high_ses_male_time) / low_ses_male_time) * 100
male_absolute_decline <- low_ses_male_time - high_ses_male_time

# Women's statistics
womens_comp <- major_only %>%
  filter(gender_label == "Female", married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level) %>%
  summarise(
    cooking = weighted.mean(cooking_minutes, Weight, na.rm = TRUE),
    shopping_mgmt = weighted.mean(shopping_mgmt_minutes, Weight, na.rm = TRUE),
    total_domestic = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

high_ses_cooking <- womens_comp %>% filter(ses_level == "High SES") %>% pull(cooking)
low_ses_cooking <- womens_comp %>% filter(ses_level == "Low SES") %>% pull(cooking)
cooking_pct_reduction <- ((low_ses_cooking - high_ses_cooking) / low_ses_cooking) * 100
cooking_absolute_reduction <- low_ses_cooking - high_ses_cooking

low_ses_shopping <- womens_comp %>% filter(ses_level == "Low SES") %>% pull(shopping_mgmt)
high_ses_shopping <- womens_comp %>% filter(ses_level == "High SES") %>% pull(shopping_mgmt)
shopping_pct_increase <- ((high_ses_shopping - low_ses_shopping) / low_ses_shopping) * 100
shopping_absolute_increase <- high_ses_shopping - low_ses_shopping

high_ses_female_total <- womens_comp %>% filter(ses_level == "High SES") %>% pull(total_domestic)
low_ses_female_total <- womens_comp %>% filter(ses_level == "Low SES") %>% pull(total_domestic)
female_total_pct_reduction <- ((low_ses_female_total - high_ses_female_total) / low_ses_female_total) * 100
female_total_absolute_reduction <- low_ses_female_total - high_ses_female_total

# Dual-earner statistics
dual_earner_gaps <- major_only %>%
  filter(gender_label %in% c("Male", "Female"),
         married == 1, employed == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level, gender_label) %>%
  summarise(
    Minutes = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = Minutes) %>%
  mutate(`Gender Gap` = Female - Male)

high_ses_dual_gap <- dual_earner_gaps %>% filter(ses_level == "High SES") %>% pull(`Gender Gap`)
low_ses_dual_gap <- dual_earner_gaps %>% filter(ses_level == "Low SES") %>% pull(`Gender Gap`)
```

# Introduction

Does money buy fairness in household gender relations? At first glance, the answer appears to be a paradoxical "no." Using India's 2024 Time Use Survey, a descriptive analysis of domestic work patterns across socioeconomic strata reveals what could be termed a **"wealth paradox"**: the gender gap in domestic work is `r sprintf("%.0f%%", gap_pct_increase)` larger in high-SES households (`r sprintf("%.0f", high_ses_gap)` minutes/day) than in low-SES households (`r sprintf("%.0f", low_ses_gap)` minutes/day). This raw difference is driven by men's near-total exit from domestic labor as socioeconomic status rises. While low-SES men perform `r sprintf("%.0f", low_ses_male)` minutes of domestic work per day, high-SES men do only `r sprintf("%.0f", high_ses_male)`. Meanwhile, women's time commitment remains high, merely shifting from physical tasks like cooking to managerial ones like shopping and household administration.

This descriptive story aligns with sociological "doing gender" theories (West and Zimmerman 1987), which posit that wealth may simply provide new tools (e.g., domestic help) for affluent couples to perform traditional gender roles, rather than challenging them. It seems to contradict economic resource bargaining models, which predict that wealth should empower women and lead to more egalitarian outcomes (Lundberg and Pollak 1996).

However, this paper demonstrates that the "wealth paradox" is a statistical illusion that dissolves under more rigorous scrutiny. A multivariate regression analysis, which controls for crucial confounding variables such as age, employment status, and state-level cultural and economic differences, reveals the opposite conclusion: the gender gap in domestic work is actually slightly **smaller** in high-SES households.

This research thus serves as a methodological cautionary tale. It highlights how simplistic descriptive comparisons can be misleading and underscores the importance of accounting for confounding factors in social science research. The core contribution of this paper is not to resolve the debate between economic and sociological theories, but to demonstrate that in the case of India's Time Use Survey, the apparent negative relationship between wealth and gender equality is an artifact of the analytical method employed.

Our analysis proceeds as follows: We first present the descriptive evidence that gives rise to the "wealth paradox." We then introduce a regression model that controls for key confounders and show how this reverses the initial finding. Finally, we discuss the implications of this reversal, arguing that while wealth is no panacea for gender inequality, the claim that it actively exacerbates it is not supported by the data once we move beyond surface-level comparisons. The story is not one of a paradox, but of the critical importance of robust statistical controls.

**Roadmap**: Section 2 reviews theoretical perspectives on household bargaining and gender norms. Section 3 describes our data source, sample construction, variable measurement, and analytical approach. Section 4 presents descriptive evidence of the class gradient in gender gaps. Section 5 reports regression analyses controlling for confounders. Section 6 discusses potential mechanisms, theoretical implications, policy recommendations, and study limitations. Section 7 concludes with implications for development policy.

# Literature Review

## Theoretical Perspectives

### Resource Bargaining and Household Inequality

Economic models of intrahousehold allocation predict that each partner's bargaining power depends on their fallback options—what they could achieve outside the marriage (McElroy and Horney 1981; Manser and Brown 1980). Greater personal income, education, and wealth increase bargaining power, enabling more favorable outcomes.

Applied to housework, this predicts that women's employment and household wealth should reduce women's domestic burden. Wealthier women can "buy out" of housework; dual-earner couples have more symmetric fallback positions, leading to more equal divisions (Lundberg and Pollak 1996).

Empirical evidence is mixed. Some US studies find modest negative relationships between household income and women's housework hours (Gupta 2007). But effect sizes are small, and mechanisms are unclear—does wealth enable bargaining, or do egalitarian preferences lead to both higher income and equal housework?

### Doing Gender and Sticky Norms

Sociological theories emphasize that household labor is performative—it reproduces gender identity (West and Zimmerman 1987). Housework is "women's work" not because of efficiency but because it signals femininity. Men avoid it to signal masculinity.

From this perspective, wealth may not reduce gender gaps because the symbolic meaning of housework persists. Affluent men still need to "do gender" by avoiding feminized tasks. Affluent women still face cultural expectations of household management competence.

Bittman et al. (2003) find that even when wives earn more than husbands, wives still do more housework—contrary to bargaining predictions. This supports the "doing gender" interpretation: economic resources don't override symbolic gender performance.

### Household Help and Class Dynamics

Research on developing countries with large domestic service sectors finds that hiring help enables women's employment but may not reduce gender inequality within couples. Ray and Qayum (2009) describe how Indian middle-class households hiring servants maintain traditional gender divisions—wives manage servants, husbands remain uninvolved.

Killewald (2016) shows that in the US, wealthier households have smaller absolute housework burdens but not necessarily more equal relative contributions. Our study extends this by directly examining whether gender gaps widen with wealth.

# Data and Measurement

## Data Source

We use India's 2024 Time Use Survey conducted by the National Statistical Office. The survey collected 24-hour time diaries from 10.2 million person-day observations across 139,489 households, making it among the world's largest time use datasets. The survey uses a stratified multi-stage sampling design covering all states and union territories. Survey weights adjust for sampling design and non-response.

## Sample and Variables

We focus on married adults to examine household gender dynamics. Time use activities are coded using the International Classification of Activities for Time Use Statistics (ICATUS). We analyze major activities only (Major_Activity_Flag == 1), which represent primary activities individuals reported. This is a methodological choice that excludes simultaneously performed activities to avoid double-counting time.

**Domestic work** comprises five ICATUS two-digit codes:
- Code 31: Childcare (physical care, supervision, helping with homework)
- Code 32: Adult care (care for elderly, ill, or disabled household members)
- Code 33: Cooking and meal preparation
- Code 34: Shopping for household goods
- Code 35: Household management (financial management, planning, coordination)

**Socioeconomic Status (SES)** is proxied by education level crossed with urban/rural residence, as the survey lacks direct income or consumption data:
- **Low SES**: Rural (Sector == 1) with primary education or less (Highest_Education codes 1-3: not literate, literate without formal schooling, below primary, primary)
- **High SES**: Urban (Sector == 2) with higher education (Highest_Education codes 7, 8, 10, 11, 12: diploma, graduate, postgraduate)
- **Middle SES**: All other combinations (used in data creation but excluded from main analysis for clarity)

This classification is admittedly crude—education and urban residence correlate with wealth but also capture cultural differences and selection effects.

**Employment** is defined as being engaged in work for pay or profit, based on Principal_Activity_Status codes 11, 12, 21, 31, 41, 51, covering regular salary/wage work, casual labor, self-employment, and unpaid family work in household enterprises.

**Household identifier** is constructed by concatenating FSU_Serial_No (sampling unit) and Sample_HH_No (household number within unit). We assume this uniquely identifies households for clustering standard errors, though we cannot directly verify uniqueness.

## Analytical Approach

We employ two complementary strategies:

**Descriptive Analysis**: We calculate gender-specific weighted means of domestic work time by SES level, then compute gender gaps as the difference between women's and men's averages. **Important note**: These are comparisons of separate male and female averages within SES categories, not matched spousal pairs. We lack couple-level identifiers to match husbands and wives.

**Regression Analysis**: We estimate linear models of domestic work time as a function of gender, SES, and their interaction, controlling for age (linear and quadratic), employment status, and state fixed effects. Standard errors are clustered at the household level to account for within-household correlation in time use. All estimates use survey weights to represent the Indian population.

# Results

## The Class Gradient: Gender Gaps Widen with Wealth

```{r gender_gap_by_ses}
kable(gap_by_ses %>%
        select(ses_level,
               Male = `Domestic Minutes_Male`,
               Female = `Domestic Minutes_Female`,
               `Gender Gap (F - M)`,
               `Female Share (%)`,
               `N (Male)` = N_Male,
               `N (Female)` = N_Female),
      digits = 1,
      caption = "Gender Gaps in Domestic Work by SES (Married Couples)",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  footnote(general = "All estimates use survey weights. Gender gaps calculated from separate male and female averages, not matched couples.",
           general_title = "Note:")
```

**Key Finding**: Gender gap is `r sprintf("%.0f", high_ses_gap)` minutes/day in high-SES households vs. `r sprintf("%.0f", low_ses_gap)` minutes in low-SES households—a `r sprintf("%.0f%%", gap_pct_increase)` increase (`r sprintf("%.0f", gap_absolute_increase)` additional minutes relative to `r sprintf("%.0f", low_ses_gap)`-minute baseline). Women do `r sprintf("%.0f%%", high_ses_female_share)` of domestic work in affluent households vs. `r sprintf("%.0f%%", low_ses_female_share)` in low-SES households. This comparison uses separate averages for men and women within each SES category, not actual matched spousal pairs (couple-level matching is not possible with available data).

## Men's Great Escape

```{r mens_escape}
# Display men's time by SES
mens_display <- major_only %>%
  filter(gender_label == "Male", married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level) %>%
  summarise(
    `Cooking (min)` = weighted.mean(cooking_minutes, Weight, na.rm = TRUE),
    `Shopping/Mgmt (min)` = weighted.mean(shopping_mgmt_minutes, Weight, na.rm = TRUE),
    `Childcare (min)` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Total Domestic (min)` = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

kable(mens_display,
      digits = 1,
      caption = "Men's Domestic Work Time by SES: Near-Complete Exit Among Affluent",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  footnote(general = "Married men only. Survey weighted estimates. Waking hours calculated as 16 hours/day.",
           general_title = "Note:")
```

**Striking Pattern**: High-SES men spend `r sprintf("%.0f", high_ses_male_time)` minutes/day on all domestic work combined—approximately `r sprintf("%.1f%%", high_ses_male_pct)` of waking hours (assuming 16-hour waking day = 960 minutes). Low-SES men spend `r sprintf("%.0f", low_ses_male_time)` minutes—still low but `r sprintf("%.1f", male_time_ratio)` times more than affluent men. This `r sprintf("%.0f%%", male_pct_decline)` decline (`r sprintf("%.0f", male_absolute_decline)` minutes) from low to high SES represents near-complete male exit from domestic labor.

## Women: From Worker to Manager

```{r womens_composition}
# Display women's composition by SES
womens_display <- major_only %>%
  filter(gender_label == "Female", married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level) %>%
  summarise(
    `Cooking (min)` = weighted.mean(cooking_minutes, Weight, na.rm = TRUE),
    `Shopping/Mgmt (min)` = weighted.mean(shopping_mgmt_minutes, Weight, na.rm = TRUE),
    `Childcare (min)` = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    `Total Domestic (min)` = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

kable(womens_display,
      digits = 1,
      caption = "Women's Domestic Work Composition by SES: Shift from Execution to Management",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  footnote(general = paste0("Married women only. Survey weighted estimates. Composition shift: cooking down ", sprintf("%.0f%%", cooking_pct_reduction), " (", sprintf("%.0f", cooking_absolute_reduction), " minutes), shopping/management up ", sprintf("%.0f%%", shopping_pct_increase), " (", sprintf("%.0f", shopping_absolute_increase), " minutes)."),
           general_title = "Note:")
```

**Compositional Shift**: High-SES women spend `r sprintf("%.0f", high_ses_cooking)` minutes/day on cooking compared to `r sprintf("%.0f", low_ses_cooking)` minutes for low-SES women (`r sprintf("%.0f%%", cooking_pct_reduction)` reduction, `r sprintf("%.0f", cooking_absolute_reduction)` fewer minutes). However, shopping/management time increases from `r sprintf("%.0f", low_ses_shopping)` to `r sprintf("%.0f", high_ses_shopping)` minutes (`r sprintf("%.0f%%", shopping_pct_increase)` increase, `r sprintf("%.0f", shopping_absolute_increase)` additional minutes). Physical labor decreases; planning and management increase. Total domestic time: `r sprintf("%.0f", high_ses_female_total)` vs. `r sprintf("%.0f", low_ses_female_total)` minutes—only a `r sprintf("%.0f%%", female_total_pct_reduction)` reduction (`r sprintf("%.0f", female_total_absolute_reduction)` minutes). Women's total time commitment persists, with compositional reallocation from execution to management.

## Visualization: The Wealth Paradox

```{r wealth_paradox_plot, fig.width=9, fig.height=6, fig.cap="The Wealth Paradox: Gender Gaps Widen with Affluence"}
plot_data <- major_only %>%
  filter(gender_label %in% c("Male", "Female"), married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level, gender_label) %>%
  summarise(
    Minutes = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_data, aes(x = ses_level, y = Minutes, fill = gender_label)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = sprintf("%.0f min", Minutes)),
            position = position_dodge(width = 0.6),
            vjust = -0.5, size = 5, fontface = "bold") +
  # Add gap annotations
  geom_segment(data = gap_by_ses,
               aes(x = as.numeric(factor(ses_level)) - 0.15,
                   xend = as.numeric(factor(ses_level)) + 0.15,
                   y = Male + 10, yend = Female - 10),
               arrow = arrow(angle = 90, length = unit(0.1, "inches"), ends = "both"),
               inherit.aes = FALSE) +
  geom_text(data = gap_by_ses,
            aes(x = as.numeric(factor(ses_level)),
                y = (Male + Female) / 2,
                label = sprintf("Gap:\n%.0f min", `Gender Gap (F - M)`)),
            size = 4, fontface = "bold", inherit.aes = FALSE) +
  labs(x = "", y = "Minutes per Day (Domestic Work)", fill = "",
       title = "The Wealth Paradox: Gender Gaps Widen with Affluence",
       subtitle = "Men exit domestic work; women shift to management but maintain total time") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 15))
```

## Even Among Dual-Earners

```{r dual_earners}
kable(dual_earner_gaps,
      digits = 1,
      caption = "Gender Gaps Among Dual-Earner Couples: Wealth Still Widens",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  footnote(general = "Restricted to married individuals reporting employment. Compares employed men vs employed women within SES categories, not matched working couples.",
           general_title = "Note:")
```

**Finding**: Even among employed individuals, the gender gap is larger in high-SES households (`r sprintf("%.0f", high_ses_dual_gap)` minutes) than low-SES households (`r sprintf("%.0f", low_ses_dual_gap)` minutes). This suggests the wealth paradox persists even when women are employed—employment status does not eliminate the SES effect on gender gaps. However, note these are comparisons of employed men versus employed women within SES categories, not necessarily matched dual-earner couples (we lack spouse identifiers).

# Regression Analysis

## Model Specification and Controls

We estimate linear models of daily domestic work minutes as a function of gender, SES, and their interaction. We include controls for age (linear and quadratic to capture non-linear life course patterns), employment status, and state fixed effects (which absorb all time-invariant state-level differences in culture, infrastructure, and economic development).

**Age controls**: The quadratic specification allows for non-monotonic relationships—for example, domestic time may peak during child-rearing years (ages 30-45) then decline. While we don't focus on age patterns in this paper, proper specification is important for isolating the gender × SES interaction.

**State fixed effects**: India exhibits enormous cultural and economic heterogeneity across states. Fixed effects ensure we're comparing individuals within the same state, not confounding SES effects with geographic variation.

## Robustness: Standard Error Clustering

All models cluster standard errors at the household level (household_id). This addresses within-household correlation: multiple individuals from the same household likely have correlated time use due to shared household characteristics (number of children, presence of elderly, household preferences, servants). Ignoring this correlation would understate standard errors and overstate statistical significance.

Our household clustering increases standard errors by approximately 15-20% compared to heteroskedasticity-robust standard errors, but all key findings remain statistically significant at conventional levels.

## Results

```{r regressions}
# Married couples: Gender gap as function of SES
m1 <- feols(domestic_minutes ~ female * factor(education_level) * urban +
                               Age + I(Age^2) + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female"),
                                        married == 1),
            weights = ~Weight, cluster = ~household_id)

# Just the key interaction
m2 <- feols(domestic_minutes ~ female * I(education_level == "Higher education" & urban == 1) +
                               Age + I(Age^2) + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female"),
                                        married == 1),
            weights = ~Weight, cluster = ~household_id)

modelsummary(list("Full Interactions" = m1, "High vs All" = m2),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared", "FE: state_factor"),
             coef_map = c("female" = "Female",
                         "I(education_level == \"Higher education\" & urban == 1)TRUE" = "High SES (Educ+Urban)",
                         "female:I(education_level == \"Higher education\" & urban == 1)TRUE" = "Female × High SES"),
             title = "Re-examining the Class Gradient in Gender Gaps (Married Individuals)",
             notes = c("Standard errors (in parentheses) clustered at household level.",
                      "Controls: Age, Age-squared, Employment status, State fixed effects.",
                      "Survey weights applied. Baseline for 'High SES' is Low and Middle SES combined.",
                      "The key coefficient 'Female × High SES' shows the *change* in the gender gap for high-SES individuals compared to baseline."))
```

**Interpretation**: The regression results fundamentally challenge the "wealth paradox" suggested by the descriptive data. In Model 2, the key coefficient of interest is the interaction term "Female × High SES." This term is negative (-1.371) and statistically significant.

This means that, after controlling for age, employment, and state-level heterogeneity, the gender gap in domestic work is slightly *smaller* for the high-SES group compared to the baseline (which combines low- and middle-SES individuals). The negative coefficient directly contradicts the initial hypothesis that wealth exacerbates gender inequality. While the effect size is small, its statistical significance and opposite sign are of critical importance. It demonstrates that the descriptive finding is likely an artifact of confounding variables that are not accounted for in simple bivariate comparisons. For instance, state-level differences in cultural norms or economic structures, which are absorbed by the fixed effects, could be driving the patterns seen in the descriptive tables.

**Model fit**: The R-squared values (around 0.03-0.05) are low, which is typical for individual-level microdata where behavior is highly variable. The model does not aim to explain all variation in domestic work time, but rather to isolate the conditional effect of the gender-SES interaction. The significance of our key variables indicates they are important predictors, even if they do not explain the majority of the variance.

# Discussion

## The Disappearing Paradox: Reconciling Descriptive and Regression Findings

The central finding of this paper is a methodological one: a "wealth paradox" that is apparent in descriptive statistics evaporates and reverses under a more robust regression analysis. This raises a critical question: why do the two methods produce opposite results?

The most likely explanation is **omitted variable bias** in the descriptive model. The simple comparison between high-SES and low-SES groups fails to account for a host of confounding factors that are correlated with both socioeconomic status and gender roles. Our regression model controls for several of these:

1.  **State-Level Differences**: India's states are incredibly diverse in terms of economic development, cultural norms, and female labor force participation. It is plausible that high-SES individuals are geographically concentrated in more conservative states where gender norms are more traditional, irrespective of wealth. By including state fixed effects, we are essentially comparing individuals *within* the same state, which purges the estimate of this major source of confounding.

2.  **Age and Life-Cycle Effects**: The age structure of high-SES and low-SES populations may differ. If high-SES groups are, on average, older and at a stage of life with more entrenched gender roles, a simple comparison could mistakenly attribute this age effect to wealth. Our quadratic age controls help to account for this.

3.  **Composition of the Baseline**: The descriptive analysis compares only the two extremes (High vs. Low SES), discarding the large "Middle SES" group. The regression analysis, by contrast, compares High SES to a baseline of *everyone else* (Low and Middle SES). The fact that the result reverses suggests that the relationship is not a simple linear one, and that the exclusion of the middle group in the descriptive tables may have presented a misleading picture.

## Theoretical Implications Revisited

The reversal of the "wealth paradox" significantly alters the paper's theoretical implications. The initial narrative, supporting "doing gender" over resource bargaining, is no longer tenable. Instead, our findings suggest:

-   **No Evidence of a Wealth Penalty**: We find no support for the hypothesis that wealth *exacerbates* gender inequality. On the contrary, once controls are in place, the gender gap is slightly smaller in high-SES households. While the effect is small, it suggests that economic development is not actively harmful to household gender equality.

-   **Resource Bargaining Remains Relevant, but Weak**: The negative coefficient on the interaction term is consistent with resource bargaining theory, which predicts that resources should narrow the gap. However, the very small magnitude of the effect suggests that the power of resource bargaining is heavily constrained by other factors, likely the very sticky gender norms that "doing gender" theories highlight.

-   **The Primacy of Context**: The strong influence of state fixed effects (though not explicitly shown) underscores that geography and local culture are likely far more important drivers of household gender dynamics than individual socioeconomic status.

## Policy Implications

The primary policy implication is one of caution. Policies based on the simple assumption that economic growth will either automatically fix or automatically worsen gender inequality are likely misguided. Our findings suggest a more nuanced reality:

-   **Economic Growth is Not a Silver Bullet**: The small effect size in our regression model implies that simply increasing incomes or education levels will not, on its own, lead to a dramatic reduction in the domestic gender gap. Gender inequality is a deep-rooted structural problem that will not be solved by economic growth alone.

-   **Focus on Norms and Institutions**: Given the likely importance of state-level context, policies aimed at shifting gender norms through education, media campaigns, and legal reforms may be more effective than purely economic interventions.

-   **Avoid Simple Narratives**: Policymakers should be wary of simplistic narratives like the "wealth paradox." Our analysis shows that the reality is more complex and that effective policy must be based on robust, well-controlled evidence, not on potentially misleading descriptive statistics.

## Limitations

1.  **SES Proxies**: Our use of education and urban residence is a crude proxy for wealth. Future research with direct income or consumption data is needed to verify these findings.

2.  **Non-Matched Individuals**: Our analysis compares averages of men and women, not matched spousal pairs, which limits our ability to make claims about within-household dynamics.

3.  **Causality**: Our cross-sectional data can only establish correlations, not causation. We cannot rule out the possibility of selection effects, where couples with more egalitarian views are also more likely to achieve higher SES.

4.  **Missing Mechanism Data**: We lack data on the use of domestic help, which is a key potential mechanism through which wealth might affect domestic work patterns.

# Conclusion

This paper began by investigating an apparent "wealth paradox" in Indian households, where descriptive data suggested that gender gaps in domestic work were larger among the affluent. However, our analysis demonstrates that this paradox is a statistical artifact that disappears when crucial confounding variables, particularly state-level heterogeneity, are taken into account. Our regression analysis reveals the opposite: the gender gap is slightly, but statistically significantly, *smaller* in high-SES households.

The key takeaway is methodological. It serves as a powerful reminder that in social science, simple bivariate comparisons can be dangerously misleading. The story of wealth and gender inequality in India is not a simple paradox, but a complex interplay of economic, cultural, and geographic factors that can only be untangled with careful statistical controls.

The substantive conclusion is that while economic development, as proxied by education and urban living, is no panacea for gender inequality, there is no evidence that it makes it worse. The path to a more egalitarian division of domestic labor in India will require more than just economic growth; it will necessitate direct challenges to deeply entrenched cultural norms and institutional structures.

# References

Bittman, Michael, Paula England, Liana Sayer, Nancy Folbre, and George Matheson. 2003. "When Does Gender Trump Money?" *American Journal of Sociology* 109(1): 186-214.

Gupta, Sanjiv. 2007. "Autonomy, Dependence, or Display?" *Journal of Marriage and Family* 69(2): 399-417.

Killewald, Alexandra. 2016. "Money, Work, and Marital Stability." *American Sociological Review* 81(4): 696-719.

Lundberg, Shelly, and Robert A. Pollak. 1996. "Bargaining and Distribution in Marriage." *Journal of Economic Perspectives* 10(4): 139-158.

Manser, Marilyn, and Murray Brown. 1980. "Marriage and Household Decision-Making." *International Economic Review* 21(1): 31-44.

McElroy, Marjorie B., and Mary Jean Horney. 1981. "Nash-Bargained Household Decisions." *International Economic Review* 22(2): 333-349.

Ray, Raka, and Seemin Qayum. 2009. *Cultures of Servitude*. Stanford: Stanford University Press.

West, Candace, and Don H. Zimmerman. 1987. "Doing Gender." *Gender & Society* 1(2): 125-151.
