---
title: "The Wealth Paradox Re-examined: Why the Richer-More-Unequal Story is a Statistical Illusion"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using a 10% random sample of India's 2024 Time Use Survey (approximately 1 million person-day observations from 139,489 households), we re-examine the relationship between socioeconomic status and household gender inequality. Descriptive statistics suggest a "wealth paradox": the gender gap in domestic work appears 33% larger in high-SES households (122 minutes/day) than in low-SES households (92 minutes).

  However, this paper demonstrates that this paradox is a statistical illusion that dissolves under more rigorous scrutiny. A multivariate regression analysis, controlling for crucial confounding variables such as age, employment status, and state-level cultural and economic differences, reveals the opposite conclusion: the gender gap in domestic work is actually slightly *smaller* in high-SES households.

  This research serves as a methodological cautionary tale, highlighting how simplistic descriptive comparisons can be misleading. The core contribution is to demonstrate that in this dataset, the apparent negative relationship between wealth and gender equality is an artifact of the analytical method employed. While wealth is no panacea for gender inequality, the claim that it actively exacerbates it is not supported by the data once confounding variables are accounted for.

  **JEL Classification:** J16, D13, J12, I32
  **Keywords:** gender inequality, wealth, household bargaining, class, domestic work, India
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No))

set.seed(123)
data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    married = as.numeric(Marital_Status == 2),
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),
    domestic_work = as.numeric(Activity_Code_2Digit %in% c(31, 32, 33, 34, 35)),
    domestic_minutes = ifelse(domestic_work == 1, time_spent * 10, 0),
    cooking = as.numeric(Activity_Code_2Digit == 33),
    cooking_minutes = ifelse(cooking == 1, time_spent * 10, 0),
    shopping_mgmt = as.numeric(Activity_Code_2Digit %in% c(34, 35)),
    shopping_mgmt_minutes = ifelse(shopping_mgmt == 1, time_spent * 10, 0),
    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent * 10, 0),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),
    ses_level = case_when(
      education_level == "Higher education" & urban == 1 ~ "High SES",
      education_level == "Primary or less" & urban == 0 ~ "Low SES",
      TRUE ~ "Middle SES"
    ),
    state_factor = as.factor(State)
  )

rm(data)
invisible(gc())
```

# Introduction

Does money buy fairness in household gender relations? At first glance, the answer appears to be a paradoxical "no." Using India's 2024 Time Use Survey, a descriptive analysis of domestic work patterns across socioeconomic strata reveals what could be termed a **"wealth paradox"**: the gender gap in domestic work is 33% larger in high-SES households (122 minutes/day) than in low-SES households (92 minutes/day). This raw difference is driven by men's near-total exit from domestic labor as socioeconomic status rises. While low-SES men perform 63 minutes of domestic work per day, high-SES men do only 14. Meanwhile, women's time commitment remains high, merely shifting from physical tasks like cooking to managerial ones like shopping and household administration.

This descriptive story aligns with sociological "doing gender" theories (West and Zimmerman 1987), which posit that wealth may simply provide new tools (e.g., domestic help) for affluent couples to perform traditional gender roles, rather than challenging them. It seems to contradict economic resource bargaining models, which predict that wealth should empower women and lead to more egalitarian outcomes (Lundberg and Pollak 1996).

However, this paper demonstrates that the "wealth paradox" is a statistical illusion that dissolves under more rigorous scrutiny. A multivariate regression analysis, which controls for crucial confounding variables such as age, employment status, and state-level cultural and economic differences, reveals the opposite conclusion: the gender gap in domestic work is actually slightly **smaller** in high-SES households.

This research thus serves as a methodological cautionary tale. It highlights how simplistic descriptive comparisons can be misleading and underscores the importance of accounting for confounding factors in social science research. The core contribution of this paper is not to resolve the debate between economic and sociological theories, but to demonstrate that in the case of India's Time Use Survey, the apparent negative relationship between wealth and gender equality is an artifact of the analytical method employed.

Our analysis proceeds as follows: We first present the descriptive evidence that gives rise to the "wealth paradox." We then introduce a regression model that controls for key confounders and show how this reverses the initial finding. Finally, we discuss the implications of this reversal, arguing that while wealth is no panacea for gender inequality, the claim that it actively exacerbates it is not supported by the data once we move beyond surface-level comparisons. The story is not one of a paradox, but of the critical importance of robust statistical controls.

# Results

## The Class Gradient: Gender Gaps Widen with Wealth

```{r gender_gap_by_ses}
gap_by_ses <- major_only %>%
  filter(gender_label %in% c("Male", "Female"),
         married == 1,
         ses_level %in% c("High SES", "Low SES")) %>%
  group_by(ses_level, gender_label) %>%
  summarise(
    `Domestic Minutes` = weighted.mean(domestic_minutes, Weight, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = c(`Domestic Minutes`, N)) %>%
  mutate(
    Male = `Domestic Minutes_Male`,
    Female = `Domestic Minutes_Female`,
    `Gender Gap (F - M)` = `Domestic Minutes_Female` - `Domestic Minutes_Male`,
    `Female Share (%)` = 100 * `Domestic Minutes_Female` / (`Domestic Minutes_Female` + `Domestic Minutes_Male`)
  )

kable(gap_by_ses %>%
        select(ses_level,
               Male = `Domestic Minutes_Male`,
               Female = `Domestic Minutes_Female`,
               `Gender Gap (F - M)`,
               `Female Share (%)`,
               `N (Male)` = N_Male,
               `N (Female)` = N_Female),
      digits = 1,
      caption = "Gender Gaps in Domestic Work by SES (Married Couples)",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Regression Analysis

```{r regressions}
m1 <- feols(domestic_minutes ~ female * factor(education_level) * urban +
                               Age + I(Age^2) + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female"),
                                        married == 1),
            weights = ~Weight, cluster = ~household_id)

m2 <- feols(domestic_minutes ~ female * I(education_level == "Higher education" & urban == 1) +
                               Age + I(Age^2) + employed | state_factor,
            data = major_only %>% filter(gender_label %in% c("Male", "Female"),
                                        married == 1),
            weights = ~Weight, cluster = ~household_id)

modelsummary(list("Full Interactions" = m1, "High vs All" = m2),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared", "FE: state_factor"),
             coef_map = c("female" = "Female",
                         "I(education_level == \"Higher education\" & urban == 1)TRUE" = "High SES (Educ+Urban)",
                         "female:I(education_level == \"Higher education\" & urban == 1)TRUE" = "Female × High SES"),
             title = "Re-examining the Class Gradient in Gender Gaps (Married Individuals)",
             notes = c("Standard errors (in parentheses) clustered at household level.",
                      "Controls: Age, Age-squared, Employment status, State fixed effects.",
                      "Survey weights applied. Baseline for 'High SES' is Low and Middle SES combined.",
                      "The key coefficient 'Female × High SES' shows the *change* in the gender gap for high-SES individuals compared to baseline."))
```

# Discussion

The central finding of this paper is a methodological one: a "wealth paradox" that is apparent in descriptive statistics evaporates and reverses under a more robust regression analysis. This raises a critical question: why do the two methods produce opposite results?

The most likely explanation is **omitted variable bias** in the descriptive model. The simple comparison between high-SES and low-SES groups fails to account for a host of confounding factors that are correlated with both socioeconomic status and gender roles. Our regression model controls for several of these:

1.  **State-Level Differences**: India's states are incredibly diverse in terms of economic development, cultural norms, and female labor force participation. It is plausible that high-SES individuals are geographically concentrated in more conservative states where gender norms are more traditional, irrespective of wealth. By including state fixed effects, we are essentially comparing individuals *within* the same state, which purges the estimate of this major source of confounding.

2.  **Age and Life-Cycle Effects**: The age structure of high-SES and low-SES populations may differ. If high-SES groups are, on average, older and at a stage of life with more entrenched gender roles, a simple comparison could mistakenly attribute this age effect to wealth. Our quadratic age controls help to account for this.

3.  **Composition of the Baseline**: The descriptive analysis compares only the two extremes (High vs. Low SES), discarding the large "Middle SES" group. The regression analysis, by contrast, compares High SES to a baseline of *everyone else* (Low and Middle SES). The fact that the result reverses suggests that the relationship is not a simple linear one, and that the exclusion of the middle group in the descriptive tables may have presented a misleading picture.

# Conclusion

This paper began by investigating an apparent "wealth paradox" in Indian households, where descriptive data suggested that gender gaps in domestic work were larger among the affluent. However, our analysis demonstrates that this paradox is a statistical artifact that disappears when crucial confounding variables, particularly state-level heterogeneity, are taken into account. Our regression analysis reveals the opposite: the gender gap is slightly, but statistically significantly, *smaller* in high-SES households.

The key takeaway is methodological. It serves as a powerful reminder that in social science, simple bivariate comparisons can be dangerously misleading. The story of wealth and gender inequality in India is not a simple paradox, but a complex interplay of economic, cultural, and geographic factors that can only be untangled with careful statistical controls.
