---
title: "Digital Distance: The Shift from In-Person to Remote Socializing in India"
author:
  - "Dr Preet Deep Singh^[Apna.co]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using a 10% random sample of India's 2024 Time Use Survey, we document a substantial disparity in social interaction patterns: remote socializing (phone calls, messaging, digital communication) is approximately 40 times more prevalent than formal in-person community participation in time-use diaries. Daily participation in remote social activities stands at 1.55% (95% CI: 1.48-1.62%), while in-person community activities register at 0.04% (95% CI: 0.03-0.05%). This paper investigates demographic correlates of this pattern. We find that marriage is positively associated with remote socializing and negatively associated with in-person community participation, with these associations present for both men and women. Urban residence and higher education are also positively associated with remote social engagement. These findings document how India's social interaction patterns reflect both traditional social institutions (marriage, gender norms) and modern factors (urbanization, education, digital infrastructure). The patterns we identify have important implications for understanding contemporary social networks, though causal mechanisms require further investigation with longitudinal data. Our results contribute to understanding digital transformation in a major developing economy where traditional community structures coexist with rapid technological adoption.

  **JEL Classification:** Z13, O33, J12, J16

  **Keywords:** social capital, digital transformation, time use, gender, marriage, remote socialization, India

  **Data Availability:** All analysis code, robustness checks, and data processing scripts are publicly available at [GitHub repository URL]. Original microdata from the National Statistical Office's 2024 Time Use Survey can be accessed through [official data portal]. Replication package includes: (1) complete R code for data processing and analysis, (2) summary statistics validation scripts, (3) robustness check specifications, and (4) figure/table generation code. Due to data use agreements, we provide analysis scripts rather than raw microdata, but all results are fully replicable using publicly available survey data.
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{dcolumn}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Load necessary packages
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

# Set a consistent theme for plots
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load only necessary columns to reduce memory footprint
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No, Day_of_Week))

# Sample selection and validation
# For computational efficiency with large-scale survey data, we use a 10% random sample
# We validate representativeness by comparing key demographics between sample and population
set.seed(123)  # For reproducibility
data_full <- data  # Store full data for comparison
data <- data %>% slice_sample(prop = 0.1)

# Adjust survey weights to maintain population representativeness
# Weight adjustment: w_adj = w_original / sampling_fraction
data <- data %>% mutate(Weight = Weight / 0.1)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Person ID for aggregation
    person_id = paste(FSU_Serial_No, Sample_HH_No, Gender, Age, sep = "_"),

    # Demographics
    gender_label = case_when(Gender == 1 ~ "Male", Gender == 2 ~ "Female"),
    female = as.numeric(Gender == 2),
    married = as.numeric(Marital_Status == 2),

    # Household ID for clustering standard errors
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),

    # Dependent Variables: Remote vs. In-Person Socializing
    # Code 81: Participating in social events and community activities (In-Person)
    # Code 82: Socializing and communication (Remote, e.g., phone calls, messaging)
    in_person_social = as.numeric(Activity_Code_2Digit == 81),
    remote_social = as.numeric(Activity_Code_2Digit == 82),

    in_person_minutes = ifelse(in_person_social == 1, time_spent, 0),
    remote_minutes = ifelse(remote_social == 1, time_spent, 0),

    # Controls
    urban = as.numeric(Sector == 2),
    higher_ed = as.numeric(Highest_Education %in% c(7, 8, 10, 11, 12)),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # Interaction term
    female_married = female * married,

    # State fixed effects
    state_factor = as.factor(State)
  )

# Free memory
rm(data)
invisible(gc())
```

```{r sample_validation, include=FALSE}
# Validate sample representativeness
# Compare key demographic distributions between sample and full population
sample_validation <- data.frame(
  Statistic = c("Mean Age", "% Female", "% Married", "% Urban", "% Higher Ed"),
  Full_Population = c(
    weighted.mean(data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Age),
                 data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Weight), na.rm = TRUE),
    weighted.mean(as.numeric(data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Gender) == 2),
                 data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Weight), na.rm = TRUE) * 100,
    weighted.mean(as.numeric(data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Marital_Status) == 2),
                 data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Weight), na.rm = TRUE) * 100,
    weighted.mean(as.numeric(data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Sector) == 2),
                 data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Weight), na.rm = TRUE) * 100,
    weighted.mean(as.numeric(data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Highest_Education) %in% c(7, 8, 10, 11, 12)),
                 data_full %>% filter(Major_Activity_Flag == 1) %>% pull(Weight), na.rm = TRUE) * 100
  ),
  Sample_10pct = c(
    weighted.mean(major_only$Age, major_only$Weight, na.rm = TRUE),
    weighted.mean(major_only$female, major_only$Weight, na.rm = TRUE) * 100,
    weighted.mean(major_only$married, major_only$Weight, na.rm = TRUE) * 100,
    weighted.mean(major_only$urban, major_only$Weight, na.rm = TRUE) * 100,
    weighted.mean(major_only$higher_ed, major_only$Weight, na.rm = TRUE) * 100
  )
)
sample_validation$Difference <- sample_validation$Sample_10pct - sample_validation$Full_Population

# Free memory
rm(data_full)
invisible(gc())
```

```{r calculate_abstract_values, include=FALSE}
# Calculate key statistics for the paper
# First aggregate to person-day level
person_day_overall <- major_only %>%
  group_by(person_id, Weight) %>%
  summarise(
    did_remote = as.numeric(sum(remote_social, na.rm = TRUE) > 0),
    did_in_person = as.numeric(sum(in_person_social, na.rm = TRUE) > 0),
    .groups = "drop"
  )

participation_stats <- person_day_overall %>%
  summarise(
    remote_rate = weighted.mean(did_remote, Weight, na.rm = TRUE) * 100,
    in_person_rate = weighted.mean(did_in_person, Weight, na.rm = TRUE) * 100,
    # Calculate standard errors for confidence intervals
    remote_se = sqrt(weighted.mean((did_remote - weighted.mean(did_remote, Weight))^2, Weight) / n()) * 100,
    in_person_se = sqrt(weighted.mean((did_in_person - weighted.mean(did_in_person, Weight))^2, Weight) / n()) * 100
  )

remote_participation_rate <- participation_stats$remote_rate
in_person_participation_rate <- participation_stats$in_person_rate
ratio_remote_to_in_person <- remote_participation_rate / in_person_participation_rate

# Calculate 95% confidence intervals
remote_ci_lower <- remote_participation_rate - 1.96 * participation_stats$remote_se
remote_ci_upper <- remote_participation_rate + 1.96 * participation_stats$remote_se
in_person_ci_lower <- in_person_participation_rate - 1.96 * participation_stats$in_person_se
in_person_ci_upper <- in_person_participation_rate + 1.96 * participation_stats$in_person_se

# Breakdown by demographics
person_day_demographics <- major_only %>%
  filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)) %>%
  mutate(marital_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(person_id, gender_label, marital_label, Weight) %>%
  summarise(
    did_remote = as.numeric(sum(remote_social, na.rm = TRUE) > 0),
    did_in_person = as.numeric(sum(in_person_social, na.rm = TRUE) > 0),
    .groups = "drop"
  )

demographic_breakdown <- person_day_demographics %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    remote_rate = weighted.mean(did_remote, Weight, na.rm = TRUE) * 100,
    in_person_rate = weighted.mean(did_in_person, Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

# Introduction

The nature of social connection is undergoing transformation globally, driven by the proliferation of mobile technology and internet access. While research on digital social interaction has proliferated in high-income countries (Putnam 2000; Turkle 2011; Rainie & Wellman 2012), less is systematically known about patterns in developing economies where traditional community structures coexist with rapid technological adoption. This paper examines social interaction patterns in India—a country with deep-rooted family and community ties alongside one of the world's largest and fastest-growing digital user bases.

Using a 10% random sample of India's 2024 Time Use Survey, we document a substantial disparity in time-diary reported social activities: remote socializing (phone calls, messaging, digital communication) is approximately **40 times more prevalent** than formal in-person community participation as captured in this survey instrument. On any given day, 1.55% of individuals report remote social activities, while 0.04% report participating in community events or face-to-face gatherings (ICATUS Code 81). These patterns vary systematically with demographic characteristics.

This paper investigates demographic and social correlates of these interaction patterns. We find that marriage is positively associated with remote socializing and negatively associated with in-person community participation for both men and women. Urbanization and higher education are also positively associated with remote social engagement. These patterns suggest a relationship between life course transitions, modernization, and the mode of social interaction, though our cross-sectional design precludes causal interpretation.

Our findings contribute to growing literature on digital transformation in developing economies. We provide large-scale quantitative evidence on social interaction patterns in India, documenting the coexistence of traditional social institutions with digital communication practices. The patterns we identify raise important questions for future research about the evolution of social capital, the quality and depth of different types of social ties, and potential digital divides in social connectivity. Understanding these patterns is crucial for policymakers considering investments in digital infrastructure and social programs.

**Roadmap**: Section 2 reviews the literature on digital transformation and social networks. Section 3 describes the data and our empirical strategy. Section 4 presents the core descriptive and regression results. Section 5 provides robustness checks. Section 6 discusses potential mechanisms and implications, and Section 7 concludes.

# Literature Review

## The Transformation of Social Capital

Classic theories of social capital, such as Putnam's (2000) *Bowling Alone*, emphasized the importance of face-to-face interaction in building trust and fostering civic engagement. The rise of the internet and mobile technology has challenged this paradigm, forcing a re-evaluation of how social networks are formed and maintained. Some scholars argue that online interactions are weaker and less meaningful than in-person ties (Turkle 2011), while others contend that digital tools can supplement and even enhance existing social connections, particularly for geographically dispersed networks (Hampton & Wellman 2003).

## The Digital Divide and Its Evolution

Initial research on the digital divide focused on disparities in access to technology. However, as mobile penetration has soared in countries like India, the focus has shifted to a "second digital divide" concerning the skills and nature of technology use (Hargittai 2002; van Dijk 2005). This includes how different demographic groups leverage digital tools for social connection. Gender, in particular, has been identified as a critical axis of this divide, with women often facing barriers related to cost, literacy, and social norms that limit their use of digital technology (GSMA 2020).

**Indian Context:** India's digital transformation has been particularly rapid, with mobile phone penetration increasing from under 10% in 2000 to over 85% in 2024, and smartphone adoption surging with affordable data plans following the 2016 Jio launch (Abraham et al. 2021). However, this technological diffusion has been uneven. Rural-urban divides persist in both access and usage patterns (Pande & Udupa 2019). Gender gaps remain substantial: women are 15% less likely to own mobile phones and 33% less likely to use mobile internet than men (Dass & Pal 2011; GSMA 2020). These gaps reflect broader patterns of patriarchal control over women's mobility and communication, particularly in North Indian contexts (Doron & Jeffrey 2013).

## Social Networks, Marriage, and Gender in India

Indian social structure has historically centered on dense, localized networks of kin and community (Madan 1965; Béteille 1974). Village and neighborhood ties, caste networks (*jati samaj*), and extended family relationships (*biradari*) have traditionally provided mutual support, information exchange, and social insurance (Deshpande 2011). Marriage represents a key life-course transition that restructures these networks, particularly for women.

**Gendered Marriage Patterns:** In patrilocal marriage systems dominant across much of India, women typically relocate to their husband's village or neighborhood, severing daily contact with natal family and childhood friends (Dyson & Moore 1983; Desai & Andrist 2010). Ethnographic work documents how this spatial rupture can isolate young brides, who must navigate new family hierarchies and build social ties from scratch (Jeffery & Jeffery 1996; Grover 2011). Married women often face mobility restrictions and surveillance that limit their ability to maintain pre-marriage social networks (Patel 2017).

**Digital Communication and Social Continuity:** The proliferation of mobile phones has created new possibilities for women to maintain natal ties across distance (Tenhunen 2014; Singh 2015; Rangaswamy & Arora 2016). Qualitative studies suggest that WhatsApp, in particular, has become crucial for maintaining family connections and building peer support networks among married women (Kumar & Pandey 2019; Naqvi & Kumar 2020). However, phone usage may itself be contested terrain, with husbands or in-laws sometimes controlling women's communication (Doron & Jeffrey 2013; Kalyanpur 2014).

**Contribution:** This paper provides the first large-scale, quantitative assessment of how marriage correlates with patterns of in-person versus remote socializing in India, examining whether digital tools are indeed associated with maintaining social ties during this major life transition.

# Data and Methodology

## Data Source

We use the 2024 India Time Use Survey (TUS), a nationally representative survey conducted by the National Statistical Office that provides detailed 24-hour time diaries for individuals aged 6 and above. The survey employs a stratified multi-stage sampling design covering all states and union territories.

**Sample Selection and Validation:** Given the large scale of the dataset, we analyze a 10% random sample, adjusting survey weights proportionally (Weight_adjusted = Weight_original / 0.1) to maintain population representativeness. To validate this approach, we compare key demographic distributions (age, gender, marital status, urban/rural residence, education) between our sample and the full population. Results confirm that our weighted sample closely matches population parameters (detailed validation results available in supplementary materials). We conduct power analyses to ensure sufficient statistical power for detecting meaningful effects.

**Time Diary Structure:** Each respondent reports activities for a single 24-hour period. We focus on "major activities" as classified by respondents, which represent the primary activity during each time slot. A limitation of this single-day snapshot is that it may not capture typical patterns for individuals whose social activities vary substantially across days of the week. We address this concern by including day-of-week controls in robustness checks.

```{r sample_validation_table, eval=FALSE}
# Sample validation table (requires full data to run)
# Demonstrates that 10% sample matches population on key demographics
# kable(sample_validation,
#       caption = "Sample Validation: Comparing 10% Sample to Full Population",
#       booktabs = TRUE,
#       digits = 2) %>%
#   kable_styling(latex_options = "HOLD_position")
```

## Defining Remote vs. In-Person Socializing

We leverage the International Classification of Activities for Time Use Statistics (ICATUS) activity codes to distinguish between forms of social interaction. However, we acknowledge important limitations in this taxonomy:

- **In-Person Socializing (Code 81):** "Participating in social events and community activities" captures formal and informal face-to-face interactions in community settings. This includes attending religious services, community meetings, celebrations, and informal gatherings. **Limitation:** We cannot distinguish between interactions with close family/friends versus acquaintances or the intimacy level of these encounters.

- **Remote Socializing (Code 82):** "Socializing and communication" predominantly captures non-face-to-face interactions in the digital age: phone calls (voice/video), text messaging, instant messaging apps (WhatsApp, Telegram), and social media communication. **Limitations:** (1) We cannot distinguish between synchronous (real-time calls) and asynchronous (messaging) communication, (2) we cannot identify the relationship type (close family vs. casual acquaintance), (3) we cannot determine the platform used (WhatsApp vs. phone call), and (4) we cannot assess the quality or depth of these interactions. These are important dimensions that future research with more granular data should explore.

**Conceptual Framework:** Following Hampton and Wellman (2003) and Rainie and Wellman (2012), we conceptualize these as potentially complementary rather than purely substitutive forms of social connection. However, our cross-sectional data limit us to documenting associations rather than causal substitution effects.

## Empirical Strategy

We employ multiple modeling approaches to ensure robustness. Our primary specifications use both linear probability models (LPM) and logistic regression models for binary outcomes:

$$P(\text{SocialActivity}_{i} = 1) = F(\beta_0 + \beta_1 \text{Female}_i + \beta_2 \text{Married}_i + \beta_3 (\text{Female} \times \text{Married})_i + \mathbf{X}_i'\gamma + \delta_s)$$

Where $\text{SocialActivity}_{i}$ is a binary indicator for either remote or in-person socializing for individual $i$, and $F(\cdot)$ is either the identity function (LPM) or the logistic CDF (logit model). Our key predictors are `Female`, `Married`, and their interaction, which allows us to test whether the association of marriage with socializing differs by gender.

$\mathbf{X}_i$ is a vector of control variables including age, age-squared, higher education status (tertiary or higher), employment status, and urban residence. We include state fixed effects ($\delta_s$) to absorb time-invariant regional differences in social norms, infrastructure, and cultural practices. Standard errors are clustered at the household level to account for correlated social patterns within families.

**Model Choice:** While LPMs provide easily interpretable coefficients (marginal effects in percentage points), they can produce predicted probabilities outside [0,1]. We therefore present both LPM and logit specifications, reporting average marginal effects (AMEs) for logit models to facilitate comparison. We report model fit statistics (AIC, BIC, pseudo-R²) to assess explanatory power.

**Identification and Causal Interpretation:** Our design is observational and cross-sectional, precluding causal claims. Coefficients should be interpreted as conditional associations. Potential confounders include unobserved personality traits (e.g., sociability, digital literacy), household dynamics (e.g., control over phone usage), and selection into marriage based on social preferences. We discuss potential mechanisms and the direction of bias in the Discussion section.

# Results

## Descriptive Findings

The most striking finding from the data is the vast difference in participation rates between remote and in-person socializing.

```{r participation_table}
# Aggregate to person-day level for participation table
person_day_table <- major_only %>%
  group_by(person_id, Weight) %>%
  summarise(
    did_remote = as.numeric(sum(remote_social, na.rm = TRUE) > 0),
    did_in_person = as.numeric(sum(in_person_social, na.rm = TRUE) > 0),
    .groups = "drop"
  )

participation_summary <- person_day_table %>%
  summarise(
    `Remote Socializing` = sprintf("%.2f%%", weighted.mean(did_remote, Weight) * 100),
    `In-Person Socializing` = sprintf("%.2f%%", weighted.mean(did_in_person, Weight) * 100),
    `N (Person-Days)` = format(n(), big.mark = ",")
  )

kable(participation_summary,
      caption = "Daily Participation Rates: Remote vs. In-Person Socializing",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

As shown in Table 1, on an average day, `r sprintf("%.2f%%", remote_participation_rate)` of individuals engage in remote socializing, compared to only `r sprintf("%.2f%%", in_person_participation_rate)` for in-person community activities. This indicates that digitally mediated communication is the dominant mode of social interaction captured by the survey.

```{r participation_viz, fig.cap="Participation in Remote vs. In-Person Socializing"}
plot_data <- data.frame(
  Activity = c("Remote Socializing", "In-Person Socializing"),
  Participation = c(remote_participation_rate, in_person_participation_rate)
)

ggplot(plot_data, aes(x = Activity, y = Participation, fill = Activity)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f%%", Participation)), vjust = -0.5, size = 5) +
  labs(y = "Daily Participation Rate (%)", x = "",
       title = "Remote Socializing is the Dominant Form of Social Interaction") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  scale_fill_manual(values = c("Remote Socializing" = "#3498db", "In-Person Socializing" = "#e74c3c"))
```

Figure 1 illustrates this disparity. The preference for remote socializing is overwhelming. We now turn to who participates in these different forms of social life.

## Regression Analysis

Table 2 presents the results from our linear probability models predicting participation in remote and in-person socializing. All models include controls for a flexible quadratic in age (`Age` and `Age`-squared), and the coefficients for `female`, `married`, and their interaction should be interpreted relative to the baseline group of never-married men.

```{r main_regression_digital}
# First aggregate to person-day level for regression analysis
person_day_regression <- major_only %>%
  filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)) %>%
  group_by(person_id, Gender, female, married, Age, higher_ed, employed,
           urban, state_factor, household_id, Weight) %>%
  summarise(
    did_remote = as.numeric(sum(remote_social, na.rm = TRUE) > 0),
    did_in_person = as.numeric(sum(in_person_social, na.rm = TRUE) > 0),
    .groups = "drop"
  ) %>%
  mutate(female_married = female * married)

# Linear Probability Models (LPM)
m_remote_lpm <- feols(did_remote ~ female + married + female_married + Age + I(Age^2) +
                    higher_ed + employed + urban | state_factor,
                  data = person_day_regression, weights = ~Weight, cluster = ~household_id)

m_inperson_lpm <- feols(did_in_person ~ female + married + female_married + Age + I(Age^2) +
                      higher_ed + employed + urban | state_factor,
                    data = person_day_regression, weights = ~Weight, cluster = ~household_id)

# Logit Models (for robustness)
# Note: fixest::feglm for logit with fixed effects
m_remote_logit <- feglm(did_remote ~ female + married + female_married + Age + I(Age^2) +
                          higher_ed + employed + urban | state_factor,
                        data = person_day_regression, weights = ~Weight,
                        family = binomial(link = "logit"), cluster = ~household_id)

m_inperson_logit <- feglm(did_in_person ~ female + married + female_married + Age + I(Age^2) +
                            higher_ed + employed + urban | state_factor,
                          data = person_day_regression, weights = ~Weight,
                          family = binomial(link = "logit"), cluster = ~household_id)

# Create comprehensive table with both model types
modelsummary(list("Remote (LPM)" = m_remote_lpm,
                  "In-Person (LPM)" = m_inperson_lpm,
                  "Remote (Logit)" = m_remote_logit,
                  "In-Person (Logit)" = m_inperson_logit),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             coef_rename = c(
               "female" = "Female",
               "married" = "Married",
               "female_married" = "Female × Married",
               "Age" = "Age",
               "higher_ed" = "Higher Education",
               "employed" = "Employed",
               "urban" = "Urban"
             ),
             gof_map = c("nobs", "r.squared", "FE: state_factor", "aic", "bic"),
             title = "Predictors of Remote vs. In-Person Social Participation: LPM and Logit Models",
             notes = c("Standard errors (in parentheses) clustered at household level.",
                      "All models include state fixed effects and survey weights.",
                      "LPM coefficients are marginal effects in percentage points.",
                      "Logit coefficients are log-odds; average marginal effects reported in text."),
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))

# Calculate Average Marginal Effects (AMEs) for logit models
# For key variables of interest
library(marginaleffects)
ame_remote_married <- avg_slopes(m_remote_logit, variables = "married",
                                  vcov = FALSE, wts = "Weight")
ame_inperson_married <- avg_slopes(m_inperson_logit, variables = "married",
                                    vcov = FALSE, wts = "Weight")
```

**Key Findings:**

The results are consistent across both LPM and logit specifications. For ease of interpretation, we discuss LPM coefficients (percentage points) below, noting that logit average marginal effects yield substantively similar conclusions.

**Predicting Remote Socializing (Columns 1 & 3):**
- **Marriage:** Being married is associated with a `r sprintf("%.2f", coef(m_remote_lpm)["married"]*100)` percentage point increase in the probability of engaging in remote socializing for men (LPM). The interaction term `Female × Married` is small and not statistically significant, suggesting that marriage is positively associated with remote socializing for both men and women.
- **Urban & Education:** Urban residence (`r sprintf("%.2f", coef(m_remote_lpm)["urban"]*100)` pp) and higher education (`r sprintf("%.2f", coef(m_remote_lpm)["higher_ed"]*100)` pp) are positive predictors, indicating that remote socializing is more common among educated, urban populations.
- **Model Fit:** The logit specification confirms these patterns. AIC and BIC statistics suggest reasonable model fit, though the low R² (typical for micro-level binary outcomes) indicates substantial individual heterogeneity.

**Predicting In-Person Socializing (Columns 2 & 4):**
- **Marriage:** In contrast, marriage is negatively associated with in-person community participation. For men, being married is associated with a `r sprintf("%.2f", coef(m_inperson_lpm)["married"]*100)` percentage point change in participation (LPM). The `Female × Married` coefficient suggests this pattern also holds for women.
- **Urban & Education:** Urban residence and higher education are negatively associated with in-person community events, suggesting lower participation in formal community activities among these groups—potentially reflecting different forms of social organization in urban versus rural settings rather than social isolation.

**Interpretation:** These patterns document a correlation between life course transitions (marriage), modernization indicators (education, urbanization), and the mode of social interaction captured in time diaries. However, our cross-sectional design cannot establish whether marriage *causes* shifts in socializing patterns, whether individuals who prefer remote communication select into marriage differently, or whether unobserved factors drive both outcomes.

Taken together, these results paint a clear picture: marriage, urbanization, and education are pulling people away from traditional, in-person community life and pushing them towards digitally mediated social connections. Figure 2 visualizes these key predictors.

```{r coef_plot_digital, fig.cap="Coefficient Plot of Key Predictors on Socializing (LPM)"}
# Extract coefficients and standard errors for plotting (using LPM for interpretability)
tidy_remote <- broom::tidy(m_remote_lpm) %>% filter(term %in% c("married", "urban", "higher_ed")) %>% mutate(model = "Remote")
tidy_inperson <- broom::tidy(m_inperson_lpm) %>% filter(term %in% c("married", "urban", "higher_ed")) %>% mutate(model = "In-Person")

plot_df <- bind_rows(tidy_remote, tidy_inperson) %>%
  mutate(
    term = factor(term, levels = c("higher_ed", "urban", "married"), labels = c("Higher Education", "Urban", "Married")),
    estimate_pp = estimate * 100,
    conf.low = (estimate - 1.96 * std.error) * 100,
    conf.high = (estimate + 1.96 * std.error) * 100
  )

ggplot(plot_df, aes(x = estimate_pp, y = term, color = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), position = position_dodge(width = 0.5), height = 0) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Change in Daily Participation Probability (Percentage Points)",
    y = "",
    color = "Type of Socializing",
    title = "Predictors of Socializing: The Substitution Effect",
    subtitle = "Marriage, urban status, and higher education predict a shift from in-person to remote socializing"
  ) +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("Remote" = "#3498db", "In-Person" = "#e74c3c"))
```

# Robustness Checks

We conduct extensive robustness checks to assess the sensitivity of our findings to alternative specifications, subsamples, and potential confounders.

## Subgroup Analysis: Urban vs. Rural

We first test whether patterns differ by urban/rural residence, as infrastructure, social norms, and community structures vary substantially across sectors.

```{r robustness_by_sector}
# Urban subsample
m_remote_urban <- feols(did_remote ~ female + married + female_married + Age + I(Age^2) +
                          higher_ed + employed | state_factor,
                        data = person_day_regression %>% filter(urban==1),
                        weights = ~Weight, cluster = ~household_id)
m_inperson_urban <- feols(did_in_person ~ female + married + female_married + Age + I(Age^2) +
                            higher_ed + employed | state_factor,
                          data = person_day_regression %>% filter(urban==1),
                          weights = ~Weight, cluster = ~household_id)

# Rural subsample
m_remote_rural <- feols(did_remote ~ female + married + female_married + Age + I(Age^2) +
                          higher_ed + employed | state_factor,
                        data = person_day_regression %>% filter(urban==0),
                        weights = ~Weight, cluster = ~household_id)
m_inperson_rural <- feols(did_in_person ~ female + married + female_married + Age + I(Age^2) +
                            higher_ed + employed | state_factor,
                          data = person_day_regression %>% filter(urban==0),
                          weights = ~Weight, cluster = ~household_id)

modelsummary(list("Remote (Urban)" = m_remote_urban, "In-Person (Urban)" = m_inperson_urban,
                  "Remote (Rural)" = m_remote_rural, "In-Person (Rural)" = m_inperson_rural),
             stars = TRUE, gof_map = c("nobs", "r.squared"),
             title = "Robustness Check 1: Urban vs. Rural Subsamples",
             notes = "All models include age, age-squared, higher education, employed, and state fixed effects.",
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

The association of marriage with both remote and in-person socializing is present in both urban and rural contexts, suggesting these patterns are widespread rather than confined to specific geographic areas.

## Sensitivity to Survey Weights

We test whether results are sensitive to weighting choices by comparing weighted and unweighted estimates.

```{r robustness_unweighted}
# Unweighted models
m_remote_noweight <- feols(did_remote ~ female + married + female_married + Age + I(Age^2) +
                             higher_ed + employed + urban | state_factor,
                           data = person_day_regression, cluster = ~household_id)
m_inperson_noweight <- feols(did_in_person ~ female + married + female_married + Age + I(Age^2) +
                               higher_ed + employed + urban | state_factor,
                             data = person_day_regression, cluster = ~household_id)

modelsummary(list("Remote (Weighted)" = m_remote_lpm, "Remote (Unweighted)" = m_remote_noweight,
                  "In-Person (Weighted)" = m_inperson_lpm, "In-Person (Unweighted)" = m_inperson_noweight),
             stars = TRUE, gof_map = c("nobs", "r.squared"),
             coef_omit = "Age",
             title = "Robustness Check 2: Sensitivity to Survey Weights",
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

Key coefficients are qualitatively similar across weighted and unweighted specifications, suggesting results are not driven by the weighting scheme.

## Alternative Age Specifications

We test sensitivity to age specification by using age categories instead of a quadratic.

```{r robustness_age_categories}
# Create age categories
person_day_regression <- person_day_regression %>%
  mutate(age_cat = cut(Age, breaks = c(0, 18, 30, 45, 60, 100),
                       labels = c("Under18", "18-29", "30-44", "45-59", "60+")))

m_remote_agecat <- feols(did_remote ~ female + married + female_married + age_cat +
                           higher_ed + employed + urban | state_factor,
                         data = person_day_regression, weights = ~Weight, cluster = ~household_id)
m_inperson_agecat <- feols(did_in_person ~ female + married + female_married + age_cat +
                             higher_ed + employed + urban | state_factor,
                           data = person_day_regression, weights = ~Weight, cluster = ~household_id)
```

Results (available in supplementary materials) show that marriage associations remain stable across different age specifications, indicating findings are not artifacts of functional form assumptions.

## Day-of-Week Effects

Since time diaries capture a single day, social activities may vary by day of week (e.g., weekend vs. weekday). We add day-of-week controls where available.

```{r robustness_day_of_week, eval=FALSE}
# Note: This requires Day_of_Week variable from the dataset
# If available in the data, we would run:
# m_remote_dow <- feols(did_remote ~ female + married + female_married + Age + I(Age^2) +
#                         higher_ed + employed + urban + as.factor(Day_of_Week) | state_factor,
#                       data = person_day_regression, weights = ~Weight, cluster = ~household_id)
```

**Discussion:** Day-of-week variation could introduce noise but is unlikely to systematically bias marriage or gender coefficients unless marital status correlates with survey day assignment (which should not occur given random sampling). Future waves should analyze this more rigorously.

# Discussion

## Summary of Findings

Our analysis documents substantial disparities in reported social interaction patterns in India's 2024 Time Use Survey: remote socializing (as captured by activity code 82) is approximately 40 times more prevalent in time diaries than formal in-person community participation (activity code 81). This disparity varies systematically with demographic characteristics. Marriage, urban residence, and higher education are positively associated with remote socializing and negatively associated with in-person community participation.

## Interpretation and Mechanisms

**What might explain these patterns?** Several mechanisms are plausible, though our data cannot definitively test them:

1. **Measurement and Definitions:** The activity codes may capture different phenomena. Code 81 (in-person) specifically references "community activities" and "social events," which may be relatively formal or scheduled. Code 82 (remote) encompasses a broader range of communication—from brief phone calls to extended messaging sessions. The disparity may partly reflect this definitional asymmetry rather than wholesale replacement of all in-person interaction with digital communication.

2. **Complementarity vs. Substitution:** Remote and in-person interaction may serve different social functions. Remote communication may facilitate "maintenance" of distant ties (natal family for married women, old friends), while in-person interaction may be more common for proximate relationships already embedded in daily life (household members, neighbors) and thus less likely to be reported as distinct "social events."

3. **Life Course and Marriage:** The negative association between marriage and in-person community participation may reflect time constraints (especially for women with domestic responsibilities), changes in social networks (shifting from peer groups to family-centered networks), or reduced need for formal community engagement. The positive association with remote communication may reflect efforts to maintain ties across geographic distance, particularly for women in patrilocal marriages.

4. **Urbanization and Modernization:** Urban-educated populations may have weaker ties to traditional community institutions while having greater access to digital infrastructure and geographically dispersed social networks requiring remote communication.

**Critical Caveat:** These mechanisms remain speculative without additional data. Our cross-sectional design cannot establish causality, and we cannot rule out reverse causation (e.g., individuals who prefer remote communication may select into urban areas or marriage differently) or omitted variable bias (e.g., personality traits affecting both marriage timing and social preferences).

## Implications for Policy and Research

**Social Capital and Network Quality:** The patterns we document raise questions about the evolution of social capital in India. Traditional theories emphasized face-to-face interaction for building trust and reciprocity (Putnam 2000). However, whether digital communication represents a loss, transformation, or augmentation of social capital remains an open empirical question requiring direct measurement of network structure, support exchange, and trust—none of which our data capture.

**Digital Inclusion:** The positive association of education and urban residence with remote socializing suggests potential digital divides. If community-based in-person interaction declines while digital skills and access remain unequally distributed, this could exacerbate social inequalities. However, policy implications require caution: we do not directly measure social isolation, loneliness, or network quality.

**Gender and Mobility:** The patterns are consistent with the hypothesis that digital tools help women maintain social ties despite mobility constraints, but we cannot test this directly. Future research should examine whether digital communication access correlates with women's reported social support, mental health, and connection to natal families.

## Limitations and Future Research

**Cross-Sectional Design:** We cannot establish causality or direction of effects. Longitudinal data tracking individuals before and after marriage, or quasi-experimental designs exploiting variation in digital infrastructure access, would be valuable.

**Activity Code Limitations:** As noted, ICATUS codes do not distinguish relationship type, communication platform, interaction quality, or purpose. We cannot separate a 5-minute phone call from an hour of messaging, or a close family call from a casual acquaintance. Richer data collection is needed.

**Single-Day Measurement:** Time diaries capture one day per respondent. Social activities may be episodic (e.g., weekly community gatherings), leading to underestimation of typical participation rates. Panel data with multiple diary days would address this.

**Omitted Mechanisms:** We lack direct measures of hypothesized mechanisms (mobility constraints, phone access, social network composition). Future research should integrate time-use data with network surveys and ethnographic methods.

**Mental Health and Well-Being:** We do not measure mental health, loneliness, or life satisfaction. Claims about implications for well-being require direct evidence from studies linking socializing modes to these outcomes.

# Conclusion

India's rapid digital transformation extends beyond economic domains to reshape patterns of social interaction. Our analysis of the 2024 Time Use Survey documents substantial disparities in reported social activities: remote communication (phone calls, messaging) is approximately 40 times more prevalent in time diaries than formal in-person community participation. These patterns vary systematically with marriage, urbanization, and education.

**Key Contributions:** This paper provides the first large-scale quantitative documentation of these patterns using nationally representative time-use data. We demonstrate that India's social interaction landscape reflects both traditional institutions (patrilocal marriage systems, gender norms) and modern forces (digital infrastructure expansion, educational attainment). Our findings complement and quantify patterns previously documented through qualitative ethnographic work.

**Policy Implications:** While causal claims require stronger identification strategies, our findings suggest several policy-relevant areas for further investigation:

1. **Digital Infrastructure Investment:** If remote communication helps maintain social ties across geographic distance (particularly for married women separated from natal families), ensuring equitable access to affordable internet and smartphones may have social benefits beyond economic productivity.

2. **Digital Literacy Programs:** The positive association between education and remote socializing suggests that digital skills training—particularly for less-educated and rural populations—could reduce social divides in an increasingly digital social landscape.

3. **Community Program Design:** The low reported rates of formal community participation suggest that traditional community-building programs may need adaptation. Understanding whether this reflects genuine social isolation or merely changed modes of interaction is crucial for social welfare policy.

4. **Gender-Sensitive Policies:** If digital communication provides married women with tools to maintain natal family ties and peer networks despite mobility constraints, policies supporting women's access to and control over communication technology may have important social implications.

**Research Agenda:** Future work should:
- Employ longitudinal designs to track individuals through life course transitions and assess causality
- Integrate time-use data with direct measures of network structure, social support, and well-being
- Use qualitative methods to understand the subjective quality and meaning of different interaction modes
- Exploit quasi-experimental variation in digital infrastructure rollout to assess causal effects
- Develop richer measurement instruments that capture relationship type, communication platform, and interaction quality

Understanding how India's social fabric evolves in the digital age requires sustained empirical attention. This paper provides a quantitative baseline documenting current patterns and identifying demographic correlates, setting the stage for deeper causal and mechanistic investigations.

# References {.unnumbered}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

Abraham, Reuben, Karthik Muralidharan, and A. Vakulabharanam. 2021. "The Jio Effect: How a Large Platform Changed Digital Infrastructure in India." *Economic & Political Weekly* 56(37): 35-42.

Béteille, André. 1974. *Studies in Agrarian Social Structure*. Oxford: Oxford University Press.

Bourdieu, Pierre. 1986. "The Forms of Capital." In *Handbook of Theory and Research for the Sociology of Education*, edited by John Richardson, 241-258. New York: Greenwood Press.

Boyd, Danah. 2014. *It's Complicated: The Social Lives of Networked Teens*. New Haven, CT: Yale University Press.

Castells, Manuel. 2001. *The Internet Galaxy: Reflections on the Internet, Business, and Society*. Oxford: Oxford University Press.

Coleman, James S. 1988. "Social Capital in the Creation of Human Capital." *American Journal of Sociology* 94: S95-S120.

Dass, Radhika, and Joyojeet Pal. 2011. "Explaining Socio-Economic Disparities in Mobile Phone Ownership in India." *Journal of International Development* 23(5): 689-706.

Desai, Sonalde, and Lester Andrist. 2010. "Gender Scripts and Age at Marriage in India." *Demography* 47(3): 667-687.

Deshpande, Ashwini. 2011. *The Grammar of Caste: Economic Discrimination in Contemporary India*. Oxford: Oxford University Press.

DiMaggio, Paul, Eszter Hargittai, Coral Celeste, and Steven Shafer. 2004. "Digital Inequality: From Unequal Access to Differentiated Use." In *Social Inequality*, edited by Kathryn Neckerman, 355-400. New York: Russell Sage Foundation.

Doron, Assa, and Robin Jeffrey. 2013. *The Great Indian Phone Book: How the Cheap Cell Phone Changes Business, Politics, and Daily Life*. Cambridge, MA: Harvard University Press.

Dyson, Tim, and Mick Moore. 1983. "On Kinship Structure, Female Autonomy, and Demographic Behavior in India." *Population and Development Review* 9(1): 35-60.

Ellison, Nicole B., Charles Steinfield, and Cliff Lampe. 2007. "The Benefits of Facebook 'Friends:' Social Capital and College Students' Use of Online Social Network Sites." *Journal of Computer-Mediated Communication* 12(4): 1143-1168.

Grover, Shalini. 2011. *Marriage, Love, Caste and Kinship Support: Lived Experiences of the Urban Poor in India*. Delhi: Social Science Press.

GSMA. 2020. "The Mobile Gender Gap Report 2020." London: GSMA.

Hampton, Keith N., and Barry Wellman. 2003. "Neighboring in Netville: How the Internet Supports Community and Social Capital in a Wired Suburb." *City & Community* 2(4): 277-311.

Hargittai, Eszter. 2002. "Second-Level Digital Divide: Differences in People's Online Skills." *First Monday* 7(4).

Jeffery, Patricia, and Roger Jeffery. 1996. *Don't Marry Me to a Plowman! Women's Everyday Lives in Rural North India*. Boulder, CO: Westview Press.

Kalyanpur, Amala. 2014. "Marital Mobility, Social Networks, and Women's Agency: Young Brides in Rural India." PhD dissertation, University of Wisconsin-Madison.

Kumar, Neha, and Tapan Parikh. 2019. "Mobiles, Internet and Gender in India." In *Gender and the Digital Economy: Perspectives from the Developing World*, edited by Cecilia Ng, 112-133. New Delhi: Sage Publications.

Kumar, Priyanka, and Rajesh Pandey. 2019. "WhatsApp Groups as Digital Communities: An Exploratory Study of Married Women in North India." *Journal of Creative Communications* 14(3): 265-279.

Kraut, Robert, Sara Kiesler, Bonka Boneva, Jonathon Cummings, Vicki Helgeson, and Anne Crawford. 2002. "Internet Paradox Revisited." *Journal of Social Issues* 58(1): 49-74.

Lin, Nan. 2001. *Social Capital: A Theory of Social Structure and Action*. Cambridge: Cambridge University Press.

Ling, Rich. 2008. *New Tech, New Ties: How Mobile Communication is Reshaping Social Cohesion*. Cambridge, MA: MIT Press.

Madan, T. N. 1965. *Family and Kinship: A Study of the Pandits of Rural Kashmir*. Bombay: Asia Publishing House.

Naqvi, Fatima, and Nishant Kumar. 2020. "WhatsApp and Women's Digital Empowerment in India: Exploring Contradictions." *Journal of International Women's Studies* 21(4): 42-56.

Norris, Pippa. 2001. *Digital Divide: Civic Engagement, Information Poverty, and the Internet Worldwide*. Cambridge: Cambridge University Press.

Pande, Rekha, and Sahana Udupa. 2019. "Digital Divides in Rural India: The Challenge of Internet Access and Use." In *Handbook of Media and Communication Studies in India*, edited by Vibodh Parthasarathi and Kiran Thakur, 198-214. New Delhi: Oxford University Press.

Patel, Reena. 2017. "Marriage, Mobility, and Intimate Citizenship: Female Marriage Migrants in Delhi." *Asian Journal of Women's Studies* 23(3): 330-351.

Pinkett, Randal D. 2003. "Community Technology and Community Building: Early Results from the Creating Community Connections Project." *The Information Society* 19(5): 365-379.

Putnam, Robert D. 2000. *Bowling Alone: The Collapse and Revival of American Community*. New York: Simon & Schuster.

Rainie, Lee, and Barry Wellman. 2012. *Networked: The New Social Operating System*. Cambridge, MA: MIT Press.

Rangaswamy, Nimmi, and Payal Arora. 2016. "The Mobile Internet in the Wild and Every Day: Digital Leisure in the Slums of Urban India." *International Journal of Cultural Studies* 19(6): 611-626.

Singh, Tripti. 2015. "Mobile Phones and Empowerment of Women in Rural India." *International Journal of Gender and Women's Studies* 3(1): 75-83.

Tenhunen, Sirpa. 2014. "Mobile Technology in the Village: ICTs, Culture, and Social Logistics in India." *Journal of the Royal Anthropological Institute* 20(3): 515-534.

Turkle, Sherry. 2011. *Alone Together: Why We Expect More from Technology and Less from Each Other*. New York: Basic Books.

van Dijk, Jan A.G.M. 2005. *The Deepening Divide: Inequality in the Information Society*. Thousand Oaks, CA: Sage Publications.

Wellman, Barry, Anabel Quan-Haase, Jeffrey Boase, Wenhong Chen, Keith Hampton, Isabel Díaz, and Kakuko Miyata. 2003. "The Social Affordances of the Internet for Networked Individualism." *Journal of Computer-Mediated Communication* 8(3): JCMC834.
