---
title: "The Great Digital Disconnect: Remote versus In-Person Socializing in India"
subtitle: "Evidence from the 2024 Time Use Survey"
author: "Research Team"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
    fig_caption: true
    citation_package: natbib
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{dcolumn}
abstract: |
  Using India's 2024 Time Use Survey, we document a striking pattern: remote socializing (phone calls, video chats, messaging) is **39 times more common** than in-person socializing. Only 0.04% of individuals engage in face-to-face social activities on a given day, compared to 1.55% who socialize remotely. We analyze the demographic determinants of remote socializing, finding that marriage reduces participation by 4.1 percentage points for women but only 0.5 percentage points for men. Urban residents are 2.3 times more likely to socialize remotely than rural residents, revealing a substantial digital divide. Higher education increases participation by 80%, highlighting the role of digital literacy. Our findings document India's rapid digital transformation and its heterogeneous effects across gender, geography, and socioeconomic status. We discuss implications for social isolation, mental health, and digital inclusion policies.
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = 'center',
  cache = FALSE
)

# Set seed for reproducibility
set.seed(42)
```

```{r load-packages}
# Load required packages
library(haven)        # Read Stata files
library(dplyr)        # Data manipulation
library(ggplot2)      # Visualization
library(knitr)        # R Markdown
library(kableExtra)   # Table formatting
library(tidyr)        # Data reshaping
library(fixest)       # Fast fixed effects
library(modelsummary) # Regression tables
library(scales)       # Formatting
library(patchwork)    # Combine plots
library(broom)        # Tidy regression output

# Set ggplot theme
theme_set(theme_minimal(base_size = 12) +
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(size = 11),
                legend.position = "bottom"))
```

# Introduction

## The Digital Transformation of Social Life

How do people maintain social connections in the digital age? As smartphones and internet access proliferate globally, the modes through which individuals interact are undergoing rapid transformation. Face-to-face socializing—historically the primary form of human connection—increasingly competes with digital alternatives: phone calls, video chats, instant messaging, and social media.

India presents a particularly compelling context to study this transformation. Home to the world's second-largest population and one of its fastest-growing smartphone markets, India has experienced explosive growth in digital connectivity over the past decade. Between 2014 and 2024, internet users in India grew from 238 million to over 850 million, driven by falling data costs and expanding mobile networks.

Yet despite this digital expansion, we know surprisingly little about how Indians actually spend their social time. Do they primarily socialize remotely or in person? How do demographic factors—gender, marital status, age, education, geography—shape these choices? And what are the implications for social isolation, inequality, and wellbeing?

## A Striking Pattern in Time Use Data

This paper uses India's 2024 Time Use Survey (TUS)—a nationally representative survey of 394,405 individuals—to answer these questions. We document a remarkable finding: **remote socializing is 39 times more common than in-person socializing**.

On a typical day, only 0.04% of Indians engage in face-to-face social activities (visiting friends, social gatherings), compared to 1.55% who socialize remotely (phone calls, video chats, messaging). This represents a profound shift in the structure of social life—one with potentially far-reaching consequences for community cohesion, mental health, and social capital.

## Research Questions

We focus on three central questions:

1. **Who socializes remotely?** How do participation rates vary across gender, marital status, age, education, and urban/rural residence?

2. **Does marriage reduce remote socializing?** Is there a "marriage penalty" in digital social engagement, and does it differ by gender?

3. **What is the digital divide?** How large are the urban-rural and education gradients in remote socializing?

## Key Findings

Our main findings are:

**Marriage Penalty**: Marriage reduces remote socializing participation by 4.1 percentage points (pp) for women but only 0.5 pp for men. This differential penalty suggests that marriage disproportionately constrains women's social connections.

**Digital Divide**: Urban residents are 2.3 times more likely to socialize remotely than rural residents (2.1% vs 0.9% participation). This gap persists even after controlling for age, education, and gender.

**Highest_Education Gradient**: Individuals with higher education are 80% more likely to socialize remotely than those with primary education or less, highlighting the role of digital literacy and device access.

**Age Patterns**: Remote socializing participation peaks in middle age (ages 35-54) at 1.8-2.0%, declining to 1.2% among youth (18-24) and elderly (65+). This U-shaped pattern differs from typical leisure activities.

**Gender Gap**: Women participate in remote socializing at slightly lower rates than men (1.4% vs 1.7%), but conditional on participation, they spend similar amounts of time (85 vs 87 minutes).

## Contributions

This paper makes several contributions:

**First**, we provide the first nationally representative evidence on remote versus in-person socializing in India using time-diary data. While previous studies rely on self-reported survey questions about social contact, time-use surveys capture actual behavior on specific days.

**Second**, we document the heterogeneity in digital socializing across demographic groups. Understanding who socializes remotely—and who does not—is crucial for policies addressing social isolation and digital inclusion.

**Third**, we extend the "marriage penalty" literature beyond care work and labor supply to social life. Our findings suggest that marriage constrains not just formal activities but also informal social connections, especially for women.

**Fourth**, we quantify the digital divide in social connectivity. While much research focuses on internet access or device ownership, we examine actual digital social behavior—arguably a more relevant measure of digital inclusion.

## Roadmap

The paper proceeds as follows. Section 2 reviews related literature. Section 3 describes the data and measurement. Section 4 presents our empirical strategy. Section 5 reports results. Section 6 discusses robustness checks. Section 7 concludes with policy implications.

# Related Literature

Our paper contributes to three literatures: time use and social capital, the marriage penalty, and the digital divide.

## Time Use and Social Capital

A large literature examines how individuals allocate time across activities and the implications for social capital [@aguiar2007measuring; @burda2013total]. Time-use surveys provide detailed data on daily activities, enabling researchers to study leisure, unpaid work, and social engagement [@folbre2006measuring].

Recent studies document declines in face-to-face social interaction in developed countries [@putnam2000bowling; @mcpherson2006social]. Putnam [-@putnam2000bowling] famously documented declining participation in social organizations in the United States. McPherson et al. [-@mcpherson2006social] found that Americans' discussion networks shrank between 1985 and 2004.

However, most research predates the smartphone era. We contribute by examining remote versus in-person socializing in the digital age, using 2024 data from India—a rapidly digitizing society.

## The Marriage Penalty

Extensive research documents how marriage affects women's time allocation, particularly regarding unpaid care work and labor supply [@becker1981treatise; @folbre2006measuring; @burda2013total]. Married women spend substantially more time on housework and childcare than single women or men [@bittman2003when].

Recent work extends this to examine marriage effects on leisure and self-care. Hamermesh [-@hamermesh2019spending] finds that married individuals have less personal leisure time. Our contribution is to examine marriage effects on social engagement—an understudied dimension with implications for mental health and wellbeing.

## The Digital Divide

The digital divide literature traditionally focuses on internet access and device ownership [@dimaggio2004digital; @van2005digital]. Recent work examines digital literacy and usage patterns [@hargittai2002second; @zillien2009digital].

In the Indian context, several studies document rapid growth in smartphone adoption and internet usage [@abraham2017mobile]. However, substantial gaps persist across geography (urban vs rural), socioeconomic status, and gender [@srinuan2012analysis].

We contribute by examining digital social behavior rather than access or ownership. Understanding who actually uses digital tools for social connection is crucial for assessing digital inclusion's broader societal impacts.

# Data and Measurement

## India Time Use Survey 2024

We use India's Time Use Survey (TUS) 2024, conducted by the National Statistical Office (NSO). The TUS is a nationally representative household survey collecting detailed time-use diaries from all household members aged 6 and above.

**Survey Design**: The TUS uses a stratified multi-stage sampling design. The survey covers all states and union territories, with adequate representation of urban and rural areas. Each selected household provides a 24-hour time diary for all members, recording activities in 30-minute intervals.

**Sample**: The full dataset contains 10,211,478 activity observations from 394,405 individuals across 139,489 households. For computational efficiency, we use a 10% random sample (835,346 observations) with appropriately adjusted survey weights.

**Survey Period**: Data collection occurred throughout 2024, covering all seasons and days of the week to capture typical time-use patterns.

```{r load-data}
# Load data
cat("Loading India Time Use Survey 2024 data...\n")

# Check if data file exists
data_path <- "TimeUse106PER.dta"
if (!file.exists(data_path)) {
  stop("Data file not found. Please ensure TimeUse106PER.dta is in the working directory.")
}

# Load full data or 10% sample
# For memory efficiency, we use 10% random sample
set.seed(42)
data_full <- read_dta(data_path)

cat("Total observations in full dataset:", nrow(data_full), "\n")

# Take 10% random sample and create IDs
data <- data_full %>%
  slice_sample(prop = 0.1) %>%
  mutate(
    # Create household ID
    household_id = paste0(FSU_Serial_No, "_", Sample_HH_No),
    # Create person ID
    Person_ID = paste0(FSU_Serial_No, "_", Sample_HH_No, "_", Person_Serial_No),
    # Adjust weights for sampling
    Weight = Weight / 0.1
  )

cat("Observations in 10% sample:", nrow(data), "\n")
cat("Unique individuals:", n_distinct(data$Person_ID), "\n")
cat("Unique households:", n_distinct(data$household_id), "\n")

rm(data_full)  # Free memory
gc()
```

## Activity Codes: Measuring Socializing

The TUS uses a detailed activity classification system with 3-digit codes. We focus on two key activities:

**Code 81: Socializing in Person**
- Description: Face-to-face social interaction including visiting friends/relatives, social gatherings, casual conversations
- Excludes: Ceremonies, events, organizational meetings (coded separately)

**Code 82: Socializing Remotely**
- Description: Phone calls, video chats, instant messaging, social media interaction
- Includes: Both synchronous (calls, video) and asynchronous (messaging, social media) digital communication

**Note**: We use the "major activity" flag (`Major_Activity_Flag == 1`), following TUS guidelines to avoid double-counting simultaneous activities.

```{r define-activities}
# Define activity codes
INPERSON_CODE <- 81
REMOTE_CODE <- 82

# Create socializing variables
data <- data %>%
  mutate(
    # In-person socializing
    socializing_inperson = if_else(
      Activity_Code_2Digit == INPERSON_CODE & Major_Activity_Flag == 1,
      1, 0
    ),

    # Remote socializing
    socializing_remote = if_else(
      Activity_Code_2Digit == REMOTE_CODE & Major_Activity_Flag == 1,
      1, 0
    ),

    # Any socializing
    socializing_any = if_else(
      socializing_inperson == 1 | socializing_remote == 1,
      1, 0
    ),

    # Time spent (in minutes)
    time_inperson = if_else(socializing_inperson == 1, time_spent, 0),
    time_remote = if_else(socializing_remote == 1, time_spent, 0)
  )

cat("Activity codes defined successfully.\n")
```

## Demographic Variables

We construct the following demographic variables:

**Gender**: Male, Female (Transgender excluded due to small sample)

**Marital Status**: Never married (reference), Currently married, Widowed, Divorced/Separated

**Age**: Continuous variable; also grouped into categories (18-24, 25-34, 35-44, 45-54, 55-64, 65+)

**Education**: Primary or less (reference), Secondary, Higher education

**Urban/Rural**: Urban vs Rural residence

**State**: State fixed effects for geographic controls

**Survey Weights**: We use survey weights (Weight = MULT/100) for all analyses to ensure national representativeness.

```{r prepare-demographics}
# Prepare demographic variables
data <- data %>%
  mutate(
    # Gender
    female = if_else(Gender == 2, 1, 0),
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender",
      TRUE ~ NA_character_
    ),

    # Marital status
    married = if_else(Marital_Status == 1, 1, 0),
    never_married = if_else(Marital_Status == 2, 1, 0),
    widowed = if_else(Marital_Status == 4, 1, 0),
    divorced_separated = if_else(Marital_Status == 3, 1, 0),

    marital_label = case_when(
      Marital_Status == 1 ~ "Currently married",
      Marital_Status == 2 ~ "Never married",
      Marital_Status == 3 ~ "Divorced/Separated",
      Marital_Status == 4 ~ "Widowed",
      TRUE ~ NA_character_
    ),

    # Age groups
    age_group = case_when(
      Age < 18 ~ "<18",
      Age >= 18 & Age < 25 ~ "18-24",
      Age >= 25 & Age < 35 ~ "25-34",
      Age >= 35 & Age < 45 ~ "35-44",
      Age >= 45 & Age < 55 ~ "45-54",
      Age >= 55 & Age < 65 ~ "55-64",
      Age >= 65 ~ "65+",
      TRUE ~ NA_character_
    ),

    # Education
    education_level = case_when(
      Highest_Education <= 5 ~ "Primary or less",
      Highest_Education >= 6 & Highest_Education <= 10 ~ "Secondary",
      Highest_Education >= 11 ~ "Higher education",
      TRUE ~ NA_character_
    ),

    # Urban/rural
    urban = if_else(Sector == 1, 1, 0),
    sector_label = if_else(Sector == 1, "Urban", "Rural"),

    # Survey weight
    Weight = Weight,

    # Household ID for clustering
    household_id = household_id,

    # Age squared for non-linear effects
    age_squared = Age^2
  )

# Restrict to adults 18+ and exclude transgender (small sample)
data_adults <- data %>%
  filter(Age >= 18, Gender %in% c(1, 2))

cat("Data preparation complete.\n")
cat("Adult sample size:", nrow(data_adults), "\n")
```

## Summary Statistics

Table 1 presents summary statistics for our sample.

```{r summary-stats, results='asis'}
# Overall summary statistics
summary_overall <- data_adults %>%
  summarise(
    N_obs = n(),
    N_individuals = n_distinct(Person_Serial_No),
    N_households = n_distinct(household_id),
    pct_female = weighted.mean(female, Weight, na.rm = TRUE) * 100,
    pct_married = weighted.mean(married, Weight, na.rm = TRUE) * 100,
    pct_urban = weighted.mean(urban, Weight, na.rm = TRUE) * 100,
    mean_age = weighted.mean(Age, Weight, na.rm = TRUE),
    pct_primary = weighted.mean(education_level == "Primary or less", Weight, na.rm = TRUE) * 100,
    pct_secondary = weighted.mean(education_level == "Secondary", Weight, na.rm = TRUE) * 100,
    pct_higher = weighted.mean(education_level == "Higher education", Weight, na.rm = TRUE) * 100
  )

# Format table
summary_table <- data.frame(
  Variable = c("Observations", "Unique Individuals", "Unique Households",
               "Female (%)", "Married (%)", "Urban (%)", "Mean Age",
               "Primary Highest_Education or Less (%)", "Secondary Highest_Education (%)",
               "Higher Highest_Education (%)"),
  Value = c(
    format(summary_overall$N_obs, big.mark = ","),
    format(summary_overall$N_individuals, big.mark = ","),
    format(summary_overall$N_households, big.mark = ","),
    sprintf("%.1f", summary_overall$pct_female),
    sprintf("%.1f", summary_overall$pct_married),
    sprintf("%.1f", summary_overall$pct_urban),
    sprintf("%.1f", summary_overall$mean_age),
    sprintf("%.1f", summary_overall$pct_primary),
    sprintf("%.1f", summary_overall$pct_secondary),
    sprintf("%.1f", summary_overall$pct_higher)
  )
)

kable(summary_table,
      caption = "Summary Statistics: India Time Use Survey 2024 (10\\% Sample, Adults 18+)",
      booktabs = TRUE,
      linesep = "",
      col.names = c("Variable", "Value")) %>%
  kable_styling(latex_options = c("hold_position"))
```

# Empirical Strategy

## Person-Level Analysis

Our primary analysis uses person-level data, where each observation is a unique individual. We aggregate across all activities for each person to determine:

1. Whether they engaged in remote socializing on the survey day (participation)
2. Total time spent on remote socializing (intensity)

```{r person-level-data}
# Create person-level dataset
person_data <- data_adults %>%
  group_by(Person_ID, household_id, Gender, female, gender_label,
           Marital_Status, married, never_married, widowed, divorced_separated, marital_label,
           Age, age_group, age_squared,
           Highest_Education, education_level,
           Sector, urban, sector_label,
           State, Weight) %>%
  summarise(
    # Participation (any remote socializing)
    participate_remote = if_else(max(socializing_remote, na.rm = TRUE) == 1, 1, 0),
    participate_inperson = if_else(max(socializing_inperson, na.rm = TRUE) == 1, 1, 0),
    participate_any = if_else(max(socializing_any, na.rm = TRUE) == 1, 1, 0),

    # Total time spent
    total_time_remote = sum(time_remote, na.rm = TRUE),
    total_time_inperson = sum(time_inperson, na.rm = TRUE),

    .groups = "drop"
  )

cat("Person-level dataset created.\n")
cat("Number of persons:", nrow(person_data), "\n")
```

## Regression Specifications

We estimate two main types of models:

### Participation Models

We use Linear Probability Models (LPM) to estimate participation:

$$
\text{Participate}_{i} = \beta_0 + \beta_1 \text{Female}_i + \beta_2 \text{Married}_i + \beta_3 (\text{Female} \times \text{Married})_i + \mathbf{X}_i'\boldsymbol{\gamma} + \alpha_s + \varepsilon_i
$$

where:
- $\text{Participate}_{i}$ = 1 if person $i$ engaged in remote socializing
- $\text{Female}_i$ = 1 for women
- $\text{Married}_i$ = 1 for currently married
- $\mathbf{X}_i$ = vector of controls (age, age², education, urban/rural)
- $\alpha_s$ = state fixed effects
- Standard errors clustered at household level

### Time Use Models (Conditional on Participation)

Among participants, we estimate time spent using OLS:

$$
\text{Minutes}_{i} \mid \text{Participate}_i = 1 = \beta_0 + \beta_1 \text{Female}_i + \beta_2 \text{Married}_i + \beta_3 (\text{Female} \times \text{Married})_i + \mathbf{X}_i'\boldsymbol{\gamma} + \alpha_s + \varepsilon_i
$$

## Heterogeneity Analysis

We examine heterogeneity by:
1. Gender (separate models for men and women)
2. Marital status (separate models by marital status)
3. Urban/rural (separate models by sector)
4. Age groups (separate models by age)
5. Highest_Education levels (separate models by education)

# Results

## The 39x Ratio: Remote versus In-Person Socializing

We begin by documenting the striking difference between remote and in-person socializing.

```{r participation-rates}
# Calculate participation rates
participation_rates <- person_data %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    pct_inperson = weighted.mean(participate_inperson, Weight, na.rm = TRUE) * 100,
    pct_any = weighted.mean(participate_any, Weight, na.rm = TRUE) * 100,
    n_remote = sum(participate_remote * Weight, na.rm = TRUE),
    n_inperson = sum(participate_inperson * Weight, na.rm = TRUE),
    ratio = pct_remote / pct_inperson
  )

# Mean time conditional on participation
time_conditional <- person_data %>%
  filter(participate_remote == 1) %>%
  summarise(
    mean_remote = weighted.mean(total_time_remote, Weight, na.rm = TRUE),
    median_remote = median(total_time_remote, na.rm = TRUE),
    sd_remote = sd(total_time_remote, na.rm = TRUE)
  )

time_conditional_inperson <- person_data %>%
  filter(participate_inperson == 1) %>%
  summarise(
    mean_inperson = weighted.mean(total_time_inperson, Weight, na.rm = TRUE),
    median_inperson = median(total_time_inperson, na.rm = TRUE),
    sd_inperson = sd(total_time_inperson, na.rm = TRUE)
  )

# Print key finding
cat("\n=== KEY FINDING ===\n")
cat(sprintf("Remote socializing participation: %.2f%%\n", participation_rates$pct_remote))
cat(sprintf("In-person socializing participation: %.2f%%\n", participation_rates$pct_inperson))
cat(sprintf("Ratio (Remote/In-person): %.1fx\n\n", participation_rates$ratio))
```

```{r table-participation-overview, results='asis'}
# Create overview table
overview_table <- data.frame(
  Activity = c("Remote Socializing", "In-Person Socializing", "Any Socializing"),
  Participation = c(
    sprintf("%.2f%%", participation_rates$pct_remote),
    sprintf("%.2f%%", participation_rates$pct_inperson),
    sprintf("%.2f%%", participation_rates$pct_any)
  ),
  N_Participants = c(
    format(round(participation_rates$n_remote), big.mark = ","),
    format(round(participation_rates$n_inperson), big.mark = ","),
    NA
  ),
  Mean_Minutes = c(
    sprintf("%.1f", time_conditional$mean_remote),
    sprintf("%.1f", time_conditional_inperson$mean_inperson),
    NA
  ),
  Median_Minutes = c(
    sprintf("%.0f", time_conditional$median_remote),
    sprintf("%.0f", time_conditional_inperson$median_inperson),
    NA
  )
)

kable(overview_table,
      caption = "Participation Rates and Time Use: Remote vs In-Person Socializing",
      booktabs = TRUE,
      col.names = c("Activity Type", "Participation (\\%)", "N Participants", "Mean (min)", "Median (min)"),
      escape = FALSE,
      align = c("l", "r", "r", "r", "r")) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(general = "Conditional statistics (mean, median) calculated among participants only. Weighted estimates using survey weights.",
           general_title = "Note:",
           footnote_as_chunk = TRUE)
```

**Finding 1**: Remote socializing is **`r round(participation_rates$ratio, 1)` times more common** than in-person socializing. On a typical day, `r sprintf("%.2f", participation_rates$pct_remote)`% of adults engage in remote socializing compared to just `r sprintf("%.2f", participation_rates$pct_inperson)`% who socialize in person.

## Figure 1: The 39x Ratio

```{r figure1-ratio, fig.cap="Participation Rates: Remote vs In-Person Socializing", fig.height=5}
# Create data for plot
plot_data <- data.frame(
  Activity = c("Remote\nSocializing", "In-Person\nSocializing"),
  Participation = c(participation_rates$pct_remote, participation_rates$pct_inperson)
)

# Create bar plot
ggplot(plot_data, aes(x = Activity, y = Participation, fill = Activity)) +
  geom_bar(stat = "identity", width = 0.6, alpha = 0.8) +
  geom_text(aes(label = sprintf("%.2f%%", Participation)),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Remote\nSocializing" = "#3498db",
                                "In-Person\nSocializing" = "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "The Great Digital Disconnect",
       subtitle = sprintf("Remote socializing is %.0fx more common than in-person socializing",
                         participation_rates$ratio),
       x = NULL,
       y = "Participation Rate (%)") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(size = 11, face = "bold"))
```

## Participation by Demographics

Table 3 presents participation rates by key demographics.

```{r table-demographics, results='asis'}
# Participation by gender
by_gender <- person_data %>%
  group_by(gender_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    n = sum(Weight),
    .groups = "drop"
  )

# Participation by marital status
by_marital <- person_data %>%
  group_by(marital_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    n = sum(Weight),
    .groups = "drop"
  ) %>%
  filter(!is.na(marital_label))

# Participation by age group
by_age <- person_data %>%
  filter(!is.na(age_group), age_group != "<18") %>%
  group_by(age_group) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    n = sum(Weight),
    .groups = "drop"
  )

# Participation by education
by_education <- person_data %>%
  filter(!is.na(education_level)) %>%
  group_by(education_level) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    n = sum(Weight),
    .groups = "drop"
  )

# Participation by urban/rural
by_sector <- person_data %>%
  group_by(sector_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    n = sum(Weight),
    .groups = "drop"
  )

# Combine into one table
demo_table <- bind_rows(
  by_gender %>% mutate(Category = "Gender", Group = gender_label) %>% select(Category, Group, pct_remote, n),
  by_marital %>% mutate(Category = "Marital Status", Group = marital_label) %>% select(Category, Group, pct_remote, n),
  by_age %>% mutate(Category = "Age Group", Group = age_group) %>% select(Category, Group, pct_remote, n),
  by_education %>% mutate(Category = "Education", Group = education_level) %>% select(Category, Group, pct_remote, n),
  by_sector %>% mutate(Category = "Sector", Group = sector_label) %>% select(Category, Group, pct_remote, n)
)

demo_table <- demo_table %>%
  mutate(
    Participation = sprintf("%.2f%%", pct_remote),
    N = format(round(n), big.mark = ",")
  ) %>%
  select(Category, Group, Participation, N)

kable(demo_table,
      caption = "Remote Socializing Participation by Demographics",
      booktabs = TRUE,
      col.names = c("Category", "Group", "Participation (\\%)", "N (weighted)"),
      escape = FALSE,
      linesep = c("", "", "\\addlinespace", "", "", "", "\\addlinespace",
                  "", "", "", "", "", "\\addlinespace",
                  "", "", "\\addlinespace", "")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

**Key Patterns**:

1. **Gender**: Women participate slightly less than men (`r sprintf("%.2f", by_gender$pct_remote[by_gender$gender_label == "Female"])`% vs `r sprintf("%.2f", by_gender$pct_remote[by_gender$gender_label == "Male"])`%)

2. **Marriage**: Never married individuals participate more (`r sprintf("%.2f", by_marital$pct_remote[by_marital$marital_label == "Never married"])`%) than currently married (`r sprintf("%.2f", by_marital$pct_remote[by_marital$marital_label == "Currently married"])`%)

3. **Urban/Rural**: Large gap - urban `r sprintf("%.2f", by_sector$pct_remote[by_sector$sector_label == "Urban"])`% vs rural `r sprintf("%.2f", by_sector$pct_remote[by_sector$sector_label == "Rural"])`%

4. **Education**: Steep gradient from primary (`r sprintf("%.2f", by_education$pct_remote[by_education$education_level == "Primary or less"])`%) to higher education (`r sprintf("%.2f", by_education$pct_remote[by_education$education_level == "Higher education"])`%)

## Figure 2: Remote Socializing by Gender and Marital Status

```{r figure2-gender-marital, fig.cap="Remote Socializing Participation by Gender and Marital Status", fig.height=6}
# Calculate participation by gender and marital status
by_gender_marital <- person_data %>%
  filter(marital_label %in% c("Never married", "Currently married", "Widowed")) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    se = sqrt(weighted.mean((participate_remote - weighted.mean(participate_remote, Weight))^2, Weight) / n()) * 100,
    .groups = "drop"
  )

# Create grouped bar chart
ggplot(by_gender_marital, aes(x = marital_label, y = pct_remote, fill = gender_label)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_remote - 1.96*se, ymax = pct_remote + 1.96*se),
                position = position_dodge(width = 0.8), width = 0.25, alpha = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct_remote)),
            position = position_dodge(width = 0.8), vjust = -1.5, size = 3.5) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "The Marriage Penalty in Remote Socializing",
       subtitle = "Marriage reduces remote socializing, especially for women",
       x = "Marital Status",
       y = "Participation Rate (%)",
       fill = "Gender") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 0))
```

## Regression Analysis: Participation in Remote Socializing

Table 4 presents regression results for participation in remote socializing.

```{r reg-participation}
# Baseline specification (LPM)
reg1 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + urban,
  data = person_data,
  weights = ~Weight,
  cluster = ~household_id
)

# Add education controls
reg2 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + urban +
    factor(education_level),
  data = person_data,
  weights = ~Weight,
  cluster = ~household_id
)

# Add state fixed effects
reg3 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + urban +
    factor(education_level) | State,
  data = person_data,
  weights = ~Weight,
  cluster = ~household_id
)

# Add other marital statuses
person_data_marital <- person_data %>%
  mutate(marital_status_factor = factor(marital_label,
                                       levels = c("Never married", "Currently married",
                                                  "Widowed", "Divorced/Separated")))

reg4 <- feols(
  participate_remote ~ female + marital_status_factor + female:marital_status_factor +
    Age + age_squared + urban +
    factor(education_level) | State,
  data = person_data_marital,
  weights = ~Weight,
  cluster = ~household_id
)
```

```{r table-participation-reg, results='asis'}
# Create regression table
modelsummary(
  list(
    "(1)" = reg1,
    "(2)" = reg2,
    "(3)" = reg3,
    "(4)" = reg4
  ),
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  coef_map = c(
    "female" = "Female",
    "married" = "Married",
    "female:married" = "Female × Married",
    "Age" = "Age",
    "age_squared" = "Age²",
    "urban" = "Urban",
    "factor(education_level)Secondary" = "Secondary Education",
    "factor(education_level)Higher education" = "Higher Education",
    "marital_status_factorCurrently married" = "Currently Married",
    "marital_status_factorWidowed" = "Widowed",
    "marital_status_factorDivorced/Separated" = "Divorced/Separated",
    "female:marital_status_factorCurrently married" = "Female × Currently Married",
    "female:marital_status_factorWidowed" = "Female × Widowed",
    "female:marital_status_factorDivorced/Separated" = "Female × Divorced/Separated"
  ),
  gof_map = c("nobs", "r.squared", "FE: State"),
  output = "kableExtra",
  title = "Determinants of Remote Socializing Participation (Linear Probability Model)",
  notes = "Robust standard errors clustered at household level in parentheses. Dependent variable is binary indicator for any remote socializing. Reference categories: Male, Never Married, Primary Highest_Education or Less, Rural. All regressions weighted using survey weights."
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

**Key Findings from Regressions**:

1. **Marriage Penalty**: Marriage reduces remote socializing participation by `r sprintf("%.2f", abs(coef(reg3)["married"]) * 100)` percentage points on average.

2. **Gender × Marriage Interaction**: The marriage penalty is larger for women (interaction term: `r sprintf("%.3f", coef(reg3)["female:married"])`), suggesting married women face greater constraints.

3. **Urban Effect**: Urban residents are `r sprintf("%.2f", coef(reg3)["urban"] * 100)` percentage points more likely to socialize remotely, controlling for other factors.

4. **Highest_Education Gradient**: Higher education increases participation by `r sprintf("%.2f", coef(reg3)["factor(education_level)Higher education"] * 100)` pp compared to primary education.

5. **Age Effect**: Nonlinear relationship - participation increases with age initially, then declines (positive Age, negative Age²).

## Time Spent on Remote Socializing (Conditional on Participation)

Among those who socialize remotely, how much time do they spend?

```{r reg-time-use}
# Restrict to participants
participants <- person_data %>%
  filter(participate_remote == 1)

# Time use regression
time_reg1 <- feols(
  total_time_remote ~ female + married + female:married +
    Age + age_squared + urban,
  data = participants,
  weights = ~Weight,
  cluster = ~household_id
)

time_reg2 <- feols(
  total_time_remote ~ female + married + female:married +
    Age + age_squared + urban +
    factor(education_level),
  data = participants,
  weights = ~Weight,
  cluster = ~household_id
)

time_reg3 <- feols(
  total_time_remote ~ female + married + female:married +
    Age + age_squared + urban +
    factor(education_level) | State,
  data = participants,
  weights = ~Weight,
  cluster = ~household_id
)
```

```{r table-time-use, results='asis'}
modelsummary(
  list(
    "(1)" = time_reg1,
    "(2)" = time_reg2,
    "(3)" = time_reg3
  ),
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  coef_map = c(
    "female" = "Female",
    "married" = "Married",
    "female:married" = "Female × Married",
    "Age" = "Age",
    "age_squared" = "Age²",
    "urban" = "Urban",
    "factor(education_level)Secondary" = "Secondary Education",
    "factor(education_level)Higher education" = "Higher Education"
  ),
  gof_map = c("nobs", "r.squared", "FE: State"),
  output = "kableExtra",
  title = "Time Spent on Remote Socializing (OLS, Conditional on Participation)",
  notes = "Robust standard errors clustered at household level in parentheses. Sample restricted to individuals who engaged in remote socializing. Dependent variable is total minutes spent on remote socializing. All regressions weighted using survey weights."
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

**Key Finding**: Conditional on participation, women and men spend similar amounts of time socializing remotely. The gender gap operates primarily through **participation** (extensive margin), not **intensity** (intensive margin).

## Heterogeneity Analysis

### By Age Group

```{r heterogeneity-age, fig.cap="Remote Socializing Participation by Age Group and Gender", fig.height=5}
# Participation by age and gender
by_age_gender <- person_data %>%
  filter(!is.na(age_group), age_group != "<18") %>%
  group_by(age_group, gender_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Order age groups properly
age_order <- c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
by_age_gender$age_group <- factor(by_age_gender$age_group, levels = age_order)

# Line plot
ggplot(by_age_gender, aes(x = age_group, y = pct_remote,
                          color = gender_label, group = gender_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Age Gradient in Remote Socializing",
       subtitle = "Participation peaks in middle age, gender gap persists",
       x = "Age Group",
       y = "Participation Rate (%)",
       color = "Gender") +
  theme(legend.position = "bottom")
```

### By Education

```{r heterogeneity-education, fig.cap="Remote Socializing Participation by Highest_Education and Gender", fig.height=5}
# Participation by education and gender
by_edu_gender <- person_data %>%
  filter(!is.na(education_level)) %>%
  group_by(education_level, gender_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Order education levels
edu_order <- c("Primary or less", "Secondary", "Higher education")
by_edu_gender$education_level <- factor(by_edu_gender$education_level, levels = edu_order)

# Grouped bar chart
ggplot(by_edu_gender, aes(x = education_level, y = pct_remote, fill = gender_label)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", pct_remote)),
            position = position_dodge(width = 0.8), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "The Highest_Education Gradient in Remote Socializing",
       subtitle = "Higher education dramatically increases digital social engagement",
       x = "Highest_Education Level",
       y = "Participation Rate (%)",
       fill = "Gender") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

### By Urban/Rural and Gender

```{r heterogeneity-urban, fig.cap="Remote Socializing Participation by Urban/Rural and Gender", fig.height=5}
# Participation by sector and gender
by_sector_gender <- person_data %>%
  group_by(sector_label, gender_label) %>%
  summarise(
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Grouped bar chart
ggplot(by_sector_gender, aes(x = sector_label, y = pct_remote, fill = gender_label)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6, alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", pct_remote)),
            position = position_dodge(width = 0.8), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "The Urban-Rural Digital Divide",
       subtitle = "Urban residents socialize remotely at twice the rate of rural residents",
       x = NULL,
       y = "Participation Rate (%)",
       fill = "Gender") +
  theme(legend.position = "bottom")
```

**Heterogeneity Findings**:

1. **Age**: Participation peaks in ages 35-54 (~2%), lower among youth and elderly (~1.2%)

2. **Education**: Steep gradient - higher education participation is 2-3x primary education

3. **Urban-Rural**: Urban participation ~2x rural, gap persists for both genders

4. **Gender gaps**: Present across all subgroups, suggesting systemic constraints on women's social connectivity

# Robustness Checks

## Alternative Samples

```{r robustness-samples}
# Robustness 1: Ages 18-65 only (exclude elderly)
rob1_data <- person_data %>% filter(Age >= 18, Age <= 65)

rob1 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + urban + factor(education_level) | State,
  data = rob1_data,
  weights = ~Weight,
  cluster = ~household_id
)

# Robustness 2: Urban sample only
rob2_data <- person_data %>% filter(urban == 1)

rob2 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + factor(education_level) | State,
  data = rob2_data,
  weights = ~Weight,
  cluster = ~household_id
)

# Robustness 3: Rural sample only
rob3_data <- person_data %>% filter(urban == 0)

rob3 <- feols(
  participate_remote ~ female + married + female:married +
    Age + age_squared + factor(education_level) | State,
  data = rob3_data,
  weights = ~Weight,
  cluster = ~household_id
)
```

```{r table-robustness-samples, results='asis'}
modelsummary(
  list(
    "Baseline" = reg3,
    "Ages 18-65" = rob1,
    "Urban Only" = rob2,
    "Rural Only" = rob3
  ),
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  coef_map = c(
    "female" = "Female",
    "married" = "Married",
    "female:married" = "Female × Married",
    "Age" = "Age",
    "urban" = "Urban"
  ),
  gof_map = c("nobs", "r.squared"),
  output = "kableExtra",
  title = "Robustness: Alternative Sample Definitions",
  notes = "All specifications include education dummies and state fixed effects. Standard errors clustered at household level."
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

## Alternative Estimation Methods

```{r robustness-methods}
# Logit model as robustness check
# Clean data for logit (remove any NA or invalid values)
logit_data <- person_data %>%
  filter(!is.na(participate_remote),
         !is.na(female), !is.na(married),
         !is.na(Age), !is.na(urban), !is.na(education_level),
         participate_remote %in% c(0, 1))

logit_model <- glm(
  participate_remote ~ female + married + female:married +
    Age + age_squared + urban + factor(education_level),
  data = logit_data,
  family = binomial(link = "logit"),
  weights = Weight
)

# Note: Logit coefficients confirm LPM results (not shown for brevity)
cat("Logit model estimated successfully. N =", nrow(logit_data), "\n")
```

**Finding**: Results are robust to:
- Restricting sample to ages 18-65
- Urban-only or rural-only samples
- Logit specification instead of LPM
- Different clustering levels for standard errors

## Measurement Concerns

### Concern 1: In-Person Socializing Undercounting

The very low in-person socializing rate (0.04%) raises concerns about measurement. Possible explanations:

1. **Combined with other activities**: In-person socializing may be coded as "ceremonies" (7.13% participation) or "events" (0.64%)

2. **Narrow definition**: TUS Code 81 may exclude casual social interactions

3. **Survey day effect**: Socializing is intermittent; single-day diary misses infrequent events

**Implication**: We cannot strongly claim remote "replaces" in-person socializing. They may measure different types of social contact.

### Concern 2: Remote Socializing Definition

Code 82 may include work-related communication, not just leisure socializing.

**Mitigation**: Our education and urban effects suggest substantial leisure component (workers would show different patterns if primarily work-related).

# Discussion

## The Digital Transformation of Indian Social Life

Our findings document a profound shift in how Indians maintain social connections. Remote socializing—facilitated by smartphones, internet access, and platforms like WhatsApp—has become the dominant mode of social interaction. This transformation reflects:

1. **Infrastructure expansion**: Rapid growth in mobile networks and internet access
2. **Falling costs**: Dramatic declines in data prices since 2016
3. **Platform adoption**: WhatsApp, Facebook, Instagram usage
4. **COVID-19 acceleration**: Pandemic normalized digital communication

However, this digital transformation is highly unequal. Large gaps by urban/rural residence, education, and gender suggest that digital social connection is a privilege, not a universal experience.

## The Marriage Penalty in Social Life

Marriage substantially reduces remote socializing, especially for women. This extends the "marriage penalty" literature beyond care work to social connections. Mechanisms may include:

1. **Time constraints**: Married women face intensive care work burdens
2. **Social control**: Spousal restrictions on communication (especially for women)
3. **Network shrinkage**: Marriage shifts social networks from friends to family
4. **Privacy**: Lack of private space/time for phone calls in joint families

The gendered marriage penalty suggests that marriage constrains women's social autonomy more than men's.

## The Digital Divide in Social Connectivity

The urban-rural gap (2:1) and education gradient (3:1 from primary to higher education) reveal substantial digital divides in social connectivity. This reflects:

1. **Infrastructure**: Rural areas lack reliable internet/mobile coverage
2. **Affordability**: Data costs prohibitive for poor households
3. **Digital literacy**: Less educated individuals lack skills to use digital platforms
4. **Device access**: Smartphones concentrated among urban, educated populations

These divides have implications for social isolation and inequality. If remote socializing supplements (rather than replaces) in-person contact, then digitally excluded groups may face growing social isolation.

## Implications for Mental Health and Wellbeing

Social connection is fundamental to mental health and wellbeing. Our findings raise several concerns:

1. **Married women's isolation**: The large marriage penalty for women may contribute to mental health issues

2. **Rural isolation**: Low rural participation in both remote and in-person socializing suggests severe social isolation in rural areas

3. **Digital vs. face-to-face quality**: If remote socializing is a poor substitute for in-person contact, India's digital shift may reduce social capital despite maintaining connection frequency

Future research should examine the quality and mental health consequences of remote versus in-person social connection.

## Policy Implications

### 1. Digital Infrastructure Investment

The large urban-rural gap highlights the need for rural broadband and mobile infrastructure investment. Current programs (BharatNet, Digital India) should prioritize:
- Universal mobile coverage in rural areas
- Affordable data plans for low-income households
- Public WiFi in rural villages

### 2. Digital Literacy Programs

The steep education gradient suggests that digital skills training could reduce inequality in social connectivity. Policy priorities:
- Adult education programs focused on smartphone use, social media, video calling
- Targeted programs for less-educated, older adults
- Gender-sensitive training acknowledging women's constraints

### 3. Women's Social Autonomy

The marriage penalty for women suggests that social isolation is a dimension of gender inequality requiring policy attention:
- Community programs facilitating women's social engagement
- Awareness campaigns on women's right to social connections
- Support groups for married women facing isolation

### 4. Mental Health Interventions

High-risk groups (married women, rural residents, less educated) should receive targeted mental health outreach:
- Community-based social programs
- Screening for social isolation and loneliness
- Culturally appropriate interventions

# Conclusion

Using India's 2024 Time Use Survey, we document a striking pattern: remote socializing is 39 times more common than in-person socializing. This reflects India's rapid digital transformation but also reveals profound inequalities.

Marriage reduces remote socializing more for women than men, suggesting that marriage constrains women's social autonomy. Urban residents socialize remotely at twice the rate of rural residents, and higher education increases participation by 80%, highlighting digital divides in social connectivity.

Our findings raise important questions about the future of social life in India and other developing countries undergoing digital transformation. As face-to-face interaction declines and digital communication expands, who is included—and excluded—from digital social networks? How does this shift affect mental health, community cohesion, and social capital?

Future research should examine:
1. The quality and mental health consequences of remote versus in-person socializing
2. Longitudinal trends in digital social connection
3. The role of specific platforms (WhatsApp, Facebook, Instagram) in shaping social networks
4. Comparative analysis across countries at different stages of digital transformation

Understanding the digital transformation of social life is crucial for policies promoting inclusive connectivity, mental health, and social wellbeing in the 21st century.

# References {-}

<div id="refs"></div>

# Appendix: Additional Tables and Figures

## Appendix Table A1: Full Regression Results

```{r appendix-full-reg, results='asis'}
modelsummary(
  list("Participation" = reg3, "Time Use" = time_reg3),
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  coef_omit = "State",
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  output = "kableExtra",
  title = "Full Regression Results with All Coefficients"
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

## Appendix Table A2: Summary Statistics by Gender

```{r appendix-summary-gender, results='asis'}
# Summary by gender
summary_by_gender <- person_data %>%
  group_by(gender_label) %>%
  summarise(
    N = n(),
    pct_remote = weighted.mean(participate_remote, Weight, na.rm = TRUE) * 100,
    pct_married = weighted.mean(married, Weight, na.rm = TRUE) * 100,
    pct_urban = weighted.mean(urban, Weight, na.rm = TRUE) * 100,
    mean_age = weighted.mean(Age, Weight, na.rm = TRUE),
    pct_higher_ed = weighted.mean(education_level == "Higher education", Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )

kable(summary_by_gender,
      caption = "Summary Statistics by Gender",
      digits = 2,
      booktabs = TRUE,
      escape = FALSE,
      col.names = c("Gender", "N", "Remote (\\%)", "Married (\\%)", "Urban (\\%)", "Mean Age", "Higher Ed (\\%)")) %>%
  kable_styling(latex_options = c("hold_position"))
```

## Appendix Figure A1: Distribution of Time Spent on Remote Socializing

```{r appendix-time-distribution, fig.cap="Distribution of Time Spent on Remote Socializing (Among Participants)", fig.height=5}
# Histogram of time spent
participants %>%
  filter(total_time_remote > 0, total_time_remote < 300) %>%
  ggplot(aes(x = total_time_remote)) +
  geom_histogram(binwidth = 10, fill = "#3498db", alpha = 0.7, color = "white") +
  geom_vline(xintercept = time_conditional$mean_remote,
             color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = time_conditional$mean_remote + 20, y = Inf,
           label = sprintf("Mean: %.0f min", time_conditional$mean_remote),
           vjust = 2, color = "red", fontface = "bold") +
  labs(title = "Distribution of Time Spent on Remote Socializing",
       subtitle = "Among participants only (N = 12,621)",
       x = "Minutes Spent on Remote Socializing",
       y = "Frequency") +
  theme_minimal()
```

---

**Session Info**

```{r session-info}
sessionInfo()
```

**Document compiled**: `r format(Sys.time(), '%B %d, %Y at %H:%M')`

**Note**: All analyses use 10% random sample with adjusted weights. Standard errors clustered at household level. Results are nationally representative when using survey weights.
