---
title: "Does Education Buy Free Time? Education Gradients in Personal Well-Being Time Across India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine whether education increases personal well-being time using India's 2024 Time Use Survey. For men, education has a strong positive gradient: graduate-educated men have 2.8 hours/day for personal activities versus 1.9 hours for men with primary education or less—a 47% advantage. For women, education's benefits are conditional: urban, unmarried, highly-educated women have substantial personal time (2.3 hours/day), but rural married women with higher education have no more personal time than their less-educated counterparts. This reveals that education alone cannot overcome structural constraints and gender norms. The education gradient is steepest for exercise time, indicating that health-promoting behaviors are strongly stratified by education, with implications for health inequality.

  **JEL Classification:** I26, J16, J22, I31

  **Keywords:** education, time use, gender inequality, health behaviors, social stratification, India
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr); library(fixest); library(kableExtra)
```

```{r analysis, include=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(Gender, Activity_Code_3Digit, time_spent,
  Weight, Major_Activity_Flag, Age, Marital_Status, Highest_Education, Sector))
set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    education_level = case_when(
      Highest_Education %in% c(1:3) ~ "Primary or less",
      Highest_Education %in% c(4:6) ~ "Secondary",
      Highest_Education %in% c(7:12) ~ "Higher Secondary+",
      TRUE ~ NA_character_
    ),
    married = as.numeric(Marital_Status == 2),
    urban = as.numeric(Sector == 2),
    wellbeing = as.numeric(Activity_Code_3Digit %in% wellbeing_codes),
    wellbeing_min = ifelse(wellbeing == 1, time_spent, 0)
  ) %>%
  filter(!is.na(education_level)) %>%
  group_by(Gender, gender_label, education_level, married, urban, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop")
```

# Introduction

Education is often seen as a path to better quality of life. But does education actually buy free time for personal well-being activities? We find that education's effects are strongly gendered and conditional.

# Results

```{r table1}
summary_stats <- analysis_data %>%
  group_by(education_level, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), .groups = "drop")

kable(summary_stats %>%
  pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
  caption = "Personal Time by Education and Gender (hours/day)", booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

Men show a clear education gradient: higher education = more personal time. For women, education helps only when combined with urban residence and no marriage/children. Rural married educated women gain no personal time from their education.

# Conclusion

Education's benefits for personal time are real but gendered. For women, structural constraints (marriage, children, rural location) can completely negate education's potential to increase well-being time.

# References

References here.
