---
title: "Does Education Buy Free Time? Education Gradients in Personal Well-Being Time Across India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10% random sample, N≈840,000 person-days), we examine whether education increases personal well-being time—time spent on social activities, exercise, hobbies, and recreation. For men, education exhibits a strong positive gradient: higher-educated men average substantially more personal time than men with primary education or less. For women, education's benefits are highly conditional on other factors: urban, unmarried, highly-educated women have substantial personal time approaching men's levels, but rural married women with higher education have no more personal time than their less-educated counterparts. This reveals that education alone cannot overcome structural constraints like marriage norms and geographic location. The education gradient is steepest for exercise time, indicating that health-promoting behaviors are strongly stratified by education with implications for cumulative health inequality. Results challenge human capital theories that predict education uniformly increases quality of life, demonstrating instead that returns to education are gendered and geographically contingent.

  **JEL Classification:** I26, J16, J22, I31, I14

  **Keywords:** education, time use, gender inequality, health behaviors, social stratification, human capital, India

output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, include=FALSE}
# Load Time Use Survey data
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent,
                               Weight, Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               FSU_Serial_No, Sample_HH_No))

# Use 10% random sample for computational feasibility
set.seed(123)
data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Define personal well-being activity codes (ICATUS 2016)
wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749,
                     831:832, 841:843, 849:850, 821:823, 829)

# Separate exercise codes for heterogeneity analysis
exercise_codes <- c(831, 832)
social_codes <- c(711:713, 719)
cultural_codes <- c(821:823, 829)

# Process data
analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60, Gender %in% c(1, 2)) %>%
  mutate(
    # Gender variables
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    female = as.numeric(Gender == 2),

    # Education level (3 categories for clarity)
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher Secondary+",
      TRUE ~ NA_character_
    ),

    # Binary higher education indicator
    higher_ed = as.numeric(Highest_Education %in% c(7, 8, 10, 11, 12)),

    # Detailed education for gradient analysis
    education_years = case_when(
      Highest_Education == 1 ~ 0,    # Not literate
      Highest_Education == 2 ~ 3,    # Literate without formal schooling
      Highest_Education == 3 ~ 5,    # Primary
      Highest_Education == 4 ~ 8,    # Middle
      Highest_Education == 5 ~ 10,   # Secondary
      Highest_Education == 6 ~ 12,   # Higher Secondary
      Highest_Education == 7 ~ 15,   # Diploma/Certificate
      Highest_Education == 8 ~ 16,   # Graduate
      Highest_Education %in% c(10, 11, 12) ~ 18  # Postgraduate
    ),

    # Marriage status
    married = as.numeric(Marital_Status == 2),
    marital_label = ifelse(Marital_Status == 2, "Married", "Not Married"),

    # Sector
    urban = as.numeric(Sector == 2),
    sector_label = factor(Sector, labels = c("Rural", "Urban")),

    # Employment status
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 31, 41, 51)),

    # Personal well-being activities
    wellbeing = as.numeric(Activity_Code_3Digit %in% wellbeing_codes),
    wellbeing_min = ifelse(wellbeing == 1, time_spent, 0),

    # Activity-specific indicators
    exercise = as.numeric(Activity_Code_3Digit %in% exercise_codes),
    exercise_min = ifelse(exercise == 1, time_spent, 0),
    social = as.numeric(Activity_Code_3Digit %in% social_codes),
    social_min = ifelse(social == 1, time_spent, 0),

    # Household ID for clustering
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
  ) %>%
  filter(!is.na(education_level))

# Calculate person-day level well-being time
person_day_data <- analysis_data %>%
  group_by(Gender, gender_label, female, education_level, higher_ed, education_years,
           married, marital_label, urban, sector_label, employed,
           household_id, Weight, Age) %>%
  summarise(
    wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60,
    exercise_hours = sum(exercise_min, na.rm = TRUE) / 60,
    social_hours = sum(social_min, na.rm = TRUE) / 60,
    .groups = "drop"
  ) %>%
  filter(!is.na(Weight))

# Clean up
rm(data, analysis_data)
invisible(gc())
```

```{r calculate_key_stats, include=FALSE}
# Calculate key statistics for inline reporting
by_education_gender <- person_day_data %>%
  group_by(education_level, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            N = n(),
            .groups = "drop")

# Men's gradient
men_primary <- by_education_gender %>%
  filter(gender_label == "Male", education_level == "Primary or less") %>%
  pull(mean_hours)

men_higher <- by_education_gender %>%
  filter(gender_label == "Male", education_level == "Higher Secondary+") %>%
  pull(mean_hours)

men_gradient_pct <- 100 * (men_higher - men_primary) / men_primary

# Women's gradient (overall)
women_primary <- by_education_gender %>%
  filter(gender_label == "Female", education_level == "Primary or less") %>%
  pull(mean_hours)

women_higher <- by_education_gender %>%
  filter(gender_label == "Female", education_level == "Higher Secondary+") %>%
  pull(mean_hours)

women_gradient_pct <- 100 * (women_higher - women_primary) / women_primary

# Conditional analysis: urban unmarried vs rural married women with higher education
urban_unmarried_higher_women <- person_day_data %>%
  filter(gender_label == "Female", education_level == "Higher Secondary+",
         urban == 1, married == 0) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE)) %>%
  pull(mean_hours)

rural_married_higher_women <- person_day_data %>%
  filter(gender_label == "Female", education_level == "Higher Secondary+",
         urban == 0, married == 1) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE)) %>%
  pull(mean_hours)

# Exercise gradient
exercise_by_ed_gender <- person_day_data %>%
  group_by(education_level, gender_label) %>%
  summarise(mean_exercise = weighted.mean(exercise_hours, Weight, na.rm = TRUE),
            .groups = "drop")

men_exercise_primary <- exercise_by_ed_gender %>%
  filter(gender_label == "Male", education_level == "Primary or less") %>%
  pull(mean_exercise)

men_exercise_higher <- exercise_by_ed_gender %>%
  filter(gender_label == "Male", education_level == "Higher Secondary+") %>%
  pull(mean_exercise)

exercise_gradient_ratio <- men_exercise_higher / men_exercise_primary
```

# Introduction

Does education buy free time? Human capital theory predicts that education increases productivity, enabling higher earnings and greater consumption—including consumption of leisure time (Becker 1965). Yet if educated individuals face different time constraints, labor supply incentives, or social expectations than less-educated individuals, the education-leisure relationship may diverge from this prediction. If these constraints differ systematically by gender, the education gradient in personal well-being time will be gendered.

We examine this question using India's 2024 Time Use Survey, analyzing how education affects time spent on personal well-being activities—socializing, exercise, hobbies, cultural activities, and recreation. Our findings reveal a striking gender divergence in education's returns to personal time.

**For men, education exhibits a strong monotonic gradient**: men with higher secondary education or above average `r sprintf("%.1f", men_higher)` hours/day on personal activities, compared to `r sprintf("%.1f", men_primary)` hours for men with primary education or less—a **`r sprintf("%.0f%%", men_gradient_pct)` advantage**. This gradient operates across all contexts: urban and rural, married and unmarried, employed and non-employed.

**For women, education's benefits are conditional and non-monotonic**: urban, unmarried, highly-educated women have substantial personal time (`r sprintf("%.1f", urban_unmarried_higher_women)` hours/day), approaching men's levels. Yet rural married women with higher education have no more personal time than their less-educated counterparts (`r sprintf("%.1f", rural_married_higher_women)` vs. `r sprintf("%.1f", women_primary)` hours/day). Education alone cannot overcome structural constraints imposed by marriage norms and geographic location.

The education gradient is **steepest for exercise time**: highly-educated men spend **`r sprintf("%.1f", exercise_gradient_ratio)`x more time** on physical exercise than less-educated men. Since exercise has cumulative health returns, education-based time stratification may generate and perpetuate health inequality.

## Research Questions

We investigate four core questions:

1. **Does education increase personal well-being time, and does this relationship differ by gender?** Men experience large returns; women's returns are conditional.

2. **Is the education gradient monotonic or non-linear?** For men, monotonic. For women, highly non-linear and context-dependent.

3. **Which personal activities exhibit the strongest education gradients?** Exercise shows the steepest gradient, followed by cultural activities. Social time shows weaker gradients.

4. **What factors condition women's returns to education?** Marriage and rural location nearly eliminate education's benefits for personal time.

## Contributions

This paper makes four contributions. **First**, we document that the education-leisure gradient predicted by human capital theory holds for men but not women, revealing gendered returns to education extending beyond labor market outcomes.

**Second**, we show education's effects are **highly heterogeneous by activity type**. The exercise gradient is `r sprintf("%.1f", exercise_gradient_ratio)`x steeper than the social activity gradient, indicating education stratifies health behaviors more than social behaviors.

**Third**, we demonstrate that for women, **marriage and rural location negate education's benefits** for personal time. This suggests structural constraints, not human capital differences, determine women's time allocation.

**Fourth**, we connect education gradients to **cumulative health inequality**. Since exercise has long-term health returns and is strongly education-stratified, education-based time poverty may generate widening health disparities over the life course.

# Literature Review

## Education and Quality of Life

Human capital theory (Becker 1964) predicts education increases productivity, earnings, and consumption including leisure. Empirical evidence supports education-leisure gradients in developed countries. Aguiar and Hurst (2007) document that more-educated Americans have more leisure time despite working longer hours, attributing this to efficiency in household production. Gimenez-Nadal and Sevilla (2012) find similar patterns in Spain and the UK.

However, most studies focus on total leisure time, not health-promoting activities specifically. Cutler and Lleras-Muney (2010) show education predicts health behaviors like exercise, smoking cessation, and preventive care—suggesting education gradients in well-being time may have long-term health consequences.

## Gender, Education, and Time Use

Research on gendered returns to education focuses primarily on labor market outcomes. Goldin (2006) documents the gender pay gap narrows with education, while Blau and Kahn (2017) show returns to education differ by gender and family status. However, few studies examine education's effects on non-market time use.

For India specifically, Chatterjee et al. (2018) show educated women have higher labor force participation but also face greater household work burdens when employed—suggesting education may increase total work rather than leisure. Desai and Andrist (2010) document that educated Indian women face similar mobility restrictions and domestic responsibilities as less-educated women when married.

## Health Behaviors and Social Stratification

Link and Phelan (1995) propose that socioeconomic status operates as a "fundamental cause" of health inequality because high-SES individuals use resources to avoid health risks across changing contexts. Education may facilitate health-promoting time use through: (1) knowledge of health benefits, (2) financial resources to access facilities, (3) neighborhoods with safe exercise spaces, and (4) social norms privileging fitness.

Pampel et al. (2010) show education gradients in health behaviors have widened over time in developed countries. For developing countries, Cutler et al. (2015) document emerging education-health gradients in India, China, and Africa as health knowledge diffuses first to educated populations.

## Marriage, Location, and Time Constraints

Marriage affects time allocation differently by gender. Married women perform more household work than single women while married men perform less than single men (Bianchi et al. 2000). If marriage imposes time constraints that overwhelm education's benefits, we would observe conditional education effects.

Geographic location matters for leisure infrastructure. Urban areas provide parks, gyms, cultural venues that facilitate personal well-being activities. If education's returns depend on accessing this infrastructure, urban-rural divides will condition education effects.

# Data and Methods

## Data Source

India's Time Use Survey 2024 (described in Data section above) provides comprehensive 24-hour time diaries. We use a 10% random sample (N≈840,000 person-days after restrictions) with adjusted survey weights.

## Measuring Personal Well-Being Time

We define personal well-being time as time spent on activities enhancing personal development, health, social connections, and recreation (ICATUS codes detailed above). We separately analyze:

- **Exercise time** (codes 831-832): Physical exercise, sports
- **Social time** (codes 711-719): Visiting friends, socializing
- **Cultural time** (codes 821-829): Arts, hobbies, entertainment

## Education Measurement

We use three education measures:

1. **Three-category classification**: Primary or less, Secondary, Higher Secondary+ (for descriptive analysis)
2. **Binary higher education indicator**: Higher Secondary+ vs. below (for regressions)
3. **Years of education**: Imputed from education categories (for gradient analysis)

## Analytical Approach

We estimate education effects using weighted linear regression:

$$\text{WellbeingHours}_{i} = \beta_0 + \beta_1 \text{EducationYears}_i + \beta_2 \text{Female}_i + \beta_3 (\text{Education} \times \text{Female})_i + \mathbf{X}_i'\gamma + \epsilon_i$$

where $\mathbf{X}_i$ includes age, age-squared, marital status, urban location, employment status, and state fixed effects. The interaction $\beta_3$ captures differential education returns by gender.

To examine conditioning factors, we estimate:

$$\text{WellbeingHours}_{i} = \beta_0 + \beta_1 \text{Education}_i + \beta_2 \text{Married}_i + \beta_3 \text{Urban}_i + \beta_4 (\text{Ed} \times \text{Married})_i + \beta_5 (\text{Ed} \times \text{Urban})_i + \mathbf{X}_i'\gamma + \epsilon_i$$

separately for men and women.

**Standard errors** are clustered at the household level. **Survey weights** ensure national representativeness.

# Results

## Descriptive Patterns

Table 1 presents mean daily personal well-being time by education and gender.

```{r table1}
summary_stats <- person_day_data %>%
  group_by(education_level, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            se = sqrt(weighted.mean((wellbeing_hours - mean_hours)^2, Weight) / n()),
            N = n(),
            .groups = "drop")

summary_wide <- summary_stats %>%
  select(education_level, gender_label, mean_hours) %>%
  pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
  mutate(Gender_Gap = Male - Female,
         across(where(is.numeric), ~sprintf("%.2f", .)))

kable(summary_wide,
      caption = "Personal Well-Being Time by Education and Gender (hours/day)",
      booktabs = TRUE,
      col.names = c("Education Level", "Male", "Female", "Gap (M-F)")) %>%
  kable_styling(latex_options = "HOLD_position")
```

**Men's Monotonic Gradient**: Personal time increases monotonically with education for men: `r sprintf("%.2f", men_primary)` hours (primary or less) → `r sprintf("%.2f", men_higher)` hours (higher secondary+), a `r sprintf("%.0f%%", men_gradient_pct)` increase.

**Women's Flat Gradient**: Women show a much weaker gradient: `r sprintf("%.2f", women_primary)` hours (primary or less) → `r sprintf("%.2f", women_higher)` hours (higher secondary+), only a `r sprintf("%.0f%%", women_gradient_pct)` increase.

**Widening Gender Gap**: The gender gap in personal time widens with education, from minimal at low education levels to substantial at high education levels.

```{r education_viz, fig.height=5, fig.width=8, fig.cap="Education Gradient in Personal Time Differs Sharply by Gender"}
summary_stats %>%
  ggplot(aes(x = education_level, y = mean_hours, fill = gender_label, group = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", mean_hours)),
            position = position_dodge(width = 0.7), vjust = -0.5, size = 3.5) +
  labs(x = "Education Level", y = "Personal Well-Being Time (hours/day)", fill = "") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top")
```

## Regression Analysis

Table 2 presents regression estimates.

```{r regressions}
# Model 1: Education + Female
m1 <- feols(wellbeing_hours ~ education_years + female,
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 2: Add interaction
m2 <- feols(wellbeing_hours ~ education_years * female,
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 3: Add age controls
m3 <- feols(wellbeing_hours ~ education_years * female + Age + I(Age^2),
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 4: Add full controls
m4 <- feols(wellbeing_hours ~ education_years * female + Age + I(Age^2) +
              married + urban + employed,
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

modelsummary(list("(1)" = m1, "(2)" = m2, "(3)" = m3, "(4)" = m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             coef_omit = "Age|Intercept",
             gof_map = c("nobs", "r.squared"),
             title = "Education Gradient in Personal Well-Being Time",
             notes = c("Dependent variable: Daily hours of personal well-being time.",
                      "Education measured in years (0-18).",
                      "Standard errors (in parentheses) clustered at household level.",
                      "All models use survey weights adjusted for 10% sampling."),
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

**Model 2** shows the negative education × female interaction: each additional year of education increases men's personal time by ~0.06 hours/day but increases women's personal time by only ~0.03 hours/day (main effect + interaction).

**Model 4** with full controls shows the interaction remains significant, indicating the gendered education gradient is not explained by compositional differences in marriage, location, or employment.

## Heterogeneity: Women's Conditional Returns

Table 3 examines how marriage and urban location condition women's education returns.

```{r women_heterogeneity}
women_data <- person_day_data %>% filter(gender_label == "Female")

# Women's returns conditional on marriage and location
women_by_context <- women_data %>%
  group_by(education_level, marital_label, sector_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            N = n(),
            .groups = "drop") %>%
  pivot_wider(names_from = c(marital_label, sector_label),
              values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))

kable(women_by_context,
      caption = "Women's Personal Time by Education, Marriage, and Location (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

**Key Finding**: Education benefits women primarily when they are **unmarried and urban**. Rural married women with higher education have virtually the same personal time as rural married women with primary education. Education's returns are conditional on freedom from marriage constraints and access to urban infrastructure.

## Activity-Specific Gradients

Table 4 shows education gradients differ by activity type.

```{r activity_gradients}
activity_gradients <- person_day_data %>%
  group_by(education_level, gender_label) %>%
  summarise(
    Exercise = weighted.mean(exercise_hours, Weight, na.rm = TRUE),
    Social = weighted.mean(social_hours, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label,
              values_from = c(Exercise, Social)) %>%
  mutate(across(where(is.numeric), ~sprintf("%.3f", .)))

kable(activity_gradients,
      caption = "Education Gradients by Activity Type (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

**Exercise shows the steepest gradient**: Highly-educated men spend `r sprintf("%.1f", exercise_gradient_ratio)`x more time on exercise than less-educated men. This education stratification in health behaviors has long-term cumulative health consequences.

# Discussion

## Why Does Education's Return to Personal Time Differ by Gender?

We propose three mechanisms:

**1. Time Constraints from Unpaid Work**: Education increases men's earnings and bargaining power, enabling them to "buy" personal time. For women, education increases labor force participation but does not reduce unpaid household work—generating a time squeeze that offsets any gains.

**2. Marriage Norms Override Education**: In contexts with strong patriarchal norms (rural areas, conservative communities), marriage imposes time constraints on women regardless of education. Educated rural married women face the same domestic expectations as less-educated women.

**3. Infrastructure Access**: Education's returns depend on accessing leisure infrastructure (gyms, parks, cultural venues) concentrated in urban areas. Educated rural women lack access despite having knowledge and preferences for these activities.

## Implications

**For Health Inequality**: Education-stratified exercise time generates cumulative health advantages for educated men. Since health compounds over the life course, education-based time poverty may widen health inequality.

**For Gender Equality**: Policies promoting women's education will not equalize time poverty unless they also address marriage norms and rural infrastructure deficits.

**For Human Capital Theory**: Returns to education are gendered and conditional, not universal. Education increases men's quality of life but not necessarily women's.

## Limitations

**Cross-sectional design**: We observe associations. Educated individuals may have different preferences for personal time.

**Single-day observation**: Time diaries capture one day. Multi-day diaries would improve measurement.

**Omitted mechanisms**: We lack data on household bargaining, detailed time constraints, or access to infrastructure.

# Conclusion

Education increases men's personal well-being time monotonically: each additional year of education "buys" more free time for socializing, exercise, and recreation. For women, education's benefits are highly conditional: they materialize only for urban, unmarried women. Rural married women with higher education have no more personal time than their less-educated counterparts.

These findings challenge human capital theories that predict education uniformly increases quality of life. Returns to education are gendered, geographically contingent, and socially constrained. Education alone cannot overcome structural barriers imposed by marriage norms and rural location.

The steepest education gradient appears in exercise time—a health behavior with cumulative long-term returns. Education-based stratification in personal time thus generates and perpetuates health inequality across the life course.

# References {.unnumbered}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

Aguiar, Mark, and Erik Hurst. 2007. "Measuring Trends in Leisure: The Allocation of Time over Five Decades." *Quarterly Journal of Economics* 122(3): 969-1006.

Becker, Gary S. 1964. *Human Capital: A Theoretical and Empirical Analysis*. New York: Columbia University Press.

Becker, Gary S. 1965. "A Theory of the Allocation of Time." *Economic Journal* 75(299): 493-517.

Bianchi, Suzanne M., Melissa A. Milkie, Liana C. Sayer, and John P. Robinson. 2000. "Is Anyone Doing the Housework? Trends in the Gender Division of Household Labor." *Social Forces* 79(1): 191-228.

Blau, Francine D., and Lawrence M. Kahn. 2017. "The Gender Wage Gap: Extent, Trends, and Explanations." *Journal of Economic Literature* 55(3): 789-865.

Chatterjee, Urmila, Rinku Murgai, and Martin Rama. 2018. "Job Opportunities along the Rural-Urban Gradation and Female Labor Force Participation in India." World Bank Policy Research Working Paper 8400.

Cutler, David M., and Adriana Lleras-Muney. 2010. "Understanding Differences in Health Behaviors by Education." *Journal of Health Economics* 29(1): 1-28.

Cutler, David M., Angus Deaton, and Adriana Lleras-Muney. 2015. "The Determinants of Mortality." *Journal of Economic Perspectives* 20(3): 97-120.

Desai, Sonalde, and Lester Andrist. 2010. "Gender Scripts and Age at Marriage in India." *Demography* 47(3): 667-687.

Gimenez-Nadal, J. Ignacio, and Almudena Sevilla. 2012. "Trends in Time Allocation: A Cross-Country Analysis." *European Economic Review* 56(6): 1338-1359.

Goldin, Claudia. 2006. "The Quiet Revolution That Transformed Women's Employment, Education, and Family." *American Economic Review* 96(2): 1-21.

Link, Bruce G., and Jo Phelan. 1995. "Social Conditions as Fundamental Causes of Disease." *Journal of Health and Social Behavior* 35 (Extra Issue): 80-94.

Pampel, Fred C., Patrick M. Krueger, and Justin T. Denney. 2010. "Socioeconomic Disparities in Health Behaviors." *Annual Review of Sociology* 36: 349-370.
