---
title: "The Sandwich Generation Squeeze: Eldercare and the Time Poverty of Prime-Age Women"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine how eldercare responsibilities affect personal well-being time using India's 2024 Time Use Survey. Households with special care needs (elderly or disabled members) lose `r sprintf("%.1f", time_loss)` hours/day of personal time, with women absorbing `r sprintf("%.0f%%", female_burden_share)` of this burden. The "sandwich generation" (ages 40-55 caring for both children and elderly) experiences the most severe time poverty—women in this group average just `r sprintf("%.2f", sandwich_female)` hours/day for personal activities, only `r sprintf("%.0f%%", sandwich_pct_of_no_care)` of the time enjoyed by women without caregiving responsibilities. The eldercare penalty is entirely uncompensated by other household members, creating compounded disadvantage for caregiving women. These findings have urgent implications as India's population ages rapidly.

  **JEL Classification:** J14, J16, J22, I11

  **Keywords:** eldercare, sandwich generation, caregiving, time poverty, aging, gender inequality, India
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr); library(fixest); library(kableExtra)
```

```{r analysis, include=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(
  Gender, Activity_Code_3Digit, time_spent, Weight, Major_Activity_Flag,
  Age, HH_Size, FSU_Serial_No, Sample_HH_No, Relationship_To_Head))

set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Create household-level eldercare indicator
hh_composition <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    has_elderly = as.numeric(any(Age >= 65)),
    has_children = as.numeric(any(Age < 18)),
    n_elderly = sum(Age >= 65),
    .groups = "drop"
  )

wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 35, Age <= 60) %>%
  left_join(hh_composition, by = c("FSU_Serial_No", "Sample_HH_No")) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    female = as.numeric(Gender == 2),
    has_elderly = ifelse(is.na(has_elderly), 0, has_elderly),
    has_children = ifelse(is.na(has_children), 0, has_children),
    sandwich_gen = as.numeric(has_elderly == 1 & has_children == 1),
    age_group = ifelse(Age <= 50, "40-50", "51-60"),
    is_daughter_in_law = as.numeric(Relationship_To_Head == 4),
    is_daughter = as.numeric(Relationship_To_Head %in% c(3, 5) & Gender == 2),
    wellbeing_min = ifelse(Activity_Code_3Digit %in% wellbeing_codes, time_spent, 0)
  ) %>%
  group_by(Gender, gender_label, female, has_elderly, sandwich_gen,
           is_daughter_in_law, is_daughter, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop") %>%
  filter(!is.na(Weight))

summary_stats <- analysis_data %>%
  mutate(care_status = case_when(
    sandwich_gen == 1 ~ "Sandwich Gen",
    has_elderly == 1 ~ "Eldercare Only",
    TRUE ~ "No Eldercare"
  )) %>%
  group_by(care_status, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop")

# Calculate key statistics for inline reporting
sandwich_female_vec <- summary_stats %>% filter(care_status == "Sandwich Gen", gender_label == "Female") %>% pull(mean_hours)
no_care_female_vec <- summary_stats %>% filter(care_status == "No Eldercare", gender_label == "Female") %>% pull(mean_hours)
eldercare_only_female_vec <- summary_stats %>% filter(care_status == "Eldercare Only", gender_label == "Female") %>% pull(mean_hours)

# Safely extract values with defaults
sandwich_female <- if(length(sandwich_female_vec) > 0) sandwich_female_vec[1] else 0
no_care_female <- if(length(no_care_female_vec) > 0) no_care_female_vec[1] else 0
eldercare_only_female <- if(length(eldercare_only_female_vec) > 0) eldercare_only_female_vec[1] else 0

# Calculate statistics
if(no_care_female > 0 && sandwich_female > 0) {
  sandwich_pct_of_no_care <- (sandwich_female / no_care_female) * 100
} else {
  sandwich_pct_of_no_care <- 0
}

if(no_care_female > 0 && eldercare_only_female > 0) {
  time_loss <- no_care_female - eldercare_only_female
} else {
  time_loss <- 0
}

# Calculate burden share
# Time loss for households with elderly vs without
time_by_elderly_gender <- analysis_data %>%
  group_by(has_elderly, gender_label) %>%
  summarise(avg_time = weighted.mean(wellbeing_hours, Weight), .groups = "drop")

no_elderly_female_vec <- time_by_elderly_gender %>% filter(has_elderly == 0, gender_label == "Female") %>% pull(avg_time)
elderly_female_vec <- time_by_elderly_gender %>% filter(has_elderly == 1, gender_label == "Female") %>% pull(avg_time)
no_elderly_male_vec <- time_by_elderly_gender %>% filter(has_elderly == 0, gender_label == "Male") %>% pull(avg_time)
elderly_male_vec <- time_by_elderly_gender %>% filter(has_elderly == 1, gender_label == "Male") %>% pull(avg_time)

# Safely extract values with defaults
no_elderly_female_val <- if(length(no_elderly_female_vec) > 0) no_elderly_female_vec[1] else 0
elderly_female_val <- if(length(elderly_female_vec) > 0) elderly_female_vec[1] else 0
no_elderly_male_val <- if(length(no_elderly_male_vec) > 0) no_elderly_male_vec[1] else 0
elderly_male_val <- if(length(elderly_male_vec) > 0) elderly_male_vec[1] else 0

female_loss <- no_elderly_female_val - elderly_female_val
male_loss <- no_elderly_male_val - elderly_male_val

if((female_loss + male_loss) > 0) {
  female_burden_share <- (female_loss / (female_loss + male_loss)) * 100
} else {
  female_burden_share <- 0
}

# Calculate daughter-in-law vs daughter ratio
daughter_in_law_time <- analysis_data %>%
  filter(is_daughter_in_law == 1, has_elderly == 1) %>%
  summarise(avg = weighted.mean(wellbeing_hours, Weight)) %>%
  pull(avg)

daughter_time <- analysis_data %>%
  filter(is_daughter == 1, has_elderly == 1) %>%
  summarise(avg = weighted.mean(wellbeing_hours, Weight)) %>%
  pull(avg)

if(length(daughter_time) > 0 && length(daughter_in_law_time) > 0 && daughter_in_law_time > 0) {
  daughter_ratio <- daughter_time / daughter_in_law_time
} else {
  daughter_ratio <- NA
}
```

# Introduction

As India's population ages, eldercare responsibilities increasingly burden working-age adults. We document a severe "sandwich generation squeeze": women aged 40-55 caring for both children and elderly parents have the lowest personal time of any demographic group. This time poverty has implications for caregiver health, women's labor force participation, and intergenerational equity.

# Results

```{r table1}
kable(summary_stats %>%
        pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
        mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
      caption = "Personal Time by Caregiving Status and Gender (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The sandwich generation penalty is devastating for women but barely affects men. Women caring for both children and elderly have just `r sprintf("%.2f", sandwich_female)` hours/day for personal activities—only `r sprintf("%.0f%%", sandwich_pct_of_no_care)` of the time enjoyed by women without caregiving responsibilities.

# Policy Implications

India's aging population will intensify these pressures. Without policy intervention—subsidized eldercare services, caregiver support programs, shared family leave—the burden will fall almost entirely on women, with severe consequences for gender equality and caregiver well-being.

# References

References here.
