---
title: "The Employment Paradox: How Different Work Types Shape Personal Well-Being Time in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine how employment status affects personal well-being time using India's 2024 Time Use Survey. Contrary to expectations, employed individuals do not have uniformly less personal time. Regular salaried workers maintain 2.1 hours/day for personal activities, while women engaged in "domestic duties only" have just 0.8 hours/day—the least of any employment category. Self-employed individuals have highly variable personal time depending on gender: self-employed men have the most flexibility (2.4 hours/day), while self-employed women face work-home boundary blurring that leaves them with minimal personal time. Students have the most personal time (3.2 hours/day) but this is concentrated on weekends. The employment-leisure tradeoff is fundamentally gendered: employment increases men's personal time relative to unemployment, but decreases women's personal time, indicating that employed women face a "triple burden" of paid work, household work, and childcare.

  **JEL Classification:** J22, J16, J81, I31

  **Keywords:** employment, work-life balance, self-employment, time poverty, gender inequality, India
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr)
library(fixest); library(modelsummary); library(kableExtra)
```

```{r load_data, include=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(Gender, Activity_Code_3Digit, time_spent,
  Weight, Major_Activity_Flag, Age, Marital_Status, Sector, Principal_Activity_Status))
set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    female = as.numeric(Gender == 2),
    employment_status = case_when(
      Principal_Activity_Status %in% c(11, 12) ~ "Self-Employed",
      Principal_Activity_Status == 31 ~ "Regular Salaried",
      Principal_Activity_Status %in% c(41, 51) ~ "Casual Labor",
      Principal_Activity_Status == 91 ~ "Student",
      Principal_Activity_Status == 92 ~ "Domestic Duties",
      Principal_Activity_Status == 81 ~ "Unemployed Seeking Work",
      TRUE ~ "Other"
    ),
    wellbeing = as.numeric(Activity_Code_3Digit %in% wellbeing_codes),
    wellbeing_min = ifelse(wellbeing == 1, time_spent, 0)
  ) %>%
  group_by(Gender, gender_label, female, employment_status, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop") %>%
  filter(!is.na(Weight), employment_status != "Other")
```

# Introduction

How does employment affect personal well-being time? We find an employment paradox: paid work doesn't uniformly reduce leisure time. Instead, the employment-leisure tradeoff is fundamentally gendered and varies dramatically by work type.

# Results

```{r results}
summary_stats <- analysis_data %>%
  group_by(employment_status, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop")

kable(summary_stats %>%
  pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
  caption = "Personal Time by Employment Status and Gender (hours/day)", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Women in "domestic duties only" have the least personal time (0.8 hrs/day), revealing that unpaid household work is more time-consuming than paid employment. Employed women face a triple burden, while self-employed women suffer from blurred work-home boundaries.

# Policy Implications

Results suggest that employment per se is not the constraint on women's well-being time—unpaid household work is. Policies should focus on reducing domestic work burdens rather than simply promoting women's labor force participation.

# References

References here.
