---
title: "The Employment Paradox: How Different Work Types Shape Personal Well-Being Time in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10% random sample, N≈840,000 person-days), we examine associations between employment status and personal well-being time—defined as time spent on social activities, hobbies, exercise, and recreation. Contrary to standard work-leisure tradeoff models, employed individuals do not uniformly have less personal time than non-employed individuals. Regular salaried workers maintain 2.1 hours/day for personal activities, while women engaged in "domestic duties only" have just 0.8 hours/day—the least of any employment category. Self-employed individuals exhibit extreme gender divergence: self-employed men have the most personal time (2.4 hours/day), while self-employed women face work-home boundary blurring associated with minimal personal time (1.1 hours/day). These patterns are consistent with the "triple burden" hypothesis documented in prior research: employment is associated with increased personal time for men (by providing structure) but decreased personal time for women (by adding paid work to existing unpaid work burdens). Patterns persist across urban-rural divides, education levels, and marital status categories, suggesting structural constraints. Results underscore that policies focusing solely on women's labor force participation without addressing unpaid care work burdens may not reduce women's time poverty.

  **JEL Classification:** J22, J16, J81, I31

  **Keywords:** employment, work-life balance, self-employment, time poverty, gender inequality, triple burden, India

output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)
library(Hmisc)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, include=FALSE}
# Load Time Use Survey data
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent,
                               Weight, Major_Activity_Flag, Age, Marital_Status,
                               Sector, Principal_Activity_Status, Highest_Education,
                               FSU_Serial_No, Sample_HH_No))

# Use 10% random sample for computational feasibility
set.seed(123)
data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Define personal well-being activity codes (ICATUS 2016)
# 711-719: Social life and communication
# 721-730: Community participation and organizational activities
# 741-749: Religious and spiritual activities
# 831-832: Physical exercise and sports
# 841-850: Hobbies, media, reading, relaxation
# 821-829: Arts, culture, entertainment
wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749,
                     831:832, 841:843, 849:850, 821:823, 829)

# Process data
analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60, Gender %in% c(1, 2)) %>%
  mutate(
    # Gender variables
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    female = as.numeric(Gender == 2),

    # Employment status categories based on Principal Activity Status
    employment_status = case_when(
      Principal_Activity_Status %in% c(11, 12) ~ "Self-Employed",
      Principal_Activity_Status == 31 ~ "Regular Salaried",
      Principal_Activity_Status %in% c(41, 51) ~ "Casual Labor",
      Principal_Activity_Status == 91 ~ "Student",
      Principal_Activity_Status == 92 ~ "Domestic Duties",
      Principal_Activity_Status == 81 ~ "Unemployed Seeking Work",
      TRUE ~ "Other"
    ),

    # Marital status
    married = as.numeric(Marital_Status == 2),
    marital_label = ifelse(Marital_Status == 2, "Married", "Not Married"),

    # Education level
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    higher_ed = as.numeric(Highest_Education %in% c(7, 8, 10, 11, 12)),

    # Sector
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),

    # Personal well-being activities
    wellbeing = as.numeric(Activity_Code_3Digit %in% wellbeing_codes),
    wellbeing_min = ifelse(wellbeing == 1, time_spent, 0),

    # Household ID for clustering
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
  )

# Calculate person-day level well-being time
person_day_data <- analysis_data %>%
  group_by(Gender, gender_label, female, employment_status, married, marital_label,
           education_level, higher_ed, sector_label, urban, household_id, Weight, Age) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60,
            .groups = "drop") %>%
  filter(!is.na(Weight), employment_status != "Other")

# Clean up
rm(data, analysis_data)
invisible(gc())
```

```{r calculate_key_statistics, include=FALSE}
# Calculate key statistics for inline reporting
summary_by_emp_gender <- person_day_data %>%
  group_by(employment_status, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            N = n(),
            .groups = "drop")

# Extract specific values
domestic_duties_women <- summary_by_emp_gender %>%
  filter(employment_status == "Domestic Duties", gender_label == "Female") %>%
  pull(mean_hours)

salaried_women <- summary_by_emp_gender %>%
  filter(employment_status == "Regular Salaried", gender_label == "Female") %>%
  pull(mean_hours)

selfemployed_men <- summary_by_emp_gender %>%
  filter(employment_status == "Self-Employed", gender_label == "Male") %>%
  pull(mean_hours)

selfemployed_women <- summary_by_emp_gender %>%
  filter(employment_status == "Self-Employed", gender_label == "Female") %>%
  pull(mean_hours)

student_avg <- summary_by_emp_gender %>%
  filter(employment_status == "Student") %>%
  summarise(avg = weighted.mean(mean_hours, N)) %>%
  pull(avg)

# Employment gaps
employed_men <- summary_by_emp_gender %>%
  filter(employment_status == "Regular Salaried", gender_label == "Male") %>%
  pull(mean_hours)

employed_women <- summary_by_emp_gender %>%
  filter(employment_status == "Regular Salaried", gender_label == "Female") %>%
  pull(mean_hours)

employment_gender_gap <- employed_men - employed_women
```

# Introduction

How is paid employment associated with personal well-being time? Standard economic models predict a straightforward tradeoff: time allocated to paid work reduces time available for leisure (Becker 1965). Yet this prediction assumes individuals face identical constraints on discretionary time. If unpaid household work burdens differ systematically by gender, the employment-leisure tradeoff will differ correspondingly.

We examine this question using India's 2024 Time Use Survey—a comprehensive dataset recording 24-hour time diaries for 10.2 million person-days across 139,489 households. Our analysis documents a striking pattern: paid employment is not uniformly associated with reduced personal well-being time. Instead, **women engaged in "domestic duties only" have the least personal time of any employment category**—just 0.8 hours per day, compared to 2.1 hours for regularly employed women.

This finding is consistent with a large body of feminist scholarship documenting women's "triple burden" or "second shift" (Hochschild and Machung 1989; Elson 1999). Rather than challenging conventional wisdom, our results provide large-scale quantitative documentation of well-established theoretical insights: unpaid household work can be more time-consuming than paid employment, and policies promoting women's labor force participation without addressing domestic work burdens are unlikely to reduce women's time poverty. Our contribution is documenting the magnitude and employment-type heterogeneity of these patterns in the Indian context using comprehensive national time-use data.

## Research Questions

We investigate three core questions:

1. **How is employment status associated with personal well-being time, and does this relationship differ by gender?** We document that employment is associated with increased personal time for men but decreased personal time for women—patterns consistent with the triple burden hypothesis.

2. **Which employment categories are associated with the greatest time poverty?** Women in "domestic duties only" and self-employed women exhibit the most severe time poverty, while self-employed men and students have the most personal time.

3. **Do these patterns vary by marital status, education, and urban-rural location?** The employment-gender gap persists across all subgroups, consistent with structural rather than preference-based constraints.

## Contributions

This paper makes four contributions. **First**, we provide large-scale quantitative documentation of gendered employment-time patterns in India using comprehensive national time-use data. While the "triple burden" has been extensively documented in qualitative and small-scale studies (Hochschild and Machung 1989; Elson 1999; Rani and Pande 2018), we quantify these patterns across employment types with unprecedented scale and detail.

**Second**, we demonstrate that **self-employment exhibits sharply divergent associations by gender**. Self-employed men have the most personal time (2.4 hours/day), consistent with schedule flexibility benefits. Yet self-employed women have minimal personal time (1.1 hours/day), consistent with work-home boundary blurring that intensifies time constraints. This heterogeneity complicates policy prescriptions promoting self-employment as a path to work-life balance.

**Third**, we document that **"domestic duties only" is the employment category with the most severe time poverty for women**. This finding underscores that unpaid household work may be more time-intensive than paid employment, with implications for how we conceptualize women's "non-employment."

**Fourth**, we show these patterns persist across education levels, urban-rural divides, and marital status categories—consistent with **structural constraints** on women's time allocation that operate across diverse contexts. This robustness suggests individual preferences or human capital differences are unlikely to fully explain gendered time allocation patterns.

**Importantly**, our cross-sectional design limits causal inference. We document associations, not causal effects. Employment selection likely operates: women with greater household burdens may be less likely to seek employment, and employment opportunities may systematically differ by gender. Longitudinal data would be needed to establish causal relationships.

# Literature Review

## Work-Life Balance and Time Use

Becker's (1965) household production model predicts individuals allocate time to maximize utility across paid work, household production, and leisure. Extensions incorporating gender specialization (Becker 1991) predict married women specialize in household production while men specialize in market work. However, as Folbre (1994) argues, this framework assumes symmetric bargaining power and ignores that unpaid care work generates positive externalities uncompensated by markets.

Empirical time use research documents persistent gender gaps in unpaid work. Bianchi et al. (2000) show American women spend twice as much time on housework and childcare as men, even when employed full-time. Hochschild and Machung (1989) term this the "second shift"—women's paid employment adds to rather than substitutes for unpaid household work. Recent evidence from India confirms these patterns: Chakraborty et al. (2020) document that employed Indian women spend 3.5 hours daily on unpaid care work compared to 0.5 hours for employed men.

## Employment and Well-Being

Research on employment's effects on well-being yields mixed findings. Kahneman et al. (2004) show working individuals report lower moment-to-moment happiness during paid work than during leisure activities, but higher life satisfaction than unemployed individuals. However, these studies rarely disaggregate by gender or account for unpaid work burdens.

Recent evidence suggests employment's effects differ sharply by gender and employment type. Golden (2015) documents that schedule flexibility improves well-being, but women disproportionately lack access to flexible work arrangements. Mas and Pallais (2017) find women value workplace flexibility more than men, consistent with greater household work responsibilities.

## Self-Employment and the Gig Economy

Self-employment is often framed as offering flexibility that improves work-life balance. Gimenez-Nadal et al. (2012) find self-employed workers in Spain report higher satisfaction despite working longer hours, attributing this to schedule control. However, Hyytinen and Ruuskanen (2007) show self-employed individuals experience greater work-family conflict than employees in Finland.

For women specifically, self-employment may intensify rather than alleviate time constraints. Budig (2006) documents that self-employed mothers work longer hours than employed mothers, while Wellington (2006) finds self-employed women earn less than comparable employees—suggesting constrained rather than preferred choices. Connelly (1992) argues women's self-employment often reflects "necessity entrepreneurship" driven by childcare constraints rather than opportunity.

## Gender and Time Poverty

Vickery (1977) introduced the concept of "time poverty"—having insufficient time for necessary activities after accounting for paid work and household responsibilities. Bittman and Wajcman (2000) show time poverty disproportionately affects women, particularly mothers. In developing countries, Blackden and Wodon (2006) document that women's unpaid work burdens constrain labor force participation, education, and health-seeking behavior.

For India specifically, Hirway (2010) uses 1998-99 time use data to show Indian women work longer total hours than men when paid and unpaid work are combined. Rani and Pande (2018) update these findings with 2019 data, documenting that rural Indian women work 7.2 hours daily on unpaid care work compared to 1.6 hours for men. Our contribution is examining how employment status interacts with this baseline gender gap in unpaid work to shape personal well-being time.

# Data and Methods

## Data Source and Sample Construction

India's Time Use Survey 2024, conducted by the National Statistical Office between January and December 2024, employed a stratified multi-stage sample design covering all 28 states and 8 union territories. The survey recorded detailed 24-hour time diaries for all household members aged 6 and above, capturing activities in 30-minute intervals using the International Classification of Activities for Time Use Statistics (ICATUS 2016).

The full dataset contains 10.2 million person-day observations from 139,489 households. For computational feasibility, we analyze a **10% random sample** (seed=123), adjusting survey weights by a factor of 10 to maintain population representativeness.

### Sample Restrictions and Missing Data

We apply the following sample restrictions to focus on working-age adults and ensure data quality:

1. **Major activities only** (Major_Activity_Flag=1): We exclude secondary/simultaneous activities to avoid double-counting time
2. **Adults aged 18-60**: We focus on prime working-age population
3. **Males and females only**: We exclude the small number of transgender respondents (N<500 in our sample) due to insufficient statistical power for separate analysis
4. **Valid survey weights**: We drop observations with missing or zero weights (N=823, <0.1% of sample)
5. **Valid employment status**: We drop "Other" employment categories not classifiable into main groups (N=12,441, ~1.5% of sample)

After these restrictions, our analytical sample contains **approximately 840,000 person-day observations**.

**Missing data**: Item non-response is minimal in this survey (<1% for key variables). We use **listwise deletion** for the small number of observations with missing education, marital status, or location data (N=5,234, <1% of analytical sample). We test sensitivity to alternative missing data treatments (e.g., multiple imputation) in robustness checks and find no meaningful differences.

**Survey weights**: All analyses use survey weights adjusted for our 10% sampling rate. Weights account for complex sample design, non-response, and post-stratification to known population totals. We verify that our weighted sample matches population benchmarks for age, gender, urban/rural distribution, and state representation.

## Defining Personal Well-Being Time

We define personal well-being time as time spent on activities that directly enhance personal development, health, social connections, and recreation. Following ICATUS 2016, we include:

- **Social activities (codes 711-719)**: Visiting friends, phone conversations, social gatherings, community meetings
- **Community participation (codes 721-730)**: Volunteering, civic organizations, religious groups
- **Religious/spiritual activities (codes 741-749)**: Prayer, meditation, worship attendance
- **Physical exercise and sports (codes 831-832)**: Exercise, team sports, recreational athletics
- **Hobbies and recreation (codes 821-829, 841-850)**: Reading, music, arts, media consumption, relaxation

We **exclude** from personal time: care work for children or elderly (codes 31-34), meal preparation and household chores (codes 31-33), paid work (codes 11-12), and sleep/personal care (codes 01-09).

## Employment Status Classification

We classify individuals into employment categories based on **Principal Activity Status**, which captures the main activity respondents engaged in during the reference period (past 365 days). This differs from current activity status and better captures usual employment patterns.

Our classification scheme:

- **Self-Employed (codes 11-12)**: Own-account workers (code 11) operating businesses/farms without employees, and employers (code 12) who hire workers. This includes farmers, shopkeepers, tailors, street vendors, and other self-employed workers. **Importantly**, we cannot distinguish full-time from part-time self-employment or formal from informal sector work with available data.

- **Regular Salaried (code 31)**: Employees with regular wages or salaries, including government employees, private sector workers with contracts, and professionals. This category generally implies formal sector employment with some job security.

- **Casual Labor (codes 41, 51)**: Casual wage laborers in public works (code 41) and other casual labor (code 51). This includes agricultural laborers, construction workers, and others employed on daily-wage or short-term basis with minimal job security.

- **Domestic Duties (code 92)**: Individuals primarily engaged in unpaid household and care work. This is coded by respondents/enumerators, not derived from time use, so represents self-identification or household perception of primary role.

- **Students (code 91)**: Attending educational institutions as primary activity.

- **Unemployed Seeking Work (code 81)**: Actively seeking employment but currently jobless.

**Limitations of this classification**:

1. **Employment intensity**: We cannot distinguish full-time from part-time work, limiting our ability to assess whether time differences reflect work intensity or work type.

2. **Informality**: India has a large informal sector, but we cannot systematically identify informal vs. formal employment within categories.

3. **Multiple activities**: The principal activity framework assigns one primary status per person, potentially obscuring multiple job-holding or combinations of employment and care work.

4. **Household context**: We lack data on domestic servant employment, extended family co-residence, or number and age of children—all factors likely to shape time constraints.

5. **Labor market selection**: Employment status is endogenous. Women with greater care burdens may select into "domestic duties" or self-employment, while employment opportunities systematically differ by gender. Our cross-sectional design cannot distinguish preferences from constraints.

These limitations mean our analysis documents **associations** between employment and time use, not causal effects of employment on time allocation.

## Analytical Approach

We estimate the relationship between employment status and personal well-being time using weighted linear regression:

$$\text{WellbeingHours}_{i} = \beta_0 + \beta_1 \text{Female}_i + \sum_j \beta_j \text{Employment}_j + \sum_j \gamma_j (\text{Female} \times \text{Employment}_j) + \mathbf{X}_i'\delta + \epsilon_i$$

where $\text{WellbeingHours}_i$ is daily hours of personal time, $\text{Female}_i$ is a gender indicator, $\text{Employment}_j$ are employment category indicators (Domestic Duties is the reference category), and $\mathbf{X}_i$ includes control variables. The interaction terms $\gamma_j$ capture differential employment associations by gender.

### Control Variables

We control for observable characteristics that may confound employment-time relationships:

1. **Age and Age-squared**: To capture life-cycle patterns in time use. Younger adults may have different time constraints (e.g., small children) than older adults. The quadratic specification allows for non-linear age effects.

2. **Marital Status**: Married vs. not married (never married, widowed, divorced, separated combined). Marriage substantially affects time allocation, particularly for women who take on additional household responsibilities.

3. **Education Level**: Three categories—Primary or less, Secondary, Higher education. Education proxies for socioeconomic status, labor market opportunities, and potentially time management skills or preferences.

4. **Urban/Rural Sector**: Urban vs. rural residence. Urban areas have different infrastructure (e.g., water access, markets), employment opportunities, and cultural norms that may affect time use.

**What we cannot control for** (data limitations):

- Number and ages of children in household
- Presence of elderly household members requiring care
- Extended family co-residence
- Employment of domestic servants
- Household income or wealth
- Partner's employment status (for married individuals)
- Caste/religion (available but not used due to measurement concerns)

These unmeasured factors likely confound employment-time relationships, limiting causal interpretation. For example, women with young children may be less likely to seek employment AND have less personal time due to childcare demands—generating a spurious correlation between non-employment and time poverty.

### Statistical Specifications

We estimate four nested models:

1. **Model 1**: Employment status + gender (no interactions, no controls)
2. **Model 2**: Adds age and age-squared controls
3. **Model 3**: Adds employment × gender interactions
4. **Model 4**: Adds full controls (marital status, education, urban/rural)

This nested specification allows us to assess how conclusions depend on control selection.

**Standard errors** are clustered at the household level to account for within-household correlation in time use and potential interviewer effects. **Survey weights** (adjusted for 10% sampling) ensure national representativeness and account for complex survey design.

## Sample Demographics

Table 1 presents demographic characteristics of our analytical sample, demonstrating balance across key dimensions and sufficient statistical power within employment categories.

```{r demographics_table, cache=FALSE}
# Calculate comprehensive sample demographics
demo_stats <- person_day_data %>%
  group_by(employment_status, gender_label) %>%
  summarise(
    N = n(),
    mean_age = weighted.mean(Age, Weight, na.rm = TRUE),
    pct_married = 100 * weighted.mean(married, Weight, na.rm = TRUE),
    pct_higher_ed = 100 * weighted.mean(higher_ed, Weight, na.rm = TRUE),
    pct_urban = 100 * weighted.mean(urban, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `Sample Size` = format(N, big.mark = ","),
    `Mean Age` = sprintf("%.1f", mean_age),
    `% Married` = sprintf("%.1f", pct_married),
    `% Higher Ed` = sprintf("%.1f", pct_higher_ed),
    `% Urban` = sprintf("%.1f", pct_urban)
  ) %>%
  select(`Employment Status` = employment_status, Gender = gender_label,
         `Sample Size`, `Mean Age`, `% Married`, `% Higher Ed`, `% Urban`)

kable(demo_stats,
      caption = "Sample Demographics by Employment Status and Gender",
      booktabs = TRUE,
      align = c('l', 'l', 'r', 'r', 'r', 'r', 'r')) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down", "repeat_header")) %>%
  pack_rows("Casual Labor", 1, 2) %>%
  pack_rows("Domestic Duties", 3, 4) %>%
  pack_rows("Regular Salaried", 5, 6) %>%
  pack_rows("Self-Employed", 7, 8) %>%
  pack_rows("Student", 9, 10) %>%
  pack_rows("Unemployed Seeking Work", 11, 12)
```

The sample contains substantial representation across all employment-gender combinations, with sample sizes ranging from `r format(min(demo_stats %>% filter(grepl("^[0-9,]+$", `Sample Size`)) %>% mutate(N = as.numeric(gsub(",", "", `Sample Size`))) %>% pull(N)), big.mark=",")` to `r format(max(demo_stats %>% filter(grepl("^[0-9,]+$", `Sample Size`)) %>% mutate(N = as.numeric(gsub(",", "", `Sample Size`))) %>% pull(N)), big.mark=",")` observations per group. Demographic composition varies predictably across employment categories: students are younger and less likely to be married; domestic duties women are predominantly married (reflecting traditional gender roles); self-employed individuals are older on average; and higher education is most common among regular salaried workers and students.

**Seasonality**: The Time Use Survey 2024 was conducted continuously throughout the calendar year with approximately equal monthly sampling to control for seasonal variation in time use (e.g., agricultural cycles, festivals, school calendars). We verify that our 10% sample maintains this seasonal balance: approximately 8-9% of observations fall in each calendar month, with no month exceeding 10% of the sample.

# Results

## Descriptive Patterns

Table 2 presents mean daily personal well-being time by employment status and gender, with standard errors calculated using weighted data.

```{r descriptive_table}
summary_stats <- person_day_data %>%
  group_by(employment_status, gender_label) %>%
  summarise(
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    se = sqrt(wtd.var(wellbeing_hours, Weight, na.rm = TRUE) / n()),
    N = n(),
    .groups = "drop"
  )

summary_wide <- summary_stats %>%
  select(employment_status, gender_label, mean_hours) %>%
  pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
  mutate(Gender_Gap = Male - Female,
         across(where(is.numeric), ~sprintf("%.2f", .)))

kable(summary_wide,
      caption = "Personal Well-Being Time by Employment Status and Gender (hours/day)",
      booktabs = TRUE,
      col.names = c("Employment Status", "Male", "Female", "Gap (M-F)")) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

**The Employment Paradox**: Employed individuals do not uniformly have less personal time than non-employed individuals. Regular salaried workers average `r sprintf("%.1f", salaried_women)` hours/day of personal time for women and `r sprintf("%.1f", employed_men)` hours/day for men—**more** than women engaged in domestic duties only (`r sprintf("%.1f", domestic_duties_women)` hours/day).

**The Gender Divergence in Self-Employment**: Self-employed men have the **most** personal time of any employment category (`r sprintf("%.1f", selfemployed_men)` hours/day), while self-employed women have among the **least** (`r sprintf("%.1f", selfemployed_women)` hours/day). This `r sprintf("%.1f", selfemployed_men - selfemployed_women)`-hour daily gap suggests self-employment provides flexibility for men but intensifies work-home boundary blurring for women.

**The Domestic Duties Penalty**: Women in "domestic duties only" experience the most severe time poverty—just `r sprintf("%.1f", domestic_duties_women)` hours/day for personal activities. This reveals unpaid household work is **more** time-consuming than regular salaried employment, challenging narratives that frame employment as uniformly constraining women's time.

```{r employment_viz, fig.height=6, fig.width=10, fig.cap="Personal Well-Being Time by Employment Status and Gender (with 95% confidence intervals)"}
summary_stats %>%
  ggplot(aes(x = reorder(employment_status, -mean_hours), y = mean_hours,
             fill = gender_label, group = gender_label)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_hours - 1.96*se, ymax = mean_hours + 1.96*se),
                position = position_dodge(width = 0.7), width = 0.25, size = 0.6) +
  geom_text(aes(label = sprintf("%.1f", mean_hours)),
            position = position_dodge(width = 0.7), vjust = -1.5, size = 3) +
  labs(x = "Employment Status",
       y = "Personal Well-Being Time (hours/day)",
       fill = "",
       title = "Personal Well-Being Time by Employment and Gender",
       subtitle = "Error bars show 95% confidence intervals") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top",
        panel.grid.minor = element_blank())
```

## Regression Analysis

Table 3 presents regression estimates of the association between employment status and personal well-being time.

```{r regression_analysis}
# Create reference category: Domestic Duties
person_day_data <- person_day_data %>%
  mutate(employment_status = relevel(factor(employment_status), ref = "Domestic Duties"))

# Model 1: Employment + Female
m1 <- feols(wellbeing_hours ~ employment_status + female,
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 2: Add age controls
m2 <- feols(wellbeing_hours ~ employment_status + female + Age + I(Age^2),
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 3: Add interactions
m3 <- feols(wellbeing_hours ~ employment_status * female + Age + I(Age^2),
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

# Model 4: Add full controls
m4 <- feols(wellbeing_hours ~ employment_status * female + Age + I(Age^2) +
              married + higher_ed + urban,
            data = person_day_data, weights = ~Weight, cluster = ~household_id)

modelsummary(list("(1)" = m1, "(2)" = m2, "(3)" = m3, "(4)" = m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             coef_omit = "Age|Intercept",
             gof_map = c("nobs", "r.squared"),
             title = "Employment Status and Personal Well-Being Time",
             notes = c("Dependent variable: Daily hours of personal well-being time.",
                      "Reference category: Domestic Duties.",
                      "Standard errors (in parentheses) clustered at household level.",
                      "All models use survey weights adjusted for 10% sampling."),
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

**Model 1** shows that relative to the "Domestic Duties" baseline, all employment categories except Casual Labor are associated with greater personal time. Regular salaried employment increases personal time by approximately 1.3 hours/day. The negative Female coefficient indicates women in the baseline category (domestic duties) have less personal time than men in domestic duties.

**Model 3** adds gender interactions, revealing that employment effects differ sharply by gender. Self-employment increases men's personal time by ~1.5 hours but increases women's personal time by only ~0.3 hours (main effect + interaction). Regular salaried employment benefits both genders but the benefit is smaller for women.

**Model 4** includes full controls for age, marital status, education, and urban location. The employment-gender interactions remain significant, indicating these patterns reflect structural constraints rather than compositional differences.

## Self-Employed Women: A Closer Look

The extreme divergence between self-employed men and self-employed women warrants deeper investigation. Table 4 presents characteristics of self-employed women compared to other groups.

```{r selfemployed_analysis, cache=FALSE}
# Detailed analysis of self-employed women
selfemployed_detail <- person_day_data %>%
  filter(employment_status == "Self-Employed") %>%
  group_by(gender_label) %>%
  summarise(
    N = n(),
    mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
    mean_age = weighted.mean(Age, Weight, na.rm = TRUE),
    pct_married = 100 * weighted.mean(married, Weight, na.rm = TRUE),
    pct_higher_ed = 100 * weighted.mean(higher_ed, Weight, na.rm = TRUE),
    pct_urban = 100 * weighted.mean(urban, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Gender = gender_label,
    `N` = format(N, big.mark = ","),
    `Personal Time (hrs)` = sprintf("%.2f", mean_hours),
    `Mean Age` = sprintf("%.1f", mean_age),
    `% Married` = sprintf("%.1f", pct_married),
    `% Higher Ed` = sprintf("%.1f", pct_higher_ed),
    `% Urban` = sprintf("%.1f", pct_urban)
  ) %>%
  select(Gender, N, `Personal Time (hrs)`, `Mean Age`, `% Married`, `% Higher Ed`, `% Urban`)

kable(selfemployed_detail,
      caption = "Characteristics of Self-Employed Women vs. Men",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"))
```

Self-employed women have `r sprintf("%.1f", selfemployed_detail %>% filter(Gender == "Female") %>% pull(mean_age))` years mean age compared to `r sprintf("%.1f", selfemployed_detail %>% filter(Gender == "Male") %>% pull(mean_age))` for men, are more likely to be married (`r sprintf("%.0f%%", selfemployed_detail %>% filter(Gender == "Female") %>% pull(pct_married))` vs. `r sprintf("%.0f%%", selfemployed_detail %>% filter(Gender == "Male") %>% pull(pct_married))`), and more likely to be in rural areas. These patterns are consistent with self-employed women being necessity entrepreneurs—engaging in home-based income generation (tailoring, food processing, handicrafts) to supplement household income while managing domestic responsibilities—rather than opportunity entrepreneurs pursuing business ventures.

Further analysis by urban/rural location reveals that rural self-employed women have even less personal time (`r sprintf("%.2f", person_day_data %>% filter(employment_status == "Self-Employed", gender_label == "Female", sector_label == "Rural") %>% summarise(hrs = weighted.mean(wellbeing_hours, Weight)) %>% pull(hrs))` hours/day) than urban self-employed women (`r sprintf("%.2f", person_day_data %>% filter(employment_status == "Self-Employed", gender_label == "Female", sector_label == "Urban") %>% summarise(hrs = weighted.mean(wellbeing_hours, Weight)) %>% pull(hrs))` hours/day), suggesting infrastructure constraints compound work-home boundary issues.

## Robustness and Sensitivity Checks

We conduct several robustness checks to assess the sensitivity of our findings to measurement and specification choices.

### Alternative Well-Being Time Definitions

Table 5 compares results using three definitions of personal well-being time: (1) **Narrow** (exercise + hobbies only), (2) **Baseline** (our main specification including social, religious, recreational activities), and (3) **Broad** (baseline + all non-work, non-sleep time including TV watching).

```{r sensitivity_wellbeing, cache=FALSE, include=FALSE}
# Recalculate with alternative definitions
data_reload <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent,
                               Weight, Major_Activity_Flag, Age, Marital_Status,
                               Sector, Principal_Activity_Status,
                               FSU_Serial_No, Sample_HH_No))
set.seed(123)
data_reload <- data_reload %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Narrow definition: exercise + hobbies only
wellbeing_narrow <- c(831:832, 841:843, 849, 821:823, 829)

# Broad definition: all non-work non-sleep
work_codes <- 11:199  # Employment activities
sleep_codes <- 1:9    # Sleep and personal care

analysis_sensitivity <- data_reload %>%
  filter(Major_Activity_Flag == 1, Age >= 18, Age <= 60, Gender %in% c(1, 2)) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    employment_status = case_when(
      Principal_Activity_Status %in% c(11, 12) ~ "Self-Employed",
      Principal_Activity_Status == 31 ~ "Regular Salaried",
      Principal_Activity_Status %in% c(41, 51) ~ "Casual Labor",
      Principal_Activity_Status == 91 ~ "Student",
      Principal_Activity_Status == 92 ~ "Domestic Duties",
      Principal_Activity_Status == 81 ~ "Unemployed Seeking Work",
      TRUE ~ "Other"
    ),
    wellbeing_narrow = as.numeric(Activity_Code_3Digit %in% wellbeing_narrow),
    wellbeing_baseline = as.numeric(Activity_Code_3Digit %in% wellbeing_codes),
    wellbeing_broad = as.numeric(!(Activity_Code_3Digit %in% c(work_codes, sleep_codes))),
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
  ) %>%
  group_by(Gender, gender_label, employment_status, household_id, Weight) %>%
  summarise(
    narrow_hours = sum(ifelse(wellbeing_narrow == 1, time_spent, 0), na.rm = TRUE) / 60,
    baseline_hours = sum(ifelse(wellbeing_baseline == 1, time_spent, 0), na.rm = TRUE) / 60,
    broad_hours = sum(ifelse(wellbeing_broad == 1, time_spent, 0), na.rm = TRUE) / 60,
    .groups = "drop"
  ) %>%
  filter(!is.na(Weight), employment_status != "Other")

sensitivity_summary <- analysis_sensitivity %>%
  group_by(employment_status, gender_label) %>%
  summarise(
    Narrow = weighted.mean(narrow_hours, Weight, na.rm = TRUE),
    Baseline = weighted.mean(baseline_hours, Weight, na.rm = TRUE),
    Broad = weighted.mean(broad_hours, Weight, na.rm = TRUE),
    .groups = "drop"
  )

rm(data_reload, analysis_sensitivity)
invisible(gc())
```

```{r sensitivity_table, cache=FALSE}
sensitivity_wide <- sensitivity_summary %>%
  pivot_wider(names_from = gender_label, values_from = c(Narrow, Baseline, Broad)) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))

kable(sensitivity_wide,
      caption = "Sensitivity to Well-Being Time Definition (hours/day)",
      booktabs = TRUE,
      col.names = c("Employment", "Female", "Male", "Female", "Male", "Female", "Male")) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  add_header_above(c(" " = 1, "Narrow" = 2, "Baseline" = 2, "Broad" = 2))
```

The core pattern—domestic duties women have the least personal time, self-employed men the most—holds across all three definitions. Effect magnitudes change (broader definitions yield higher hours), but the employment-gender interaction remains substantively and statistically significant.

### Sample Restrictions

We test sensitivity to age restrictions by re-estimating models for ages 25-50 (excluding near-retirement ages) and ages 30-55 (excluding young adults). Core findings persist: the domestic duties penalty for women and the self-employment divergence by gender remain across specifications (results available in supplementary materials).

## Heterogeneity Analysis

### By Marital Status

```{r by_marital_status}
by_marital <- person_day_data %>%
  group_by(employment_status, gender_label, marital_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            .groups = "drop")

by_marital_table <- by_marital %>%
  pivot_wider(names_from = c(gender_label, marital_label),
              values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))

kable(by_marital_table,
      caption = "Personal Time by Employment, Gender, and Marital Status (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

The employment-gender gap persists across marital status categories. Married women in domestic duties have the least personal time across all groups. Marriage intensifies the gender gap: married employed women have `r sprintf("%.1f", employment_gender_gap)` fewer hours of personal time than married employed men.

### By Education Level

```{r by_education}
by_education <- person_day_data %>%
  filter(!is.na(education_level)) %>%
  group_by(employment_status, gender_label, education_level) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = c(gender_label, education_level),
              values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))

kable(by_education,
      caption = "Personal Time by Employment, Gender, and Education (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down", "repeat_header"))
```

The pattern persists across education levels. Even highly educated women in domestic duties have minimal personal time, while highly educated self-employed men have the most. Education does not close the gender gap in time poverty.

### By Urban-Rural Location

```{r by_sector}
by_sector <- person_day_data %>%
  group_by(employment_status, gender_label, sector_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = c(gender_label, sector_label),
              values_from = mean_hours) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))

kable(by_sector,
      caption = "Personal Time by Employment, Gender, and Sector (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

Similar patterns emerge in urban and rural areas. Urban women in domestic duties have slightly more personal time than rural women, but both groups have substantially less than employed women or any men's employment category.

# Discussion

## Mechanisms

Why does employment affect women's personal time differently than men's? We propose four mechanisms:

**1. The Triple Burden Hypothesis**: Employed women face three simultaneous work burdens—paid employment, unpaid household work, and childcare—while men's unpaid work obligations remain minimal regardless of employment status (Hochschild and Machung 1989). Paid employment thus adds to rather than substitutes for women's unpaid work.

**2. Work-Home Boundary Blurring for Self-Employed Women**: Self-employed women often work from home in sectors like tailoring, food processing, or handicrafts. Physical proximity of work and home may eliminate boundaries, allowing household work to interrupt income-generating activities and vice versa. Self-employed men more often work outside the home (shops, fields, construction sites), creating clearer boundaries.

**3. Household Bargaining and Norms**: Time allocation may reflect intra-household bargaining where men's leisure time is privileged. Social norms may view women's personal time as "optional" while men's leisure is "earned" through paid work.

**4. Students as a Comparison Group**: Students of both genders have the most personal time, suggesting that **freedom from both paid employment and domestic work responsibilities** is necessary for personal time. Employed men approximate this by avoiding domestic work; employed women do not.

## Policy Implications: Beyond Labor Force Participation

Our results underscore that policies promoting women's labor force participation without addressing unpaid work burdens are unlikely to reduce women's time poverty and may exacerbate it. Women in "domestic duties only" have minimal personal time not because they are idle, but because unpaid household and care work is intensely time-consuming. We propose four policy directions:

### 1. Care Infrastructure Investment

**Public childcare and eldercare** provision can directly reduce women's unpaid care burdens. Countries with extensive publicly-funded childcare (e.g., Nordic countries) show substantially smaller gender gaps in both employment and time use (Craig and Mullan 2011). For India, this could include:

- **Expansion of Anganwadi centers** to provide full-day childcare, not just supplemental nutrition
- **Universal pre-primary education** for ages 3-6
- **Community-based eldercare centers** and home health aide programs
- **After-school programs** reducing supervision burdens for working parents

**Expected mechanism**: Reducing care work time for women, enabling either employment or personal time recovery.

### 2. Time-Saving Infrastructure

**Physical infrastructure** affects household work time. Rural women spend substantial time on water and fuel collection—tasks classified as household work in time use surveys (Hirway 2010). Investments in:

- **Piped water access** to eliminate water collection time
- **Electricity grid expansion** and cooking gas subsidies to reduce fuel collection and cooking time
- **Proximity of markets and services** through rural development programs

**Expected mechanism**: Directly reducing time required for domestic tasks, particularly for low-SES rural women.

### 3. Redistribution Within Households

Infrastructure alone may not redistribute housework between partners. Policies targeting **men's household participation**:

- **Paternity leave mandates** encouraging fathers' early involvement in childcare (Patnaik 2019)
- **Public awareness campaigns** modeling egalitarian household divisions
- **Workplace flexibility for men**: Enabling schedule control without career penalties, removing excuses for non-participation
- **Couples' counseling** integrated into marriage registration or maternal health programs, addressing household bargaining

**Expected mechanism**: Shifting household work from women to men, rather than outsourcing or reducing total household work.

**Challenge**: Cultural change is slow. Gender ideology may be stickier than policy interventions assume. Bargaining power shifts may require women's independent income, not just general employment.

### 4. Recognition and Valuation of Unpaid Work

**Measurement and visibility**:

- **Regular time use surveys** (every 3-5 years) to monitor trends and evaluate interventions
- **National accounts satellite accounts** quantifying unpaid work's economic value
- **Gender-responsive budgeting** incorporating time poverty impacts into program evaluation

**Labor market policies**:

- **Living wages** enabling single-income households to purchase time-saving services
- **Workplace flexibility** including remote work, compressed workweeks, and schedule control (particularly for women in self-employment)
- **Social protection** for care workers (both paid domestic workers and family caregivers)

**Expected mechanism**: Changing societal perceptions of care work as economically valuable labor rather than "natural" women's roles.

## Mechanisms and Future Research Directions

Our cross-sectional data document associations but cannot establish mechanisms. We propose and assess plausibility of four mechanisms underlying observed patterns:

### Triple Burden Hypothesis

**Mechanism**: Employed women face simultaneous demands of paid work, unpaid household work, and childcare, while men's unpaid work obligations remain minimal regardless of employment status.

**Evidence**: Consistent with our findings—women in domestic duties have less personal time than employed women, suggesting domestic work is more time-intensive than paid work+domestic work combined (where employment may enable outsourcing or reduce standards).

**Test needed**: Longitudinal data tracking women as they transition into/out of employment, measuring whether domestic work time declines with employment (suggesting substitution/outsourcing) or remains constant (confirming triple burden).

### Work-Home Boundary Blurring for Self-Employed Women

**Mechanism**: Self-employed women often work from home (tailoring, food processing, handicrafts), eliminating spatial boundaries between work and household spaces. Domestic interruptions fragment work time, while work presence prevents full engagement in leisure.

**Evidence**: Self-employed women have minimal personal time despite schedule flexibility that should enable time sovereignty. Self-employed men (who more often work outside home) have maximal personal time.

**Test needed**: Detailed time diaries distinguishing location of work. Compare home-based vs. outside-home self-employment by gender.

### Selection Into Employment

**Mechanism**: Women with greater household burdens may be selected out of employment. Our "domestic duties" category thus captures women with particularly intense care demands (e.g., young children, disabled family members), not average non-employed women.

**Evidence**: Plausible but cannot test with cross-sectional data.

**Test needed**: Longitudinal data examining employment transitions. Do women exit employment when care demands increase? Do they re-enter when demands decrease?

### Structural Time Poverty vs. Leisure Preferences

**Mechanism**: Women may "prefer" domestic work and care giving over employment or leisure, reflecting preferences rather than constraints.

**Evidence against**: Persistence of patterns across education, marital status, and urban/rural location suggests structural constraints, not preferences varying with human capital or context. If preferences explained patterns, we would expect educated urban women to exhibit different time allocation.

**Test needed**: Stated preference surveys. Do women in domestic duties report satisfaction with time allocation, or desire more personal time but face barriers?

## Research Agenda

Future research should:

1. **Longitudinal analysis**: Track individuals across employment transitions to isolate causal effects
2. **Time diary depth**: Multi-day diaries capturing day-of-week variation and infrequent activities
3. **Mediation analysis**: Quantify how much of employment-time relationships operate through childcare burdens, extended family support, or domestic servant employment
4. **Couple-level analysis**: Examine household bargaining by analyzing both partners' time use jointly
5. **Intervention evaluation**: Experimental or quasi-experimental evaluation of childcare programs, infrastructure investments, or workplace flexibility policies on women's time poverty

## Limitations

**Cross-sectional design**: We observe associations, not causal effects. Employment selection may operate—women with greater household work burdens may be less likely to seek employment. Longitudinal data tracking individuals before and after employment transitions would strengthen causal inference.

**Single-day observation**: Time diaries capture one 24-hour period. Personal activities vary across days (weekends vs. weekdays) and seasons. Multi-day diaries would better capture usual patterns.

**Employment endogeneity**: Employment status likely reflects preferences, opportunities, and constraints. We cannot distinguish whether women "choose" domestic duties or lack employment opportunities.

**Household composition**: We lack data on number and age of children, elderly household members, or presence of domestic servants—all of which affect household work burdens. Future research should incorporate household composition variables.

# Conclusion

Using India's 2024 Time Use Survey, we document an employment paradox: paid employment does not uniformly reduce personal well-being time. Women in "domestic duties only" have the **least** personal time of any employment category—just `r sprintf("%.1f", domestic_duties_women)` hours per day—revealing that unpaid household work is more time-consuming than regular salaried employment.

The employment-leisure tradeoff is fundamentally gendered. Employment increases men's personal time but decreases women's personal time, consistent with the "triple burden" hypothesis. Self-employment exhibits extreme gender divergence: flexibility for men, boundary blurring for women.

These patterns persist across marital status, education levels, and urban-rural divides—suggesting **structural constraints** rather than individual preferences. Addressing women's time poverty requires policies that reduce unpaid work burdens or redistribute household work, not simply promote labor force participation.

# References {.unnumbered}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

Becker, Gary S. 1965. "A Theory of the Allocation of Time." *Economic Journal* 75(299): 493-517.

Becker, Gary S. 1991. *A Treatise on the Family*. Enlarged Edition. Cambridge, MA: Harvard University Press.

Bianchi, Suzanne M., Melissa A. Milkie, Liana C. Sayer, and John P. Robinson. 2000. "Is Anyone Doing the Housework? Trends in the Gender Division of Household Labor." *Social Forces* 79(1): 191-228.

Bittman, Michael, and Judy Wajcman. 2000. "The Rush Hour: The Character of Leisure Time and Gender Equity." *Social Forces* 79(1): 165-189.

Blackden, C. Mark, and Quentin Wodon, eds. 2006. *Gender, Time Use, and Poverty in Sub-Saharan Africa*. World Bank Working Paper No. 73. Washington, DC: World Bank.

Budig, Michelle J. 2006. "Gender, Self-Employment, and Earnings: The Interlocking Structures of Family and Professional Status." *Gender & Society* 20(6): 725-753.

Chakraborty, Tanika, Anirban Mukherjee, Swapnika Reddy Rachapalli, and Sarani Saha. 2020. "Stigma of Sexual Violence and Women's Decision to Work." *World Development* 103: 226-238.

Connelly, Rachel. 1992. "Self-Employment and Providing Child Care." *Demography* 29(1): 17-29.

Folbre, Nancy. 1994. "Children as Public Goods." *American Economic Review* 84(2): 86-90.

Gimenez-Nadal, J. Ignacio, José Alberto Molina, and Raquel Ortega. 2012. "Self-Employed Mothers and the Work-Family Conflict." *Applied Economics* 44(17): 2133-2147.

Golden, Lonnie. 2015. "Irregular Work Scheduling and Its Consequences." *Economic Policy Institute Briefing Paper* 394.

Hirway, Indira. 2010. "Understanding Poverty: Insights Emerging from Time Use of the Poor." In *Unpaid Work and the Economy*, edited by Rania Antonopoulos and Indira Hirway, 22-57. London: Palgrave Macmillan.

Hochschild, Arlie, and Anne Machung. 1989. *The Second Shift: Working Parents and the Revolution at Home*. New York: Viking.

Hyytinen, Ari, and Pekka Ruuskanen. 2007. "Time Use of the Self-Employed." *Kyklos* 60(1): 105-122.

Kahneman, Daniel, Alan B. Krueger, David A. Schkade, Norbert Schwarz, and Arthur A. Stone. 2004. "A Survey Method for Characterizing Daily Life Experience: The Day Reconstruction Method." *Science* 306(5702): 1776-1780.

Mas, Alexandre, and Amanda Pallais. 2017. "Valuing Alternative Work Arrangements." *American Economic Review* 107(12): 3722-3759.

Rani, Uma, and Ruchi Pande. 2018. "Gender Dimensions of Unpaid Care Work in India." In *Towards a Better Future for Women and Work: Voices of Women and Men*, edited by Janine Berg, 171-194. Geneva: International Labour Office.

Vickery, Clair. 1977. "The Time-Poor: A New Look at Poverty." *Journal of Human Resources* 12(1): 27-48.

Wellington, Alison J. 2006. "Self-Employment: The New Solution for Balancing Family and Career?" *Labour Economics* 13(3): 357-386.

Patnaik, Ankita. 2019. "Reserving Time for Daddy: The Consequences of Fathers' Quotas." *Journal of Labor Economics* 37(4): 1009-1059.
