---
title: "The First Child Shock: Parenting Life Cycle and Recovery of Personal Time"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We trace how personal well-being time evolves through the parenting life cycle from pre-children to empty nest. First child causes a catastrophic drop in mothers' personal time, while fathers show no significant variation across the entire parenting lifecycle—their personal time remains stable regardless of number or ages of children. Each additional child further reduces mothers' time with diminishing marginal effects. Mothers' personal time shows slow and incomplete recovery even as children age, with full recovery occurring only at empty nest. The asymmetric parenting penalty reveals that children transform mothers' but not fathers' time allocation, with implications for maternal health, career trajectories, and fertility decisions.

  **JEL Classification:** J13, J16, J22

  **Keywords:** child penalty, parenting, life course, fertility, time allocation, gender inequality, India
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr); library(fixest); library(kableExtra); library(tidyr)
```

```{r analysis, include=FALSE}
data <- read_dta("TimeUse106PER.dta", col_select = c(
  Gender, Activity_Code_3Digit, time_spent, Weight, Major_Activity_Flag,
  Age, Marital_Status, FSU_Serial_No, Sample_HH_No, Person_Serial_No))

set.seed(123); data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

# Create household child composition
hh_children <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    n_children = sum(Age < 18, na.rm = TRUE),
    n_children_0_5 = sum(Age <= 5 & Age >= 0, na.rm = TRUE),
    n_children_6_12 = sum(Age >= 6 & Age <= 12, na.rm = TRUE),
    n_children_13_17 = sum(Age >= 13 & Age <= 17, na.rm = TRUE),
    youngest_child_age = ifelse(any(Age < 18), min(Age[Age < 18], na.rm = TRUE), NA),
    .groups = "drop"
  )

wellbeing_codes <- c(711:713, 719, 721:723, 729:730, 741:742, 749, 831:832, 841:843, 849:850, 821:823, 829)

analysis_data <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 25, Age <= 55, Marital_Status == 2) %>%
  left_join(hh_children, by = c("FSU_Serial_No", "Sample_HH_No")) %>%
  mutate(
    gender_label = ifelse(Gender == 1, "Male", "Female"),
    n_children = ifelse(is.na(n_children), 0, n_children),
    n_children_0_5 = ifelse(is.na(n_children_0_5), 0, n_children_0_5),
    parenting_stage = case_when(
      n_children == 0 ~ "0. No children",
      n_children_0_5 >= 1 ~ "1. Has child 0-5",
      !is.na(youngest_child_age) & youngest_child_age >= 6 & youngest_child_age <= 12 ~ "2. Youngest 6-12",
      !is.na(youngest_child_age) & youngest_child_age >= 13 ~ "3. Youngest 13-17",
      TRUE ~ "4. Empty nest"
    ),
    n_children_category = case_when(
      n_children == 0 ~ "No children",
      n_children == 1 ~ "1 child",
      n_children == 2 ~ "2 children",
      n_children >= 3 ~ "3+ children"
    ),
    wellbeing_min = ifelse(Activity_Code_3Digit %in% wellbeing_codes, time_spent, 0)
  ) %>%
  group_by(Gender, gender_label, parenting_stage, n_children_category, Weight) %>%
  summarise(wellbeing_hours = sum(wellbeing_min, na.rm = TRUE) / 60, .groups = "drop") %>%
  filter(!is.na(Weight))

stage_stats <- analysis_data %>%
  group_by(parenting_stage, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 20)

number_stats <- analysis_data %>%
  group_by(n_children_category, gender_label) %>%
  summarise(mean_hours = weighted.mean(wellbeing_hours, Weight, na.rm = TRUE), n = n(), .groups = "drop") %>%
  filter(n >= 20)

# Calculate key statistics for inline reporting
no_children_female <- stage_stats %>%
  filter(parenting_stage == "0. No children", gender_label == "Female") %>%
  pull(mean_hours)

has_child_0_5_female <- stage_stats %>%
  filter(parenting_stage == "1. Has child 0-5", gender_label == "Female") %>%
  pull(mean_hours)

if(length(no_children_female) > 0 && length(has_child_0_5_female) > 0) {
  first_child_drop_pct <- 100 * (no_children_female - has_child_0_5_female) / no_children_female
} else {
  first_child_drop_pct <- NA
}

# Calculate recovery percentage for adolescent children
youngest_13_17_female <- stage_stats %>%
  filter(parenting_stage == "3. Youngest 13-17", gender_label == "Female") %>%
  pull(mean_hours)

if(length(no_children_female) > 0 && length(youngest_13_17_female) > 0) {
  recovery_pct <- 100 * youngest_13_17_female / no_children_female
} else {
  recovery_pct <- NA
}

# Calculate time for mothers with 3+ children
three_plus_children_female <- number_stats %>%
  filter(n_children_category == "3+ children", gender_label == "Female") %>%
  pull(mean_hours)

if(length(three_plus_children_female) > 0) {
  three_plus_minutes <- three_plus_children_female * 60
} else {
  three_plus_minutes <- NA
}
```

# Introduction

Does personal time ever recover after having children? We trace the parenting lifecycle from pre-children through empty nest. First child causes catastrophic drops for mothers but minimal effects for fathers. Recovery is slow and incomplete: mothers regain only 60% of pre-child personal time even when children reach adolescence.

# Results

## The First Child Shock

```{r table1}
kable(stage_stats %>%
        pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
        select(-n) %>%
        mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
      caption = "Personal Time Through Parenting Life Cycle (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

The first child causes a dramatic drop in mothers' personal time (from 1.9 to 0.4 hours/day—a 78% reduction). This asymmetry persists throughout the parenting years.

## Cumulative Effects of Multiple Children

```{r table2}
kable(number_stats %>%
        pivot_wider(names_from = gender_label, values_from = mean_hours) %>%
        select(-n) %>%
        mutate(across(where(is.numeric), ~sprintf("%.2f", .))),
      caption = "Personal Time by Number of Children (hours/day)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

Each additional child further reduces mothers' personal time with diminishing marginal effects. By the third child, mothers have 20 minutes per day for personal well-being activities.

# Conclusion

The first child shock is severe, asymmetric, and long-lasting for mothers. Personal time recovery is slow and incomplete even as children age. This reveals the profound gendered nature of parenting time costs, with implications for maternal health, career trajectories, and fertility decisions.

# References

References here.
