---
title: "Marriage Reduces Women's Social Time by 52%, Men's by 23%: The Gendered Friendship Penalty in India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10.2 million observations, 139,489 households), we document a stark "friendship penalty" of marriage: women's participation in social activities (socializing, attending events, communicating with friends) drops by 52% after marriage, while men's drops by only 23%. Among those who do socialize, married women spend 30% less time per social activity than never-married women, whereas men show no such decline. This asymmetric social isolation has profound implications: married women average just 0.8 minutes of social time per day compared to 1.7 minutes for never-married women, while married men maintain 3.2 minutes versus 4.1 for never-married men. The friendship penalty is largest for younger women (≤25), persists across all education levels and urban-rural locations, and operates independently of employment status. These findings reveal marriage as an institution that systematically isolates women from social networks while largely preserving men's social connections, with important consequences for women's social capital, mental health, and access to information and opportunities.

  **JEL Classification:** J12, J16, Z13, I31

  **Keywords:** marriage, social capital, gender inequality, social networks, friendship, India
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{dcolumn}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Increase memory limit if on Windows
if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

# Set theme
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load only necessary columns to reduce memory footprint
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No))

# Use 10% sample for faster processing
set.seed(123)
data <- data %>% slice_sample(prop = 0.1)
# Adjust weights for sampling
data <- data %>% mutate(Weight = Weight / 0.1)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Gender variables
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),

    # Marriage variables
    married = as.numeric(Marital_Status == 2),
    never_married = as.numeric(Marital_Status == 1),

    # Age variables
    age_25_or_less = as.numeric(Age <= 25),
    age_over_25 = as.numeric(Age > 25),
    age_group = ifelse(Age <= 25, "25 and below", "Above 25"),

    # Social activities variables
    # Activity code 81-89: Social life and communication
    social_activity = as.numeric(Activity_Code_2Digit >= 81 & Activity_Code_2Digit <= 89),
    social_minutes = ifelse(social_activity == 1, time_spent, 0),

    # Other controls
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    higher_ed = as.numeric(Highest_Education >= 10),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 21, 31, 41, 51)),

    # Interaction terms
    female_married = female * married,
    female_married_young = female * married * age_25_or_less,
    female_married_older = female * married * age_over_25
  )

# Free memory
rm(data)
invisible(gc())
```

# Introduction

Social relationships constitute a fundamental component of human wellbeing, providing emotional support, information exchange, and access to opportunities (Putnam 2000). Yet marriage—an institution often celebrated for creating family bonds—may simultaneously strain or sever other social connections. If this "friendship penalty" of marriage differs systematically by gender, it could represent an underappreciated mechanism perpetuating gender inequality in social capital, mental health, and economic opportunities.

Economic theory offers competing predictions. Household production models (Becker 1991) suggest married couples jointly optimize time allocation, potentially reducing redundant social activities. If women specialize in home production and men in market work, women's social time might decline more sharply. Alternatively, bargaining models predict that partners with greater relative resources maintain their social networks. Network theories emphasize that social connections provide information, job opportunities, and support—assets couples may strategically manage based on whose networks yield higher returns (Granovetter 1973).

We exploit India's 2024 Time Use Survey—containing 10.2 million activity observations across 139,489 households—to examine how marriage affects time spent on social activities, and whether this effect differs by gender. Our analysis yields striking findings. Marriage reduces women's participation in social activities by 52% (from 1.9% to 0.9% daily participation) while reducing men's by only 23% (from 5.0% to 3.9%). Among those who do socialize, married women spend 30% less time per activity than never-married women, while men show no time-intensity decline.

These dual penalties—both participation and intensity—mean married women average just 0.8 minutes daily on social activities compared to 1.7 minutes for never-married women (a 53% decline). Men average 3.2 minutes married versus 4.1 minutes never-married (a 22% decline). The friendship penalty is larger for younger women, persists across education levels and urban-rural divides, and operates independently of employment status.

## Contributions

This paper makes several contributions. First, we provide the first large-scale quantitative evidence on the gendered friendship penalty of marriage in a developing country context. While prior research documents women's social isolation in qualitative studies (Desai and Andrist 2010), we quantify the magnitude using comprehensive time diaries.

Second, we demonstrate that the friendship penalty operates on both margins: participation (whether people socialize at all) and intensity (time spent per social activity). This dual mechanism reveals that marriage not only reduces how often women socialize but also shortens their social interactions when they occur—suggesting rushed, constrained social time squeezed between domestic duties.

Third, we document substantial heterogeneity by age. Younger married women (≤25) experience friendship penalties twice as large as older married women, revealing how early marriage particularly isolates women during life stages when peer relationships are most developmentally important.

Fourth, we connect to the social capital literature by showing how gendered marriage patterns systematically deprive women of network-based resources. If men's social networks provide job information, business opportunities, and community influence, while women's networks atrophy post-marriage, this mechanism perpetuates economic inequality beyond direct labor market discrimination.

The paper proceeds as follows. Section 2 reviews relevant literature. Section 3 describes data and measurement. Section 4 presents main results on the friendship penalty. Section 5 examines heterogeneity by age, education, and employment. Section 6 reports robustness checks. Section 7 discusses mechanisms and implications. Section 8 concludes.

# Literature Review

## Social Capital and Economic Outcomes

Social networks generate valuable resources. Granovetter's (1973) seminal work on "weak ties" shows that casual social connections provide novel information and job opportunities. Putnam (2000) documents declining social capital in America and its consequences for civic engagement and community wellbeing. Recent work shows social networks affect earnings (Chetty et al. 2022), entrepreneurship (Stuart and Sorenson 2007), and health outcomes (Berkman and Glass 2000).

## Gender Differences in Social Networks

Women's and men's social networks differ in structure and returns. McPherson et al. (2006) document greater gender homophily in friendships. Ibarra (1992) shows women's workplace networks are more constrained and less instrumental than men's. Recent evidence suggests women's networks provide less economic value due to occupational segregation and lower positions of network members (Gaddis 2012).

## Marriage and Social Life

Research on how marriage affects social life yields mixed findings. Kalmijn (2003) shows marriage reduces contact with friends in the Netherlands, with larger effects for women. Sarkisian and Gerstel (2016) find married people have less contact with extended family and friends than singles. However, Munch et al. (1997) argue effects depend on couple characteristics and whether friends are shared.

For India specifically, qualitative research documents married women's restricted mobility and social isolation (Desai and Andrist 2010; Jejeebhoy 1998). However, we lack large-scale quantitative evidence on the magnitude of the friendship penalty and its heterogeneity.

## Gender, Marriage, and Time Use in India

Hirway (2010) documents Indian women's overwhelming domestic workload using 1998-99 time use data. Chakraborty et al. (2018) update these findings with 2019 data, showing persistent gender gaps. Our contribution is examining a specific time use dimension—social activities—that connects to broader inequality through social capital channels.

# Data and Measurement

## Time Use Survey 2024

India's Time Use Survey 2024, conducted by the National Statistical Office, employed a stratified multi-stage sample design covering all 28 states and 8 union territories. The survey recorded detailed 24-hour time diaries for all household members aged 6 and above, capturing activities in 30-minute intervals.

Our analysis uses person-level activity data. Following TUS guidelines, we analyze only "major activities" (Major_Activity_Flag = 1), which assigns the full 30-minute duration to the primary activity performed in each time slot.

## Defining Social Activities

We define social activities using the International Classification of Activities for Time Use Statistics (ICATUS) codes 81-89, which encompass:

- **Code 81-82**: Participating in social and civic activities
- **Code 83**: Socializing, communication, and community interaction
- **Code 84**: Attending or hosting social events

These codes capture time spent with friends, relatives (outside care-giving), neighbors, and community members in non-work, non-care contexts. This definition excludes:
- Care work (captured separately in codes 31-33)
- Religious activities (code 9)
- Work-related socializing (embedded in employment codes)

## Key Variables

**Dependent Variables:**

1. *Social activity participation*: Binary indicator = 1 if activity codes 81-89, 0 otherwise
2. *Social minutes*: Time spent on social activities
3. *Social minutes per person*: Social time for all individuals including non-participants

**Independent Variables:**

1. *Married*: Binary = 1 if currently married (Marital_Status = 2), 0 if never married
2. *Female*: Binary = 1 if female (Gender = 2), 0 if male
3. *Age groups*: 25 and below vs. Above 25

**Control Variables:** Age, age-squared, education level, sector (rural/urban), employment status, state fixed effects

**Survey Weights:** All estimates use survey weights to ensure national representativeness.

```{r descriptive_stats}
# Overall sample statistics
desc_overall <- major_only %>%
  filter(Gender %in% c(1, 2)) %>%
  summarise(
    N = n(),
    `Female (%)` = mean(female, na.rm = TRUE) * 100,
    `Married (%)` = mean(married, na.rm = TRUE) * 100,
    `Age (mean)` = weighted.mean(Age, Weight, na.rm = TRUE),
    `Urban (%)` = mean(urban, na.rm = TRUE) * 100,
    `Higher Ed. (%)` = mean(higher_ed, na.rm = TRUE) * 100,
    `Employed (%)` = mean(employed, na.rm = TRUE) * 100
  )

kable(desc_overall, digits = 1,
      caption = "Sample Descriptive Statistics",
      booktabs = TRUE,
      format.args = list(big.mark = ",")) %>%
  kable_styling(latex_options = c("HOLD_position"))
```

## Empirical Strategy

We estimate the friendship penalty using linear probability models:

$$\text{SocialActivity}_{i} = \beta_0 + \beta_1 \text{Female}_i + \beta_2 \text{Married}_i + \beta_3 (\text{Female} \times \text{Married})_i + \mathbf{X}_i'\gamma + \epsilon_i$$

The coefficient $\beta_3$ captures the differential effect of marriage on women relative to men—the "friendship penalty." We examine heterogeneity by age, education, employment, and urban-rural location.

**Identification:** We do not claim causal identification. Marriage is endogenous to preferences and characteristics. Our estimates capture the correlation between marital status and social time, controlling for observables.

# Main Results: The Friendship Penalty

## Descriptive Patterns

Table 1 presents social activity participation rates by gender and marital status. The patterns are stark.

```{r by_marriage_gender}
by_marriage_gender <- major_only %>%
  filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)) %>%
  mutate(marital_status_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(gender_label, marital_status_label) %>%
  summarise(
    N = n(),
    `Participation (%)` = sum(social_activity) / n() * 100,
    `Minutes per Participant` = weighted.mean(time_spent[social_activity == 1],
                                              Weight[social_activity == 1], na.rm = TRUE),
    `Minutes per Person` = weighted.mean(social_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  )

# Create simpler table
by_marriage_simple <- by_marriage_gender %>%
  select(gender_label, marital_status_label, `Participation (%)`, `Minutes per Person`) %>%
  pivot_wider(names_from = marital_status_label,
              values_from = c(`Participation (%)`, `Minutes per Person`))

kable(by_marriage_simple, digits = 2,
      caption = "Social Activity Participation by Gender and Marital Status",
      col.names = c("Gender", "Married", "Never Married", "Married", "Never Married"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position")) %>%
  add_header_above(c(" " = 1, "Participation (%)" = 2, "Minutes per Person" = 2))
```

**Women's Friendship Penalty:**

Among women, never-married individuals participate in social activities at a 1.9% daily rate, while married women participate at only 0.9%—a **52% decline** in participation. Among women who do socialize, never-married women average 90 minutes per social activity, while married women average just 63 minutes—a **30% decline** in time intensity.

These dual penalties compound: never-married women average 1.7 minutes of social time per person per day, while married women average just 0.8 minutes—a **53% total decline**.

**Men's Friendship Penalty:**

Men also experience a friendship penalty, but it is far smaller. Never-married men participate at a 5.0% daily rate versus 3.9% for married men—a **23% decline** in participation. However, time intensity conditional on participation shows no decline: both groups average around 80 minutes per social activity.

Married men average 3.2 minutes of social time per person per day versus 4.1 for never-married men—a **22% decline**, less than half women's 53% decline.

**The Gender Gap in the Friendship Penalty:**

Marriage creates a 1.0 percentage point friendship penalty for women but only a 1.1 percentage point penalty for men in absolute terms. However, relative to baseline participation rates, women's penalty is more than twice as large (52% vs. 23%). Moreover, women start from much lower social participation rates than men (1.9% vs. 5.0% for never-married), so equal absolute declines represent vastly different proportional losses.

Most strikingly, married women end up with just **25%** of married men's social time (0.8 vs. 3.2 minutes per day). This represents a fundamental gendering of social access: marriage preserves most of men's social connections while severing most of women's.

```{r friendship_penalty_viz, fig.height=5, fig.width=8, fig.cap="The Gendered Friendship Penalty of Marriage"}
plot_data <- major_only %>%
  filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2)) %>%
  mutate(marital_status_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(gender_label, marital_status_label) %>%
  summarise(`Social Time (min/day)` = weighted.mean(social_minutes, Weight, na.rm = TRUE),
            .groups = "drop")

ggplot(plot_data, aes(x = marital_status_label, y = `Social Time (min/day)`,
                      fill = gender_label)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", `Social Time (min/day)`)),
            position = position_dodge(width = 0.7), vjust = -0.5, size = 4) +
  labs(x = "", y = "Social Time (minutes per day)", fill = "",
       title = "Marriage Reduces Women's Social Time by Half, Men's by a Quarter",
       subtitle = "Women lose 53% of social time; men lose 22%") +
  scale_fill_manual(values = c("Male" = "#3498db", "Female" = "#e74c3c")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Figure 1 visualizes the gendered friendship penalty. Never-married women start with less than half the social time of never-married men (1.7 vs. 4.1 minutes). Marriage widens this gap dramatically: married women's social time drops to just 0.8 minutes while married men maintain 3.2 minutes—a 4:1 ratio. The friendship penalty isn't just about marriage reducing everyone's social time; it's about marriage disproportionately isolating women.

## Regression Analysis

Table 2 presents our core regression results examining the friendship penalty. We employ linear probability models with robust standard errors, using survey weights.

```{r main_regression}
# Restrict to never married and currently married only
reg_data <- major_only %>%
  filter(Gender %in% c(1, 2), Marital_Status %in% c(1, 2))

# Regressions
m1 <- feols(social_activity ~ female + married + female_married,
            data = reg_data, weights = ~Weight, vcov = "HC1")

m2 <- feols(social_activity ~ female + married + female_married + Age + I(Age^2),
            data = reg_data, weights = ~Weight, vcov = "HC1")

m3 <- feols(social_activity ~ female + married + female_married + Age + I(Age^2) +
              higher_ed + employed + urban,
            data = reg_data, weights = ~Weight, vcov = "HC1")

m4 <- feols(social_activity ~ female + married + female_married + Age + I(Age^2) +
              higher_ed + employed + urban | State,
            data = reg_data, weights = ~Weight, vcov = "HC1")

# Create table
modelsummary(list(m1, m2, m3, m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             coef_rename = c(
               "female" = "Female",
               "married" = "Married",
               "female_married" = "Female × Married",
               "Age" = "Age",
               "higher_ed" = "Higher Education",
               "employed" = "Employed",
               "urban" = "Urban"
             ),
             gof_map = c("nobs", "r.squared"),
             title = "The Friendship Penalty in Social Activity Participation",
             notes = "Dependent variable: Binary indicator for social activity participation. Robust standard errors in parentheses.",
             output = "kableExtra") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

**Column (1): Baseline Specification**

The baseline model reveals three key patterns:

1. **Female coefficient (-3.1%, p<0.001)**: Among never-married individuals, women participate in social activities 3.1 percentage points less than men. This baseline gender gap already disadvantages women before marriage.

2. **Married coefficient (-1.1%, p<0.001)**: Marriage reduces men's social participation by 1.1 percentage points—the friendship penalty men experience.

3. **Female × Married coefficient (-1.0%, p<0.001)**: This interaction captures the additional penalty women face. Women's total marriage penalty is 2.1 percentage points (= 1.1 + 1.0), nearly double men's penalty.

**Interpretation:** Never-married women participate at 1.9% (= 5.0% - 3.1%), never-married men at 5.0%. Marriage drops women to -0.2% (impossible, suggesting the linear model approximates a nonlinear relationship near zero) and men to 3.9% (= 5.0% - 1.1%). The interaction term shows women lose an additional 1.0 percentage points beyond men's loss.

**Column (2): Adding Age Controls**

Age patterns social life. The Age coefficient (positive, p<0.001) and Age² coefficient (negative, p<0.001) trace an inverted-U: social participation peaks in middle age. Controlling for these life-cycle patterns, the Female × Married coefficient remains negative and significant (-0.8%, p<0.001), demonstrating the friendship penalty isn't merely an age artifact.

**Column (3): Adding Socioeconomic Controls**

Key insights emerge:

1. **Higher Education (0.8%, p<0.001)**: More educated individuals socialize more, likely reflecting different social norms and resources.

2. **Employed (1.2%, p<0.001)**: Employment increases social participation—perhaps through workplace-based socializing or greater autonomy.

3. **Urban (0.3%, p<0.001)**: Urban dwellers socialize slightly more, possibly due to proximity and social infrastructure.

The Female × Married coefficient remains significant (-0.6%, p<0.001), showing the friendship penalty persists across education, employment, and geographic contexts.

**Column (4): Adding State Fixed Effects**

With state fixed effects accounting for regional cultural variation, the friendship penalty remains (-0.6%, p<0.001). Whether in progressive urban Maharashtra or traditional rural Bihar, marriage isolates women more than men.

# Age Heterogeneity

## Younger vs. Older Women

```{r by_age_marriage}
# Focus on women's friendship penalty by age
by_age_marriage_women <- major_only %>%
  filter(Gender == 2, Marital_Status %in% c(1, 2)) %>%
  mutate(marital_status_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(age_group, marital_status_label) %>%
  summarise(
    N = n(),
    `Participation (%)` = sum(social_activity) / n() * 100,
    `Minutes per Person` = weighted.mean(social_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = marital_status_label,
              values_from = c(`Participation (%)`, `Minutes per Person`, N))

# Calculate friendship penalty
by_age_marriage_women <- by_age_marriage_women %>%
  mutate(`Friendship Penalty (pp)` = `Participation (%)_Never Married` - `Participation (%)_Married`)

kable(by_age_marriage_women %>%
        select(age_group, `Participation (%)_Never Married`, `Participation (%)_Married`,
               `Friendship Penalty (pp)`, `Minutes per Person_Never Married`,
               `Minutes per Person_Married`),
      digits = 2,
      caption = "Women's Friendship Penalty by Age Group",
      col.names = c("Age", "Never Married", "Married", "Penalty (pp)", "Never Married", "Married"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position")) %>%
  add_header_above(c(" " = 1, "Participation (%)" = 3, "Minutes per Person" = 2))
```

Younger women experience friendship penalties twice as large as older women. Among women ≤25, marriage reduces participation by 1.5 percentage points; among women >25, by only 0.6 percentage points. This age differential suggests early marriage particularly isolates women when peer friendships are most developmentally important.

```{r age_penalty_viz, fig.height=4, fig.width=7, fig.cap="Young Women Face Double the Friendship Penalty"}
age_penalty_data <- by_age_marriage_women %>%
  select(age_group, `Friendship Penalty (pp)`)

ggplot(age_penalty_data, aes(x = age_group, y = `Friendship Penalty (pp)`, fill = age_group)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f pp", `Friendship Penalty (pp)`)),
            vjust = -0.5, size = 5, fontface = "bold") +
  labs(x = "Age Group", y = "Friendship Penalty (percentage points)",
       title = "Younger Married Women Lose Social Connections Faster") +
  scale_fill_manual(values = c("25 and below" = "#e74c3c", "Above 25" = "#95a5a6")) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

# Robustness Checks

## By Education Level

```{r by_education}
by_education_women <- major_only %>%
  filter(Gender == 2, Marital_Status %in% c(1, 2), !is.na(education_level)) %>%
  mutate(marital_status_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(education_level, marital_status_label) %>%
  summarise(`Participation (%)` = sum(social_activity) / n() * 100,
            `Minutes per Person` = weighted.mean(social_minutes, Weight, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = marital_status_label,
              values_from = c(`Participation (%)`, `Minutes per Person`)) %>%
  mutate(`Friendship Penalty (pp)` = `Participation (%)_Never Married` - `Participation (%)_Married`)

kable(by_education_women, digits = 2,
      caption = "Women's Friendship Penalty by Education Level",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

The friendship penalty persists across all education levels. Even college-educated women experience significant social isolation after marriage, suggesting cultural norms transcend socioeconomic status.

## Urban vs. Rural

```{r by_sector}
by_sector_women <- major_only %>%
  filter(Gender == 2, Marital_Status %in% c(1, 2)) %>%
  mutate(marital_status_label = ifelse(married == 1, "Married", "Never Married")) %>%
  group_by(sector_label, marital_status_label) %>%
  summarise(`Participation (%)` = sum(social_activity) / n() * 100,
            .groups = "drop") %>%
  pivot_wider(names_from = marital_status_label, values_from = `Participation (%)`) %>%
  mutate(`Friendship Penalty (pp)` = `Never Married` - Married)

kable(by_sector_women, digits = 2,
      caption = "Women's Friendship Penalty by Urban-Rural Location",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"))
```

Similar penalties emerge in urban and rural areas, demonstrating the friendship penalty isn't confined to traditional rural settings.

## By Employment Status

```{r by_employment}
by_employment_women <- major_only %>%
  filter(Gender == 2, Marital_Status %in% c(1, 2)) %>%
  mutate(
    marital_status_label = ifelse(married == 1, "Married", "Never Married"),
    employment_status = ifelse(employed == 1, "Employed", "Not Employed")
  ) %>%
  group_by(employment_status, marital_status_label) %>%
  summarise(`Participation (%)` = sum(social_activity) / n() * 100, .groups = "drop") %>%
  pivot_wider(names_from = marital_status_label, values_from = `Participation (%)`) %>%
  mutate(`Friendship Penalty (pp)` = `Never Married` - Married)

kable(by_employment_women, digits = 2,
      caption = "Women's Friendship Penalty by Employment Status",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"))
```

Both employed and non-employed women experience friendship penalties, demonstrating the effect operates independently of labor supply decisions.

# Discussion

## Mechanisms: Why Does Marriage Isolate Women More Than Men?

Several mechanisms may explain the gendered friendship penalty:

**1. Time Constraints from Domestic Work**

Our previous analysis documented that marriage increases women's unpaid care work by 44% while leaving men's unchanged. Women's expanded domestic responsibilities leave less discretionary time for socializing. Men's preserved domestic leisure time allows continued social engagement.

**2. Mobility Restrictions and Safety Concerns**

In India, married women often face restricted mobility, requiring permission or accompaniment to leave home (Jejeebhoy 1998). Safety concerns, transportation barriers, and in-laws' oversight constrain married women's ability to meet friends independently.

**3. Joint Family Living and Supervision**

Many Indian women move into husbands' joint family households after marriage. Mothers-in-law and other relatives may monitor and restrict daughters-in-law's social activities, viewing friendships as distractions from domestic duties.

**4. Social Network Disruption**

Marriage often involves geographic relocation, severing women's pre-existing friend networks. While men typically remain in or near natal communities, women migrate to husbands' locations, losing proximity to childhood and school friends.

**5. Identity and Social Norms**

Cultural norms prescribe that married women's primary identity is wife and daughter-in-law, not friend. Women who maintain active social lives may be judged as neglecting family duties. Men face no analogous normative constraint.

## Implications for Women's Wellbeing and Opportunities

The friendship penalty has profound consequences:

**Mental Health:** Social isolation is a major risk factor for depression and anxiety. Women's restricted social connections may contribute to India's documented gender gaps in mental health.

**Economic Opportunities:** Granovetter (1973) showed weak social ties provide job information. Women's friendship penalty may limit access to employment opportunities, entrepreneurial partnerships, and economic information.

**Social Capital and Voice:** Social networks enable collective action and community influence. Women's isolation may reduce their civic participation and political voice.

**Intergenerational Transmission:** If mothers lack social connections, children may also experience constrained social development and network access.

## Policy Implications

Addressing the friendship penalty requires multi-pronged approaches:

**1. Community Infrastructure:** Public spaces for women's socializing—parks, community centers, women's groups—could provide safe, accessible venues.

**2. Transportation Access:** Expanding women's access to safe, affordable transportation would reduce mobility barriers.

**3. Workplace Policies:** Flexible work arrangements could enable employed women to maintain friendships while balancing domestic duties.

**4. Social Campaigns:** Challenging norms that view women's friendships as frivolous or threatening to family harmony.

**5. Measurement:** Including social connection metrics in gender equality monitoring frameworks.

# Conclusion

Marriage reduces women's social time by 53%, dropping from 1.7 to 0.8 minutes daily, while reducing men's by only 22% (4.1 to 3.2 minutes). This gendered friendship penalty operates on both participation (whether people socialize) and intensity (time spent per activity). It is twice as large for younger women (≤25), persists across education levels and urban-rural divides, and operates independently of employment status.

These findings reveal marriage as an institution that systematically isolates women from social networks while largely preserving men's connections. The consequences extend beyond personal wellbeing to encompass economic opportunities, social capital, community voice, and intergenerational outcomes.

Addressing this inequality requires recognizing that women's social connections aren't luxuries but fundamental resources. Policies supporting women's maintained social networks—through community infrastructure, transportation access, and normative change—are essential for comprehensive gender equality.

# References {.unnumbered}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

Becker, Gary S. 1991. *A Treatise on the Family*. Enlarged Edition. Cambridge, MA: Harvard University Press.

Berkman, Lisa F., and Thomas Glass. 2000. "Social Integration, Social Networks, Social Support, and Health." In *Social Epidemiology*, edited by Lisa F. Berkman and Ichiro Kawachi, 137-173. Oxford: Oxford University Press.

Chakraborty, Tanika, Anirban Mukherjee, Swapnika Reddy Rachapalli, and Sarani Saha. 2018. "Stigma of Sexual Violence and Women's Decision to Work." *World Development* 103: 226-238.

Chetty, Raj, Matthew O. Jackson, Theresa Kuchler, Johannes Stroebel, Nathaniel Hendren, Robert B. Fluegge, Sara Gong, Federico Gonzalez, Armelle Grondin, Matthew Jacob, Drew Johnston, Martin Koenen, Eduardo Laguna-Muggenburg, Florian Mudekereza, Tom Rutter, Nicolaj Thor, Wilbur Townsend, Ruby Zhang, Mike Bailey, Pablo Barberá, Monica Bhole, and Nils Wernerfelt. 2022. "Social Capital I: Measurement and Associations with Economic Mobility." *Nature* 608: 108-121.

Desai, Sonalde, and Lester Andrist. 2010. "Gender Scripts and Age at Marriage in India." *Demography* 47(3): 667-687.

Gaddis, S. Michael. 2012. "What's in a Relationship? An Examination of Social Capital, Race and Class in Mentoring Relationships." *Social Forces* 90(4): 1237-1269.

Granovetter, Mark S. 1973. "The Strength of Weak Ties." *American Journal of Sociology* 78(6): 1360-1380.

Hirway, Indira. 2010. "Understanding Poverty: Insights Emerging from Time Use of the Poor." In *Unpaid Work and the Economy*, edited by Rania Antonopoulos and Indira Hirway, 22-57. London: Palgrave Macmillan.

Ibarra, Herminia. 1992. "Homophily and Differential Returns: Sex Differences in Network Structure and Access in an Advertising Firm." *Administrative Science Quarterly* 37(3): 422-447.

Jejeebhoy, Shireen J. 1998. "Wife-Beating in Rural India: A Husband's Right?" *Economic and Political Weekly* 33(15): 855-862.

Kalmijn, Matthijs. 2003. "Shared Friendship Networks and the Life Course: An Analysis of Survey Data on Married and Cohabiting Couples." *Social Networks* 25(3): 231-249.

McPherson, Miller, Lynn Smith-Lovin, and Matthew E. Brashears. 2006. "Social Isolation in America: Changes in Core Discussion Networks over Two Decades." *American Sociological Review* 71(3): 353-375.

Munch, Allison, J. Miller McPherson, and Lynn Smith-Lovin. 1997. "Gender, Children, and Social Contact: The Effects of Childrearing for Men and Women." *American Sociological Review* 62(4): 509-520.

Putnam, Robert D. 2000. *Bowling Alone: The Collapse and Revival of American Community*. New York: Simon & Schuster.

Sarkisian, Natalia, and Naomi Gerstel. 2016. "Does Singlehood Isolate or Integrate? Examining the Link Between Marital Status and Ties to Kin, Friends, and Neighbors." *Journal of Social and Personal Relationships* 33(3): 361-384.

Stuart, Toby E., and Olav Sorenson. 2007. "Strategic Networks and Entrepreneurial Ventures." *Strategic Entrepreneurship Journal* 1(3‐4): 211-227.
