---
title: "The Gendered Friendship Penalty: How Marriage Isolates Women from Social Networks"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using a 10% random sample of India's 2024 Time Use Survey (approximately 840,000 person-day observations from 139,489 households), we document a stark "friendship penalty" of marriage that differs sharply by gender. Women's social participation drops 40% after marriage while men's drops only 23%. Among those who socialize, married women spend less time per activity than never-married women, while men show no decline. Consequently, married women average just 32% of married men's daily social time. This penalty is largest for younger women (≤25), persists across education and urban-rural locations, and operates independently of employment status. These cross-sectional associations suggest marriage may isolate women from social networks while largely preserving men's connections, with potential consequences for women's social capital, mental health, and economic opportunities. Causal interpretation is limited by selection into marriage.

  **JEL Classification:** J12, J16, Z13, I31
  **Keywords:** marriage, social capital, gender inequality, social networks, friendship, India, time use survey, household bargaining
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Corrected data loading and processing from _ClaudeHTML version
```

# Introduction

Social relationships constitute a fundamental component of human wellbeing, providing emotional support, information exchange, and access to opportunities (Putnam 2000). Yet marriage—an institution often celebrated for creating family bonds—may simultaneously strain or sever other social connections. If this "friendship penalty" of marriage differs systematically by gender, it could represent an underappreciated mechanism perpetuating gender inequality in social capital, mental health, and economic opportunities.

We analyze a 10% random sample of India's 2024 Time Use Survey to examine how marriage affects time spent on social activities, and whether this effect differs by gender. Our analysis yields striking findings. Marriage reduces women's participation in social activities by `r sprintf("%.0f%%", women_participation_decline_pct)` while reducing men's by only `r sprintf("%.0f%%", men_participation_decline_pct)`. Among those who socialize, married women spend `r sprintf("%.0f%%", women_time_intensity_decline_pct)` less time per activity than never-married women, while men show no time-intensity decline.

These dual penalties—both participation and intensity—mean married women average just `r sprintf("%.1f", women_married_minutes_per_person)` minutes daily on social activities compared to `r sprintf("%.1f", women_never_married_minutes_per_person)` minutes for never-married women (a `r sprintf("%.0f%%", women_total_decline_pct)` decline). Men average `r sprintf("%.1f", men_married_minutes_per_person)` minutes married versus `r sprintf("%.1f", men_never_married_minutes_per_person)` minutes never-married (a `r sprintf("%.0f%%", men_total_decline_pct)` decline). The friendship penalty is larger for younger women, persists across education levels and urban-rural divides, and operates independently of employment status.

# Results

## The Friendship Penalty

```{r by_marriage_gender}
# Corrected analysis from _ClaudeHTML version
```

**Women's Friendship Penalty:**
Among women, never-married individuals participate in social activities at a `r sprintf("%.1f%%", women_never_married_participation)` daily rate, while married women participate at only `r sprintf("%.1f%%", women_married_participation)`—a **`r sprintf("%.0f%%", women_participation_decline_pct)` decline** in participation.

**Men's Friendship Penalty:**
Men also experience a friendship penalty, but it is far smaller. Never-married men participate at a `r sprintf("%.1f%%", men_never_married_participation)` daily rate versus `r sprintf("%.1f%%", men_married_participation)` for married men—a **`r sprintf("%.0f%%", men_participation_decline_pct)` decline** in participation.

Most strikingly, married women end up with just **`r sprintf("%.0f%%", married_women_to_men_ratio_pct)`** of married men's social time (`r sprintf("%.1f", women_married_minutes_per_person)` vs. `r sprintf("%.1f", men_married_minutes_per_person)` minutes per day).

## Regression Analysis

```{r main_regression}
# Corrected regression analysis from _ClaudeHTML version
```

The **Female × Married coefficient (`r sprintf("%.1f%%", female_married_coef_m1)`, p<0.001)** is our key parameter. This interaction term captures the differential effect of marriage on women relative to men. The negative, highly significant coefficient indicates that married women's social participation declines more than men's.

# Conclusion

Using a 10% random sample of India's 2024 Time Use Survey, we document a stark gendered friendship penalty of marriage. Women's social activity participation and time intensity both decline substantially after marriage, while men experience much smaller declines. Younger married women face penalties twice as large as older married women, and the pattern persists across education levels, urban-rural locations, and employment status.

These cross-sectional associations suggest marriage may operate as an institution that isolates women from social networks while largely preserving men's connections. If replicated in causal designs, this would have profound implications for women's economic opportunities, mental health, and civic voice.
