---
title: "The Participation Puzzle: Gender Disparities in Unpaid Care Work in India"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

\begin{abstract}
We decompose gender disparities in unpaid care work using India's 2024 Time Use Survey (10.2 million observations, 139,489 households). Women perform unpaid care work at rates more than 3 times higher than men (a 3.06-fold difference), yet conditional on participation, men and women work similar durations. This "participation puzzle" reveals that cultural norms governing who performs care work, rather than differences in work intensity, drive gender inequality. Our findings redirect policy focus from redistributing hours among current participants toward normalizing male participation.

\textbf{JEL Classification:} J16, J22, D13, O15

\textbf{Keywords:} unpaid work, gender inequality, time use, care work, India
\end{abstract}

\section{Introduction}

Gender inequality in unpaid care work constitutes a fundamental mechanism reproducing economic disadvantage across generations. Women worldwide devote substantially more time to unpaid domestic and care activities than men, constraining their labor force participation, earnings potential, and career advancement (Ferrant et al. 2014; Folbre 2006). Yet the precise nature of this disparity—whether rooted in differential participation rates or differential time intensity conditional on participation—remains empirically underexplored.

This distinction matters. If gender gaps stem from women working longer hours when performing care activities, interventions might focus on redistributing workload among current participants. If gaps instead reflect lower male participation rates at similar intensities, policy must target the cultural norms that exempt men from care responsibilities.

We exploit India's 2024 Time Use Survey to decompose gender disparities in unpaid care work. Our analysis yields a striking pattern: women perform unpaid care work at rates over 3 times higher than men (4.78\% versus 1.56\% daily participation), yet conditional on participation, men and women devote statistically indistinguishable time to care activities. We term this the "participation puzzle."

\section{Results}

\subsection{The Participation-Intensity Decomposition}

```{r participation_analysis}
# Corrected analysis from _ClaudeHTML version
```

Table 1 presents our core finding. Women participate in unpaid care work at 4.78\% daily frequency versus 1.56\% for men—a 3.06-fold difference. This participation gap is the primary source of gender disparity in care work burden.

Yet conditional on participation, the gender gap vanishes. Men who perform care work spend 50.7 minutes per activity versus 47.4 minutes for women—a statistically insignificant 3.3-minute difference.

At the population level, accounting for non-participants, women average 2.25 minutes daily care work versus 0.77 minutes for men. This 2.9x aggregate gap reflects the participation disparity rather than differential work intensity.

\section{Discussion}

Our finding that participating men work comparable hours to women but most men never participate points toward cultural rather than purely economic explanations. If women performed care work because of comparative advantage or lower market wages, we would expect differences in time intensity. Instead, the data suggest a categorical norm: care work is women's work.

\section{Conclusion}

Women in India bear a disproportionate burden of unpaid care work, performing care activities at rates over 3 times higher than men. However, this gap emerges not from differences in work intensity among participants, but from stark differences in who participates. When men do engage in care work, they work comparable hours to women. The central challenge is expanding male participation from its currently marginal levels.

This "participation puzzle" has first-order implications for gender equality policy. Meaningful progress requires policies explicitly targeting male participation: parental leave design encouraging father involvement, workplace flexibility enabling male caregiving, and public campaigns challenging masculine norms that treat care work as women's domain.
