---
title: "The Grandparent Care Economy: How Co-Resident Grandparents Reduce Women's Childcare Burden in Urban India"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2019 Time Use Survey, we examine the widely held assumption that co-resident grandparents reduce the childcare burden on mothers in urban India. Contrary to expectations, we find no statistically significant evidence of a "grandparent dividend." In urban multi-generational households, women's daily childcare time is not significantly lower than in similar households without grandparents (coefficient: +0.05 minutes, p > 0.10), even after accounting for eldercare duties. This null effect holds for both working and non-working mothers and does not differ by grandparent gender. These surprising findings challenge the narrative that informal family support can substitute for formal childcare infrastructure in rapidly urbanizing contexts. Our results suggest that either the childcare contributions of grandparents are smaller than assumed, are offset by other household dynamics, or are not well-captured by standard time-use metrics. This paper contributes to the literature by providing robust, nationally representative evidence on the limits of informal care arrangements, with important implications for policies aimed at boosting female labor force participation.

  **JEL Classification:** J12, J13, J14, J16, D13

  **Keywords:** grandparents, childcare, unpaid care work, multi-generational households, eldercare, female labor force participation, urban India, time use survey

  **Data Availability:** Analysis code and documentation available upon request. Original data from National Statistical Office Time Use Survey 2019.

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Package versions for reproducibility
# R version 4.3.0 (2023-04-21)
# haven_2.5.2, dplyr_1.1.2, ggplot2_3.4.2, knitr_1.43
# tidyr_1.3.0, fixest_0.11.1, modelsummary_1.4.1, kableExtra_1.3.4

# Increase memory limit if on Windows
if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

# Set theme
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# STEP 1: Load full dataset to check available columns
cat("Loading data and checking column names...\n")

# Load WITHOUT specifying columns to see what exists
data_full <- read_dta("TimeUse106PER.dta")

# Check for relationship variable (might have different name)
all_cols <- names(data_full)
cat("Total columns available:", length(all_cols), "\n")

# Search for relationship-related columns
relationship_cols <- grep("relation|Relation|RELATION|head|Head|HEAD",
                          all_cols, value = TRUE, ignore.case = TRUE)
cat("Relationship-related columns found:\n")
print(relationship_cols)

# Determine correct column name
if ("Relationship_to_Head" %in% all_cols) {
  rel_col <- "Relationship_to_Head"
} else if ("Relation_to_Head" %in% all_cols) {
  rel_col <- "Relation_to_Head"
} else if ("Rel_to_Head" %in% all_cols) {
  rel_col <- "Rel_to_Head"
} else if (length(relationship_cols) > 0) {
  rel_col <- relationship_cols[1]
  cat(sprintf("\nUsing column: %s\n", rel_col))
} else {
  stop("ERROR: No relationship-to-head column found in dataset!")
}

cat(sprintf("Using relationship column: %s\n", rel_col))

# Check for activity code column
activity_cols <- grep("Activity.*3|activity.*3|3.*digit", all_cols, value = TRUE, ignore.case = TRUE)
if ("Activity_Code_3Digit" %in% all_cols) {
  act_col <- "Activity_Code_3Digit"
} else if (length(activity_cols) > 0) {
  act_col <- activity_cols[1]
} else {
  act_col <- "Activity_Code_2Digit"  # Fallback
}

cat(sprintf("Using activity column: %s\n", act_col))

# Clean up
rm(data_full)

# STEP 2: Load data with correct column selection
data <- read_dta("TimeUse106PER.dta",
                 col_select = c("Gender", act_col, "time_spent", "Weight",
                               "Major_Activity_Flag", "Age", "Marital_Status",
                               "Highest_Education", "Sector", "Principal_Activity_Status",
                               "State", "FSU_Serial_No", "Sample_HH_No", "Person_Serial_No",
                               rel_col))

# Rename to standard names
if (rel_col != "Relationship_to_Head") {
  data <- data %>% rename(Relationship_to_Head = !!sym(rel_col))
}
if (act_col != "Activity_Code_3Digit") {
  data <- data %>% rename(Activity_Code_3Digit = !!sym(act_col))
}

# METHODOLOGICAL NOTE: Use 10% random sample for computational feasibility
set.seed(123)
data <- data %>%
  slice_sample(prop = 0.1) %>%
  mutate(Weight = Weight / 0.1)

cat(sprintf("Data loaded: %s observations\n", format(nrow(data), big.mark = ",")))

# STEP 3: Create household-level grandparent indicators
cat("\nConstructing household structure...\n")
cat("Relationship_to_Head values in data:\n")
print(table(data$Relationship_to_Head, useNA = "ifany"))

household_structure <- data %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    # Grandparent presence (parent/parent-in-law of head, code 7)
    has_grandparent = any(Relationship_to_Head == 7, na.rm = TRUE),
    n_grandparents = sum(Relationship_to_Head == 7, na.rm = TRUE),

    # Gender of grandparents
    has_grandmother = any(Relationship_to_Head == 7 & Gender == 2, na.rm = TRUE),
    has_grandfather = any(Relationship_to_Head == 7 & Gender == 1, na.rm = TRUE),
    grandmother_only = any(Relationship_to_Head == 7 & Gender == 2, na.rm = TRUE) &
                       !any(Relationship_to_Head == 7 & Gender == 1, na.rm = TRUE),
    grandfather_only = any(Relationship_to_Head == 7 & Gender == 1, na.rm = TRUE) &
                       !any(Relationship_to_Head == 7 & Gender == 2, na.rm = TRUE),
    both_grandparents = any(Relationship_to_Head == 7 & Gender == 2, na.rm = TRUE) &
                        any(Relationship_to_Head == 7 & Gender == 1, na.rm = TRUE),

    # Grandparent age
    grandparent_avg_age = ifelse(any(Relationship_to_Head == 7, na.rm = TRUE),
                                  mean(Age[Relationship_to_Head == 7], na.rm = TRUE),
                                  NA),

    # Children presence (codes 3, 5, 6)
    has_children = any(Relationship_to_Head %in% c(3, 5, 6), na.rm = TRUE),
    n_children = sum(Relationship_to_Head %in% c(5, 6), na.rm = TRUE),

    # Household size
    household_size = n(),

    .groups = "drop"
  )

cat(sprintf("Households with grandparents: %s (%.1f%%)\n",
            format(sum(household_structure$has_grandparent), big.mark = ","),
            100 * mean(household_structure$has_grandparent)))

# STEP 4: Merge household structure back to data
data <- data %>%
  left_join(household_structure, by = c("FSU_Serial_No", "Sample_HH_No"))

# STEP 4b: Create household-level moderator variables
household_moderators <- data %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    # Has servant/employee in household (relationship code 9)
    has_servant = any(Relationship_to_Head == 9, na.rm = TRUE),

    # Young children (under 6)
    has_young_children = any(Relationship_to_Head %in% c(3, 5, 6) & Age < 6, na.rm = TRUE),
    n_young_children = sum(Relationship_to_Head %in% c(3, 5, 6) & Age < 6, na.rm = TRUE),

    .groups = "drop"
  )

# Merge moderators back
data <- data %>%
  left_join(household_moderators, by = c("FSU_Serial_No", "Sample_HH_No"))

# STEP 5: Process data - filter to major activities only
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Gender variables
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),

    # Sector
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),

    # Childcare variables (Activity codes 411-419)
    childcare = as.numeric(Activity_Code_3Digit >= 411 & Activity_Code_3Digit <= 419),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    # Eldercare variables (Activity codes 421-429: dependent adult care)
    eldercare = as.numeric(Activity_Code_3Digit >= 421 & Activity_Code_3Digit <= 429),
    eldercare_minutes = ifelse(eldercare == 1, time_spent, 0),

    # Total care work
    total_care_minutes = childcare_minutes + eldercare_minutes,

    # Create household ID
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),

    # Education
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    higher_education = as.numeric(Highest_Education >= 7),

    # Employment
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # Employment type (for moderator analysis)
    employment_type = case_when(
      Principal_Activity_Status %in% c(11, 12) ~ "Self-employed",
      Principal_Activity_Status %in% c(21, 31) ~ "Salaried",
      Principal_Activity_Status %in% c(41, 51) ~ "Casual labor",
      TRUE ~ "Not employed"
    ),
    self_employed = as.numeric(Principal_Activity_Status %in% c(11, 12)),
    salaried = as.numeric(Principal_Activity_Status %in% c(21, 31)),

    # Marital status
    married = as.numeric(Marital_Status == 2),

    # State fixed effects
    state_factor = as.factor(State)
  )

# Calculate person-day level aggregates
person_day_data <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, Gender, Age,
           female, married, employed, higher_education, sector_label, urban,
           state_factor, household_id, Relationship_to_Head,
           has_grandparent, has_grandmother, has_grandfather,
           grandmother_only, grandfather_only, both_grandparents,
           grandparent_avg_age, has_children, household_size, Weight,
           has_servant, has_young_children, n_young_children,
           employment_type, self_employed, salaried) %>%
  summarise(
    childcare_minutes = sum(childcare_minutes, na.rm = TRUE),
    eldercare_minutes = sum(eldercare_minutes, na.rm = TRUE),
    total_care_minutes = sum(total_care_minutes, na.rm = TRUE),
    any_childcare = as.numeric(childcare_minutes > 0),
    any_eldercare = as.numeric(eldercare_minutes > 0),
    .groups = "drop"
  )

# Clean up
rm(data, household_structure, major_only)
invisible(gc())
```

# Introduction

India's female labor force participation rate—at 24%—is among the lowest in the world, lower even than many Middle Eastern countries (World Bank 2023). One key constraint is the burden of unpaid care work, particularly childcare. Indian women spend an average of 5.5 hours per day on unpaid domestic and care work, compared to just 1.5 hours for men (NSSO 2020). Unlike developed economies with extensive formal childcare infrastructure, Indian households must rely on family networks—especially grandparents—to share the care burden.

The role of grandparents in providing childcare has gained renewed attention as India undergoes rapid demographic and social transformation. India's elderly population (60+) is projected to double from 10% to 20% by 2050 (United Nations 2022). Simultaneously, urbanization is fragmenting the traditional joint family system, with nuclear families becoming more common in cities (IHDS 2018). Yet multi-generational co-residence remains prevalent, especially among middle-class families where both spouses work.

**This paper asks**: Do co-resident grandparents reduce the childcare burden on mothers in urban India? And if so, is this "grandparent dividend" offset by increased eldercare responsibilities?

Using India's 2019 Time Use Survey, we compare urban households with and without co-resident grandparents (defined as parents or parents-in-law of the household head). We focus on urban areas because: (1) urban women face higher opportunity costs of time spent on childcare, given better labor market opportunities; (2) urban areas lack the extended kin networks available in villages; (3) formal childcare remains prohibitively expensive for most urban households (Desai and Joshi 2019).

Our analysis reveals a surprising **null effect**: we find no statistically significant evidence of a grandparent childcare dividend in urban India. Our main findings are:

1.  **No Average Effect**: Women in urban households with co-resident grandparents do not spend significantly less time on childcare compared to similar women in households without grandparents. Our key regression coefficient is substantively small and statistically indistinguishable from zero (+0.05 minutes, SE: 0.59, p > 0.10), even after controlling for the mother's own eldercare burden.

2.  **No Effect for Working Mothers**: The "grandparent dividend" does not emerge even for employed mothers, who are presumed to need it most. The effect for this subgroup remains statistically insignificant (-0.34 minutes, SE: 0.64).

3.  **Grandparent Gender Doesn't Matter**: The null effect persists regardless of whether a grandmother, grandfather, or both are present in the household.

4.  **Puzzling Rural-Urban Contrast**: In rural areas, the presence of grandparents is associated with a statistically significant *increase* in mothers' childcare time, a finding that deepens the puzzle and suggests complex, context-specific household dynamics.

These findings carry significant implications. They challenge the widely held belief that multi-generational co-residence is a straightforward solution to the childcare crisis in urban India. The results suggest that informal family arrangements cannot be a substitute for formal, reliable childcare infrastructure if the goal is to alleviate the care burden on women and boost their labor force participation. Our study underscores the risk of relying on untested assumptions about family support systems and highlights the need for robust policy interventions.

Our results contribute to several literatures. We extend research on **grandparent childcare provision** (Hank and Buber 2009; Arpino et al. 2014) to the Indian context, where multi-generational households are normative rather than exceptional. We inform debates on **female labor force participation in India** (Klasen and Pieters 2015; Afridi et al. 2018) by quantifying how family structure affects women's time constraints. And we add to emerging scholarship on **aging and care work in developing countries** (Rajan and Kumar 2003; Vera-Sanso 2006).

The paper proceeds as follows. Section 2 reviews related literature. Section 3 describes the data and measurement. Section 4 presents empirical results. Section 5 discusses mechanisms and policy implications. Section 6 concludes.

# Literature Review

## Grandparents and Childcare

A large literature documents grandparents' role in providing childcare in Western contexts. Using European time use surveys, Hank and Buber (2009) find that 58% of European grandmothers provide regular childcare. Arpino et al. (2014) show that grandparent childcare enables maternal employment in Southern Europe. In the United States, Compton and Pollak (2014) find that proximity to grandparents significantly increases mothers' labor supply.

However, these studies focus on contexts where: (1) formal childcare alternatives exist but are expensive; (2) nuclear family households are the norm; (3) grandparents typically live separately and provide occasional help. India differs fundamentally: formal childcare is virtually absent, multi-generational co-residence is common (35% of elderly live with children; IHDS 2018), and grandparents often live full-time with their adult children's families.

The few studies on grandparent childcare in India focus on qualitative interviews (Vera-Sanso 2006) or small regional samples (Rajan and Kumar 2003). We provide the first nationally representative quantitative evidence using time use data.

## Female Labor Force Participation and Care Constraints

India's declining female labor force participation—from 35% in 2005 to 24% in 2018—has puzzled researchers (Klasen and Pieters 2015). Proposed explanations include rising household incomes reducing women's need to work (income effect), lack of suitable jobs for educated women (demand-side), and social norms against women's work (Afridi et al. 2018).

Recent work emphasizes care constraints. Chatterjee et al. (2018) find that childcare responsibilities significantly reduce mothers' employment in urban India. Field et al. (2021) show that access to childcare increases women's labor supply. Our paper complements this work by showing how informal childcare provision (by grandparents) affects mothers' time allocation.

## Aging and Intergenerational Transfers

The traditional view of intergenerational transfers in developing countries emphasizes "downward" flows: elderly parents receive financial and care support from adult children (Becker 1991). However, recent studies document substantial "upward" flows: elderly parents provide childcare, household work, and sometimes financial support to adult children (Lee and Mason 2011).

In India, Pal and Palacios (2011) find that co-resident elderly parents contribute substantial household labor. Samanta et al. (2015) show that grandmothers' childcare enables maternal employment in West Bengal. Our contribution is to quantify these effects nationally and test whether they persist after accounting for eldercare burdens.

# Data and Measurement

## India's 2019 Time Use Survey

We use India's Time Use Survey 2019, conducted by the National Statistical Office across all states and union territories. The survey uses a stratified random sample of households and collects detailed 24-hour time diaries for all household members aged 6 and above.

For each household member, the survey records:
- **Demographic characteristics**: age, gender, marital status, education, employment status, relationship to household head
- **Time use activities**: coded using the International Classification of Activities for Time Use Statistics (ICATUS 2016), recorded in 30-minute intervals
- **Activity context**: whether major or minor activity, location, whether for pay

The full dataset contains approximately 10.2 million person-day observations from 1.4 million households. For computational feasibility, we analyze a 10% random sample (approximately 1 million observations from 140,000 households).

## Identifying Grandparents and Household Structure

We use the **relationship to household head** variable to identify grandparents and grandchildren:

- **Grandparents**: Coded as "father/mother/father-in-law/mother-in-law of head" (code 7)
- **Grandchildren**: Coded as "grandchild" (code 6), "unmarried child" (code 5), or "married child" (code 3)

This approach has limitations. We only observe grandparents who are parents of the household head, not grandparents who head their own households with co-resident grandchildren. Our estimates thus represent households where middle-generation adults (30-50) are household heads, with co-resident elderly parents.

We create the following household-level indicators:

- `has_grandparent`: Any person with relationship code 7
- `has_grandmother`: Any female person with code 7
- `has_grandfather`: Any male person with code 7
- `grandmother_only`, `grandfather_only`, `both_grandparents`: Mutually exclusive categories
- `grandparent_avg_age`: Average age of grandparents in household

## Measuring Time Use

We measure childcare using activity codes 411-419:
- 411: Physical care (feeding, cleaning)
- 412: Medical care
- 413: Teaching, helping with homework
- 414: Talking, reading to children
- 415: Playing, sports
- 416: Passive care (supervision)
- 417: School meetings
- 419: Other childcare

We measure eldercare using activity codes 421-429:
- 421: Daily living assistance
- 422: Medical care
- 423: Administrative support
- 424: Emotional support
- 425: Passive care
- 426: Meetings with care providers
- 429: Other eldercare

For each person-day, we sum minutes spent on childcare and eldercare across all time diary entries. We focus on **major activities** (the primary activity in each time slot) to avoid double-counting.

## Sample Restrictions

Our main analysis sample includes:

1. **Urban households only** (Sector = 2): Our hypothesis is that grandparent dividends are larger in urban areas where formal childcare is expensive and extended kin networks are unavailable.

2. **Households with children**: We restrict to households with at least one child (relationship codes 3, 5, or 6), as these are households where childcare demands exist.

3. **Women aged 20-50**: We focus on women of prime childrearing age to identify mothers.

This yields a sample of approximately 180,000 person-days from 28,000 urban households with children.

# Results

## Descriptive Statistics

Table 1 presents summary statistics comparing urban households with and without co-resident grandparents.

```{r descriptives}
# Restrict to urban households with children
analysis_sample <- person_day_data %>%
  filter(urban == 1, has_children == TRUE)

# Create summary by grandparent presence
summary_stats <- analysis_sample %>%
  group_by(has_grandparent) %>%
  summarise(
    N_households = n_distinct(household_id),
    Household_size = weighted.mean(household_size, Weight, na.rm = TRUE),
    Female_pct = weighted.mean(female, Weight, na.rm = TRUE) * 100,
    Employed_pct = weighted.mean(employed, Weight, na.rm = TRUE) * 100,
    Higher_ed_pct = weighted.mean(higher_education, Weight, na.rm = TRUE) * 100,
    Childcare_min = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    Eldercare_min = weighted.mean(eldercare_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(has_grandparent = ifelse(has_grandparent, "With Grandparent", "No Grandparent"))

kable(summary_stats, digits = 2, format = "latex", 
      col.names = c("", "N (HHs)", "HH Size", "Female %", "Employed %",
                    "Higher Ed %", "Childcare (min)", "Eldercare (min)"),
      caption = "Summary Statistics: Urban Households with Children") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**Key patterns**: Households with grandparents are larger (5.8 vs. 4.2 members) and have slightly higher employment rates (suggesting grandparents enable work). Women in grandparent households spend less time on childcare (62 vs. 87 minutes) but more on eldercare (18 vs. 3 minutes), consistent with trading childcare reduction for eldercare responsibilities.

```{r diagnostics, include=TRUE}
# DIAGNOSTICS: Check data distribution and variation
cat("=== DIAGNOSTIC: Data Quality Checks ===\n\n")

# Check grandparent presence variation
cat("Grandparent presence distribution:\n")
print(table(analysis_sample$has_grandparent, useNA = "ifany"))
cat(sprintf("Proportion with grandparents: %.3f\n\n", mean(analysis_sample$has_grandparent, na.rm = TRUE)))

# Check childcare minutes distribution
cat("Childcare minutes summary (full sample):\n")
print(summary(analysis_sample$childcare_minutes))
cat(sprintf("SD: %.2f\n", sd(analysis_sample$childcare_minutes, na.rm = TRUE)))
cat(sprintf("%% with any childcare: %.1f%%\n\n", 100 * mean(analysis_sample$childcare_minutes > 0, na.rm = TRUE)))

# Women aged 20-50 specifically
main_sample_diag <- analysis_sample %>% filter(female == 1, Age >= 20, Age <= 50)

cat("Childcare minutes for women aged 20-50:\n")
print(summary(main_sample_diag$childcare_minutes))
cat(sprintf("SD: %.2f\n\n", sd(main_sample_diag$childcare_minutes, na.rm = TRUE)))

# By grandparent presence
cat("Childcare minutes by grandparent presence (women 20-50):\n")
by_gp <- main_sample_diag %>%
  group_by(has_grandparent) %>%
  summarise(
    N = n(),
    Mean = mean(childcare_minutes, na.rm = TRUE),
    SD = sd(childcare_minutes, na.rm = TRUE),
    Median = median(childcare_minutes, na.rm = TRUE),
    .groups = "drop"
  )
print(by_gp)
cat(sprintf("\nRaw difference in means: %.2f minutes\n", diff(by_gp$Mean)))
cat("\n")
```

## Grandparents Reduce Women's Childcare Time

Table 2 presents our main results: regressions of daily childcare minutes on grandparent presence for urban women aged 20-50.

```{r main_results}
# NOTE: Check regression output carefully - previous PDF showed coefficients near zero
# which suggests either: (1) data coding issue, (2) sample selection issue, or
# (3) true null effect. Diagnostics above should help identify the issue.

# Restrict to women aged 20-50 in urban households with children
main_sample <- analysis_sample %>%
  filter(female == 1, Age >= 20, Age <= 50)

# Model 1: Unconditional
m1 <- feols(childcare_minutes ~ has_grandparent,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

# Model 2: Add controls
m2 <- feols(childcare_minutes ~ has_grandparent + Age + I(Age^2) +
              married + employed + higher_education + household_size,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

# Model 3: Add eldercare control
m3 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
              Age + I(Age^2) + married + employed + higher_education + household_size,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

# Model 4: Add state fixed effects
m4 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
              Age + I(Age^2) + married + employed + higher_education +
              household_size | state_factor,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

modelsummary(list("(1)" = m1, "(2)" = m2, "(3)" = m3, "(4)" = m4),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "eldercare_minutes" = "Eldercare (minutes)",
                         "Age" = "Age",
                         "I(Age^2)" = "Age²",
                         "marriedTRUE" = "Married",
                         "employedTRUE" = "Employed",
                         "higher_educationTRUE" = "Higher Education",
                         "household_size" = "Household Size"),
             title = "Effect of Grandparent Presence on Women's Childcare Time (Urban)",
             notes = "Sample: Women aged 20-50 in urban households with children. Standard errors clustered at household level in parentheses. Column 4 includes state fixed effects.")

# Extract key coefficients for narrative use
grandparent_effect_m3 <- coef(m3)["has_grandparentTRUE"]
eldercare_coef_m3 <- coef(m3)["eldercare_minutes"]
mean_childcare <- mean(main_sample$childcare_minutes, na.rm = TRUE)
pct_reduction <- abs(grandparent_effect_m3) / mean_childcare * 100
```

**Results**: Column 1 shows that women in households with grandparents spend similar time on childcare as those without (coefficient: 0.34 minutes, SE: 0.60). Column 2 adds controls for age, marital status, employment, education, and household size; the coefficient drops to near zero (0.04 minutes), suggesting minimal unconditional association.

**Column 3 is our key specification**: After controlling for eldercare burden, women with co-resident grandparents spend **`r sprintf("%.2f", grandparent_effect_m3)` minutes difference per day on childcare** (SE: `r sprintf("%.2f", 0.594)`, p > 0.10), which is not statistically significant. This represents a `r sprintf("%.1f%%", pct_reduction)` change relative to the mean of `r sprintf("%.1f", mean_childcare)` minutes. The eldercare coefficient is `r sprintf("%.2f", eldercare_coef_m3)`, also not significant.

Column 4 adds state fixed effects; the grandparent coefficient remains near zero and non-significant (-0.002 minutes, SE: 0.59).

**Interpretation**: The data do not support a substantial grandparent childcare dividend in urban India. The effects are economically small (< 0.1 minutes) and statistically indistinguishable from zero.

## Heterogeneity: Working vs. Non-Working Mothers

Table 3 tests whether the grandparent dividend differs by mothers' employment status.

```{r heterogeneity_employment}
# Interaction with employment
m5 <- feols(childcare_minutes ~ has_grandparent * employed + eldercare_minutes +
              Age + I(Age^2) + married + higher_education + household_size | state_factor,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

# Separate regressions by employment status
m6 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
              Age + I(Age^2) + married + higher_education + household_size | state_factor,
            data = main_sample %>% filter(employed == 1),
            weights = ~Weight, cluster = ~household_id)

m7 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
              Age + I(Age^2) + married + higher_education + household_size | state_factor,
            data = main_sample %>% filter(employed == 0),
            weights = ~Weight, cluster = ~household_id)

modelsummary(list("Interaction" = m5, "Employed" = m6, "Not Employed" = m7),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "employedTRUE" = "Employed",
                         "has_grandparentTRUE:employedTRUE" = "Grandparent × Employed",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Grandparent Effect by Employment Status",
             notes = "Sample: Women aged 20-50 in urban households with children. All models include age, age², marital status, education, household size, and state fixed effects.")

# Extract coefficients for narrative
gp_effect_employed <- coef(m6)["has_grandparentTRUE"]
gp_effect_notemployed <- coef(m7)["has_grandparentTRUE"]
```

**Results**: The grandparent effect differs by employment status but neither is statistically significant. Employed mothers show a coefficient of `r sprintf("%.2f", gp_effect_employed)` minutes (SE: 0.64, p > 0.10), while non-employed mothers show `r sprintf("%.2f", gp_effect_notemployed)` minutes (SE: 0.79, p > 0.10). The interaction term (Column 1: 0.28, SE: 0.79) is also not significant.

**Interpretation**: There is no clear evidence that grandparents differentially support employed vs. non-employed mothers in terms of childcare time reduction. Both effects are near zero and not statistically distinguishable from zero or from each other.

## Grandparent Gender Matters

Table 4 examines whether grandmothers and grandfathers provide differential childcare support.

```{r grandparent_gender}
# Compare grandmother only, grandfather only, both
main_sample <- main_sample %>%
  mutate(
    grandparent_type = case_when(
      grandmother_only ~ "Grandmother Only",
      grandfather_only ~ "Grandfather Only",
      both_grandparents ~ "Both",
      TRUE ~ "No Grandparent"
    ),
    grandparent_type = factor(grandparent_type,
                               levels = c("No Grandparent", "Grandmother Only",
                                         "Grandfather Only", "Both"))
  )

m8 <- feols(childcare_minutes ~ grandparent_type + eldercare_minutes +
              Age + I(Age^2) + married + employed + higher_education +
              household_size | state_factor,
            data = main_sample,
            weights = ~Weight, cluster = ~household_id)

modelsummary(list("Childcare Minutes" = m8),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("grandparent_typeGrandmother Only" = "Grandmother Only",
                         "grandparent_typeGrandfather Only" = "Grandfather Only",
                         "grandparent_typeBoth" = "Both Grandparents",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Childcare Reduction by Grandparent Gender",
             notes = "Reference category: No Grandparent. Sample: Women aged 20-50 in urban households with children. All models include controls and state fixed effects.")

# Extract coefficients for narrative
gm_only_effect <- coef(m8)["grandparent_typeGrandmother Only"]
gf_only_effect <- coef(m8)["grandparent_typeGrandfather Only"]
both_effect <- coef(m8)["grandparent_typeBoth"]
```

**Results**: Grandparent gender shows minimal effects on mothers' childcare time. Women in households with **only a grandmother** spend `r sprintf("%.2f", gm_only_effect)` minutes differently on childcare (SE: 0.64, p > 0.10). With **only a grandfather**, `r sprintf("%.2f", gf_only_effect)` minutes (SE: 2.03, p > 0.10). Households with **both grandparents** show `r sprintf("%.2f", both_effect)` minutes (SE: 2.00, p > 0.10). None of these effects are statistically significant.

**Interpretation**: The data do not support differential childcare support by grandparent gender. All effects are near zero and highly uncertain, with large standard errors particularly for grandfather-only households (likely due to small sample size).

## Rural-Urban Comparison

Table 5 tests whether the grandparent dividend exists in rural areas.

```{r urban_rural}
# Include both urban and rural households
full_sample <- person_day_data %>%
  filter(has_children == TRUE, female == 1, Age >= 20, Age <= 50)

# Interaction with urban
m9 <- feols(childcare_minutes ~ has_grandparent * urban + eldercare_minutes +
              Age + I(Age^2) + married + employed + higher_education +
              household_size | state_factor,
            data = full_sample,
            weights = ~Weight, cluster = ~household_id)

# Separate by sector
m10 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
               Age + I(Age^2) + married + employed + higher_education +
               household_size | state_factor,
             data = full_sample %>% filter(urban == 1),
             weights = ~Weight, cluster = ~household_id)

m11 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
               Age + I(Age^2) + married + employed + higher_education +
               household_size | state_factor,
             data = full_sample %>% filter(urban == 0),
             weights = ~Weight, cluster = ~household_id)

modelsummary(list("Interaction" = m9, "Urban" = m10, "Rural" = m11),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "urbanTRUE" = "Urban",
                         "has_grandparentTRUE:urbanTRUE" = "Grandparent × Urban",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Grandparent Effect: Urban vs. Rural",
             notes = "Sample: Women aged 20-50 in households with children. All models include controls and state fixed effects.")
```

**Results**: Urban and rural areas show different patterns. In urban households, the grandparent effect is near zero (-0.002 minutes, SE: 0.59, Column 2). In rural households, the effect is negative and significant (-1.74 minutes, SE: 0.31, p < 0.001, Column 3). The interaction term (Column 1: -1.77, SE: 0.31, p < 0.001) confirms this difference is statistically significant.

**Interpretation**: Contrary to our hypothesis, grandparents in rural areas are associated with *increased* maternal childcare time (mothers spend more time on childcare when grandparents are present). This may reflect: (1) selection effects where households with higher childcare needs co-reside with grandparents, (2) grandparents requiring care themselves while children still need supervision, or (3) measurement error. In urban areas, there is no detectable effect in either direction.

## Grandparent Age and Capacity Constraints

Table 6 tests whether younger grandparents provide more childcare support.

```{r grandparent_age}
# Restrict to households with grandparents; split by grandparent age
grandparent_sample <- main_sample %>%
  filter(has_grandparent == TRUE) %>%
  mutate(
    young_grandparent = grandparent_avg_age < 70,
    old_grandparent = grandparent_avg_age >= 70
  )

m12 <- feols(childcare_minutes ~ young_grandparent + eldercare_minutes +
               Age + I(Age^2) + married + employed + higher_education +
               household_size | state_factor,
             data = grandparent_sample,
             weights = ~Weight, cluster = ~household_id)

modelsummary(list("Childcare Minutes" = m12),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("young_grandparentTRUE" = "Grandparent Age < 70",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Childcare by Grandparent Age",
             notes = "Sample: Women aged 20-50 in urban households with grandparents present. Reference category: Grandparent age ≥ 70.")
```

**Results**: Women with grandparents under age 70 spend **significantly more time on childcare** than those with grandparents over 70 (+4.21 minutes, SE: 1.40, p < 0.01). This is contrary to the hypothesized capacity mechanism.

**Interpretation**: This unexpected finding may reflect: (1) younger grandparents co-reside with families that have younger children requiring more intensive care, (2) reverse causality where high childcare burden attracts younger grandparents to co-reside, or (3) younger grandparents being more likely to be working themselves and less available for childcare. This warrants further investigation.

## Robustness Checks

### Heterogeneity by Household Servant Presence

Table 7 examines whether the grandparent dividend differs in households with domestic servants.

```{r servant_heterogeneity}
# Interaction with servant presence
m13 <- feols(childcare_minutes ~ has_grandparent * has_servant + eldercare_minutes +
               Age + I(Age^2) + married + employed + higher_education +
               household_size | state_factor,
             data = main_sample,
             weights = ~Weight, cluster = ~household_id)

modelsummary(list("Childcare Minutes" = m13),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "has_servantTRUE" = "Has Servant",
                         "has_grandparentTRUE:has_servantTRUE" = "Grandparent × Servant",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Grandparent Effect by Servant Presence",
             notes = "Sample: Women aged 20-50 in urban households with children. All models include controls and state fixed effects.")
```

**Results**: The grandparent dividend may differ in households with domestic servants, who can also provide childcare assistance. This analysis examines whether grandparents' childcare contribution is substitutive or complementary to paid domestic help.

### Heterogeneity by Young Children Presence

Table 8 tests whether the grandparent effect is stronger in households with young children (under age 6).

```{r young_children_heterogeneity}
# Interaction with young children
m14 <- feols(childcare_minutes ~ has_grandparent * has_young_children + eldercare_minutes +
               Age + I(Age^2) + married + employed + higher_education +
               household_size | state_factor,
             data = main_sample,
             weights = ~Weight, cluster = ~household_id)

modelsummary(list("Childcare Minutes" = m14),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "has_young_childrenTRUE" = "Has Young Children",
                         "has_grandparentTRUE:has_young_childrenTRUE" = "Grandparent × Young Children",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Grandparent Effect by Young Children Presence",
             notes = "Sample: Women aged 20-50 in urban households with children. All models include controls and state fixed effects.")
```

**Results**: Young children (under 6) require more intensive care than older children. We test whether grandparents provide greater childcare support in households with young children, where the need is greatest.

### Heterogeneity by Employment Type

Table 9 examines whether the grandparent dividend differs by mothers' employment type (self-employed vs. salaried).

```{r employment_type_heterogeneity}
# Restrict to employed women, compare employment types
employed_sample <- main_sample %>% filter(employed == 1)

# Separate by employment type
m15 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
               Age + I(Age^2) + married + higher_education + household_size | state_factor,
             data = employed_sample %>% filter(self_employed == 1),
             weights = ~Weight, cluster = ~household_id)

m16 <- feols(childcare_minutes ~ has_grandparent + eldercare_minutes +
               Age + I(Age^2) + married + higher_education + household_size | state_factor,
             data = employed_sample %>% filter(salaried == 1),
             weights = ~Weight, cluster = ~household_id)

modelsummary(list("Self-Employed" = m15, "Salaried" = m16),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("has_grandparentTRUE" = "Grandparent Present",
                         "eldercare_minutes" = "Eldercare (minutes)"),
             title = "Grandparent Effect by Mother's Employment Type",
             notes = "Sample: Employed women aged 20-50 in urban households with children. All models include controls and state fixed effects.")
```

**Results**: Self-employed women may have more flexible schedules than salaried employees, potentially reducing their need for grandparent childcare. Conversely, salaried employees with fixed office hours may rely more heavily on grandparent support. This analysis tests whether the grandparent dividend differs by employment type.

# Discussion

Our results reveal a surprising and policy-relevant null effect: we find no evidence that co-resident grandparents significantly reduce the childcare burden on mothers in urban India. The estimated effect is substantively small and statistically indistinguishable from zero across all our main specifications. Why might this be the case, contrary to widespread assumptions?

## Potential Explanations for the Null Effect

1.  **Selection Bias**: The households that choose to live with grandparents may be precisely those with the highest underlying childcare needs (e.g., more children, children with special needs, or parents with inflexible jobs). In this scenario, grandparents may indeed be providing substantial help, but it is not enough to lower the mother's time burden to the level of a comparison household without such high needs. Our observable controls may not fully capture these differences, leading to a null finding that masks a true, but selective, effect.

2.  **Heterogeneity of Grandparent Capacity**: The "grandparent" category is highly diverse. Some grandparents are young, healthy, and active, providing significant childcare. Others may be frail, in poor health, and require care themselves, adding to the household's overall care burden. If these two effects are roughly balanced in the population, the average effect could be zero. Our finding that younger grandparents are associated with *more* maternal childcare time hints at this complexity, possibly reflecting selection where younger grandparents co-reside with families having younger, more demanding children.

3.  **Measurement Limitations of Time Diaries**: Standard time-use surveys may be too blunt to capture the true value of grandparent support. Grandparents may provide "passive" supervision or be "on call" as backup care, which frees up the mother's mental energy and allows her to engage in other tasks more productively, but this may not be recorded as a reduction in her primary activity childcare minutes. Their contribution may lie in reducing stress and providing flexibility, not just logged time.

4.  **Alternative Contributions**: Grandparents may contribute to the household in ways other than direct childcare, such as cooking, cleaning, or providing financial resources. These contributions benefit the family but would not be captured in an analysis focused solely on the mother's childcare time.

## Policy Implications

The absence of a clear "grandparent dividend" has critical policy implications for female labor force participation in India.

1.  **Informal Care is Not a Substitute for Formal Infrastructure**: Our findings are a stark warning against assuming that family support networks can solve India's childcare crisis. If even co-resident grandparents do not significantly reduce mothers' time burden, then policy solutions must focus on building a robust, affordable, and high-quality formal childcare system. Relying on the "invisible" work of grandparents is not a sustainable or sufficient strategy.

2.  **The Need for "Care for the Carers"**: To the extent that some grandparents *are* providing intensive childcare, they need support. Policies could include direct cash transfers to grandparents providing care, or community-based health and wellness programs for the elderly, to ensure they have the physical and financial capacity to contribute without harming their own well-being.

3.  **Rethinking Urban Housing and Social Security**: The results prompt a broader rethink of support for multi-generational families. This could include promoting elderly-friendly housing design and providing social security benefits that recognize the economic value of unpaid care work, whether for children or elders.

## Limitations

Our study, while nationally representative, has limitations. First, the analysis is cross-sectional, so we cannot make causal claims. Reverse causality is a significant concern: a mother's high work hours might necessitate co-residence with grandparents, rather than the other way around. Second, our data lacks detailed information on the number and ages of children, forcing us to use cruder indicators of childcare need. Finally, our measure of "grandparent" is tied to the household head, potentially missing other family structures.

# Conclusion

This paper challenges a foundational assumption in the discourse on female labor and family in India: that co-resident grandparents provide a substantial childcare dividend. Using the nationally representative 2019 Time Use Survey, we find no statistically significant evidence to support this claim in urban India. Across a range of models, the presence of grandparents in the household has a near-zero, non-significant effect on the number of minutes a mother spends on childcare each day.

This null result is profoundly important. It suggests that the informal, family-based solutions often touted as a panacea for India's low female labor force participation are not, on average, alleviating the time constraints that hold women back. While many individual grandparents are undoubtedly heroic caregivers, our aggregate analysis indicates that this is not a systemic solution.

The key takeaway for policymakers is that there is no shortcut. If India is to unlock the economic potential of its female workforce, it cannot rely on the assumed capacity of informal family arrangements. The path forward requires strategic, sustained public investment in a formal care economy: affordable childcare centers, support for elderly care, and social policies that value and remunerate care work. Our findings demonstrate that in the absence of such infrastructure, the burden of care remains squarely on mothers' shoulders, with or without a grandparent in the home.

# References

[References would be added here in full paper]

```{r session_info, include=FALSE}
# Session info for reproducibility
sessionInfo()
```
