---
title: "Structure Over Culture: How Economic and Demographic Constraints Dominate Religious Identity in Shaping Daily Life"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  We examine intersectionality across multiple dimensions—income, religion, region, household size, employment status, and age—in time use patterns using India's 2024 Time Use Survey. Our central finding challenges conventional wisdom about religious and cultural determinism: **structural economic and demographic variables create far more similarity in daily life than religion creates difference**. Lower-income Hindus, Muslims, Sikhs, and Christians allocate their time remarkably similarly—the standard deviation in domestic work across religions within the lowest income tercile is only 14 minutes per day. In contrast, within-religion variation across income levels averages 89 minutes per day, a 6.4-fold difference. This pattern extends beyond income: household size, employment status, and life stage (age cohorts) all create strong cross-religious convergence. Our variance decomposition reveals that structural variables (income 30%, employment 25%, household size 18%, age 12%) collectively explain 85% of variance, while religion explains only 7%. Women in large households, employed women, and women in peak childbearing years (30-44) show similar time use patterns regardless of religion. These findings have profound implications for policy and social theory, suggesting that universal economic and structural interventions may be more effective than culture-specific programs in addressing time poverty. **Shared economic circumstances create more similarity than shared religious identity**.

  **JEL Classification:** J16, J22, Z12, Z13, I32, J13

  **Keywords:** intersectionality, income inequality, religion, class, time use, household structure, employment, India, cultural economics, structural constraints
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)
library(viridis)
library(RColorBrewer)
library(reshape2)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load India Time Use Survey 2024 data
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, Religion, FSU_Serial_No, Sample_HH_No,
                               Household_Size))

# Use 10% random sample with weight adjustment for computational efficiency
set.seed(123)
data <- data %>%
  slice_sample(prop = 0.1) %>%
  mutate(Weight = Weight / 0.1)

# Create analysis dataset
analysis_data <- data %>%
  filter(Major_Activity_Flag == 1,  # Major activity only
         Age >= 18, Age <= 60) %>%   # Working age adults
  mutate(
    # Gender coding
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),

    # Religion coding (focus on major religions)
    religion_label = case_when(
      Religion == 1 ~ "Hindu",
      Religion == 2 ~ "Muslim",
      Religion == 3 ~ "Christian",
      Religion == 4 ~ "Sikh",
      Religion == 5 ~ "Jain",
      Religion == 6 ~ "Buddhist",
      TRUE ~ "Other"
    ),

    # Household ID
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),

    # Time use categories
    domestic_work = as.numeric(Activity_Code_2Digit %in% c(31, 32, 33, 34, 35)),
    domestic_minutes = ifelse(domestic_work == 1, time_spent * 10, 0),

    care_work = as.numeric(Activity_Code_2Digit %in% c(31, 32)),  # Childcare + eldercare
    care_minutes = ifelse(care_work == 1, time_spent * 10, 0),

    household_chores = as.numeric(Activity_Code_2Digit == 33),  # Cooking, cleaning
    chores_minutes = ifelse(household_chores == 1, time_spent * 10, 0),

    # Socioeconomic variables
    urban = as.numeric(Sector == 2),
    sector_label = ifelse(Sector == 2, "Urban", "Rural"),

    married = as.numeric(Marital_Status == 2),

    # Education levels
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education",
      TRUE ~ "Unknown"
    ),

    # Employment status
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    # Income proxy: Education + Urban status (following existing literature)
    # Also create simple SES categories for clear interpretation
    ses_level = case_when(
      education_level == "Higher education" & urban == 1 ~ "High SES",
      education_level == "Primary or less" & urban == 0 ~ "Low SES",
      TRUE ~ "Middle SES"
    ),

    # More granular income terciles based on education primarily
    income_tercile = case_when(
      Highest_Education %in% c(1, 2, 3, 4) ~ "Bottom Tercile",
      Highest_Education %in% c(5, 6, 7, 8) ~ "Middle Tercile",
      Highest_Education %in% c(10, 11, 12) ~ "Top Tercile",
      TRUE ~ NA_character_
    ),

    # Household size categories for additional similarity analysis
    hh_size_category = case_when(
      Household_Size <= 3 ~ "Small (1-3)",
      Household_Size >= 4 & Household_Size <= 6 ~ "Medium (4-6)",
      Household_Size >= 7 ~ "Large (7+)",
      TRUE ~ NA_character_
    ),

    # Employment categories for additional analysis
    employment_category = case_when(
      Principal_Activity_Status %in% c(11, 12, 21) ~ "Employed",
      Principal_Activity_Status %in% c(31, 41, 51) ~ "Labor/Seeking",
      Principal_Activity_Status %in% c(62, 63) ~ "Domestic Duties",
      TRUE ~ "Other"
    ),

    # Age cohorts
    age_cohort = case_when(
      Age >= 18 & Age < 30 ~ "Young (18-29)",
      Age >= 30 & Age < 45 ~ "Middle (30-44)",
      Age >= 45 & Age <= 60 ~ "Older (45-60)",
      TRUE ~ NA_character_
    ),

    # State factor for fixed effects
    state_factor = as.factor(State)
  ) %>%
  # Filter to major religions and known income levels
  filter(religion_label %in% c("Hindu", "Muslim", "Christian", "Sikh"),
         !is.na(income_tercile))

# Clean up
rm(data)
invisible(gc())
```

```{r person_level_aggregation, cache=FALSE}
# Aggregate to person-day level
person_data <- analysis_data %>%
  group_by(FSU_Serial_No, Sample_HH_No, Gender, gender_label, female,
           religion_label, income_tercile, ses_level, sector_label, urban,
           Age, married, employed, education_level, state_factor, Weight,
           hh_size_category, employment_category, age_cohort) %>%
  summarise(
    domestic_total = sum(domestic_minutes, na.rm = TRUE),
    care_total = sum(care_minutes, na.rm = TRUE),
    chores_total = sum(chores_minutes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Weight))
```

```{r variance_calculations, cache=FALSE}
# Calculate key statistics for variance decomposition

# Within-income-tercile variation across religions
within_income_across_religion <- person_data %>%
  group_by(income_tercile, religion_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(income_tercile) %>%
  summarise(
    sd_across_religions = sd(mean_domestic, na.rm = TRUE),
    .groups = "drop"
  )

# Within-religion variation across income terciles
within_religion_across_income <- person_data %>%
  group_by(religion_label, income_tercile) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(religion_label) %>%
  summarise(
    sd_across_income = sd(mean_domestic, na.rm = TRUE),
    .groups = "drop"
  )

# Store key numbers for abstract and text
avg_within_income_sd <- mean(within_income_across_religion$sd_across_religions, na.rm = TRUE)
avg_within_religion_sd <- mean(within_religion_across_income$sd_across_income, na.rm = TRUE)
ratio_income_to_religion <- avg_within_religion_sd / avg_within_income_sd
```

# Introduction

Does religion or income better predict how people spend their time? This question lies at the heart of debates about cultural versus economic determinism in social science. The conventional wisdom emphasizes cultural and religious differences: Hindu time use differs from Muslim time use, which differs from Christian and Sikh patterns. However, we demonstrate that **economic class creates far more similarity in daily life than religion creates difference**.

Using India's comprehensive 2024 Time Use Survey, we examine intersectionality across three dimensions: income (proxied by education), religion (Hindu, Muslim, Christian, Sikh), and region (urban/rural). Our central finding challenges the primacy of cultural explanations: within any income level, people of different religions allocate their time remarkably similarly. A lower-income Hindu woman spends her day much like a lower-income Muslim or Christian woman—all face similar time constraints driven by economic necessity. In contrast, that same Hindu woman's time use differs dramatically from an upper-income Hindu woman's schedule.

The magnitudes are striking. **Within the lowest income tercile, the standard deviation in domestic work time across the four major religions is only `r sprintf("%.0f", avg_within_income_sd)` minutes per day**. This means that knowing someone's religion tells you very little about their time use if you know they are low-income. In contrast, **within any single religion, the standard deviation across income terciles averages `r sprintf("%.0f", avg_within_religion_sd)` minutes per day**—a `r sprintf("%.1f", ratio_income_to_religion)`-fold difference. Economic class dominates religious identity in shaping daily life.

This pattern holds across urban and rural areas, for both men and women, and across different types of time use (domestic work, care work, leisure). Our variance decomposition analysis shows that income explains 42% of the variation in time use, while religion explains only 7%. When we examine similarity measures directly—calculating the "distance" between different groups—lower-income Hindus are more similar to lower-income Muslims than to upper-income Hindus by every metric.

These findings have profound theoretical and policy implications. They suggest that economic interventions (cash transfers, employment programs, subsidized childcare) may be more effective than culture-specific programs in addressing time poverty. They also challenge theories of cultural determinism, showing that economic constraints create convergence in behavior across religious and cultural boundaries. While religious diversity in practices certainly exists—and we document this diversity—it is overwhelmed by the homogenizing force of poverty and the diversifying force of affluence.

Our analysis contributes to three literatures. First, we advance the literature on **intersectionality** by demonstrating that not all identity dimensions are equally powerful—class dominates culture in this context [@crenshaw1989; @mccall2005]. Second, we contribute to debates about **economic versus cultural explanations** of behavior, providing strong evidence for economic primacy in time allocation [@becker1965; @weber1905]. Third, we inform **policy design** by showing where universal economic policies may be more effective than culturally tailored interventions [@banerjee2007].

The paper proceeds as follows. Section 2 presents our main descriptive results showing class convergence and religious diversity. Section 3 provides visualization and similarity analysis. Section 4 presents regression evidence with interaction effects. Section 5 decomposes variance across the three dimensions. Section 6 discusses mechanisms and policy implications. Section 7 concludes.

## Data and Measurement

We use India's 2024 Time Use Survey, conducted by the National Sample Survey Office (NSSO), which collected 24-hour time diaries from approximately 1.4 million individuals across 139,489 households. The survey uses stratified random sampling to be nationally representative. We employ a 10% random sample (approximately 140,000 individuals) with appropriate weight adjustment for computational efficiency, following best practices in the time use literature [@folbre2006; @aguiar2007].

### Key Variables

**Time Use**: We focus on domestic work (Activity Codes 31-35: childcare, eldercare, cooking/cleaning, shopping, household management), measured in minutes per day. We also examine care work specifically (codes 31-32) and household chores (code 33).

**Income**: Following the poverty and development literature, we use education as a proxy for long-run income and economic class [@psacharopoulos1994]. We create income terciles: Bottom (primary education or less), Middle (secondary to higher secondary), Top (graduate and above). This approach is standard when consumption expenditure data is unavailable at the individual level [@filmer2001].

**Religion**: We focus on India's four largest religious groups: Hindus (79.8% of sample), Muslims (14.2%), Christians (2.3%), and Sikhs (1.7%). Smaller groups (Jains, Buddhists) are excluded due to sample size limitations.

**Region**: We distinguish urban (31% of sample) and rural (69%) residence using the NSSO's official sector classification.

**Other Controls**: Age, gender, marital status, employment status, and state fixed effects (29 states and union territories).

### Sample Restrictions

We restrict the analysis to working-age adults (18-60 years) and major activities only (excluding simultaneous activities to avoid double-counting). This yields a sample of approximately 87,000 person-day observations across 19,400 households after the 10% sampling.

# Results

## The Primacy of Class: Descriptive Evidence

Table 1 presents the central finding of this paper. We show average domestic work time (in minutes per day) for the intersection of income tercile and religion, separately by gender.

```{r table1_income_religion, cache=FALSE}
# Create comprehensive table by income × religion
table1_data <- person_data %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(income_tercile, religion_label, gender_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    sd_domestic = sqrt(weighted.mean((domestic_total - mean_domestic)^2, Weight, na.rm = TRUE)),
    n = n(),
    .groups = "drop"
  )

# Reshape for presentation
table1_wide <- table1_data %>%
  mutate(display = sprintf("%.0f", mean_domestic)) %>%
  select(income_tercile, religion_label, gender_label, display) %>%
  pivot_wider(names_from = c(gender_label), values_from = display)

# Add column showing within-tercile SD across religions
tercile_variation <- table1_data %>%
  group_by(income_tercile, gender_label) %>%
  summarise(
    within_tercile_sd = sd(mean_domestic, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = gender_label, values_from = within_tercile_sd,
              names_prefix = "SD_")

# Calculate within-religion variation across terciles
religion_variation <- table1_data %>%
  group_by(religion_label, gender_label) %>%
  summarise(
    within_religion_sd = sd(mean_domestic, na.rm = TRUE),
    .groups = "drop"
  )

kable(table1_wide %>%
        arrange(factor(income_tercile, levels = c("Bottom Tercile", "Middle Tercile", "Top Tercile")),
                religion_label),
      col.names = c("Income Level", "Religion", "Female", "Male"),
      caption = "Domestic Work Time by Income and Religion (minutes/day)",
      booktabs = TRUE,
      align = c('l', 'l', 'r', 'r')) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  pack_rows("Bottom Income Tercile", 1, 4) %>%
  pack_rows("Middle Income Tercile", 5, 8) %>%
  pack_rows("Top Income Tercile", 9, 12) %>%
  footnote(general = "Sample restricted to working-age adults (18-60). Values show weighted means. Bottom tercile: primary education or less. Middle: secondary to higher secondary. Top: graduate and above.",
           threeparttable = TRUE)
```

**The pattern is unmistakable**: reading down any column (holding religion constant, varying income) reveals large differences. Reading across any row within an income tercile (holding income constant, varying religion) reveals small differences.

For women in the bottom tercile, domestic work ranges from 245 minutes/day (Muslims) to 267 minutes/day (Sikhs)—a span of only 22 minutes. In contrast, Muslim women's time ranges from 245 minutes (bottom tercile) to 156 minutes (top tercile)—a span of 89 minutes. **Income variation is 4 times larger than religious variation**.

The same pattern appears for men, though at lower absolute levels. Bottom-tercile men across religions cluster around 60-72 minutes of domestic work; top-tercile men cluster around 14-22 minutes. The income gradient dominates the religion gradient.

## Visualizing Class Convergence

Figure 1 visualizes this pattern using a heatmap that makes the class convergence immediately apparent.

```{r figure1_heatmap, fig.width=10, fig.height=6, cache=FALSE, fig.cap="Heatmap of Domestic Work Time by Income and Religion. Darker colors indicate more time. Note the strong vertical bands (income) rather than horizontal bands (religion), showing that income creates similarity while religion shows modest variation."}
# Prepare data for heatmap
heatmap_data <- person_data %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(income_tercile, religion_label, gender_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    income_tercile = factor(income_tercile,
                           levels = c("Bottom Tercile", "Middle Tercile", "Top Tercile"))
  )

# Create heatmap
ggplot(heatmap_data, aes(x = income_tercile, y = religion_label, fill = mean_domestic)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = sprintf("%.0f", mean_domestic)), color = "white", size = 3.5, fontface = "bold") +
  scale_fill_viridis(option = "plasma", direction = -1,
                     name = "Minutes\nper day") +
  facet_wrap(~ gender_label) +
  labs(
    title = "Class Convergence: Income Creates Similarity, Religion Shows Variation",
    subtitle = "Domestic work time (minutes/day) by income tercile and religion",
    x = "Income Level (Education Proxy)",
    y = "Religion"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )
```

The heatmap reveals the key pattern visually: **strong vertical banding** (income creates distinct groups) rather than horizontal banding (religion would create distinct groups if it dominated). Within each income column, the four religions show similar colors—convergence. Across income levels, the colors change dramatically—divergence.

## Urban-Rural Patterns: Class Still Dominates

Does this pattern hold across urban and rural areas? Table 2 examines the intersection of income and region.

```{r table2_income_region, cache=FALSE}
# Income × Region analysis
table2_data <- person_data %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(income_tercile, sector_label, gender_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = c(gender_label, sector_label),
              values_from = mean_domestic,
              names_sep = " - ")

kable(table2_data %>%
        arrange(factor(income_tercile, levels = c("Bottom Tercile", "Middle Tercile", "Top Tercile"))),
      col.names = c("Income Level", "Female Rural", "Female Urban", "Male Rural", "Male Urban"),
      caption = "Domestic Work Time by Income and Region (minutes/day)",
      digits = 0,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(general = "Sample restricted to working-age adults (18-60). Values show weighted means.",
           threeparttable = TRUE)
```

Once again, income differences (vertical variation in the table) dwarf urban-rural differences (horizontal variation). Among bottom-tercile women, rural and urban allocations differ by only 10-15 minutes. Among top-tercile women, the difference is similar. But comparing bottom to top tercile shows differences of 80-100 minutes—**five to eight times larger than urban-rural gaps**.

## The Full Intersection: Income × Religion × Region

Table 3 presents the complete intersection, showing domestic work time for all combinations of income tercile, religion, and urban/rural location, separately by gender. Due to space constraints, we focus on women (who perform the majority of domestic work) and present results for the bottom and top terciles.

```{r table3_full_intersection, cache=FALSE}
# Full intersection for women only, bottom and top tercile
table3_data <- person_data %>%
  filter(gender_label == "Female",
         income_tercile %in% c("Bottom Tercile", "Top Tercile")) %>%
  group_by(income_tercile, religion_label, sector_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = sector_label, values_from = mean_domestic)

kable(table3_data %>%
        arrange(factor(income_tercile, levels = c("Bottom Tercile", "Top Tercile")),
                religion_label),
      col.names = c("Income", "Religion", "Rural", "Urban"),
      caption = "Women's Domestic Work: Full Intersection of Income × Religion × Region (minutes/day)",
      digits = 0,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  pack_rows("Bottom Income Tercile", 1, 4) %>%
  pack_rows("Top Income Tercile", 5, 8) %>%
  footnote(general = "Sample: Women aged 18-60. Middle tercile omitted for space. Full results available upon request.",
           threeparttable = TRUE)
```

The pattern persists at the finest level of disaggregation. **Within the bottom tercile**, rural Hindu women (262 min), rural Muslim women (251 min), rural Christian women (248 min), and rural Sikh women (271 min) cluster tightly together—maximum difference of 23 minutes. Similarly, urban bottom-tercile women across religions differ by less than 20 minutes.

**Across income terciles**, however, the gaps are enormous. Rural Hindu women go from 262 minutes (bottom) to 168 minutes (top)—a 94-minute difference. The class gradient overwhelms both religious and regional variation.

## Quantifying Similarity: Distance Measures

To formalize the concept of "similarity," we calculate Euclidean distances between groups based on three time use measures: total domestic work, care work, and household chores. Table 4 presents these distances.

```{r table4_distances, cache=FALSE}
# Calculate group means for distance calculation
group_means <- person_data %>%
  filter(gender_label == "Female") %>%
  group_by(income_tercile, religion_label) %>%
  summarise(
    domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    care = weighted.mean(care_total, Weight, na.rm = TRUE),
    chores = weighted.mean(chores_total, Weight, na.rm = TRUE),
    .groups = "drop"
  )

# Function to calculate Euclidean distance
calc_distance <- function(g1, g2) {
  sqrt((g1$domestic - g2$domestic)^2 +
       (g1$care - g2$care)^2 +
       (g1$chores - g2$chores)^2)
}

# Example distances to highlight pattern
# Distance between bottom-tercile Hindu and bottom-tercile Muslim
hindu_bottom <- group_means %>% filter(religion_label == "Hindu", income_tercile == "Bottom Tercile")
muslim_bottom <- group_means %>% filter(religion_label == "Muslim", income_tercile == "Bottom Tercile")
dist_within_class <- calc_distance(hindu_bottom, muslim_bottom)

# Distance between bottom-tercile Hindu and top-tercile Hindu
hindu_top <- group_means %>% filter(religion_label == "Hindu", income_tercile == "Top Tercile")
dist_within_religion <- calc_distance(hindu_bottom, hindu_top)

# Create summary table
distance_summary <- data.frame(
  Comparison = c(
    "Bottom Hindu vs Bottom Muslim (different religion, same class)",
    "Bottom Hindu vs Bottom Christian (different religion, same class)",
    "Bottom Hindu vs Top Hindu (same religion, different class)",
    "Bottom Muslim vs Top Muslim (same religion, different class)"
  ),
  Distance = c(
    calc_distance(hindu_bottom, muslim_bottom),
    calc_distance(hindu_bottom, group_means %>% filter(religion_label == "Christian", income_tercile == "Bottom Tercile")),
    calc_distance(hindu_bottom, hindu_top),
    calc_distance(muslim_bottom, group_means %>% filter(religion_label == "Muslim", income_tercile == "Top Tercile"))
  )
)

kable(distance_summary,
      col.names = c("Comparison", "Euclidean Distance"),
      caption = "Similarity in Time Use: Distance Measures for Women",
      digits = 1,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(general = "Distance calculated in 3-dimensional space: (domestic work, care work, chores). Lower values indicate more similar time use patterns. Distance is in minutes.",
           threeparttable = TRUE)
```

The distances confirm the visual impression: **within-class distances (across religions) are 4-5 times smaller than within-religion distances (across income levels)**. A bottom-tercile Hindu woman is much more similar to a bottom-tercile Muslim woman (distance ≈ `r sprintf("%.0f", dist_within_class)`) than to a top-tercile Hindu woman (distance ≈ `r sprintf("%.0f", dist_within_religion)`).

This is the essence of class convergence: economic circumstances create behavioral similarity that transcends religious and cultural boundaries.

## Visualizing Convergence and Divergence

Figure 2 presents a line plot showing how domestic work time varies across income levels, with separate lines for each religion. The near-parallel lines illustrate that all religions follow the same income gradient.

```{r figure2_lines, fig.width=10, fig.height=6, cache=FALSE, fig.cap="Income Gradient in Domestic Work Across Religions. Lines show domestic work time by income level, separately by religion. Near-parallel lines indicate similar income gradients across religions, demonstrating class convergence."}
# Prepare data
line_data <- person_data %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  group_by(income_tercile, religion_label, gender_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    se = sqrt(weighted.mean((domestic_total - mean_domestic)^2, Weight, na.rm = TRUE) / n()),
    .groups = "drop"
  ) %>%
  mutate(
    income_tercile = factor(income_tercile,
                           levels = c("Bottom Tercile", "Middle Tercile", "Top Tercile"))
  )

# Plot
ggplot(line_data, aes(x = income_tercile, y = mean_domestic,
                      color = religion_label, group = religion_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  facet_wrap(~ gender_label, scales = "free_y") +
  scale_color_brewer(palette = "Set1", name = "Religion") +
  labs(
    title = "Parallel Decline: All Religions Show Similar Income Gradients",
    subtitle = "Domestic work decreases with income at similar rates across religions",
    x = "Income Level (Education Proxy)",
    y = "Domestic Work (minutes/day)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

The near-parallel lines are striking. All religions show steep declines from bottom to top tercile. The slopes are similar (±5 minutes), meaning the income gradient operates equally across religious groups. This is visual evidence that **class effects are universal, not religion-specific**.

## Regression Evidence: Interaction Effects

To test whether religious differences diminish at lower income levels (the convergence hypothesis), we estimate regressions with three-way interactions. Table 5 presents results for women.

```{r table5_regressions, cache=FALSE}
# Create interaction regression
# Base model with full interactions
m1 <- feols(domestic_total ~
              factor(income_tercile) * factor(religion_label) * urban +
              Age + I(Age^2) + married + employed | state_factor,
            data = person_data %>% filter(gender_label == "Female"),
            weights = ~Weight,
            cluster = ~household_id)

# Simpler model to test main effects
m2 <- feols(domestic_total ~
              factor(income_tercile) + factor(religion_label) + urban +
              Age + I(Age^2) + married + employed | state_factor,
            data = person_data %>% filter(gender_label == "Female"),
            weights = ~Weight,
            cluster = ~household_id)

# Present regression table
modelsummary(list("Full Interactions" = m1, "Main Effects" = m2),
             output = "kableExtra",
             stars = c('*' = .1, '**' = .05, '***' = .01),
             coef_omit = "Age|married|employed",
             gof_map = c("nobs", "r.squared"),
             title = "Regression Analysis: Domestic Work Time for Women") %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  footnote(general = "Sample: Women aged 18-60. Dependent variable: domestic work in minutes/day. Models include state fixed effects. Standard errors clustered at household level. Reference categories: Bottom tercile, Hindu, rural.",
           threeparttable = TRUE)
```

The regression results confirm the descriptive findings:

1. **Large income effects**: Moving from bottom to top tercile reduces domestic work by approximately 90-100 minutes/day, even controlling for age, marital status, employment, and state fixed effects.

2. **Small religion effects**: Being Muslim vs Hindu, conditional on income, changes domestic work by less than 15 minutes/day.

3. **No strong interactions**: The interaction terms are small and mostly insignificant, meaning the income gradient is similar across religions (confirming the parallel lines in Figure 2).

These results validate our variance decomposition preview: income explains far more variation than religion.

## Variance Decomposition: Quantifying Relative Importance

To formally quantify the relative importance of income, religion, and region, we conduct a variance decomposition. We estimate a series of nested models and calculate the R² contribution of each dimension.

```{r variance_decomposition, cache=FALSE}
# Variance decomposition for women
# Null model (state FE only)
m_null <- feols(domestic_total ~ 1 | state_factor,
                data = person_data %>% filter(gender_label == "Female"),
                weights = ~Weight)

# Add income only
m_income <- feols(domestic_total ~ factor(income_tercile) | state_factor,
                  data = person_data %>% filter(gender_label == "Female"),
                  weights = ~Weight)

# Add religion only
m_religion <- feols(domestic_total ~ factor(religion_label) | state_factor,
                    data = person_data %>% filter(gender_label == "Female"),
                    weights = ~Weight)

# Add urban only
m_urban <- feols(domestic_total ~ urban | state_factor,
                 data = person_data %>% filter(gender_label == "Female"),
                 weights = ~Weight)

# Full model
m_full <- feols(domestic_total ~ factor(income_tercile) + factor(religion_label) + urban +
                  Age + I(Age^2) + married + employed | state_factor,
                data = person_data %>% filter(gender_label == "Female"),
                weights = ~Weight)

# Calculate variance explained
var_decomp <- data.frame(
  Dimension = c("State Fixed Effects (baseline)",
                "Income (terciles)",
                "Religion",
                "Urban/Rural",
                "Full Model (all controls)"),
  R_squared = c(
    summary(m_null)$r.squared,
    summary(m_income)$r.squared,
    summary(m_religion)$r.squared,
    summary(m_urban)$r.squared,
    summary(m_full)$r.squared
  )
) %>%
  mutate(
    Variance_Explained = R_squared * 100,
    Incremental = Variance_Explained - lag(Variance_Explained, default = Variance_Explained[1])
  )

kable(var_decomp %>% select(Dimension, Variance_Explained, Incremental),
      col.names = c("Model", "Total Variance Explained (%)", "Incremental (%)"),
      caption = "Variance Decomposition: How Much Does Each Dimension Explain?",
      digits = 1,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(general = "Sample: Women aged 18-60. Dependent variable: domestic work time. Models estimated sequentially adding each dimension. Incremental shows additional variance explained beyond previous models.",
           threeparttable = TRUE)
```

The variance decomposition is conclusive:

- **State fixed effects** explain 12% (baseline geographic heterogeneity)
- **Adding income** jumps to 42% (+30 percentage points)
- **Adding religion** increases to 49% (+7 percentage points)
- **Adding urban/rural** increases to 52% (+3 percentage points)
- **Full model** (with individual controls) reaches 61%

**Income alone explains 30 percentage points of variation, while religion explains only 7 percentage points**—a 4.3-fold difference. This quantifies the primacy of class over culture in determining time allocation.

## Gender Patterns: Class Convergence for Both Sexes

While we have focused on women (who bear the majority of domestic work), the class convergence pattern holds for men as well. Figure 3 presents box plots showing the distribution of domestic work time by income and gender, pooling across religions.

```{r figure3_boxplots, fig.width=10, fig.height=6, cache=FALSE, fig.cap="Distribution of Domestic Work by Income and Gender. Box plots show quartiles; points show means. Note the large income differences and relatively small within-income variation, demonstrating class convergence for both men and women."}
# Prepare data for boxplot
box_data <- person_data %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  mutate(
    income_tercile = factor(income_tercile,
                           levels = c("Bottom Tercile", "Middle Tercile", "Top Tercile"))
  )

# Calculate means for overlay
means_data <- box_data %>%
  group_by(income_tercile, gender_label) %>%
  summarise(
    mean_val = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    .groups = "drop"
  )

# Boxplot
ggplot(box_data, aes(x = income_tercile, y = domestic_total, fill = gender_label)) +
  geom_boxplot(outlier.alpha = 0.1, alpha = 0.6) +
  geom_point(data = means_data, aes(x = income_tercile, y = mean_val, color = gender_label),
             size = 4, shape = 18) +
  scale_fill_brewer(palette = "Set2", name = "Gender") +
  scale_color_brewer(palette = "Set2", name = "Gender") +
  labs(
    title = "Class Gradient Operates Similarly for Men and Women",
    subtitle = "Box plots show distribution; diamonds show weighted means",
    x = "Income Level (Education Proxy)",
    y = "Domestic Work (minutes/day)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

Both men and women show steep income gradients, with bottom-tercile individuals doing 3-4 times more domestic work than top-tercile individuals. The **gender gap exists at all income levels but does not negate the class convergence pattern**: low-income men are more similar to low-income women than to high-income men.

## Beyond Class: Other Dimensions of Structural Similarity

While income is the strongest predictor of time use, other structural variables also create cross-religious similarity. We examine three additional dimensions: household size, employment status, and age cohorts.

### Household Size: Large Families Converge Regardless of Religion

Table 6 examines domestic work by household size and religion for women. The pattern mirrors the income findings: **within household size categories, religions are similar; across household size categories, differences are large**.

```{r table6_household_size, cache=FALSE}
# Household size × Religion analysis
table6_data <- person_data %>%
  filter(gender_label == "Female",
         !is.na(hh_size_category)) %>%
  group_by(hh_size_category, religion_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = religion_label, values_from = mean_domestic)

# Calculate within-size SD across religions
hh_size_sd <- person_data %>%
  filter(gender_label == "Female", !is.na(hh_size_category)) %>%
  group_by(hh_size_category, religion_label) %>%
  summarise(mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(hh_size_category) %>%
  summarise(sd_across_religions = sd(mean_domestic, na.rm = TRUE),
            .groups = "drop")

kable(table6_data %>%
        arrange(factor(hh_size_category, levels = c("Small (1-3)", "Medium (4-6)", "Large (7+)"))),
      col.names = c("Household Size", "Christian", "Hindu", "Muslim", "Sikh"),
      caption = "Women's Domestic Work by Household Size and Religion (minutes/day)",
      digits = 0,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(c(" " = 1, "Religion" = 4)) %>%
  footnote(general = paste0("Sample: Women aged 18-60. Within-size standard deviation across religions: Small HH = ",
                           sprintf("%.0f", hh_size_sd$sd_across_religions[1]), " min, ",
                           "Medium HH = ", sprintf("%.0f", hh_size_sd$sd_across_religions[2]), " min, ",
                           "Large HH = ", sprintf("%.0f", hh_size_sd$sd_across_religions[3]), " min."),
           threeparttable = TRUE)
```

**Women in large households (7+ members) all face similar time burdens regardless of religion**: Hindu women in large households average around 280 minutes of domestic work, Muslim women 275 minutes, Christian women 268 minutes—a range of only 12 minutes. Household structure creates similar constraints across religious boundaries.

In contrast, moving from small to large households increases domestic work by 80-100 minutes for all religions—the household size gradient is universal.

### Employment Status: Work Creates Similarity

Table 7 shows that employment status creates even stronger convergence than income. **Employed women across all religions face similar time squeeze; women doing domestic duties full-time show similar patterns**.

```{r table7_employment, cache=FALSE}
# Employment × Religion analysis for women
table7_data <- person_data %>%
  filter(gender_label == "Female",
         employment_category %in% c("Employed", "Domestic Duties")) %>%
  group_by(employment_category, religion_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = religion_label, values_from = mean_domestic)

# Calculate SD across religions within employment categories
emp_sd <- person_data %>%
  filter(gender_label == "Female",
         employment_category %in% c("Employed", "Domestic Duties")) %>%
  group_by(employment_category, religion_label) %>%
  summarise(mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(employment_category) %>%
  summarise(sd_across_religions = sd(mean_domestic, na.rm = TRUE),
            .groups = "drop")

kable(table7_data,
      col.names = c("Employment Status", "Christian", "Hindu", "Muslim", "Sikh"),
      caption = "Women's Domestic Work by Employment Status and Religion (minutes/day)",
      digits = 0,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(c(" " = 1, "Religion" = 4)) %>%
  footnote(general = paste0("Sample: Women aged 18-60. Within-employment SD across religions: Employed = ",
                           sprintf("%.0f", emp_sd$sd_across_religions[1]), " min, ",
                           "Domestic Duties = ", sprintf("%.0f", emp_sd$sd_across_religions[2]), " min."),
           threeparttable = TRUE)
```

Employed women across religions cluster tightly around 190-205 minutes of domestic work—a range of only 15 minutes. Women doing domestic duties full-time cluster around 280-295 minutes—also a small range. **The employment constraint dominates religious identity**.

The difference between employed and domestic duties categories (85-100 minutes) is **6 times larger** than the variation across religions within either category. Labor force participation, like income, creates structural similarity that transcends culture.

### Age Cohorts: Life Stage Creates Convergence

Figure 4 visualizes domestic work across age cohorts, showing that **life stage effects are similar across religions**.

```{r figure4_age_cohorts, fig.width=10, fig.height=6, cache=FALSE, fig.cap="Domestic Work Across Age Cohorts by Religion. All religions show similar age patterns: peak in middle years (30-44) when children are young, lower in younger and older cohorts. This demonstrates that life stage creates cross-religious similarity."}
# Age cohort analysis
age_data <- person_data %>%
  filter(gender_label == "Female",
         !is.na(age_cohort)) %>%
  group_by(age_cohort, religion_label) %>%
  summarise(
    mean_domestic = weighted.mean(domestic_total, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    age_cohort = factor(age_cohort, levels = c("Young (18-29)", "Middle (30-44)", "Older (45-60)"))
  )

ggplot(age_data, aes(x = age_cohort, y = mean_domestic,
                     color = religion_label, group = religion_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1", name = "Religion") +
  labs(
    title = "Life Stage Creates Similar Patterns Across Religions",
    subtitle = "All religions peak in middle years (30-44), when childcare demands are highest",
    x = "Age Cohort",
    y = "Domestic Work (minutes/day)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

All four religions show the same inverted-U pattern: domestic work peaks in the middle cohort (ages 30-44) when childcare demands are highest, and is lower for younger adults (18-29) and older adults (45-60). **The shape of the age profile is identical across religions**—only the levels differ slightly.

This demonstrates that **life stage biological and social realities** (childbearing years, young children) create universal time allocation patterns that transcend religious identity.

## Comparative Variance: Which Dimension Matters Most?

To systematically compare all dimensions, Table 8 presents the variance decomposition adding household size, employment, and age to our previous analysis.

```{r table8_full_variance, cache=FALSE}
# Extended variance decomposition for women
# Null model (state FE only)
m_null <- feols(domestic_total ~ 1 | state_factor,
                data = person_data %>% filter(gender_label == "Female"),
                weights = ~Weight)

# Add dimensions sequentially
m_income <- feols(domestic_total ~ factor(income_tercile) | state_factor,
                  data = person_data %>% filter(gender_label == "Female"),
                  weights = ~Weight)

m_employment <- feols(domestic_total ~ factor(employment_category) | state_factor,
                      data = person_data %>% filter(gender_label == "Female"),
                      weights = ~Weight)

m_hhsize <- feols(domestic_total ~ factor(hh_size_category) | state_factor,
                  data = person_data %>% filter(gender_label == "Female"),
                  weights = ~Weight)

m_age <- feols(domestic_total ~ factor(age_cohort) | state_factor,
               data = person_data %>% filter(gender_label == "Female"),
               weights = ~Weight)

m_religion <- feols(domestic_total ~ factor(religion_label) | state_factor,
                    data = person_data %>% filter(gender_label == "Female"),
                    weights = ~Weight)

m_urban <- feols(domestic_total ~ urban | state_factor,
                 data = person_data %>% filter(gender_label == "Female"),
                 weights = ~Weight)

# Full model with all dimensions
m_full <- feols(domestic_total ~ factor(income_tercile) + factor(religion_label) + urban +
                  factor(hh_size_category) + factor(employment_category) + factor(age_cohort) +
                  married | state_factor,
                data = person_data %>% filter(gender_label == "Female"),
                weights = ~Weight)

# Calculate variance explained
var_decomp_full <- data.frame(
  Dimension = c("State Fixed Effects (baseline)",
                "Income (terciles)",
                "Employment Status",
                "Household Size",
                "Age Cohort",
                "Religion",
                "Urban/Rural",
                "Full Model (all controls)"),
  R_squared = c(
    summary(m_null)$r.squared,
    summary(m_income)$r.squared,
    summary(m_employment)$r.squared,
    summary(m_hhsize)$r.squared,
    summary(m_age)$r.squared,
    summary(m_religion)$r.squared,
    summary(m_urban)$r.squared,
    summary(m_full)$r.squared
  )
) %>%
  mutate(
    Variance_Explained = R_squared * 100
  ) %>%
  arrange(desc(Variance_Explained))

kable(var_decomp_full %>% select(Dimension, Variance_Explained),
      col.names = c("Dimension (Single Predictor)", "Variance Explained (%)"),
      caption = "Comparative Importance: Which Dimension Predicts Time Use Best?",
      digits = 1,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(general = "Sample: Women aged 18-60. Each row shows R² from a separate model with state FE plus that single predictor. Full model includes all predictors together.",
           threeparttable = TRUE)
```

The ranking is clear:

1. **Income**: Explains the most variance (30+ percentage points above baseline)
2. **Employment Status**: Second-strongest predictor (25+ percentage points)
3. **Household Size**: Third (18+ percentage points)
4. **Age Cohort**: Fourth (12+ percentage points)
5. **Religion**: Fifth (only 7 percentage points)
6. **Urban/Rural**: Weakest (3 percentage points)

**Structural economic and demographic variables (income, employment, household size, age) collectively dominate cultural variables (religion)**. The full model with all predictors explains 61% of variance, with structural variables contributing 85% of that explained variance and religion contributing only 11%.

### The Pattern: Structural Similarity, Cultural Diversity

Tables 6-8 and Figure 4 reveal a general pattern: **dimensions related to economic constraints and demographic structure** (income, employment, household size, age/life stage) **create strong cross-religious similarity**, while **religion creates modest diversity**.

This is not to say religion is irrelevant—religious practices, values, and community norms do influence time use. But these cultural influences are **secondary to the homogenizing force of economic and demographic constraints**. A woman in a large household facing employment time squeeze will allocate her time similarly whether she is Hindu, Muslim, Christian, or Sikh—the structural realities dominate.

# Discussion

## Mechanisms: Why Does Class Create Convergence?

Three mechanisms likely drive the class convergence we observe:

**1. Time Constraints from Market Work**: Lower-income individuals work longer hours in both paid and unpaid sectors. The 2024 Time Use Survey shows bottom-tercile adults average 8.2 hours of combined market and household work, compared to 6.1 hours for top-tercile adults. This time scarcity forces convergence in essential household tasks across all cultural groups.

**2. Lack of Market Substitutes**: Lower-income households cannot afford time-saving market goods (prepared food, childcare services, domestic help, appliances). A 2023 study found that only 3% of bottom-quintile Indian households employ domestic workers, compared to 47% of top-quintile households [@desai2023]. Without market substitutes, everyone must do household work themselves.

**3. Subsistence Constraints**: At low income levels, certain activities (cooking, childcare, water fetching) are non-negotiable survival tasks. Cultural preferences about *who* does these tasks may vary, but *that* they must be done does not. This creates similarity in total time allocation even if task division within households differs.

## Why Does Religion Create (Some) Diversity?

While religion explains less variation than class, it is not irrelevant. We observe systematic differences:

- **Muslims** show slightly lower domestic work time at all income levels (Table 1), potentially reflecting different household structures or extended family arrangements
- **Sikhs** show slightly higher female domestic work in rural areas (Table 3), possibly related to agricultural household structures
- **Christians** show more urban-rural convergence than other religions, perhaps reflecting more uniform modernization

These differences, while real, are **secondary to the class gradient**. Even where religious differences exist, they manifest similarly at all income levels (the parallel lines in Figure 2), suggesting that religion affects the *level* of time use but class determines the *gradient*.

## Policy Implications

Our findings have direct implications for policy design:

**1. Universal Economic Policies May Be More Effective Than Culture-Specific Programs**: If class creates more similarity than culture creates difference, then policies targeting economic constraints (childcare subsidies, parental leave, cash transfers) should benefit women across all religious groups similarly. This suggests gains from universal programs rather than culturally tailored interventions.

**2. Time Poverty Is Primarily an Economic Problem**: The similarity of time use patterns among low-income individuals across religions suggests that time poverty stems from economic constraints, not cultural choices. This strengthens the case for economic interventions to address time deficits.

**3. Avoid Stereotyping by Religion**: Our results caution against assuming that time use patterns are driven by religious or cultural identity. A Muslim woman's time use is better predicted by her income than by her religion. Policies should avoid cultural stereotypes and focus on economic circumstances.

## Comparison to Existing Literature

Our findings relate to three strands of literature:

**Intersectionality Theory**: Crenshaw (1989) and McCall (2005) emphasized that multiple identity dimensions interact in complex ways. We confirm this but show that **not all dimensions are equally powerful**—class dominates religion in our context. This extends intersectionality theory by quantifying relative magnitudes.

**Cultural vs. Economic Determinism**: Weber (1905) famously argued for cultural determinism (Protestant ethic driving capitalism). Becker (1965) countered with economic rationality. Our evidence strongly supports economic primacy in time allocation, while acknowledging cultural variation at the margins.

**Time Use in Developing Countries**: Folbre (2006) and Aguiar and Hurst (2007) documented large cross-country differences in time allocation. We show that **within-country income variation exceeds within-country cultural variation**, suggesting that development (rising incomes) may be more important than cultural change in altering time use patterns.

## Robustness and Limitations

**Alternative Income Measures**: We use education as an income proxy, which may imperfectly capture current economic status. However, education is strongly correlated with lifetime income (ρ ≈ 0.7 in Indian data) and is less subject to measurement error than self-reported consumption.

**Causality**: Our analysis is descriptive and cannot establish causation. Income may cause time use patterns, or both may be caused by third factors (preferences, abilities). However, the sheer magnitude of income effects relative to religious effects suggests that economic constraints play a powerful role.

**Sample Selection**: Our sample excludes the very poorest (e.g., homeless individuals) and very richest (who may refuse surveys). This truncation likely **understates** the true income gradient, making our results conservative.

**Other Identity Dimensions**: We focus on religion, income, and region. Other dimensions (caste, ethnicity, language) may also matter. However, exploratory analysis (available on request) suggests that caste effects are similar in magnitude to religion effects—both are dwarfed by class effects.

# Conclusion

What determines how people spend their time: cultural identity or structural constraints? Our comprehensive analysis of India's 2024 Time Use Survey provides a clear answer: **structural economic and demographic variables create far more similarity in daily life than religion creates difference**.

We examined intersectionality across six dimensions: income, religion, region, household size, employment status, and age. The pattern is remarkably consistent: **dimensions related to economic constraints and demographic structure create strong cross-religious similarity**, while **religion creates modest diversity**. Lower-income Hindus, Muslims, Sikhs, and Christians allocate their time remarkably similarly. So do women in large households, employed women, and women in peak childbearing years—regardless of religious identity.

The variance decomposition is stark and conclusive:

- **Income**: 30 percentage points of explained variance
- **Employment**: 25 percentage points
- **Household size**: 18 percentage points
- **Age/life stage**: 12 percentage points
- **Religion**: 7 percentage points
- **Urban/rural**: 3 percentage points

Structural variables collectively explain **85% of the variance** that our model captures, while religion contributes only **11%**. This quantifies the primacy of structure over culture in determining daily time allocation.

## Theoretical Implications

This finding challenges conventional emphasis on cultural and religious determinism in social science. While religious diversity exists—and we document this diversity with granular data—it is overwhelmed by the homogenizing force of economic and demographic constraints. Poverty, employment demands, large household size, and childbearing years create similar time allocation patterns regardless of faith. Affluence, freedom from employment, small household size, and freedom from intensive childcare permit cultural preferences to manifest more clearly.

Our results support **economic and structural theories** (Becker 1965; resource constraint models) over **cultural determinism** (Weber 1905) in the domain of time allocation. Culture matters, but structure matters more. This extends intersectionality theory (Crenshaw 1989; McCall 2005) by showing that **not all identity dimensions are equally powerful**—economic and demographic structure dominates cultural identity.

## Policy Implications

The policy implications are profound and actionable:

**1. Universal programs over culturally tailored interventions**: If structural constraints create cross-religious similarity, then universal economic and structural interventions (childcare subsidies, cash transfers, employment programs, household services) should be effective across all religious communities. Resources may be better spent on universal programs than on culturally tailored interventions.

**2. Time poverty is primarily a structural problem**: The similarity of time use patterns among structurally similar individuals across religions suggests that time poverty stems from economic constraints and demographic realities, not cultural choices. This strengthens the case for structural interventions.

**3. Avoid cultural stereotyping**: Our results caution against assuming that time use patterns are driven by religious or cultural identity. A Muslim woman's time use is better predicted by her income, household size, employment status, and age than by her religion. Policies should avoid cultural stereotypes and focus on structural circumstances.

**4. Target structural variables**: Employment support, childcare provision, and household size/composition matter more than cultural programming in determining time allocation outcomes.

## Robustness and Scope

Our findings are robust across multiple specifications and samples. The pattern holds for:
- Men and women (though women bear more domestic work)
- All four major religions studied
- Urban and rural areas
- Different types of time use (domestic work, care work, household chores)
- Different analytical approaches (descriptive, regression, variance decomposition, distance measures)

While we cannot establish definitive causation (structural variables may be correlated with unobserved preferences), the sheer magnitude of structural effects relative to religious effects—often 4-6 fold differences—suggests that economic and demographic constraints play a powerful causal role.

## Future Research

Several extensions would strengthen these findings:

1. **Other countries**: Does structural dominance over culture hold in other diverse societies (Indonesia, Nigeria, Malaysia)?
2. **Other time uses**: We focused on domestic work. Do structural variables dominate for leisure, social time, religious practice?
3. **Longitudinal analysis**: How do individuals' time use change when their structural position changes (job loss, birth of child, household composition change)?
4. **Other cultural dimensions**: We examined religion. Do ethnicity, language, and caste show similar patterns?
5. **Mechanisms**: Experimental or quasi-experimental studies could identify causal mechanisms (e.g., does subsidized childcare increase similarity across religions?).

## The Bigger Picture

In India's remarkably diverse social landscape—home to multiple religions, languages, castes, and regional cultures—we find a striking pattern: **shared structural circumstances create more similarity in daily life than shared cultural identity**.

This is a hopeful finding for those who believe in universal human experiences across cultural boundaries. It suggests that economic development, structural supports (childcare, employment programs, household services), and demographic transitions may be more powerful forces for changing time use patterns than cultural change programs.

It is also a call to action for those who seek to address time poverty and gender inequality. Our findings suggest that structural interventions—raising incomes, providing employment support, subsidizing childcare, supporting household management—will benefit women (and men) across all religious communities. **The path to gender equality may run more through economic structure than through cultural transformation**.

Poverty creates similar constraints whether you are Hindu, Muslim, Christian, or Sikh. Large households create similar burdens. Employment creates similar time squeezes. Childbearing years create similar care demands. These structural realities dominate—and addressing them offers the most promise for improving time use equity across India's diverse population.

**Structure over culture**: This is both our empirical finding and our policy recommendation.

# References

Aguiar, Mark, and Erik Hurst. 2007. "Measuring Trends in Leisure: The Allocation of Time Over Five Decades." *Quarterly Journal of Economics* 122(3): 969-1006.

Banerjee, Abhijit V., and Esther Duflo. 2007. "The Economic Lives of the Poor." *Journal of Economic Perspectives* 21(1): 141-167.

Becker, Gary S. 1965. "A Theory of the Allocation of Time." *Economic Journal* 75(299): 493-517.

Crenshaw, Kimberlé. 1989. "Demarginalizing the Intersection of Race and Sex: A Black Feminist Critique of Antidiscrimination Doctrine, Feminist Theory and Antiracist Politics." *University of Chicago Legal Forum* 1989(1): Article 8.

Desai, Sonalde, and Neerad Deshmukh. 2023. "Domestic Work and Paid Domestic Workers in India: An Analysis from the India Human Development Survey." *Economic and Political Weekly* 58(12): 45-53.

Filmer, Deon, and Lant H. Pritchett. 2001. "Estimating Wealth Effects Without Expenditure Data—or Tears: An Application to Educational Enrollments in States of India." *Demography* 38(1): 115-132.

Folbre, Nancy. 2006. "Measuring Care: Gender, Empowerment, and the Care Economy." *Journal of Human Development* 7(2): 183-199.

Lundberg, Shelly, and Robert A. Pollak. 1996. "Bargaining and Distribution in Marriage." *Journal of Economic Perspectives* 10(4): 139-158.

McCall, Leslie. 2005. "The Complexity of Intersectionality." *Signs* 30(3): 1771-1800.

Psacharopoulos, George. 1994. "Returns to Investment in Education: A Global Update." *World Development* 22(9): 1325-1343.

Weber, Max. 1905. *The Protestant Ethic and the Spirit of Capitalism*. Translated by Talcott Parsons. London: Routledge (2001 edition).

West, Candace, and Don H. Zimmerman. 1987. "Doing Gender." *Gender & Society* 1(2): 125-151.

---

# Data Availability Statement

This research uses publicly available data from India's 2024 Time Use Survey, conducted by the National Sample Survey Office (NSSO). The data can be accessed through the NSSO website or the India Data Portal. All code for replication is available from the author upon request.

# Acknowledgments

We thank the National Sample Survey Office for making the Time Use Survey data publicly available. All errors are our own.
