---
title: "The Marriage Care Penalty: How Marriage Disproportionately Increases Women's Unpaid Work"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using a 10% random sample of India's 2024 Time Use Survey (approximately 836,000 person-day observations from 9,969 households), we document a stark marriage penalty in unpaid care work. While marriage is associated with a small, statistically significant increase in men's participation in unpaid care, it prompts a dramatically larger increase for women. This creates a substantial "marriage penalty" for women, which is significantly larger for younger women (≤25) than older women (>25), and persists across all education levels, urban-rural locations, and employment status. We find no gender difference in time intensity conditional on participation, indicating the disparity operates entirely on the participation margin. These findings reveal marriage as a critical institution activating gender-specific norms about domestic responsibilities, with important implications for policies targeting gender equality in household production.

  **JEL Classification:** J12, J16, J22, D13

  **Keywords:** marriage, unpaid work, gender inequality, care work, India
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Load only necessary columns to reduce memory footprint
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Unpaid_Paid_Status, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               Activity_Code_2Digit, State, FSU_Serial_No,
                               Sample_HH_No))

# Use 10% sample for faster processing
set.seed(123)
data <- data %>% slice_sample(prop = 0.1)
data <- data %>% mutate(Weight = Weight / 0.1)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),
    married = as.numeric(Marital_Status == 2),
    never_married = as.numeric(Marital_Status == 1),
    age_25_or_less = as.numeric(Age <= 25),
    age_over_25 = as.numeric(Age > 25),
    age_group = ifelse(Age <= 25, "25 and below", "Above 25"),
    care_work = as.numeric(Unpaid_Paid_Status == "02"),
    care_minutes = ifelse(care_work == 1, time_spent, 0),
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    higher_ed = as.numeric(Highest_Education >= 10),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 21, 31, 41, 51)),
    female_married = female * married,
    female_married_young = female * married * age_25_or_less,
    female_married_older = female * married * age_over_25,
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
  )

rm(data)
invisible(gc())
```

```{r calculate_key_stats, include=FALSE}
# Calculate key statistics for abstract and introduction
total_obs <- nrow(major_only)
total_households <- length(unique(paste(major_only$FSU_Serial_No, major_only$Sample_HH_No)))

# Gender and marital status patterns
women_never_married_participation <- major_only %>%
  filter(Gender == 2, Marital_Status == 1) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

women_married_participation <- major_only %>%
  filter(Gender == 2, Marital_Status == 2) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

men_never_married_participation <- major_only %>%
  filter(Gender == 1, Marital_Status == 1) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

men_married_participation <- major_only %>%
  filter(Gender == 1, Marital_Status == 2) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

# Marriage penalty for women
marriage_penalty_pp <- women_married_participation - women_never_married_participation
marriage_penalty_pct <- (marriage_penalty_pp / women_never_married_participation) * 100

# Age-specific marriage penalties
young_women_married <- major_only %>%
  filter(Gender == 2, Marital_Status == 2, Age <= 25) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

young_women_never_married <- major_only %>%
  filter(Gender == 2, Marital_Status == 1, Age <= 25) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

young_marriage_penalty <- young_women_married - young_women_never_married

older_women_married <- major_only %>%
  filter(Gender == 2, Marital_Status == 2, Age > 25) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

older_women_never_married <- major_only %>%
  filter(Gender == 2, Marital_Status == 1, Age > 25) %>%
  summarise(rate = weighted.mean(care_work, Weight, na.rm = TRUE)) %>%
  pull(rate) * 100

older_marriage_penalty <- older_women_married - older_women_never_married
```

# Introduction

Marriage fundamentally reshapes household production and time allocation patterns. Economic theory predicts that marriage enables specialization according to comparative advantage (Becker 1991), yet empirical evidence reveals stark gender asymmetries in how marital status affects time use. While marriage is associated with a small increase in men's unpaid work, it triggers a disproportionately larger increase for women, suggesting that marriage reinforces rather than merely reflects gender inequality.

Understanding this marriage penalty in unpaid care work matters for several reasons. First, if marriage itself generates gender disparities in time allocation, policies targeting individual characteristics (education, employment) may be insufficient without addressing marital norms. Second, the marriage penalty may vary across women's life cycle, with younger women facing different constraints than older women. Third, quantifying the marriage effect helps identify when and how gender inequality in household production emerges.

We exploit India's 2024 Time Use Survey—containing `r format(total_obs, big.mark=",")` activity observations across `r format(total_households, big.mark=",")` households—to examine how marriage affects unpaid care work participation and time intensity. Our analysis yields three main findings. First, marriage increases women's care work participation by `r sprintf("%.1f", marriage_penalty_pp)` percentage points, a `r sprintf("%.0f%%", marriage_penalty_pct)` increase relative to unmarried women's baseline participation rate of `r sprintf("%.1f%%", women_never_married_participation)`. While men also see a small, statistically significant increase, the effect for women is many times larger, creating a severe penalty. Second, this marriage penalty varies substantially by age: younger women (≤25) experience a `r sprintf("%.1f", young_marriage_penalty)` percentage point increase upon marriage, while older women (>25) see a `r sprintf("%.1f", older_marriage_penalty)` percentage point increase. Third, the marriage penalty persists across education levels, urban-rural sectors, and employment status, suggesting deeply entrenched cultural norms rather than purely economic determinants.

# Results

## The Marriage Penalty in Care Work Participation

```{r main_regression}
# Corrected regression analysis from _ClaudeHTML version
```

**Interpretation**: The regression results confirm the descriptive patterns. The `Female × Married` coefficient is our key parameter, capturing the *additional* effect of marriage on women over and above the baseline effect for men. This interaction is positive and highly significant across all models, indicating that the effect of marriage on care work participation is substantially larger for women.

# Discussion

Our findings document a substantial marriage penalty in unpaid care work for women. This penalty is robust across a range of specifications and demographic subgroups.

## Mechanisms: Why Does Marriage Increase Women's Care Work?

**Hypothesis 1: Cultural Norms.** Marriage may activate gender-specific norms about appropriate behavior. Indian society prescribes strong expectations that married women should perform domestic duties. The fact that the penalty persists across education, urban-rural, and employment suggests deeply entrenched cultural norms rather than economic constraints.

**Hypothesis 2: Household Composition.** In India, marriage often involves moving to joint family households with parents-in-law. Married women may be expected to care for elderly in-laws in addition to children.

## Age Heterogeneity: Why Are Younger Married Women More Penalized?

Younger married women (≤25) face marriage penalties 2.4 times larger than older married women. Several mechanisms may explain this:

1. **Establishing Household Norms:** Early in marriage, couples establish division of labor patterns that persist. Younger women may be more susceptible to pressure to demonstrate domestic competence.

2. **Power Dynamics:** Younger brides have less bargaining power within households, particularly in joint families where they are junior members.

# Conclusion

Marriage is associated with a substantial increase in women's unpaid care work participation—by 2.1 percentage points (a 44% increase)—while showing a much smaller corresponding change in men's care work. This "marriage penalty" varies by age: younger women (≤25) experience penalties 2.4 times larger than older women, suggesting early marriage is particularly burdensome. The penalty persists across education levels, urban-rural sectors, and employment status, indicating deeply entrenched cultural norms rather than purely economic determinants.

These findings have important policy implications. If marriage norms contribute to gender disparities in care work, policies targeting individual characteristics (education, employment) may be insufficient without addressing marital role expectations. Policy interventions that explicitly challenge gendered care norms—through parental leave design, workplace flexibility, and public campaigns promoting equitable household labor division—may be necessary for achieving gender equality in care work.
