---
title: "Master Descriptive Statistics Reference"
subtitle: "India Time Use Survey 2024 - Comprehensive Activity and Demographic Patterns"
author: "Research Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: true
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

\newpage

# Executive Summary

This document provides comprehensive descriptive statistics for India's Time Use Survey 2024 to inform research question selection. It includes:

- **Participation rates** for all major activity codes
- **Time spent** conditional on participation (mean, median, range)
- **Sample sizes** for key demographic subgroups
- **Household composition** patterns
- **Cross-tabulations** by gender, marital status, age, education, and urban/rural status

---

# Data Overview

```{r load_data}
# STEP 1: Load full dataset to check available columns
cat("Loading data and checking column names...\n")

# Load WITHOUT specifying columns to see what exists
data_full <- read_dta("TimeUse106PER.dta")

# Check for relationship variable (might have different name)
all_cols <- names(data_full)
cat("Total columns available:", length(all_cols), "\n")

# Detect relationship column name
relationship_cols <- grep("relation|Relation|RELATION|head|Head|HEAD",
                          all_cols, value = TRUE, ignore.case = TRUE)

if ("Relationship_to_Head" %in% all_cols) {
  rel_col <- "Relationship_to_Head"
} else if ("Relation_to_Head" %in% all_cols) {
  rel_col <- "Relation_to_Head"
} else if ("Rel_to_Head" %in% all_cols) {
  rel_col <- "Rel_to_Head"
} else if (length(relationship_cols) > 0) {
  rel_col <- relationship_cols[1]  # Use first match
} else {
  rel_col <- NULL  # Will create dummy variable
  cat("WARNING: No relationship-to-head column found. Some analyses will be limited.\n")
}

# Detect household size column
hh_size_cols <- grep("household.*size|Household.*Size|HH.*Size|hhsize",
                     all_cols, value = TRUE, ignore.case = TRUE)

if ("Household_Size" %in% all_cols) {
  hh_size_col <- "Household_Size"
} else if ("household_size" %in% all_cols) {
  hh_size_col <- "household_size"
} else if (length(hh_size_cols) > 0) {
  hh_size_col <- hh_size_cols[1]
} else {
  hh_size_col <- NULL
  cat("WARNING: No household size column found.\n")
}

# Build column selection vector
cols_to_select <- c("FSU_Serial_No", "Sample_HH_No", "Person_Serial_No",
                    "Gender", "Age", "Marital_Status", "Highest_Education",
                    "Sector", "Principal_Activity_Status",
                    "Activity_Code_2Digit", "Activity_Code_3Digit",
                    "time_spent", "Weight", "Major_Activity_Flag",
                    "Unpaid_Paid_Status", "Day_of_Week", "State")

if (!is.null(rel_col)) cols_to_select <- c(cols_to_select, rel_col)
if (!is.null(hh_size_col)) cols_to_select <- c(cols_to_select, hh_size_col)

# Clean up
rm(data_full)
gc()

# Load data with correct column selection
set.seed(12345)
data <- read_dta("TimeUse106PER.dta", col_select = cols_to_select)

# Rename to standard names if needed
if (!is.null(rel_col) && rel_col != "Relationship_to_Head") {
  data <- data %>% rename(Relationship_to_Head = !!sym(rel_col))
}
if (!is.null(hh_size_col) && hh_size_col != "Household_Size") {
  data <- data %>% rename(Household_Size = !!sym(hh_size_col))
}

# Create dummy variables if columns don't exist
if (is.null(rel_col)) {
  data <- data %>% mutate(Relationship_to_Head = NA_real_)
}
if (is.null(hh_size_col)) {
  data <- data %>% mutate(Household_Size = NA_real_)
}

# 10% random sample
sampled_data <- data %>%
  mutate(sample_id = row_number()) %>%
  filter(sample_id %% 10 == 1) %>%
  mutate(Weight = Weight / 0.1)  # Adjust weights

# Filter to major activities only
major_only <- sampled_data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    # Create standard labels
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    marital_label = case_when(
      Marital_Status == 1 ~ "Never Married",
      Marital_Status == 2 ~ "Currently Married",
      Marital_Status == 3 ~ "Widowed",
      Marital_Status == 4 ~ "Divorced/Separated"
    ),
    sector_label = ifelse(Sector == 1, "Rural", "Urban")
  )

rm(sampled_data, data)
gc()
```

\newpage

# Activity Code Participation Rates

## All Major Activity Categories

```{r activity_participation}
# Code from _ClaudeHTML version
```

\newpage

# Demographic Breakdowns

## Overall Demographics

```{r overall_demographics}
# Code from _ClaudeHTML version
```

\newpage

## Activity Participation by Gender and Marital Status

### Educational Activities (21-29) by Demographics

```{r education_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_edu <- major_only %>%
  mutate(edu_activity = as.numeric(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_edu = max(edu_activity, na.rm = TRUE),
    total_edu_min = sum(time_spent * edu_activity, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

edu_act_demo <- person_level_edu %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_edu),
    Participation_pct = weighted.mean(any_edu, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_edu_min[any_edu == 1],
        Weight[any_edu == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(edu_act_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Educational Activities by Gender and Marital Status")
```

### Physical Exercise (71) by Demographics

```{r exercise_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_exercise <- major_only %>%
  mutate(exercise = as.numeric(Activity_Code_2Digit == 71)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_exercise = max(exercise, na.rm = TRUE),
    total_exercise_min = sum(time_spent * exercise, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

exercise_demo <- person_level_exercise %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_exercise),
    Participation_pct = weighted.mean(any_exercise, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_exercise_min[any_exercise == 1],
        Weight[any_exercise == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)


kable(exercise_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Physical Exercise by Gender and Marital Status")
```

### Religious Activities (9) by Demographics

```{r religious_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_religious <- major_only %>%
  mutate(religious = as.numeric(Activity_Code_2Digit == 9)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_religious = max(religious, na.rm = TRUE),
    total_religious_min = sum(time_spent * religious, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

religious_demo <- person_level_religious %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_religious),
    Participation_pct = weighted.mean(any_religious, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_religious_min[any_religious == 1],
        Weight[any_religious == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(religious_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Religious Activities by Gender and Marital Status")
```

---
... (rest of the corrected file)
---
