---
title: "Master Descriptive Statistics Reference"
subtitle: "India Time Use Survey 2024 - Comprehensive Activity and Demographic Patterns"
author: "Research Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: true
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

\newpage

# Executive Summary

This document provides comprehensive descriptive statistics for India's Time Use Survey 2024 to inform research question selection. It includes:

- **Participation rates** for all major activity codes
- **Time spent** conditional on participation (mean, median, range)
- **Sample sizes** for key demographic subgroups
- **Household composition** patterns
- **Cross-tabulations** by gender, marital status, age, education, and urban/rural status

---

# Data Overview

```{r load_data}
# Corrected data loading and processing from _ClaudeHTML version
```

\newpage

# Activity Code Participation Rates

## All Major Activity Categories

```{r activity_participation}
# Code from _ClaudeHTML version
```

\newpage

# Demographic Breakdowns

## Overall Demographics

```{r overall_demographics}
# Code from _ClaudeHTML version
```

\newpage

## Activity Participation by Gender and Marital Status

### Educational Activities (21-29) by Demographics

```{r education_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_edu <- major_only %>%
  mutate(edu_activity = as.numeric(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_edu = max(edu_activity, na.rm = TRUE),
    total_edu_min = sum(time_spent * edu_activity, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

edu_act_demo <- person_level_edu %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_edu),
    Participation_pct = weighted.mean(any_edu, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_edu_min[any_edu == 1],
        Weight[any_edu == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(edu_act_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Educational Activities by Gender and Marital Status")
```

### Physical Exercise (71) by Demographics

```{r exercise_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_exercise <- major_only %>%
  mutate(exercise = as.numeric(Activity_Code_2Digit == 71)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_exercise = max(exercise, na.rm = TRUE),
    total_exercise_min = sum(time_spent * exercise, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

exercise_demo <- person_level_exercise %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_exercise),
    Participation_pct = weighted.mean(any_exercise, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_exercise_min[any_exercise == 1],
        Weight[any_exercise == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)


kable(exercise_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Physical Exercise by Gender and Marital Status")
```

### Religious Activities (9) by Demographics

```{r religious_demographics}
# --- FIX: Correct participation rate calculation ---
person_level_religious <- major_only %>%
  mutate(religious = as.numeric(Activity_Code_2Digit == 9)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_religious = max(religious, na.rm = TRUE),
    total_religious_min = sum(time_spent * religious, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  )

religious_demo <- person_level_religious %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = sum(any_religious),
    Participation_pct = weighted.mean(any_religious, Weight, na.rm = TRUE) * 100,
    Mean_min = weighted.mean(
        total_religious_min[any_religious == 1],
        Weight[any_religious == 1],
        na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(religious_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Religious Activities by Gender and Marital Status")
```

---
... (rest of the corrected file)
---
