---
title: "Master Descriptive Statistics Reference"
subtitle: "India Time Use Survey 2024 - Comprehensive Activity and Demographic Patterns"
author: "Research Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: true
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6,
                      cache = TRUE)

# Load packages
library(haven)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

\newpage

# Executive Summary

This document provides comprehensive descriptive statistics for India's Time Use Survey 2024 to inform research question selection. It includes:

- **Participation rates** for all major activity codes
- **Time spent** conditional on participation (mean, median, range)
- **Sample sizes** for key demographic subgroups
- **Household composition** patterns
- **Cross-tabulations** by gender, marital status, age, education, and urban/rural status

**Key Use**: Consult this document BEFORE designing a research question to assess:
1. Whether sufficient people participate in the activity of interest
2. What demographic variation exists
3. What sample sizes are available for subgroup analyses

---

# Data Overview

```{r load_data}
# Load data with 10% random sample for computational feasibility
set.seed(12345)

data <- read_dta("TimeUse106PER.dta",
                 col_select = c(
                   "FSU_Serial_No", "Sample_HH_No", "Person_Serial_No",
                   "Gender", "Age", "Marital_Status", "Highest_Education",
                   "Sector", "Principal_Activity_Status",
                   "Activity_Code_2Digit", "Activity_Code_3Digit",
                   "time_spent", "Weight", "Major_Activity_Flag",
                   "Unpaid_Paid_Status", "Day_of_Week",
                   "Relationship_to_Head", "State", "Household_Size"
                 ))

# 10% random sample
sampled_data <- data %>%
  mutate(sample_id = row_number()) %>%
  filter(sample_id %% 10 == 1) %>%
  mutate(Weight = Weight / 0.1)  # Adjust weights

# Filter to major activities only
major_only <- sampled_data %>%
  filter(Major_Activity_Flag == 1)

rm(sampled_data, data)
gc()
```

```{r sample_overview}
overview_stats <- data.frame(
  Metric = c("Total Observations (10% sample)",
             "Unique Households",
             "Unique Individuals",
             "Average observations per person",
             "Survey Period"),
  Value = c(
    format(nrow(major_only), big.mark = ","),
    format(length(unique(paste(major_only$FSU_Serial_No, major_only$Sample_HH_No))), big.mark = ","),
    format(length(unique(paste(major_only$FSU_Serial_No, major_only$Sample_HH_No,
                               major_only$Person_Serial_No))), big.mark = ","),
    round(nrow(major_only) / length(unique(paste(major_only$FSU_Serial_No,
                                                  major_only$Sample_HH_No,
                                                  major_only$Person_Serial_No))), 1),
    "January-December 2024"
  )
)

kable(overview_stats, booktabs = TRUE,
      caption = "Dataset Overview (10\\% Random Sample with Adjusted Weights)") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

# Activity Code Participation Rates

## All Major Activity Categories

```{r activity_participation}
# Create activity category labels based on ICATUS 2016
activity_labels <- data.frame(
  code = c(1, 2, 3, 4, 5, 6, 7, 8, 9,
           21, 22, 23, 24, 25, 26, 27, 28, 29,
           31, 32, 33, 34, 35,
           41, 42, 43, 44, 45, 46, 47, 48, 49,
           51, 52, 53, 54, 55, 56, 57, 58, 59,
           61, 62, 63, 64, 65, 66, 67, 68, 69,
           71, 72, 73, 74, 75, 76, 77, 78, 79,
           81, 82, 83, 84, 85, 86, 87, 88, 89,
           91, 92, 93, 94, 95, 96, 97, 98, 99),
  label = c(
    "Employment", "Production/processing for own use", "Unpaid trainee/apprentice",
    "Unpaid work business/farm", "Job search", "Starting business",
    "Study/training for employment", "Improving job skills", "Other employment-related",

    "Formal education/schooling", "Homework/assignments", "Additional classes",
    "Library/learning resources", "Learning new skills", "Education-related travel",
    "Discussions with teachers", "Educational admin", "Other education",

    "Food preparation", "Household upkeep", "Care for textiles/clothing",
    "Shopping/procurement", "Household management",

    "Physical childcare", "Teaching/supervising children", "Accompanying children",
    "Transporting children", "Childcare travel", "Reading to children",
    "Playing with children", "Healthcare for children", "Other childcare",

    "Physical care adults", "Healthcare support adults", "Administrative help adults",
    "Accompanying adults", "Transporting adults", "Travel for adult care",
    "Emotional support adults", "Supervising adults", "Other adult care",

    "Participating in culture", "Attending cultural events", "Visiting heritage sites",
    "Hobbies/crafts", "Games/puzzles", "Collecting items", "Reading for leisure",
    "Watching TV/videos", "Other culture/leisure",

    "Physical exercise/sports", "Outdoor recreation", "Hunting/fishing",
    "Sports events attendance", "Sports travel", "Walking for leisure",
    "Yoga/meditation", "Sports organizing", "Other sports/outdoor",

    "Socializing in person", "Socializing remotely", "Participating in events",
    "Attending ceremonies", "Civic participation", "Organizational meetings",
    "Helping others", "Volunteer work", "Other social activities",

    "Personal hygiene", "Eating/drinking", "Sleeping/napping", "Healthcare",
    "Receiving personal services", "Private activities", "Resting",
    "Waiting/queuing", "Other personal care"
  )
)

# Calculate participation for all 2-digit codes
participation_2digit <- major_only %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_observations = n(),
    N_people_participating = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_rate_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_minutes = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_minutes = median(rep(time_spent, times = pmax(1, round(Weight)))),
    Min_minutes = min(time_spent, na.rm = TRUE),
    Max_minutes = max(time_spent, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code")) %>%
  arrange(desc(Participation_rate_pct))

kable(participation_2digit %>%
        select(Activity_Code_2Digit, label, Participation_rate_pct,
               N_people_participating, Mean_minutes, Median_minutes),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Participation (%)", "N People",
                    "Mean (min)", "Median (min)"),
      caption = "Participation Rates for All Activity Codes (Sorted by Participation Rate)") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

\newpage

## Key Activity Codes for Research Questions

### Educational Activities (Codes 21-29)

```{r education_activities}
education_detail <- major_only %>%
  filter(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29) %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_obs = n(),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    P25 = quantile(rep(time_spent, times = pmax(1, round(Weight))), 0.25),
    P75 = quantile(rep(time_spent, times = pmax(1, round(Weight))), 0.75),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code"))

kable(education_detail %>%
        select(Activity_Code_2Digit, label, Participation_pct, N_obs,
               Mean_min, Median_min, P25, P75),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Part. (%)", "N",
                    "Mean", "Median", "P25", "P75"),
      caption = "Educational Activities Detail (Codes 21-29)") %>%
  kable_styling(latex_options = c("hold_position"))
```

### Physical Exercise and Sports (Code 71)

```{r exercise_activities}
exercise_detail <- major_only %>%
  filter(Activity_Code_2Digit == 71) %>%
  summarise(
    N_obs = n(),
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    SD_min = sd(rep(time_spent, times = pmax(1, round(Weight)))),
    Min_min = min(time_spent),
    Max_min = max(time_spent)
  )

kable(t(exercise_detail), booktabs = TRUE, digits = 2,
      col.names = "Value",
      caption = "Physical Exercise and Sports (Code 71) - Summary Statistics") %>%
  kable_styling(latex_options = c("hold_position"))
```

### Religious and Spiritual Activities (Code 9)

```{r religious_activities}
religious_detail <- major_only %>%
  filter(Activity_Code_2Digit == 9) %>%
  summarise(
    N_obs = n(),
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    SD_min = sd(rep(time_spent, times = pmax(1, round(Weight)))),
    Min_min = min(time_spent),
    Max_min = max(time_spent)
  )

kable(t(religious_detail), booktabs = TRUE, digits = 2,
      col.names = "Value",
      caption = "Religious and Spiritual Activities (Code 9) - Summary Statistics") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

# Demographic Breakdowns

## Overall Demographics

```{r overall_demographics}
major_only <- major_only %>%
  mutate(
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender",
      TRUE ~ "Other"
    ),
    marital_label = case_when(
      Marital_Status == 1 ~ "Never married",
      Marital_Status == 2 ~ "Currently married",
      Marital_Status == 3 ~ "Widowed",
      Marital_Status == 4 ~ "Divorced/Separated",
      TRUE ~ "Unknown"
    ),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education",
      TRUE ~ "Unknown"
    ),
    age_group = case_when(
      Age < 18 ~ "<18",
      Age >= 18 & Age < 25 ~ "18-24",
      Age >= 25 & Age < 35 ~ "25-34",
      Age >= 35 & Age < 45 ~ "35-44",
      Age >= 45 & Age < 55 ~ "45-54",
      Age >= 55 & Age < 65 ~ "55-64",
      Age >= 65 ~ "65+",
      TRUE ~ "Unknown"
    ),
    urban = as.numeric(Sector == 2)
  )

# Gender distribution
gender_dist <- major_only %>%
  group_by(gender_label) %>%
  summarise(
    N = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight)) * 100,
    .groups = "drop"
  )

kable(gender_dist, booktabs = TRUE, digits = 2,
      caption = "Gender Distribution") %>%
  kable_styling(latex_options = c("hold_position"))

# Marital status distribution
marital_dist <- major_only %>%
  group_by(marital_label) %>%
  summarise(
    N = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight)) * 100,
    .groups = "drop"
  )

kable(marital_dist, booktabs = TRUE, digits = 2,
      caption = "Marital Status Distribution") %>%
  kable_styling(latex_options = c("hold_position"))

# Age group distribution
age_dist <- major_only %>%
  group_by(age_group) %>%
  summarise(
    N = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight)) * 100,
    .groups = "drop"
  )

kable(age_dist, booktabs = TRUE, digits = 2,
      caption = "Age Group Distribution") %>%
  kable_styling(latex_options = c("hold_position"))

# Education distribution
edu_dist <- major_only %>%
  filter(education_level != "Unknown") %>%
  group_by(education_level) %>%
  summarise(
    N = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight[major_only$education_level != "Unknown"])) * 100,
    .groups = "drop"
  )

kable(edu_dist, booktabs = TRUE, digits = 2,
      caption = "Education Level Distribution") %>%
  kable_styling(latex_options = c("hold_position"))

# Urban/Rural distribution
sector_dist <- major_only %>%
  group_by(urban) %>%
  summarise(
    Sector = ifelse(first(urban) == 1, "Urban", "Rural"),
    N = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight)) * 100,
    .groups = "drop"
  )

kable(sector_dist %>% select(Sector, N, Percent), booktabs = TRUE, digits = 2,
      caption = "Urban/Rural Distribution") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

## Activity Participation by Gender and Marital Status

### Educational Activities (21-29) by Demographics

```{r education_demographics}
edu_act_demo <- major_only %>%
  filter(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$gender_label == first(gender_label) &
      major_only$marital_label == first(marital_label)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(edu_act_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Educational Activities by Gender and Marital Status") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

### Physical Exercise (71) by Demographics

```{r exercise_demographics}
exercise_demo <- major_only %>%
  filter(Activity_Code_2Digit == 71) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$gender_label == first(gender_label) &
      major_only$marital_label == first(marital_label)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(exercise_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Physical Exercise by Gender and Marital Status") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

### Religious Activities (9) by Demographics

```{r religious_demographics}
religious_demo <- major_only %>%
  filter(Activity_Code_2Digit == 9) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$gender_label == first(gender_label) &
      major_only$marital_label == first(marital_label)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(religious_demo, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Religious Activities by Gender and Marital Status") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

\newpage

# Household Composition Patterns

## Relationship to Household Head

```{r relationship_head}
relationship_labels <- data.frame(
  code = 1:9,
  label = c("Self (Household Head)", "Spouse of Head", "Married Child",
            "Spouse of Married Child", "Unmarried Child", "Grandchild",
            "Father/Mother/Father-in-law/Mother-in-law", "Other Relative",
            "Servant/Employee/Other Non-Relative")
)

rel_dist <- major_only %>%
  group_by(Relationship_to_Head) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Percent = (sum(Weight) / sum(major_only$Weight)) * 100,
    .groups = "drop"
  ) %>%
  left_join(relationship_labels, by = c("Relationship_to_Head" = "code"))

kable(rel_dist %>% select(label, N_people, Percent),
      booktabs = TRUE, digits = 2,
      col.names = c("Relationship to Head", "N People", "Percent (%)"),
      caption = "Distribution of Household Members by Relationship to Head") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Household Head Characteristics

```{r household_head_characteristics}
# Filter to household heads only
heads <- major_only %>%
  filter(Relationship_to_Head == 1) %>%
  distinct(FSU_Serial_No, Sample_HH_No, Person_Serial_No, .keep_all = TRUE)

head_gender <- heads %>%
  group_by(gender_label) %>%
  summarise(
    N_households = n(),
    Percent = (sum(Weight) / sum(heads$Weight)) * 100,
    .groups = "drop"
  )

kable(head_gender, booktabs = TRUE, digits = 2,
      col.names = c("Gender of Head", "N Households", "Percent (%)"),
      caption = "Gender Distribution of Household Heads") %>%
  kable_styling(latex_options = c("hold_position"))

head_age <- heads %>%
  summarise(
    Mean_age = weighted.mean(Age, Weight, na.rm = TRUE),
    Median_age = median(rep(Age, times = pmax(1, round(Weight)))),
    Min_age = min(Age, na.rm = TRUE),
    Max_age = max(Age, na.rm = TRUE)
  )

kable(t(head_age), booktabs = TRUE, digits = 1,
      col.names = "Value",
      caption = "Age Distribution of Household Heads") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Children in Households

```{r children_statistics}
# Identify children (age 6-18, relationship codes 5 or 6)
children <- major_only %>%
  filter(Age >= 6 & Age <= 18, Relationship_to_Head %in% c(5, 6)) %>%
  distinct(FSU_Serial_No, Sample_HH_No, Person_Serial_No, .keep_all = TRUE)

child_stats <- data.frame(
  Metric = c("Total children (age 6-18)",
             "Children in female-headed households",
             "Children in male-headed households",
             "Percent female-headed (among children's HH)",
             "Average age of children"),
  Value = c(
    nrow(children),
    NA,  # Will calculate below
    NA,
    NA,
    round(weighted.mean(children$Age, children$Weight, na.rm = TRUE), 1)
  )
)

# Need to merge with household head gender
children_with_head <- major_only %>%
  filter(Age >= 6 & Age <= 18, Relationship_to_Head %in% c(5, 6)) %>%
  distinct(FSU_Serial_No, Sample_HH_No, Person_Serial_No, .keep_all = TRUE) %>%
  left_join(
    heads %>% select(FSU_Serial_No, Sample_HH_No, head_gender = gender_label),
    by = c("FSU_Serial_No", "Sample_HH_No")
  )

child_by_head_gender <- children_with_head %>%
  group_by(head_gender) %>%
  summarise(
    N_children = n(),
    Percent = (sum(Weight, na.rm = TRUE) / sum(children_with_head$Weight, na.rm = TRUE)) * 100,
    .groups = "drop"
  )

kable(child_by_head_gender, booktabs = TRUE, digits = 2,
      col.names = c("Gender of Household Head", "N Children (6-18)", "Percent (%)"),
      caption = "Children (Age 6-18) by Gender of Household Head") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

# Cross-Tabulations for Research Questions

## Q1: Children's Educational Activities by Household Head Gender

```{r children_education_by_head}
children_edu <- major_only %>%
  filter(Age >= 6 & Age <= 18, Relationship_to_Head %in% c(5, 6),
         Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29) %>%
  left_join(
    heads %>% select(FSU_Serial_No, Sample_HH_No, head_gender = gender_label),
    by = c("FSU_Serial_No", "Sample_HH_No")
  ) %>%
  group_by(head_gender) %>%
  summarise(
    N_children_in_edu = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Total_observations = n(),
    Mean_minutes = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_minutes = median(rep(time_spent, times = pmax(1, round(Weight)))),
    .groups = "drop"
  )

kable(children_edu, booktabs = TRUE, digits = 2,
      col.names = c("Gender of HH Head", "N Children", "N Obs", "Mean (min)", "Median (min)"),
      caption = "Children's Educational Activities by Household Head Gender") %>%
  kable_styling(latex_options = c("hold_position"))

# Participation rate
children_edu_participation <- major_only %>%
  filter(Age >= 6 & Age <= 18, Relationship_to_Head %in% c(5, 6)) %>%
  left_join(
    heads %>% select(FSU_Serial_No, Sample_HH_No, head_gender = gender_label),
    by = c("FSU_Serial_No", "Sample_HH_No")
  ) %>%
  mutate(edu_activity = as.numeric(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, head_gender) %>%
  summarise(
    any_edu = max(edu_activity, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  ) %>%
  group_by(head_gender) %>%
  summarise(
    N_children = n(),
    Participation_pct = weighted.mean(any_edu, Weight, na.rm = TRUE) * 100,
    .groups = "drop"
  )

kable(children_edu_participation, booktabs = TRUE, digits = 2,
      col.names = c("Gender of HH Head", "N Children", "Participation (%)"),
      caption = "Children's Educational Activity Participation Rate by HH Head Gender") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Q2: Exercise by Gender and Marital Status (Marriage Penalty Analysis)

```{r exercise_marriage_penalty}
exercise_marriage <- major_only %>%
  filter(gender_label %in% c("Male", "Female"),
         marital_label %in% c("Never married", "Currently married")) %>%
  mutate(exercise = as.numeric(Activity_Code_2Digit == 71)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_exercise = max(exercise, na.rm = TRUE),
    total_exercise_min = sum(time_spent * exercise, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  ) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = n(),
    Participation_pct = weighted.mean(any_exercise, Weight, na.rm = TRUE) * 100,
    Mean_minutes = weighted.mean(total_exercise_min, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(exercise_marriage, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Exercise Participation: Marriage Penalty Analysis") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Q3: Religious Activities by Gender and Marital Status

```{r religious_marriage_analysis}
religious_marriage <- major_only %>%
  filter(gender_label %in% c("Male", "Female"),
         marital_label %in% c("Never married", "Currently married")) %>%
  mutate(religious = as.numeric(Activity_Code_2Digit == 9)) %>%
  group_by(FSU_Serial_No, Sample_HH_No, Person_Serial_No, gender_label, marital_label) %>%
  summarise(
    any_religious = max(religious, na.rm = TRUE),
    total_religious_min = sum(time_spent * religious, na.rm = TRUE),
    Weight = first(Weight),
    .groups = "drop"
  ) %>%
  group_by(gender_label, marital_label) %>%
  summarise(
    N_people = n(),
    Participation_pct = weighted.mean(any_religious, Weight, na.rm = TRUE) * 100,
    Mean_minutes = weighted.mean(total_religious_min, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(gender_label, marital_label)

kable(religious_marriage, booktabs = TRUE, digits = 2,
      col.names = c("Gender", "Marital Status", "N People", "Part. (%)", "Mean (min)"),
      caption = "Religious Activities: Gender and Marriage Patterns") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

# Additional Activity Breakdowns

## Leisure and Recreation Activities (61-69)

```{r leisure_activities}
leisure_detail <- major_only %>%
  filter(Activity_Code_2Digit >= 61 & Activity_Code_2Digit <= 69) %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_obs = n(),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code"))

kable(leisure_detail %>%
        select(Activity_Code_2Digit, label, Participation_pct, N_obs, Mean_min, Median_min),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Part. (%)", "N", "Mean", "Median"),
      caption = "Leisure and Recreation Activities (Codes 61-69)") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

## Social Activities (81-89)

```{r social_activities}
social_detail <- major_only %>%
  filter(Activity_Code_2Digit >= 81 & Activity_Code_2Digit <= 89) %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_obs = n(),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code"))

kable(social_detail %>%
        select(Activity_Code_2Digit, label, Participation_pct, N_obs, Mean_min, Median_min),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Part. (%)", "N", "Mean", "Median"),
      caption = "Social Life and Communication Activities (Codes 81-89)") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

## Volunteer and Community Service (51-59)

```{r volunteer_activities}
volunteer_detail <- major_only %>%
  filter(Activity_Code_2Digit >= 51 & Activity_Code_2Digit <= 59) %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_obs = n(),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code"))

kable(volunteer_detail %>%
        select(Activity_Code_2Digit, label, Participation_pct, N_obs, Mean_min, Median_min),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Part. (%)", "N", "Mean", "Median"),
      caption = "Volunteer and Community Service Activities (Codes 51-59)") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

\newpage

# Age-Specific Patterns

## Educational Activities by Age Group

```{r education_by_age}
edu_by_age <- major_only %>%
  filter(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29) %>%
  group_by(age_group) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$age_group == first(age_group)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_group)

kable(edu_by_age, booktabs = TRUE, digits = 2,
      col.names = c("Age Group", "N People", "Part. (%)", "Mean (min)"),
      caption = "Educational Activities by Age Group") %>%
  kable_styling(latex_options = c("hold_position"))
```

## Exercise by Age Group and Gender

```{r exercise_by_age_gender}
exercise_age_gender <- major_only %>%
  filter(Activity_Code_2Digit == 71, gender_label %in% c("Male", "Female")) %>%
  group_by(age_group, gender_label) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$age_group == first(age_group) &
      major_only$gender_label == first(gender_label)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_group, gender_label)

kable(exercise_age_gender, booktabs = TRUE, digits = 2,
      col.names = c("Age Group", "Gender", "N People", "Part. (%)", "Mean (min)"),
      caption = "Exercise by Age Group and Gender") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

## Religious Activities by Age Group and Gender

```{r religious_by_age_gender}
religious_age_gender <- major_only %>%
  filter(Activity_Code_2Digit == 9, gender_label %in% c("Male", "Female")) %>%
  group_by(age_group, gender_label) %>%
  summarise(
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight[
      major_only$age_group == first(age_group) &
      major_only$gender_label == first(gender_label)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_group, gender_label)

kable(religious_age_gender, booktabs = TRUE, digits = 2,
      col.names = c("Age Group", "Gender", "N People", "Part. (%)", "Mean (min)"),
      caption = "Religious Activities by Age Group and Gender") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

\newpage

# Urban-Rural Patterns

## Key Activities by Urban/Rural Status

```{r urban_rural_comparison}
urban_rural_activities <- data.frame(
  activity_name = character(),
  codes = character(),
  Rural_Part = numeric(),
  Urban_Part = numeric(),
  Rural_Mean = numeric(),
  Urban_Mean = numeric()
)

# Educational activities
edu_ur <- major_only %>%
  filter(Activity_Code_2Digit >= 21 & Activity_Code_2Digit <= 29) %>%
  group_by(urban) %>%
  summarise(
    Part_pct = (sum(Weight) / sum(major_only$Weight[major_only$urban == first(urban)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  )

urban_rural_activities <- rbind(urban_rural_activities,
  data.frame(
    activity_name = "Educational activities",
    codes = "21-29",
    Rural_Part = edu_ur$Part_pct[edu_ur$urban == 0],
    Urban_Part = edu_ur$Part_pct[edu_ur$urban == 1],
    Rural_Mean = edu_ur$Mean_min[edu_ur$urban == 0],
    Urban_Mean = edu_ur$Mean_min[edu_ur$urban == 1]
  )
)

# Exercise
exercise_ur <- major_only %>%
  filter(Activity_Code_2Digit == 71) %>%
  group_by(urban) %>%
  summarise(
    Part_pct = (sum(Weight) / sum(major_only$Weight[major_only$urban == first(urban)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  )

urban_rural_activities <- rbind(urban_rural_activities,
  data.frame(
    activity_name = "Physical exercise",
    codes = "71",
    Rural_Part = exercise_ur$Part_pct[exercise_ur$urban == 0],
    Urban_Part = exercise_ur$Part_pct[exercise_ur$urban == 1],
    Rural_Mean = exercise_ur$Mean_min[exercise_ur$urban == 0],
    Urban_Mean = exercise_ur$Mean_min[exercise_ur$urban == 1]
  )
)

# Religious activities
religious_ur <- major_only %>%
  filter(Activity_Code_2Digit == 9) %>%
  group_by(urban) %>%
  summarise(
    Part_pct = (sum(Weight) / sum(major_only$Weight[major_only$urban == first(urban)])) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    .groups = "drop"
  )

urban_rural_activities <- rbind(urban_rural_activities,
  data.frame(
    activity_name = "Religious activities",
    codes = "9",
    Rural_Part = religious_ur$Part_pct[religious_ur$urban == 0],
    Urban_Part = religious_ur$Part_pct[religious_ur$urban == 1],
    Rural_Mean = religious_ur$Mean_min[religious_ur$urban == 0],
    Urban_Mean = religious_ur$Mean_min[religious_ur$urban == 1]
  )
)

kable(urban_rural_activities, booktabs = TRUE, digits = 2,
      col.names = c("Activity", "Codes", "Rural Part. (%)", "Urban Part. (%)",
                    "Rural Mean", "Urban Mean"),
      caption = "Key Activities by Urban/Rural Status") %>%
  kable_styling(latex_options = c("hold_position"))
```

\newpage

# Recommendations for Research Question Selection

## Summary of Feasibility

Based on the descriptive statistics, here are recommendations for the three proposed research questions:

### 1. Children's Educational Activities & Household Head Gender

**Feasibility**: Consult the tables in Section 6.1 for:
- Sample sizes of children in female-headed vs male-headed households
- Participation rates in educational activities
- Mean time spent when participating

**Decision criteria**:
- If participation rate > 10%: HIGH FEASIBILITY
- If participation rate 5-10%: MODERATE FEASIBILITY (proceed with caution)
- If participation rate < 5%: LOW FEASIBILITY (consider alternative specification)

### 2. Exercise and Marriage Penalty

**Feasibility**: Consult the tables in Section 6.2 for:
- Exercise participation by gender and marital status
- Sample sizes for never-married vs married individuals
- Gender gaps in participation

**Decision criteria**:
- If participation rate > 2%: FEASIBLE for marriage penalty analysis
- If participation rate 1-2%: MARGINAL (consider combining with broader recreation codes)
- If participation rate < 1%: NOT FEASIBLE (too few observations)

### 3. Religious Activities and Gender

**Feasibility**: Consult the tables in Section 6.3 for:
- Religious activity participation by gender and marital status
- Sample sizes for key comparisons
- Age and education patterns

**Decision criteria**:
- If participation rate > 5%: HIGH FEASIBILITY
- If clear gender differences exist: STRONG research question
- If age/education/marital variation exists: Multiple testable hypotheses

---

## General Guidelines

**Minimum participation rates for robust analysis**:
- Main effects (gender, marital status): 2-3% participation minimum
- Interaction effects (gender × marital × age): 5-7% participation recommended
- Triple interactions: 10%+ participation ideal

**Sample size considerations**:
- For regression analysis with state fixed effects: 10,000+ observations ideal
- For subgroup analysis: 1,000+ observations per subgroup
- For rare activities: May need to pool age groups or simplify specifications

**Statistical power**:
- Larger participation rates = more precise estimates
- Lower participation rates = need larger effect sizes to detect

---

\newpage

# Appendix: Full Activity Code List

```{r full_activity_list}
# All activity codes with participation
all_activities <- major_only %>%
  group_by(Activity_Code_2Digit) %>%
  summarise(
    N_obs = n(),
    N_people = length(unique(paste(FSU_Serial_No, Sample_HH_No, Person_Serial_No))),
    Participation_pct = (sum(Weight) / sum(major_only$Weight)) * 100,
    Mean_min = weighted.mean(time_spent, Weight, na.rm = TRUE),
    Median_min = median(rep(time_spent, times = pmax(1, round(Weight)))),
    SD_min = sd(rep(time_spent, times = pmax(1, round(Weight)))),
    .groups = "drop"
  ) %>%
  left_join(activity_labels, by = c("Activity_Code_2Digit" = "code")) %>%
  arrange(Activity_Code_2Digit)

kable(all_activities %>%
        select(Activity_Code_2Digit, label, Participation_pct, N_obs, N_people,
               Mean_min, Median_min),
      booktabs = TRUE, digits = 2,
      col.names = c("Code", "Activity", "Part. (%)", "N Obs", "N People",
                    "Mean", "Median"),
      caption = "Complete Activity Code Reference (All Codes)") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

---

**Document compiled**: `r format(Sys.time(), '%B %d, %Y at %H:%M')`

**Note**: All statistics use 10% random sample with adjusted weights. Percentages represent population estimates. Sample sizes (N) refer to the 10% sample and should be multiplied by 10 for full dataset estimates.
