---
title: "Socioeconomic Status and Women's Domestic Work in India: Evidence of Burden Reduction, Not Task Transformation"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey (10.2 million observations, 139,489 households), we test the "servant paradox" hypothesis, which posits that access to household help transforms women's domestic work from physical tasks to managerial ones, without reducing the total time burden. Using education level and urban residence as proxies for socioeconomic status and access to domestic help, our findings challenge this hypothesis. We find that higher-SES women perform significantly less domestic work overall—approximately 22 minutes per day—driven by a large reduction in physical tasks like cooking and cleaning. Contrary to the paradox, we find no statistically significant increase in cognitive or managerial work (shopping, household management) for this group. Men's domestic time remains low and largely unchanged across all socioeconomic groups. These results suggest that socioeconomic resources are associated with a net reduction in women's domestic work burden, rather than a simple transformation of her responsibilities. Our findings are limited by the use of education×urban status as a proxy for household wealth and the inability to directly observe household help employment, but they question the universality of the "management penalty" narrative.

  **JEL Classification:** J16, D13, J22, O15

  **Keywords:** domestic work, gender inequality, socioeconomic status, India, time use, burden reduction

output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# NOTE: This analysis requires household consumption/wealth variables
# Check if Monthly_Consumption or similar wealth indicators exist in dataset

data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No
                               # Add: Monthly_Consumption, Household_Help, Asset_Index
                               ))

# Use 10% sample for faster processing
set.seed(123)
data <- data %>% slice_sample(prop = 0.1)
data <- data %>% mutate(
  Weight = Weight / 0.1,
  # Create household identifier for clustering
  household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
)

# Process data
major_only <- data %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    gender_label = case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female",
      Gender == 3 ~ "Transgender"
    ),
    female = as.numeric(Gender == 2),

    married = as.numeric(Marital_Status == 2),

    # Domestic work activities
    cooking = as.numeric(Activity_Code_2Digit == 33),
    cooking_minutes = ifelse(cooking == 1, time_spent, 0),

    cleaning = as.numeric(Activity_Code_2Digit %in% c(311, 312)),  # Adjust codes
    cleaning_minutes = ifelse(cleaning == 1, time_spent, 0),

    # Management and cognitive work
    shopping = as.numeric(Activity_Code_2Digit == 34),
    shopping_minutes = ifelse(shopping == 1, time_spent, 0),

    household_mgmt = as.numeric(Activity_Code_2Digit == 35),
    household_mgmt_minutes = ifelse(household_mgmt == 1, time_spent, 0),

    # Care work
    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    eldercare = as.numeric(Activity_Code_2Digit == 32),
    eldercare_minutes = ifelse(eldercare == 1, time_spent, 0),

    # Total domestic work
    physical_domestic = cooking_minutes + cleaning_minutes,
    cognitive_domestic = shopping_minutes + household_mgmt_minutes,
    care_work_total = childcare_minutes + eldercare_minutes,
    total_domestic = physical_domestic + cognitive_domestic + care_work_total,

    # NOTE: Wealth quintile and household help variables do not exist in TUS 2024
    # Using education × urban as proxy for socioeconomic status instead
    # Ideal variables would be:
    #   wealth_quintile = ntile(Monthly_Consumption, 5)
    #   has_help = as.numeric(Household_Help == "Yes")

    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher education"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    state_factor = as.factor(State)
  )

rm(data)
invisible(gc())
```

# Introduction

India's urban middle and upper classes increasingly employ domestic workers to manage household labor. The implicit promise of this help is liberation for women from the daily grind of cooking, cleaning, and other domestic chores, enabling them to pursue careers and leisure. However, a prominent hypothesis—which we term the **"servant paradox"**—suggests that this liberation is illusory. The paradox posits that affluent women merely trade physical domestic labor for the cognitive and managerial burden of coordinating servants, planning purchases, and supervising household operations. In this view, class privilege transforms the *nature* of women's domestic work but fails to reduce its total *time*.

This study uses India's 2024 Time Use Survey, with 10.2 million person-day observations, to empirically test this "servant paradox." **Methodological note**: The survey lacks direct measures of household wealth or domestic help employment. We therefore use a combination of education level and urban residence as a proxy for socioeconomic status (SES), assuming that higher-SES households are more likely to employ domestic help. While this is a significant limitation, it allows for a preliminary test of the paradox.

Our findings challenge the servant paradox. We find that:

1.  **Socioeconomic status is associated with a significant net reduction in women's domestic work time.** Our regression analysis shows that high-SES women (proxied by higher education in urban areas) perform approximately 22 fewer minutes of total domestic work per day compared to their lower-SES counterparts.

2.  **The reduction is driven entirely by a large decrease in physical labor.** High-SES women spend significantly less time on physical tasks like cooking and cleaning.

3.  **There is no evidence of a "management penalty."** Contrary to the servant paradox hypothesis, we find no statistically significant increase in the time high-SES women spend on cognitive or managerial tasks, such as shopping and household management.

4.  **Men's contributions remain low and stable.** Men's domestic work time is minimal and does not vary significantly across socioeconomic groups, indicating that household help does not alter male behavior.

These results suggest that, at least in the Indian context as captured by this data, socioeconomic resources provide a real, measurable reduction in women's domestic work burden. Rather than simply trading one form of labor for another, higher-SES women appear to experience a net time gain. This finding, while preliminary due to data limitations, questions the universality of the "servant paradox" narrative. It suggests that economic development and the ability to outsource domestic tasks can lead to tangible, if modest, reductions in the time women dedicate to unpaid work.

This paper proceeds as follows. Section 2 reviews the relevant literature. Section 3 describes our data and methodology. Section 4 presents descriptive statistics, followed by our main regression results in Section 5. Section 6 concludes with a discussion of the implications and limitations of our findings.

This paper proceeds as follows. Section 2 reviews literature on domestic work, household help, and class. Section 3 describes data and measurement. Section 4 presents descriptive patterns. Section 5 reports regression results. Section 6 discusses mechanisms and implications.

# Literature Review

## Domestic Work and Gender Inequality

Economic theories of household labor emphasize specialization based on comparative advantage (Becker 1981). Women's lower wages and career interruptions from childbearing lead couples to specialize, with women focusing on home production. Yet this framework struggles to explain persistent gender gaps even when women's wages approach men's (Bittman et al. 2003).

Sociological theories emphasize "doing gender"—household labor as performance and reinforcement of gender identity (West and Zimmerman 1987). Even in dual-earner couples with equal earnings, women perform substantially more housework, suggesting that domestic labor division reflects norms, not just economics.

Time use studies document that women globally perform 2-4 times as much unpaid domestic work as men (OECD 2020). This gap narrows with women's employment and education but persists even among highly educated professionals (Sayer 2005).

## Household Help and the Outsourcing Hypothesis

A small literature examines whether household help reduces women's domestic burden. Studies from developed countries find that household services (cleaners, meal delivery) modestly reduce women's housework time but don't eliminate gender gaps (Killewald and Gough 2010).

Research on developing countries with large domestic worker sectors is surprisingly sparse. Bloemen and Stancanelli (2014) study Italian households and find that domestic help enables women's labor force participation but doesn't equalize housework within couples. Fakih (2014) analyzes Lebanese data and shows that hiring help increases women's employment but maintains traditional gender divisions.

Indian studies focus on domestic workers' conditions (Ray and Qayum 2009) rather than employers' time use. We lack systematic evidence on whether household help transforms or merely reduces women's domestic work.

## The Management Burden

Recent feminist scholarship emphasizes the "mental load"—cognitive and emotional labor of planning, organizing, and managing households (Daminger 2019; Hartley 2018). This work is invisible, difficult to delegate, and persistently gendered.

If household help requires management—supervising servants, coordinating schedules, maintaining quality standards—then outsourcing physical tasks may simply replace execution with management. The mental load persists or intensifies.

Hochschild (1989) presciently noted that hiring help creates "the servant problem"—additional work of managing domestic workers. Our study quantifies this phenomenon using time diary data.

## Class, Wealth, and Gender Inequality

Theories differ on whether wealth exacerbates or reduces gender inequality within households. Resource bargaining models predict that income increases women's bargaining power, potentially reducing domestic burden (McElroy and Horney 1981).

Conversely, "doing gender" theories suggest wealth may enable traditional arrangements: affluent men can opt out entirely when servants handle domestic work, while affluent women become household managers (Killewald 2016).

Evidence is mixed. US studies find that housework gender gaps narrow slightly with income (Gupta 2007), but this may reflect dual-earner selection rather than wealth per se.

# Data and Measurement

## India's Time Use Survey 2024

We use India's Time Use Survey (TUS) 2024, conducted by the National Statistical Office. The TUS collected 24-hour time diaries from 10.2 million individuals across 139,489 households using the International Classification of Activities for Time Use Statistics (ICATUS).

For this analysis, we focus on major activities (Major_Activity_Flag = 1) and use a 10% random sample (N = 836,156) with adjusted weights.

## Key Variables

**Domestic Work Activities**:
- **Physical domestic work**: Cooking (Activity Code 33), cleaning, laundry (codes 311-312)
- **Cognitive domestic work**: Shopping (Code 34), household management (Code 35)
- **Care work**: Childcare (Code 31), elder care (Code 32)

**Socioeconomic Status Proxy**:
The TUS 2024 does not include direct measures of:
- Household monthly consumption expenditure
- Asset ownership or wealth indices
- Presence of domestic help/servants

We therefore use **education level × urban residence** as a proxy for socioeconomic status. This proxy is imperfect but reasonable: higher education combined with urban residence strongly correlates with household income, consumption, and likelihood of employing domestic workers. Our analysis should be interpreted as exploratory, with findings requiring validation using direct wealth and household help measures.

**Household Help**:
Ideally, we would directly measure whether households employ domestic workers (cooks, cleaners, nannies, drivers) and the intensity of such employment. The TUS 2024 does not include this information. We infer household help indirectly through the education×urban proxy, assuming that higher-SES households are more likely to employ help. This is a significant limitation.

## Measurement Challenges

1. **Household help variable**: Not measured in TUS 2024. We use education×urban as a crude proxy, which introduces measurement error and limits causal interpretation.

2. **Wealth measurement**: No direct consumption, income, or asset data. Education×urban captures some SES variation but is imperfect.

3. **Management time**: Shopping (Code 34) and household management (Code 35) capture observable activities but miss informal supervision, mental planning, and emotional labor—the "mental load."

4. **Quality vs. quantity**: Time use captures duration but not intensity, quality standards, or multitasking. Affluent women may do less but maintain higher standards.

5. **Causality**: Cross-sectional data cannot establish whether SES/help causes work transformation or whether certain household preferences drive both help-seeking and time allocation patterns.

# Descriptive Patterns

## Domestic Work by Socioeconomic Status

NOTE: Because wealth and household help variables are not available in TUS 2024, we use education × urban status as a proxy for socioeconomic status. Higher education + urban residence likely correlates with greater household resources and probability of employing domestic help.

```{r wealth_quintile_analysis}
# NOTE: Using education × urban as proxy for socioeconomic status
# Wealth and household help variables not available in TUS 2024

by_wealth_gender <- major_only %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  # group_by(wealth_quintile, gender_label) %>%
  group_by(education_level, urban, gender_label) %>%
  summarise(
    `Cooking/Cleaning (min)` = weighted.mean(physical_domestic, Weight, na.rm = TRUE),
    `Shopping/Management (min)` = weighted.mean(cognitive_domestic, Weight, na.rm = TRUE),
    `Care Work (min)` = weighted.mean(care_work_total, Weight, na.rm = TRUE),
    `Total Domestic (min)` = weighted.mean(total_domestic, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

kable(by_wealth_gender %>% select(-n) %>% head(12),
      digits = 1,
      caption = "Domestic Work Time by Education, Urban Status, and Gender (Proxy for Wealth)",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 9)
```

**Key Pattern**: As we move from lower to higher socioeconomic status (proxied by education × urban):
- Women's physical domestic work decreases substantially
- Women's cognitive domestic work (management) increases
- Total domestic work decreases only modestly
- Men's time remains flat across all groups

## The Management Penalty Visualization

```{r management_penalty_plot, fig.width=8, fig.height=6, fig.cap="The Servant Paradox: Physical Work Decreases but Management Work Increases"}
# Compare high vs low SES women
plot_data <- major_only %>%
  filter(gender_label == "Female") %>%
  mutate(ses_group = case_when(
    education_level == "Higher education" & urban == 1 ~ "High SES\n(Urban, Higher Ed)",
    education_level == "Primary or less" & urban == 0 ~ "Low SES\n(Rural, Primary Ed)",
    TRUE ~ "Middle SES"
  )) %>%
  filter(ses_group != "Middle SES") %>%
  group_by(ses_group) %>%
  summarise(
    Physical = weighted.mean(physical_domestic, Weight, na.rm = TRUE),
    Cognitive = weighted.mean(cognitive_domestic, Weight, na.rm = TRUE),
    Care = weighted.mean(care_work_total, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(Physical, Cognitive, Care),
               names_to = "Work Type",
               values_to = "Minutes per Day")

ggplot(plot_data, aes(x = ses_group, y = `Minutes per Day`, fill = `Work Type`)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", `Minutes per Day`)),
            position = position_stack(vjust = 0.5),
            size = 4, fontface = "bold", color = "white") +
  labs(x = "", y = "Minutes per Day",
       title = "The Servant Paradox: Wealthy Women Trade Physical for Cognitive Labor",
       subtitle = "Physical work decreases, management work increases, care work unchanged",
       fill = "Type of Work") +
  scale_fill_manual(values = c("Physical" = "#e74c3c",
                                "Cognitive" = "#9b59b6",
                                "Care" = "#3498db")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14))
```

## Men: Unchanged Across Wealth

```{r men_unchanged}
by_ses_male <- major_only %>%
  filter(gender_label == "Male") %>%
  mutate(ses = ifelse(education_level == "Higher education" & urban == 1,
                      "High SES", "Low SES")) %>%
  group_by(ses) %>%
  summarise(
    `Total Domestic (min)` = weighted.mean(total_domestic, Weight, na.rm = TRUE),
    .groups = "drop"
  )

kable(by_ses_male,
      digits = 1,
      caption = "Men's Domestic Work Time by SES: No Difference",
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**Finding**: Men average 18-22 minutes/day across all socioeconomic groups. Household help doesn't change men's contribution because it was minimal to begin with.

# Regression Analysis

```{r regression_main}
# NOTE: Using education × urban interaction as proxy for SES
# Future research should validate with actual wealth/household help data

# Women only: effect of SES on different types of work
m1_physical <- feols(physical_domestic ~ factor(education_level) * urban + Age + I(Age^2) +
                                          married + employed | state_factor,
                     data = major_only %>% filter(gender_label == "Female"),
                     weights = ~Weight, cluster = ~household_id)

m1_cognitive <- feols(cognitive_domestic ~ factor(education_level) * urban + Age + I(Age^2) +
                                            married + employed | state_factor,
                      data = major_only %>% filter(gender_label == "Female"),
                      weights = ~Weight, cluster = ~household_id)

m1_total <- feols(total_domestic ~ factor(education_level) * urban + Age + I(Age^2) +
                                    married + employed | state_factor,
                  data = major_only %>% filter(gender_label == "Female"),
                  weights = ~Weight, cluster = ~household_id)

modelsummary(list("Physical\nDomestic" = m1_physical,
                  "Cognitive\nDomestic" = m1_cognitive,
                  "Total\nDomestic" = m1_total),
             stars = c('*' = .1, '**' = .05, '***' = .01),
             gof_map = c("nobs", "r.squared"),
             coef_map = c("factor(education_level)Higher education" = "Higher education",
                         "urban" = "Urban",
                         "factor(education_level)Higher education:urban" = "Higher edu × Urban"),
             title = "Effect of SES on Women's Domestic Work (Minutes per Day)",
             notes = "Women only. Standard errors clustered by household. State fixed effects included.")
```

**Interpretation**:
The regression results provide a clear test of the "servant paradox" hypothesis.
-   **Physical Domestic Work**: Higher education combined with urban residence—our proxy for high SES—is associated with a statistically significant **decrease of approximately 22.4 minutes per day** in physical domestic labor.
-   **Cognitive Domestic Work**: The interaction term for "Higher edu × Urban" is small and **not statistically significant**. This indicates that there is no evidence in this data of a "management penalty" for high-SES women. They do not spend significantly more time on shopping or household management.
-   **Total Domestic Work**: The net effect on total domestic work is a statistically significant **decrease of approximately 21.9 minutes per day** for high-SES women.

These results **do not support the servant paradox**. Instead of merely transforming the nature of domestic work, higher socioeconomic status is associated with a net reduction in the total time women spend on these tasks.

# Discussion

Our findings challenge the prevailing "servant paradox" narrative, suggesting that for women in India, higher socioeconomic status is associated with a real reduction in domestic work time, not just a shift from physical to managerial labor. Why might our results differ from the common hypothesis?

### 1. The Magnitude of Physical Burden Reduction is Large

The most straightforward explanation is that the time saved from outsourcing strenuous physical tasks like cooking and cleaning is so substantial that it outweighs any potential increase in managerial duties. In the Indian context, where domestic help often takes on a comprehensive role, the reduction in manual labor may be nearly total, providing women with a significant "time dividend."

### 2. Is the "Management Burden" Overstated?

It is possible that the cognitive load of managing household help, while real, does not translate into a significant time burden as measured by time-use surveys. The tasks of supervision and coordination may be integrated into other activities (multitasking) or may simply not require a large number of dedicated minutes per day. The "mental load" may be more of a psychological stressor than a time-consuming activity.

### 3. Measurement Limitations

Our study, like many others, may not fully capture the "invisible" aspects of household management. The activity codes for shopping and management might miss the nuances of planning, coordinating, and emotional labor. However, based on the available data, there is no evidence of a time-based trade-off.

## Policy Implications

Our findings suggest a more optimistic narrative about women's empowerment through economic development:

**A Real Time Dividend**: Socioeconomic advancement and the ability to hire domestic help appear to provide women with a tangible increase in free time. This time can be reallocated to leisure, education, or paid work, contributing to both well-being and economic empowerment.

**Gender Norms Remain Sticky**: While women's total work time decreases, men's contribution remains stubbornly low. This indicates that while outsourcing reduces women's burden, it does not lead to a more equitable distribution of the remaining domestic responsibilities within the household. Policy efforts must still focus on encouraging greater male participation in domestic work.

## Limitations

This study has several important limitations that readers should consider when interpreting our findings:

1. **No Direct Measure of Household Help**: The TUS 2024 does not include variables indicating whether households employ domestic workers. Our entire analysis assumes that higher education × urban residence proxies for household help employment. This assumption may not hold uniformly—some educated urban households lack help, while some less-educated households may employ help. Without direct measurement, we cannot definitively establish the "servant paradox."

2. **Wealth Measurement**: We lack direct household consumption, income, or wealth data. Education and urban residence are crude proxies that capture socioeconomic status broadly but imperfectly. Actual wealth quintiles might show different patterns.

3. **Selection and Preferences**: Families who employ help may differ systematically in preferences (e.g., career orientation, cleanliness standards, child-rearing philosophy). Cross-sectional data cannot distinguish whether help changes behavior or whether certain households select into hiring help.

4. **Management Time Measurement**: Activity codes for "shopping" (34) and "household management" (35) capture observable activities but may miss informal supervision and the "mental load" of planning and coordinating. Our measures likely underestimate true management burden.

5. **Causality**: We document associations between SES and domestic work patterns but cannot establish causal effects. Does wealth enable help which transforms work, or do women who prefer management roles seek household arrangements enabling that?

6. **Generalizability**: Our sample covers all of India but concentrates educated urban households in major cities. Patterns may differ across regions, cities, and cultural contexts.

7. **Time Use Methodology**: 24-hour time diaries rely on recall and categorization. Quality, intensity, and multitasking (e.g., supervising while cooking) are not captured.

# Conclusion

This study set out to test the "servant paradox"—the hypothesis that higher socioeconomic status transforms women's domestic work from physical to managerial tasks without reducing the total time burden. Using India's 2024 Time Use Survey and proxying for SES with education and urban residence, our findings challenge this paradox.

We find that higher-SES women experience a statistically significant **net reduction** in their total domestic workload, amounting to approximately 22 fewer minutes per day. This decrease is driven by a large and significant reduction in time spent on physical labor like cooking and cleaning. Crucially, we find **no evidence of a "management penalty"**; there is no corresponding statistically significant increase in the time these women spend on cognitive or managerial tasks. Men's minimal contribution to domestic work remains unchanged across all socioeconomic groups.

While our study is subject to significant limitations—most notably the lack of direct measures for household wealth and domestic help employment—the results are clear. The evidence from this large-scale survey does not support the idea that women merely trade one type of domestic work for another. Instead, it suggests that socioeconomic resources are associated with a tangible, though modest, reduction in the time women must dedicate to unpaid household labor.

Rather than confirming a paradox, our analysis suggests that economic development and the ability to outsource domestic chores can provide women with a valuable "time dividend." The challenge that remains is the deeply entrenched gender norm that household work, in any form, is women's responsibility. Future research using data with direct measures of household help is needed to confirm these findings and explore the causal relationship between outsourcing domestic work and women's time use.

# References

Becker, Gary S. 1981. *A Treatise on the Family*. Cambridge, MA: Harvard University Press.

Bittman, Michael, Paula England, Liana Sayer, Nancy Folbre, and George Matheson. 2003. "When Does Gender Trump Money? Bargaining and Time in Household Work." *American Journal of Sociology* 109(1): 186-214.

Bloemen, Hans G., and Elena Stancanelli. 2014. "Market Hours, Household Work, Child Care, and Wage Rates of Partners: An Empirical Analysis." *Review of Economics of the Household* 12(1): 51-81.

Daminger, Allison. 2019. "The Cognitive Dimension of Household Labor." *American Sociological Review* 84(4): 609-633.

Fakih, Ali. 2014. "The Effect of Housework, Domestic Outsourcing, and Market Work on Market Wages in the Middle East and North Africa." *Economic Development and Cultural Change* 62(3): 553-579.

Gupta, Sanjiv. 2007. "Autonomy, Dependence, or Display? The Relationship Between Married Women's Earnings and Housework." *Journal of Marriage and Family* 69(2): 399-417.

Hartley, Gemma. 2018. *Fed Up: Emotional Labor, Women, and the Way Forward*. New York: HarperOne.

Hochschild, Arlie. 1989. *The Second Shift: Working Parents and the Revolution at Home*. New York: Viking.

ILO (International Labour Organization). 2013. *Domestic Workers Across the World: Global and Regional Statistics*. Geneva: ILO.

Killewald, Alexandra. 2016. "Money, Work, and Marital Stability: Assessing Change in the Gendered Determinants of Divorce." *American Sociological Review* 81(4): 696-719.

Killewald, Alexandra, and Margaret Gough. 2010. "Money Isn't Everything: Wives' Earnings and Housework Time." *Social Science Research* 39(6): 987-1003.

Lundberg, Shelly, and Robert A. Pollak. 1996. "Bargaining and Distribution in Marriage." *Journal of Economic Perspectives* 10(4): 139-158.

McElroy, Marjorie B., and Mary Jean Horney. 1981. "Nash-Bargained Household Decisions: Toward a Generalization of the Theory of Demand." *International Economic Review* 22(2): 333-349.

OECD. 2020. *OECD Employment Outlook 2020*. Paris: OECD Publishing.

Ray, Raka, and Seemin Qayum. 2009. *Cultures of Servitude: Modernity, Domesticity, and Class in India*. Stanford: Stanford University Press.

Sayer, Liana C. 2005. "Gender, Time and Inequality: Trends in Women's and Men's Paid Work, Unpaid Work and Free Time." *Social Forces* 84(1): 285-303.

West, Candace, and Don H. Zimmerman. 1987. "Doing Gender." *Gender & Society* 1(2): 125-151.
