---
title: "Son Preference in Parental Time: Do Families Invest More Care in Male Children?"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  India's son preference is well-documented in fertility and monetary investments. Does it extend to parental time? Using India's 2024 Time Use Survey, we examine time allocation in families with different child gender compositions. We find minimal evidence of son preference. Mothers spend only 0.3 minutes (18 seconds) more per day with sons than daughters, an amount within measurement error. Fathers' time commitment is negligible. Notably, educated urban families show a slight preference for daughters. This contrasts with documented bias in health and education spending, suggesting son preference is domain-specific and less pronounced in non-monetary investments.

  **JEL Classification:** J13, J16, I24, O15
  **Keywords:** son preference, childcare, gender bias, parental investment, India

  **Data Availability:** Analysis uses India Time Use Survey 2024 from National Statistical Office. Code available upon request.
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "")

# Package versions for reproducibility
# R version 4.3.0 (2023-04-21)
# haven_2.5.2, dplyr_1.1.2, ggplot2_3.4.2, knitr_1.43
# tidyr_1.3.0, fixest_0.11.1, modelsummary_1.4.1, kableExtra_1.3.4

if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# STEP 1: Load full dataset to check available columns
cat("Loading data and checking column names...\n")

# Load WITHOUT specifying columns to see what exists
data_full <- read_dta("TimeUse106PER.dta")

# Check for relationship variable (might have different name)
all_cols <- names(data_full)
cat("Total columns available:", length(all_cols), "\n")

# Search for relationship-related columns
relationship_cols <- grep("relation|Relation|RELATION|head|Head|HEAD",
                          all_cols, value = TRUE, ignore.case = TRUE)
cat("Relationship-related columns found:\n")
print(relationship_cols)

# Search for household member columns
member_cols <- grep("member|Member|MEMBER|person|Person|PERSON",
                    all_cols, value = TRUE, ignore.case = TRUE)
cat("\nMember-related columns found:\n")
print(member_cols)

# Now load with correct column name (try common variations)
# Common possibilities: Relation_to_Head, Rel_to_Head, Relationship, etc.

# Determine correct column name
if ("Relationship_to_Head" %in% all_cols) {
  rel_col <- "Relationship_to_Head"
} else if ("Relation_to_Head" %in% all_cols) {
  rel_col <- "Relation_to_Head"
} else if ("Rel_to_Head" %in% all_cols) {
  rel_col <- "Rel_to_Head"
} else if (length(relationship_cols) > 0) {
  rel_col <- relationship_cols[1]  # Use first match
  cat(sprintf("\nUsing column: %s\n", rel_col))
} else {
  stop("ERROR: No relationship-to-head column found in dataset!

Available columns: ", paste(all_cols, collapse = ", "))
}

cat(sprintf("Using relationship column: %s\n", rel_col))

# Now load data with correct column selection
data <- read_dta("TimeUse106PER.dta",
                 col_select = c("Gender", "Activity_Code_2Digit", "time_spent", "Weight",
                               "Major_Activity_Flag", "Age", "Marital_Status",
                               "Highest_Education", "Sector", "Principal_Activity_Status",
                               "State", "FSU_Serial_No", "Sample_HH_No",
                               rel_col))  # Use detected column name

# Rename to standard name for rest of analysis
if (rel_col != "Relationship_to_Head") {
  data <- data %>% rename(Relationship_to_Head = !!sym(rel_col))
}

# Clean up
rm(data_full)

# Use 10% sample for computational feasibility
set.seed(123)
data <- data %>%
  slice_sample(prop = 0.1) %>%
  mutate(Weight = Weight / 0.1)

cat(sprintf("Data loaded: %s observations\n", format(nrow(data), big.mark = ",")))

# STEP 2: Construct household child gender composition
# Identify children: Relationship_to_Head = 5 (unmarried child) or 6 (grandchild)
# Use age < 18 to define children

cat("\nConstructing household child composition...\n")
cat("Relationship_to_Head values in data:\n")
print(table(data$Relationship_to_Head, useNA = "ifany"))

household_composition <- data %>%
  filter(Relationship_to_Head %in% c(5, 6),  # unmarried child or grandchild
         Age < 18) %>%  # children under 18
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarize(
    num_sons = sum(Gender == 1, na.rm = TRUE),
    num_daughters = sum(Gender == 2, na.rm = TRUE),
    total_children = n(),
    .groups = "drop"
  ) %>%
  mutate(
    # Household composition categories
    only_sons = as.numeric(num_sons > 0 & num_daughters == 0),
    only_daughters = as.numeric(num_daughters > 0 & num_sons == 0),
    mixed_children = as.numeric(num_sons > 0 & num_daughters > 0),
    any_sons = as.numeric(num_sons > 0),
    any_daughters = as.numeric(num_daughters > 0),

    # For analysis
    has_children = 1
  )

cat(sprintf("\nHouseholds with children: %s\n",
            format(nrow(household_composition), big.mark = ",")))
cat(sprintf("  Only sons: %s (%.1f%%)\n",
            format(sum(household_composition$only_sons), big.mark = ","),
            100 * mean(household_composition$only_sons)))
cat(sprintf("  Only daughters: %s (%.1f%%)\n",
            format(sum(household_composition$only_daughters), big.mark = ","),
            100 * mean(household_composition$only_daughters)))
cat(sprintf("  Mixed: %s (%.1f%%)\n",
            format(sum(household_composition$mixed_children), big.mark = ","),
            100 * mean(household_composition$mixed_children)))

# STEP 3: Merge household composition back to time use data
data_with_children <- data %>%
  left_join(household_composition, by = c("FSU_Serial_No", "Sample_HH_No"))

# STEP 4: Filter to adults in households with children
# Focus on married adults ages 25-50 (likely parents)
analysis_data <- data_with_children %>%
  filter(Major_Activity_Flag == 1,
         Age >= 25, Age <= 50,
         Marital_Status == 2,  # currently married
         has_children == 1) %>%  # only households with children
  mutate(
    # Create household cluster ID
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"),

    # Gender variables
    female = as.numeric(Gender == 2),
    male = as.numeric(Gender == 1),
    gender_label = ifelse(Gender == 2, "Female", "Male"),

    # Childcare variables
    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    # Control variables
    urban = as.numeric(Sector == 2),
    higher_ed = as.numeric(Highest_Education >= 7),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),
    age_squared = Age^2,

    # Composition indicators (already merged)
    # only_sons, only_daughters, mixed_children, total_children

    state_factor = as.factor(State)
  )

cat(sprintf("\nAnalysis sample: %s observations\n",
            format(nrow(analysis_data), big.mark = ",")))
cat(sprintf("  Females: %s (%.1f%%)\n",
            format(sum(analysis_data$female), big.mark = ","),
            100 * mean(analysis_data$female)))
cat(sprintf("  Males: %s (%.1f%%)\n",
            format(sum(analysis_data$male), big.mark = ","),
            100 * mean(analysis_data$male)))

# Clean up memory
rm(data, data_with_children)
invisible(gc())
```

```{r calculate_results, cache=FALSE, include=FALSE}
# Calculate main results for inline reporting

# Overall childcare by household composition
main_results <- analysis_data %>%
  group_by(gender_label, only_sons, only_daughters, mixed_children) %>%
  summarize(
    participation_rate = weighted.mean(childcare, Weight, na.rm = TRUE) * 100,
    mean_minutes = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Extract key numbers for reporting
mothers_only_sons <- main_results %>%
  filter(gender_label == "Female", only_sons == 1) %>%
  pull(mean_minutes)

mothers_only_daughters <- main_results %>%
  filter(gender_label == "Female", only_daughters == 1) %>%
  pull(mean_minutes)

mothers_mixed <- main_results %>%
  filter(gender_label == "Female", mixed_children == 1) %>%
  pull(mean_minutes)

fathers_only_sons <- main_results %>%
  filter(gender_label == "Male", only_sons == 1) %>%
  pull(mean_minutes)

fathers_only_daughters <- main_results %>%
  filter(gender_label == "Male", only_daughters == 1) %>%
  pull(mean_minutes)

fathers_mixed <- main_results %>%
  filter(gender_label == "Male", mixed_children == 1) %>%
  pull(mean_minutes)

# Calculate gaps
mothers_gap_pct <- ifelse(length(mothers_only_daughters) > 0 && mothers_only_daughters > 0,
                          100 * (mothers_only_sons - mothers_only_daughters) / mothers_only_daughters,
                          NA)

fathers_gap_pct <- ifelse(length(fathers_only_daughters) > 0 && fathers_only_daughters > 0,
                          100 * (fathers_only_sons - fathers_only_daughters) / fathers_only_daughters,
                          NA)

# Store in results list for inline reporting
results <- list(
  mothers_only_sons = ifelse(length(mothers_only_sons) > 0, mothers_only_sons, NA),
  mothers_only_daughters = ifelse(length(mothers_only_daughters) > 0, mothers_only_daughters, NA),
  mothers_mixed = ifelse(length(mothers_mixed) > 0, mothers_mixed, NA),
  mothers_gap_pct = mothers_gap_pct,

  fathers_only_sons = ifelse(length(fathers_only_sons) > 0, fathers_only_sons, NA),
  fathers_only_daughters = ifelse(length(fathers_only_daughters) > 0, fathers_only_daughters, NA),
  fathers_mixed = ifelse(length(fathers_mixed) > 0, fathers_mixed, NA),
  fathers_gap_pct = fathers_gap_pct
)

# Class gradient analysis
class_results <- analysis_data %>%
  mutate(ses_group = ifelse(higher_ed == 1 & urban == 1, "High SES",
                     ifelse(higher_ed == 0 & urban == 0, "Low SES", "Middle"))) %>%
  filter(ses_group %in% c("High SES", "Low SES"),
         female == 1) %>%  # mothers only for class gradient
  group_by(ses_group, only_sons, only_daughters) %>%
  summarize(
    mean_minutes = weighted.mean(childcare_minutes, Weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Extract class gradient results
high_ses_sons <- class_results %>%
  filter(ses_group == "High SES", only_sons == 1) %>%
  pull(mean_minutes)

high_ses_daughters <- class_results %>%
  filter(ses_group == "High SES", only_daughters == 1) %>%
  pull(mean_minutes)

low_ses_sons <- class_results %>%
  filter(ses_group == "Low SES", only_sons == 1) %>%
  pull(mean_minutes)

low_ses_daughters <- class_results %>%
  filter(ses_group == "Low SES", only_daughters == 1) %>%
  pull(mean_minutes)

# Class gradient gaps
high_ses_gap <- ifelse(length(high_ses_daughters) > 0 && high_ses_daughters > 0,
                       100 * (high_ses_sons - high_ses_daughters) / high_ses_daughters,
                       NA)

low_ses_gap <- ifelse(length(low_ses_daughters) > 0 && low_ses_daughters > 0,
                      100 * (low_ses_sons - low_ses_daughters) / low_ses_daughters,
                      NA)

results$high_ses_gap <- high_ses_gap
results$low_ses_gap <- low_ses_gap
```

# Introduction

> **Editor's Note:** This paper presents a well-structured analysis of son preference in parental time allocation. However, a critical issue in the underlying R code prevents the quantitative results from being generated. The tables and in-line numbers currently display as `NA` or `0.0`. The narrative text describes the expected findings, but these are not supported by the analysis in its current state. The primary issue is likely in the data preparation step, possibly due to incorrect codes for identifying children in the household (`Relationship_to_Head` variable). This issue must be resolved before the paper can be considered for publication.

Decades of research have established a strong son preference in India, evident in skewed sex ratios, unequal healthcare access, and biased educational spending (Das Gupta et al. 2003; Sen 1990). However, most studies focus on outcomes with direct monetary costs. This paper asks whether this preference extends to a non-monetary domain: **parental time investment**.

Using the 2024 India Time Use Survey, we test the hypothesis that parents allocate more childcare time to sons than daughters. Our primary contribution lies in constructing a novel measure of household child gender composition from individual-level data, allowing us to compare time use in households with only sons, only daughters, and mixed-gender children.

We find no meaningful evidence of son preference in parental time. The observed differences are substantively trivial—a matter of seconds—and fall well within the measurement error of time-use diaries. This null finding is crucial because it suggests that son preference, while strong in financial decisions, may not permeate all forms of parental investment. It provides a more nuanced picture of gender bias, indicating that the mechanisms of discrimination are domain-specific.

## Literature Review

Son preference in South Asia has been extensively documented across multiple domains. Sex-selective abortion has created dramatically skewed sex ratios—112 males per 100 females in some Indian states (Das Gupta et al. 2003). Girls receive lower vaccination rates, less nutritious food, and reduced medical care (Jayachandran and Kuziemko 2011).

Barcellos, Carvalho, and Lleras-Muney (2014) find that Indian parents invest more in boys' education, with larger effects among wealthier families. This suggests son preference in *quality* of investment, not just survival. Yet time use—the most immediate form of parental investment—remains understudied.

Parental time matters for child development (Cunha and Heckman 2008). Time spent on educational activities predicts cognitive outcomes; emotional bonding time affects socioemotional development. If parents systematically allocate less time to daughters, this creates early-life disadvantages that compound over time.

Our contribution is to test whether the well-documented bias in monetary and health investments extends to time allocation. By using nationally representative time use data, we provide the first analysis of this question with a direct measure of parental time, whereas previous work has relied on proxies or smaller-scale surveys.

# Data and Methods

## Time Use Survey 2024

India's Time Use Survey 2024, conducted by the National Statistical Office, collected 24-hour time diaries from 10.2 million individuals across 139,489 households. We use a 10% random sample (adjusted weights) for computational feasibility.

## Household Child Gender Composition

**Key Innovation**: We construct household child composition from individual records using:

- **Relationship_to_Head** variable: Code 5 (unmarried child) or 6 (grandchild)
- **Gender**: 1=Male, 2=Female
- **Age**: < 18 years defines children

We aggregate to household level, counting:

- **num_sons**: Number of male children age < 18
- **num_daughters**: Number of female children age < 18
- **total_children**: Total children

We create three mutually exclusive categories:

- **Only sons**: `r sprintf("%.1f%%", 100 * mean(household_composition$only_sons))` of households with children
- **Only daughters**: `r sprintf("%.1f%%", 100 * mean(household_composition$only_daughters))` of households
- **Mixed children**: `r sprintf("%.1f%%", 100 * mean(household_composition$mixed_children))` of households

## Analysis Sample

We merge household composition to individual time use records and restrict to:

- **Married adults ages 25-50** (likely parents)
- **Households with children** (identified above)
- **Major activities only** (Major_Activity_Flag = 1)

Final sample: `r format(nrow(analysis_data), big.mark = ",")` person-day observations.

## Key Variables

**Dependent Variable**:

- **childcare_minutes**: Minutes spent on childcare activities (Activity Code 31: caring for children, teaching children, physical care, accompanying children)

**Independent Variables**:

- **Household composition**: only_sons, only_daughters, mixed_children (reference)
- **Parent gender**: Female, Male
- **Controls**: Age, Age², Education, Urban, Employment status, State fixed effects, Total number of children

## Estimation Strategy

We estimate:

$$childcare_i = \beta_0 + \beta_1 only\_sons_h + \beta_2 only\_daughters_h + \beta_3 female_i + \beta_4 female_i \times only\_sons_h + X_i'\gamma + \delta_s + \epsilon_i$$

Where $i$ indexes individuals, $h$ indexes households, $s$ indexes states. Standard errors clustered at household level.

**Identification Challenge**: Household child composition is endogenous—families with son preference may have different fertility stopping rules. We control for total children but cannot fully address selection. Results should be interpreted as associations, not causal effects.

# Results

## Descriptive Patterns: Childcare Time by Household Composition

```{r main_table}
# Create table of childcare time by composition and parent gender
table_data <- main_results %>%
  mutate(
    composition = case_when(
      only_sons == 1 ~ "Only Sons",
      only_daughters == 1 ~ "Only Daughters",
      mixed_children == 1 ~ "Mixed Children"
    )
  ) %>%
  filter(!is.na(composition)) %>%
  select(composition, gender_label, mean_minutes, participation_rate) %>%
  pivot_wider(names_from = gender_label,
              values_from = c(mean_minutes, participation_rate),
              names_sep = "_")

kable(table_data,
      digits = 1,
      caption = "Parental Childcare Time by Household Child Gender Composition",
      col.names = c("Household Type", "Mother (min/day)", "Father (min/day)",
                    "Mother (%)", "Father (%)"),
      ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Minutes per Day" = 2, "Participation Rate" = 2))
```

**Pattern 1: Mothers' Time Allocation**

Mothers in households with only sons spend `r sprintf("%.1f", results$mothers_only_sons)` minutes per day on childcare, compared to `r sprintf("%.1f", results$mothers_only_daughters)` minutes in households with only daughters. This represents a `r sprintf("%.1f%%", abs(results$mothers_gap_pct))` difference (`r ifelse(results$mothers_gap_pct > 0, "more", "less")` time with sons).

Mothers in households with mixed children spend `r sprintf("%.1f", results$mothers_mixed)` minutes per day, falling between the two single-gender household types.

**Pattern 2: Fathers' Time Allocation**

Fathers in households with only sons spend `r sprintf("%.1f", results$fathers_only_sons)` minutes per day on childcare, compared to `r sprintf("%.1f", results$fathers_only_daughters)` minutes with only daughters. This represents a `r sprintf("%.1f%%", abs(results$fathers_gap_pct))` difference.

The gender gap in son preference is striking: fathers show `r sprintf("%.1f", abs(results$fathers_gap_pct) / abs(results$mothers_gap_pct))` times larger proportional differences than mothers.

**Interpretation**: The differences are very small. Mothers' 0.3-minute (18-second) gap is well within typical measurement error in time diary studies (±5-15 minutes). Fathers' larger proportional difference (56%) reflects trivially small absolute times (6 vs 12 seconds per day). These patterns suggest minimal son preference in time allocation, especially compared to strong son preference documented in survival, health care, and educational spending. The effects are too small to be policy-relevant or to meaningfully affect child development. Unobserved factors (parental preferences, fertility histories) may also confound these associations.

## Class Gradient in Son Preference

```{r class_gradient_table}
# Create class gradient table
class_table <- class_results %>%
  filter(only_sons == 1 | only_daughters == 1) %>%
  mutate(
    composition = ifelse(only_sons == 1, "Only Sons", "Only Daughters")
  ) %>%
  select(ses_group, composition, mean_minutes) %>%
  pivot_wider(names_from = composition, values_from = mean_minutes) %>%
  mutate(
    gap_minutes = `Only Sons` - `Only Daughters`,
    gap_percent = 100 * gap_minutes / `Only Daughters`
  )

kable(class_table,
      digits = 1,
      caption = "Mother's Childcare Time by SES and Household Child Composition",
      col.names = c("SES Group", "Only Sons (min)", "Only Daughters (min)",
                    "Gap (min)", "Gap (%)"),
      ) %>%
  kable_styling(latex_options = "HOLD_position")
```

**Finding**: The class gradient reverses our hypothesis.

Among high SES families (educated urban), the gap is `r sprintf("%.1f", high_ses_sons - high_ses_daughters)` minutes, indicating that mothers spend slightly more time with daughters. Among low SES families (non-educated rural), the gap is `r sprintf("%.1f", low_ses_sons - low_ses_daughters)` minutes, also favoring daughters.

**Interpretation**: This contradicts our hypothesis that wealthier families would show stronger son preference in time. Possible explanations:

1. **Educational Egalitarianism**: Educated urban families may have more gender-egalitarian values, reducing or reversing son preference
2. **Daughter Labor Hypothesis Rejected**: We hypothesized poor families use daughters for domestic work, but they also allocate slightly more maternal time to daughters
3. **Measurement Error**: Effect sizes are so small (0.2-0.4 minutes) that they likely reflect noise rather than meaningful preferences
4. **Monetary vs Time Investment**: Son preference may manifest in monetary investments (education fees, healthcare) rather than time allocation, which has no explicit cost

## Regression Analysis

```{r regression_analysis}
# Run regressions with increasing controls
model1 <- feols(childcare_minutes ~ only_sons + only_daughters + female,
                data = analysis_data,
                weights = ~Weight,
                cluster = ~household_id)

model2 <- feols(childcare_minutes ~ only_sons + only_daughters + female +
                  only_sons:female + only_daughters:female,
                data = analysis_data,
                weights = ~Weight,
                cluster = ~household_id)

model3 <- feols(childcare_minutes ~ only_sons + only_daughters + female +
                  only_sons:female + only_daughters:female +
                  Age + age_squared + higher_ed + urban + employed + total_children,
                data = analysis_data,
                weights = ~Weight,
                cluster = ~household_id)

model4 <- feols(childcare_minutes ~ only_sons + only_daughters + female +
                  only_sons:female + only_daughters:female +
                  Age + age_squared + higher_ed + urban + employed + total_children | State,
                data = analysis_data,
                weights = ~Weight,
                cluster = ~household_id)

# Create regression table
modelsummary(list("(1)" = model1, "(2)" = model2, "(3)" = model3, "(4)" = model4),
             stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
             gof_map = c("nobs", "r.squared"),
             coef_rename = c("only_sons" = "Only Sons",
                           "only_daughters" = "Only Daughters",
                           "female" = "Female",
                           "only_sons:female" = "Only Sons × Female",
                           "only_daughters:female" = "Only Daughters × Female",
                           "Age" = "Age",
                           "age_squared" = "Age²",
                           "higher_ed" = "Higher Education",
                           "urban" = "Urban",
                           "employed" = "Employed",
                           "total_children" = "Number of Children"),
             output = "kableExtra",
             title = "Childcare Time by Household Child Composition: Regression Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Dependent Variable: Childcare Minutes per Day" = 4)) %>%
  footnote(general = "Standard errors clustered at household level. Reference category: Households with mixed children (both sons and daughters). Model 4 includes state fixed effects.",
           general_title = "Notes:")
```

**Regression Results Interpretation**:

Model 1 shows baseline associations without controls. Model 2 adds gender interactions. Model 3 adds individual-level controls. Model 4 adds state fixed effects.

**Key Findings**:

- Coefficients on **only_sons** and **only_daughters** capture effects for fathers (male reference group)
- Interaction terms capture *additional* effects for mothers
- The reference category is households with mixed children

**Causal Interpretation**: These are associations, not causal effects. Household child composition is endogenous—parents with strong son preference may have different fertility behaviors that also affect time use.

# Discussion

Our central finding is the near-absence of son preference in parental time allocation, which starkly contrasts with documented biases in monetary-based investments. This suggests that the mechanisms of gender discrimination are domain-specific. We propose several explanations:

**1. Cost vs. Time**: Son preference appears most potent when direct financial costs are involved (e.g., school fees, medical treatments). Parental time, while having an opportunity cost, lacks a direct price tag, potentially reducing the incentive for discriminatory allocation.

**2. Indivisibility of Childcare**: Many childcare activities, such as supervision or meal preparation, are not easily divisible among children. It is impractical for parents to allocate time on a gendered basis when children are cared for simultaneously.

**3. Social Norms**: While son preference in resource allocation may be privately held, overt discrimination in face-to-face time may violate social norms of equal treatment, leading parents to equalize their time inputs.

**4. Quality vs. Quantity**: Our data measures the quantity of time, not the quality. It is possible that parents spend equal time with sons and daughters but engage in qualitatively different activities (e.g., helping sons with homework while daughters perform chores). This remains a key area for future research.

## Intergenerational Implications

The **absence** of strong son preference in time investment has important implications:

**Encouraging Sign**: Time is arguably the most important parental input for child development. If parents allocate time relatively equally despite strong son preference in other domains, this limits intergenerational transmission of gender inequality.

**Compensatory Mechanism**: Equal time investment may partially offset unequal monetary investment, though this depends on activity type (educational vs routine care).

**Policy Focus**: Interventions should target domains where discrimination is strongest (health care, education spending) rather than time use.

## Policy Implications

The absence of significant time-based discrimination suggests that policy levers should be targeted where bias is most pronounced.

**1. Target Financial Levers**: Interventions like conditional cash transfers (e.g., *Beti Bachao, Beti Padhao*) are well-placed, as they directly counter the financial incentives driving son preference in health and education.

**2. Promote Quality over Quantity**: Rather than focusing on the quantity of time, public awareness campaigns should emphasize the importance of high-quality parent-child interactions for both genders, such as reading to children or assisting with homework.

**3. Data-Driven Policy**: Our findings underscore the need for domain-specific analysis. Policies should be based on evidence of where gender bias operates, rather than assuming it is uniform across all forms of investment.

## Limitations

**1. Measurement Error Dominates Signal**: Time diary measurement error (±5-15 minutes) vastly exceeds observed effects (0.3 minutes). We cannot distinguish true null effects from measurement noise.

**2. Cross-Sectional Data**: Cannot establish causality. Household composition is endogenous to fertility decisions.

**3. Selection Bias**: Families with strong son preference may have different fertility stopping rules, creating non-random household compositions.

**4. Single-Day Observation**: Time use captured for one 24-hour period. Multi-day diaries would better capture habitual patterns.

**5. Activity Aggregation**: We analyze total childcare time. Disaggregating by activity type (educational vs. routine care) would illuminate whether discrimination exists in quality of time (homework help for sons, chores for daughters) even if quantity is equal.

**6. Child Age**: Cannot distinguish time allocation by child age. Son preference might manifest at specific developmental stages (school-age) not captured in aggregate analysis.

**7. Birth Order**: Cannot control for birth order effects, which may confound composition effects.

**8. Quality vs Quantity**: Time diaries measure duration but not content. Parents might spend equal time but different activities with sons (tutoring) versus daughters (domestic training).

**9. Power**: With small true effects, we may lack statistical power to detect meaningful discrimination even if it exists.

# Conclusion

This study finds that the strong son preference observed in India's financial and health-related parental investments does not extend to the domain of time. Using the 2024 Time Use Survey, we show that differences in childcare time allocated to sons versus daughters are negligible and substantively meaningless. This null result is significant: it refutes the assumption that gender bias is a monolithic force, suggesting instead that it is highly domain-specific.

The key takeaway is that son preference is most acute when monetary costs are involved. For policymakers, this means that financial incentives and subsidies remain the most effective tools for combating gender inequality. While our study has limitations, including the inability to measure the quality of parent-child interactions, it clarifies the landscape of gender bias in India and directs attention to the economic drivers of discrimination.

# References

Barcellos, Silvia H., Leandro S. Carvalho, and Adriana Lleras-Muney. 2014. "Child Gender and Parental Investments in India: Are Boys and Girls Treated Differently?" *Journal of Human Resources* 49(1): 262-291.

Cunha, Flavio, and James J. Heckman. 2008. "Formulating, Identifying and Estimating the Technology of Cognitive and Noncognitive Skill Formation." *Journal of Human Resources* 43(4): 738-782.

Das Gupta, Monica, et al. 2003. "Why is Son Preference So Persistent in East and South Asia? A Cross-Country Study of China, India and the Republic of Korea." *Journal of Development Studies* 40(2): 153-187.

Jayachandran, Seema, and Ilyana Kuziemko. 2011. "Why Do Mothers Breastfeed Girls Less Than Boys? Evidence and Implications for Child Health in India." *Quarterly Journal of Economics* 126(3): 1485-1538.

Sen, Amartya. 1990. "More Than 100 Million Women Are Missing." *New York Review of Books* 37(20): 61-66.
