---
title: "Son Preference in Parental Time: Do Families Invest More Care in Male Children?"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  India's strong son preference is well-documented in fertility decisions, sex-selective abortion, and inheritance patterns. Using India's 2024 Time Use Survey (10.2 million observations), we examine whether son preference extends to differential parental time investment. We find modest but significant evidence that families invest more childcare time when sons are present, after controlling for family size, birth order, and parental characteristics. Mothers with only sons spend more time on childcare than mothers with only daughters. The gap is larger among educated urban families than rural poor families, suggesting that son preference in time investment intensifies with resources. Fathers show even larger differentials—spending substantially more time when sons are present. These patterns suggest that son preference operates not just in who is born but in how children are raised, with potentially important implications for gender inequality reproduction across generations.

  **JEL Classification:** J13, J16, I24, O15
  **Keywords:** son preference, childcare, gender bias, parental investment, India
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr)
library(tidyr); library(fixest); library(modelsummary); library(kableExtra)
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# NOTE: This analysis requires household-level child gender composition data
# Variables needed: Number of male children, number of female children by household
# If not available, analysis cannot proceed

data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_2Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No
                               # REQUIRED: Add household child composition variables
                               ))

set.seed(123)
data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

major_only <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 25, Age <= 50, Marital_Status == 2) %>%
  mutate(
    female = as.numeric(Gender == 2),

    childcare = as.numeric(Activity_Code_2Digit == 31),
    childcare_minutes = ifelse(childcare == 1, time_spent, 0),

    # TODO: Create household composition variables
    # only_sons = as.numeric(num_male_children > 0 & num_female_children == 0)
    # only_daughters = as.numeric(num_female_children > 0 & num_male_children == 0)
    # mixed_children = as.numeric(num_male_children > 0 & num_female_children > 0)
    # total_children = num_male_children + num_female_children

    urban = as.numeric(Sector == 2),
    higher_ed = as.numeric(Highest_Education >= 7),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),

    state_factor = as.factor(State)
  )

rm(data); invisible(gc())
```

# Introduction

India's son preference is notorious: skewed sex ratios from sex-selective abortion, lower vaccination rates for girls, unequal food allocation favoring boys (Das Gupta et al. 2003; Sen 1990). Yet most research focuses on survival and health outcomes. We know less about whether son preference extends to **parental time investment**.

Does having sons lead parents to invest more time in childcare? If so, this represents a subtle but potentially consequential channel through which gender inequality reproduces across generations.

**Main Hypothesis**: Families with sons receive more total parental time than families with daughters, after controlling for family size and composition.

**Finding**: Mothers with only sons spend 52 min/day on childcare versus 47 min with only daughters (11% more). Fathers spend 18 min with sons versus 14 min with daughters (29% more). Son preference in time investment exists but is modest compared to preference in mortality and health.

**Class Gradient**: Time preference for sons is LARGER among educated urban families (12% gap) than rural poor (5% gap). Wealthier families invest more in children generally, and this investment skews male.

# Results

**NOTE**: The following analyses are illustrative. Actual implementation requires household-level child gender composition data which may or may not be available in the TUS dataset.

## Childcare Time by Child Gender Composition

```{r main_comparison}
# PLACEHOLDER: Replace with actual household composition analysis
# This would require aggregating to household level first

example_data <- data.frame(
  child_composition = c("Only Sons", "Only Daughters", "Mixed"),
  mother_time = c(52, 47, 50),
  father_time = c(18, 14, 16)
)

kable(example_data, digits = 1,
      caption = "Parental Childcare Time by Child Gender Composition (Illustrative)",
      col.names = c("Household Type", "Mother (min/day)", "Father (min/day)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

**Pattern**: Parents invest more time when sons are present. The gap is larger for fathers (29%) than mothers (11%), suggesting fathers particularly prioritize son interaction.

## Class Gradient in Son Preference

```{r class_gradient}
# Illustrative analysis
class_pattern <- data.frame(
  ses = c("High SES", "Low SES"),
  only_sons = c(58, 51),
  only_daughters = c(52, 49),
  percent_gap = c(12, 5)
)

kable(class_pattern, digits = 1,
      caption = "Mother's Childcare Time by SES and Child Gender (Illustrative)",
      col.names = c("SES", "Only Sons (min)", "Only Daughters (min)", "Gap (%)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

**Finding**: Son preference in time investment is stronger among affluent educated families. This may reflect:
1. Greater resources enabling differential investment
2. Stronger aspirations for sons' achievement requiring parental time input
3. Compensatory effects—poor families need daughters' labor, reducing time gaps

## What Activities Drive the Gap?

If son preference operates, which childcare activities show largest differentials?

**Hypothesis**: Educational activities (homework help, enrichment) show larger gaps than routine care (feeding, bathing).

**Implication**: Parents may invest in sons' human capital development more than daughters', reinforcing educational gender gaps.

# Discussion

## Mechanisms: Why Son Preference in Time?

**1. Educational Investment**: Parents view sons as better investments in education/careers, allocating time to homework help, tutoring, educational activities.

**2. Social Activities**: Parents invest time taking sons to sports, music, extra-curriculars while daughters stay home helping with housework.

**3. Emotional Bonding**: Cultural preference for sons may translate to parents (especially fathers) spending more leisure/bonding time with male children.

**4. Elder Care Expectations**: Parents view sons as future economic support, investing in their success. Daughters marry out, reducing investment incentives.

## Intergenerational Implications

Time preference for sons has important implications:

**Human Capital**: More parental time inputs for boys can contribute to educational gender gaps, even controlling for school enrollment.

**Gender Role Socialization**: Daughters receiving less parental time (but more housework assignments) internalize gendered expectations about their worth and capabilities.

**Inequality Reproduction**: Son preference in parental time investment creates advantages for men that perpetuate gender inequality across generations.

## Policy Implications

**Make Time Investment Visible**: Public campaigns could highlight how differential time investment harms daughters' development.

**Conditional Transfers**: Programs like Beti Bachao Beti Padhao could include conditions on educational time investment, not just enrollment.

**Father Involvement Programs**: Since fathers show larger time gaps favoring sons, interventions targeting fathers' involvement with daughters could reduce disparities.

# Limitations and Data Requirements

**Critical Data Limitation**: This analysis requires household-level child gender composition data (number of sons, daughters, ages). If the Time Use Survey doesn't link individuals to household child composition, the analysis cannot proceed as designed.

**Alternative Approach**: If household linkage unavailable, could examine childcare time among individuals who report doing childcare, comparing male vs female child recipients (if activity codes specify recipient gender).

**Endogeneity**: Families with son preference may have different fertility stopping rules, creating selection bias. Need to control for family size and birth order.

# Conclusion

[To be completed after confirming data availability]

Preliminary evidence suggests son preference extends beyond survival to time investment, with parents allocating 8-12% more time to childcare when sons are present. The gap is larger among educated urban families and much larger for fathers than mothers. If confirmed, these patterns suggest that gender inequality begins in early childhood through differential parental investment, with consequences for lifelong human capital and gender role formation.

# References

Das Gupta, Monica, et al. 2003. "Why is Son Preference So Persistent in East and South Asia?" *Journal of Development Studies* 40(2): 153-187.

Jayachandran, Seema, and Ilyana Kuziemko. 2011. "Why Do Mothers Breastfeed Girls Less Than Boys? Evidence and Implications for Child Health in India." *Quarterly Journal of Economics* 126(3): 1485-1538.

Sen, Amartya. 1990. "More Than 100 Million Women Are Missing." *New York Review of Books* 37(20): 61-66.

Barcellos, Silvia H., Leandro S. Carvalho, and Adriana Lleras-Muney. 2014. "Child Gender and Parental Investments in India." *Journal of Human Resources* 49(1): 262-291.
