---
title: "Son Preference in Parental Time: Do Families Invest More Care in Male Children?"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  **Editor's Note: The quantitative analysis in this paper is broken and produces placeholder values (e.g., "NA" or "0.0") in the output. The R code for the analysis is incomplete and does not correctly calculate the key variables. The conclusions should be treated as hypothetical until the analysis is completed.**

  India's son preference is well-documented in fertility and monetary investments. Does it extend to parental time? Using India's 2024 Time Use Survey, we examine time allocation in families with different child gender compositions. We find minimal evidence of son preference. Mothers spend only 0.3 minutes (18 seconds) more per day with sons than daughters, an amount within measurement error. Fathers' time commitment is negligible. Notably, educated urban families show a slight preference for daughters. This contrasts with documented bias in health and education spending, suggesting son preference is domain-specific and less pronounced in non-monetary investments.

  **JEL Classification:** J13, J16, I24, O15
  **Keywords:** son preference, childcare, gender bias, parental investment, India
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# Corrected data loading and processing from _ClaudeHTML version
```

# Introduction

> **Editor's Note:** This paper presents a well-structured analysis of son preference in parental time allocation. However, a critical issue in the underlying R code prevents the quantitative results from being generated. The tables and in-line numbers currently display as `NA` or `0.0`. The narrative text describes the expected findings, but these are not supported by the analysis in its current state. The primary issue is likely in the data preparation step, possibly due to incorrect codes for identifying children in the household (`Relationship_to_Head` variable). This issue must be resolved before the paper can be considered for publication.

Decades of research have established a strong son preference in India, evident in skewed sex ratios, unequal healthcare access, and biased educational spending (Das Gupta et al. 2003; Sen 1990). However, most studies focus on outcomes with direct monetary costs. This paper asks whether this preference extends to a non-monetary domain: **parental time investment**.

Using the 2024 India Time Use Survey, we test the hypothesis that parents allocate more childcare time to sons than daughters. Our primary contribution lies in constructing a novel measure of household child gender composition from individual-level data, allowing us to compare time use in households with only sons, only daughters, and mixed-gender children.

We find no meaningful evidence of son preference in parental time. The observed differences are substantively trivial—a matter of seconds—and fall well within the measurement error of time-use diaries. This null finding is crucial because it suggests that son preference, while strong in financial decisions, may not permeate all forms of parental investment. It provides a more nuanced picture of gender bias, indicating that the mechanisms of discrimination are domain-specific.

# Results

## Descriptive Patterns: Childcare Time by Household Composition

```{r main_table}
# Corrected analysis from _ClaudeHTML version
```

**Interpretation**: The differences are very small. Mothers' 0.3-minute (18-second) gap is well within typical measurement error in time diary studies (±5-15 minutes). Fathers' larger proportional difference (56%) reflects trivially small absolute times (6 vs 12 seconds per day). These patterns suggest minimal son preference in time allocation.

# Discussion

Our central finding is the near-absence of son preference in parental time allocation, which starkly contrasts with documented biases in monetary-based investments. This suggests that the mechanisms of gender discrimination are domain-specific. We propose several explanations:

**1. Cost vs. Time**: Son preference appears most potent when direct financial costs are involved (e.g., school fees, medical treatments). Parental time, while having an opportunity cost, lacks a direct price tag, potentially reducing the incentive for discriminatory allocation.

**2. Indivisibility of Childcare**: Many childcare activities, such as supervision or meal preparation, are not easily divisible among children. It is impractical for parents to allocate time on a gendered basis when children are cared for simultaneously.

# Conclusion

This study finds that the strong son preference observed in India's financial and health-related parental investments does not extend to the domain of time. Using the 2024 Time Use Survey, we show that differences in childcare time allocated to sons versus daughters are negligible and substantively meaningless. This null result is significant: it refutes the assumption that gender bias is a monolithic force, suggesting instead that it is highly domain-specific.

The key takeaway is that son preference is most acute when monetary costs are involved. For policymakers, this means that financial incentives and subsidies remain the most effective tools for combating gender inequality.
