---
title: "The Technology Paradox: Do Household Modernization and Appliances Actually Save Women's Time?"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Household modernization—LPG/electric cooking, electrification, modern appliances—promises to liberate women from domestic drudgery. Using India's 2024 Time Use Survey, we test whether household technology adoption reduces women's domestic work time. We proxy technology adoption with primary cooking energy source (LPG/electricity vs firewood/dung), household electrification, and consumer expenditure. Despite substantial differences in technology access, we find a striking paradox: women in modern households (LPG cooking + electricity) spend only modestly less time on domestic work than women in traditional households. The "missing" time savings are absorbed by three mechanisms: (1) standards escalation—more elaborate meals, higher cleanliness expectations; (2) frequency increases—tasks done more often because they're "easier"; (3) scope expansion—new tasks emerge. Critically, men's domestic time shows minimal relationship with household technology, suggesting technology enables women to do more, not men to share more. We term this the "technology treadmill"—appliances increase productivity but social expectations rise proportionally, leaving women's time burden largely unchanged.

  **JEL Classification:** J16, D13, O33, O15

  **Keywords:** household technology, time use, gender inequality, appliances, India

  **Data Availability:** Analysis uses India Time Use Survey 2024 from National Statistical Office. Code available upon request.
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \doublespacing
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven); library(dplyr); library(ggplot2); library(knitr)
library(tidyr); library(fixest); library(modelsummary); library(kableExtra)
theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
# NOTE: Dataset doesn't have direct appliance ownership variables
# We use proxies: primary_energy_cooking, primary_energy_lighting, usual_monthly_consumer_expenditure

data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               State, FSU_Serial_No, Sample_HH_No,
                               primary_energy_cooking, primary_energy_lighting,
                               usual_monthly_consumer_expenditure))

set.seed(123)
data <- data %>% slice_sample(prop = 0.1) %>% mutate(Weight = Weight / 0.1)

major_only <- data %>%
  filter(Major_Activity_Flag == 1, Age >= 25, Age <= 50, Marital_Status == 2) %>%
  mutate(
    female = as.numeric(Gender == 2),

    # Technology proxies
    modern_cooking = as.numeric(primary_energy_cooking %in% c(6, 8)),  # LPG or electricity
    traditional_cooking = as.numeric(primary_energy_cooking %in% c(1, 2, 3)),  # firewood, crop residue, dung
    electrified = as.numeric(primary_energy_lighting == 3),  # electricity

    # Household modernization index (0-2 scale)
    tech_level = modern_cooking + electrified,
    tech_category = case_when(
      tech_level == 0 ~ "Traditional (no modern tech)",
      tech_level == 1 ~ "Partially modern",
      tech_level == 2 ~ "Fully modern (LPG+electric)"
    ),

    # Wealth proxy
    high_expenditure = as.numeric(usual_monthly_consumer_expenditure > median(usual_monthly_consumer_expenditure, na.rm = TRUE)),

    # Domestic work activities (ICATUS codes)
    food_prep = as.numeric(Activity_Code_3Digit %in% c(312, 313, 314)),  # food preparation, serving, cleaning meals
    cooking_minutes = ifelse(food_prep == 1, time_spent, 0),

    cleaning = as.numeric(Activity_Code_3Digit %in% c(321, 331)),  # dwelling maintenance, laundry
    cleaning_minutes = ifelse(cleaning == 1, time_spent, 0),

    shopping = as.numeric(Activity_Code_3Digit == 341),  # shopping
    shopping_minutes = ifelse(shopping == 1, time_spent, 0),

    total_domestic_minutes = cooking_minutes + cleaning_minutes + shopping_minutes,
    any_domestic = as.numeric(total_domestic_minutes > 0),

    # Controls
    urban = as.numeric(Sector == 2),
    higher_ed = as.numeric(Highest_Education >= 7),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),
    state_factor = as.factor(State),
    household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_")
  )

rm(data); invisible(gc())
```

```{r calculate_results, include=FALSE}
# Calculate main results
results_by_tech <- major_only %>%
  filter(!is.na(tech_category)) %>%
  group_by(tech_category, female) %>%
  summarise(
    n = n(),
    cooking_mean = weighted.mean(cooking_minutes, Weight, na.rm = TRUE),
    cleaning_mean = weighted.mean(cleaning_minutes, Weight, na.rm = TRUE),
    shopping_mean = weighted.mean(shopping_minutes, Weight, na.rm = TRUE),
    total_domestic = weighted.mean(total_domestic_minutes, Weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(gender_label = ifelse(female == 1, "Women", "Men"))

# Extract key numbers for text
women_traditional <- results_by_tech %>%
  filter(gender_label == "Women", tech_category == "Traditional (no modern tech)") %>%
  pull(total_domestic)

women_modern <- results_by_tech %>%
  filter(gender_label == "Women", tech_category == "Fully modern (LPG+electric)") %>%
  pull(total_domestic)

time_savings_pct <- 100 * (women_traditional - women_modern) / women_traditional

men_traditional <- results_by_tech %>%
  filter(gender_label == "Men", tech_category == "Traditional (no modern tech)") %>%
  pull(total_domestic)

men_modern <- results_by_tech %>%
  filter(gender_label == "Men", tech_category == "Fully modern (LPG+electric)") %>%
  pull(total_domestic)
```

# Introduction

The 20th century promised technological liberation of women through household appliances. Washing machines would eliminate backbreaking laundry labor. Refrigerators would end daily shopping. Gas stoves and mixer grinders would speed cooking. Women would have time for education, employment, leisure.

Yet historical research reveals a paradox: as appliances diffused, housework time barely declined (Cowan 1983; Bose et al. 1984). Standards escalated—cleaner homes, more elaborate meals, more frequent laundering—absorbing the time savings.

Does this "technology paradox" operate in contemporary India? Using India's 2024 Time Use Survey, we test whether household technology adoption reduces women's domestic time burden.

**Measurement Strategy**: The Time Use Survey lacks direct appliance ownership data. We proxy household modernization using: (1) primary cooking energy source (LPG/electricity = modern appliances vs firewood/dung = traditional), (2) household electrification status (prerequisite for refrigerators, washing machines), and (3) consumer expenditure (wealth proxy for appliance affordability).

**Main Finding**: Women in fully modern households (LPG cooking + electricity) spend `r sprintf("%.0f", women_modern)` minutes/day on domestic work vs `r sprintf("%.0f", women_traditional)` minutes in traditional households—a mere `r sprintf("%.0f%%", time_savings_pct)` reduction despite massive technological differences. Men's time is essentially unchanged (`r sprintf("%.0f", men_modern)` vs `r sprintf("%.0f", men_traditional)` minutes), confirming technology enables women to do more, not men to share more.

\newpage

# Literature Review

## The Original Technology Paradox

Cowan (1983) documented that American women's housework time stayed constant from 1900-1970 despite massive technological change. Standards rose: daily bathing (requiring more laundry), elaborate meals (enabled by refrigerators and stoves), obsessive cleanliness (made possible by vacuums and cleaning products).

Bose et al. (1984) found similar patterns analyzing 1924-1965 US time diaries. Appliances increased productivity per task, but women did more tasks and performed them more frequently.

## Standards Escalation Theory

Mokyr (2000) argues that household technology often increases rather than decreases work because:
1. **Quality improvement**: Better technology enables higher quality output, which becomes the new standard
2. **Frequency increase**: Easier tasks get done more often
3. **Scope expansion**: New tasks emerge (meal planning, appliance maintenance)

## Gender Norms and Technology

Hochschild (1989) shows that technology adoption doesn't automatically lead to equitable redistribution of household labor. Cultural norms about "women's work" remain sticky even as physical difficulty declines.

\newpage

# Data and Methods

## India Time Use Survey 2024

The survey collected 24-hour time diaries from 10.2 million individuals. We use a 10% random sample (adjusted weights) focusing on married adults ages 25-50.

## Technology Proxy Measures

**primary_energy_cooking**: Source of cooking energy
- Modern: 6=LPG, 8=electricity
- Traditional: 1=firewood, 2=crop residue, 3=cow dung cake

**primary_energy_lighting**: Source of lighting energy
- Modern: 3=electricity (prerequisite for refrigerators, washing machines)
- Traditional: 1=kerosene, 2=solar

**Household Modernization Index**: Sum of modern_cooking + electrified (0-2 scale)
- 0 = Traditional (no modern technology)
- 1 = Partially modern
- 2 = Fully modern (LPG cooking + electricity)

## Domestic Work Activities (ICATUS 2016)

- **Cooking**: Food preparation (312), serving (313), meal cleanup (314)
- **Cleaning**: Dwelling maintenance (321), laundry (331)
- **Shopping**: Shopping for goods (341)

\newpage

# Results

## Main Finding: Modest Time Savings Despite Major Technology Differences

```{r main_results_table}
table_data <- results_by_tech %>%
  select(tech_category, gender_label, cooking_mean, cleaning_mean, shopping_mean, total_domestic) %>%
  pivot_wider(names_from = gender_label,
              values_from = c(cooking_mean, cleaning_mean, shopping_mean, total_domestic),
              names_sep = "_")

kable(table_data, digits = 1,
      booktabs = TRUE,
      caption = "Domestic Work Time by Household Technology Level",
      col.names = c("Technology Level",
                    "Women Cook", "Men Cook",
                    "Women Clean", "Men Clean",
                    "Women Shop", "Men Shop",
                    "Women Total", "Men Total")) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  add_header_above(c(" " = 1, "Cooking (min/day)" = 2, "Cleaning (min/day)" = 2,
                     "Shopping (min/day)" = 2, "Total Domestic (min/day)" = 2))
```

Women in fully modern households spend `r sprintf("%.0f", women_modern)` minutes/day on domestic work compared to `r sprintf("%.0f", women_traditional)` minutes in traditional households—only a `r sprintf("%.0f%%", time_savings_pct)` reduction despite moving from firewood cooking and no electricity to LPG stoves and electrification.

Men's domestic time is essentially flat across technology levels (`r sprintf("%.0f", men_traditional)` → `r sprintf("%.0f", men_modern)` minutes), confirming that technology doesn't enable male participation.

## Decomposing the Paradox: Where Does Saved Time Go?

The modest overall reduction masks offsetting changes:

**Time Saved**:
- Cooking per meal faster with LPG (vs firewood gathering/lighting)
- Clothes washing faster (if washing machine access via electrification)

**Time Added Back**:
- More elaborate meals (LPG enables complex cooking techniques)
- Higher frequency (washing clothes more often)
- Higher standards (cleaner homes, fresher food)
- New tasks (shopping for variety enabled by refrigerators)

\newpage

# Discussion

## Mechanisms: Why Technology Doesn't Liberate

### 1. Standards Escalation (Cowan Effect)

Better technology enables better output, which becomes the new minimum standard:
- LPG stoves enable elaborate multi-dish meals → social expectations for varied cooking
- Electrification enables food storage → expectations for fresh ingredients and variety
- Modern cleaning tools enable spotless homes → raised cleanliness expectations

### 2. Frequency Increases

Easier tasks get done more often:
- Laundry: From once/week (manual) → multiple times/week (machine-accessible)
- Cooking: More elaborate meals more frequently
- Shopping: More trips for fresh ingredients (enabled by better storage)

### 3. Gender Norms Unchanged

Technology changes **what** women do, not **who** does what. Men don't increase contributions when appliances make tasks easier. This reveals that the barrier to male involvement isn't physical difficulty but gendered norms about whose responsibility it is.

## Policy Implications

**1. Technology Alone Won't Achieve Equality**: Subsidizing LPG, electrification, or appliances will improve quality of life but won't substantially reduce women's time burden without addressing norms.

**2. Focus on Employment Facilitation**: Technology's main benefit may be enabling women's labor force participation (by reducing per-task time enough to fit employment), not reducing total domestic time.

**3. Target Cultural Norms**: Public campaigns should emphasize equitable sharing of domestic work, not just easier work.

\newpage

# Conclusion

Household modernization—moving from firewood to LPG cooking, from kerosene to electricity—dramatically changes the quality and nature of domestic work but barely reduces women's total time burden. Standards escalate, frequencies increase, and new tasks emerge, absorbing the productivity gains. Meanwhile, men's contributions remain unchanged regardless of technology availability, revealing that gendered norms, not task difficulty, drive inequality.

These findings suggest that the promise of technological liberation was naive. Without challenging cultural norms about whose responsibility household work is, technology just changes what women do, not whether they remain primarily responsible. The path to gender equality requires not just better appliances but better norms about who should use them.

# References

Bose, Christine E., Philip L. Bereano, and Mary Malloy. 1984. "Household Technology and the Social Construction of Housework." *Technology and Culture* 25(1): 53-82.

Cowan, Ruth Schwartz. 1983. *More Work for Mother: The Ironies of Household Technology from the Open Hearth to the Microwave*. New York: Basic Books.

Hochschild, Arlie Russell. 1989. *The Second Shift: Working Parents and the Revolution at Home*. New York: Viking.

Mokyr, Joel. 2000. "Why 'More Work for Mother?' Knowledge and Household Behavior, 1870-1945." *Journal of Economic History* 60(1): 1-41.
