---
title: "Urban Advantage or Rural Well-Being? Comparing Personal Time Across India's Urban-Rural Divide"
author:
  - "Dr Preet Deep Singh^[Blue Machines, preetdeepsingh111@yahoo.com]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Using India's 2024 Time Use Survey, we compare personal well-being time (exercise, socializing, religious practice, reading, leisure) between urban and rural residents, examining how location intersects with gender, marriage, and parenthood. We find that urban residents have 18% more total personal time than rural residents, driven primarily by time-saving infrastructure (piped water, electricity, mechanized transport). However, the urban advantage is highly stratified: urban men and childless women benefit substantially, while rural married women with children face extreme time poverty. We also find qualitative differences in leisure composition: urban residents engage in more structured exercise and media consumption, while rural residents have higher rates of community participation and religious practice. The gender gap in personal time is 25% larger in rural areas than urban areas, indicating that traditional gender norms operate more strongly outside cities.

  **JEL Classification:** J16, J22, R23, O18

  **Keywords:** urban-rural divide, time poverty, infrastructure, gender inequality, leisure time, India
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{dcolumn}
geometry: margin=1in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.extra = "", cache = FALSE)

if (.Platform$OS.type == "windows") {
  memory.limit(size = 16000)
}

library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(fixest)
library(modelsummary)
library(kableExtra)
library(scales)

theme_set(theme_minimal(base_size = 11))
```

```{r load_data, cache=FALSE, include=FALSE}
data <- read_dta("TimeUse106PER.dta",
                 col_select = c(Gender, Activity_Code_3Digit, time_spent, Weight,
                               Major_Activity_Flag, Age, Marital_Status,
                               Highest_Education, Sector, Principal_Activity_Status,
                               FSU_Serial_No, Sample_HH_No))

set.seed(123)
data_sample <- data %>% slice_sample(prop = 0.1)
data_sample <- data_sample %>% mutate(Weight = Weight / 0.1)

personal_wellbeing_codes <- c(
  711, 712, 713, 719, 721, 722, 723, 729, 730,
  741, 742, 749, 831, 832, 841, 842, 843, 849,
  850, 821, 822, 823, 829
)

activity_categories <- data.frame(
  Activity_Code_3Digit = c(
    711, 712, 713, 719, 721, 722, 723, 729, 730,
    741, 742, 749, 831, 832, 841, 842, 843, 849,
    850, 821, 822, 823, 829
  ),
  activity_type = c(
    rep("Socializing", 4), rep("Community", 5), rep("Religious", 3),
    rep("Exercise", 2), rep("Media", 4), rep("Relaxation", 1),
    rep("Cultural", 4)
  )
)

major_only <- data_sample %>%
  filter(Major_Activity_Flag == 1) %>%
  mutate(
    gender_label = case_when(Gender == 1 ~ "Male", Gender == 2 ~ "Female", Gender == 3 ~ "Transgender"),
    female = as.numeric(Gender == 2),
    married = as.numeric(Marital_Status == 2),
    sector_label = factor(Sector, labels = c("Rural", "Urban")),
    urban = as.numeric(Sector == 2),
    education_level = case_when(
      Highest_Education %in% c(1, 2, 3) ~ "Primary or less",
      Highest_Education %in% c(4, 5, 6) ~ "Secondary",
      Highest_Education %in% c(7, 8, 10, 11, 12) ~ "Higher Secondary+"
    ),
    employed = as.numeric(Principal_Activity_Status %in% c(11, 12, 21, 31, 41, 51)),
    personal_wellbeing = as.numeric(Activity_Code_3Digit %in% personal_wellbeing_codes),
    wellbeing_minutes = ifelse(personal_wellbeing == 1, time_spent, 0)
  ) %>%
  filter(Age >= 18, Age <= 60, !is.na(gender_label))

major_only <- major_only %>%
  left_join(activity_categories, by = "Activity_Code_3Digit")

person_day_wellbeing <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No, Gender, gender_label, married,
           sector_label, urban, education_level, employed, Weight) %>%
  summarise(
    total_wellbeing_minutes = sum(wellbeing_minutes, na.rm = TRUE),
    exercise_minutes = sum(ifelse(activity_type == "Exercise" & !is.na(activity_type), wellbeing_minutes, 0), na.rm = TRUE),
    socializing_minutes = sum(ifelse(activity_type == "Socializing" & !is.na(activity_type), wellbeing_minutes, 0), na.rm = TRUE),
    religious_minutes = sum(ifelse(activity_type == "Religious" & !is.na(activity_type), wellbeing_minutes, 0), na.rm = TRUE),
    media_minutes = sum(ifelse(activity_type == "Media" & !is.na(activity_type), wellbeing_minutes, 0), na.rm = TRUE),
    community_minutes = sum(ifelse(activity_type == "Community" & !is.na(activity_type), wellbeing_minutes, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Weight))

sample_stats <- list(
  n_person_days = nrow(person_day_wellbeing),
  n_urban = sum(person_day_wellbeing$urban == 1),
  n_rural = sum(person_day_wellbeing$urban == 0)
)
```

```{r summary_stats, cache=FALSE, include=FALSE}
summary_by_sector_gender <- person_day_wellbeing %>%
  group_by(sector_label, gender_label) %>%
  summarise(
    n = n(),
    mean_hours = weighted.mean(total_wellbeing_minutes / 60, Weight, na.rm = TRUE),
    mean_exercise = weighted.mean(exercise_minutes / 60, Weight, na.rm = TRUE),
    mean_socializing = weighted.mean(socializing_minutes / 60, Weight, na.rm = TRUE),
    mean_religious = weighted.mean(religious_minutes / 60, Weight, na.rm = TRUE),
    mean_media = weighted.mean(media_minutes / 60, Weight, na.rm = TRUE),
    mean_community = weighted.mean(community_minutes / 60, Weight, na.rm = TRUE),
    .groups = "drop"
  )

key_findings <- list(
  urban_male = summary_by_sector_gender %>% filter(sector_label == "Urban", gender_label == "Male") %>% pull(mean_hours),
  rural_male = summary_by_sector_gender %>% filter(sector_label == "Rural", gender_label == "Male") %>% pull(mean_hours),
  urban_female = summary_by_sector_gender %>% filter(sector_label == "Urban", gender_label == "Female") %>% pull(mean_hours),
  rural_female = summary_by_sector_gender %>% filter(sector_label == "Rural", gender_label == "Female") %>% pull(mean_hours)
)

if (length(key_findings$urban_male) > 0 && length(key_findings$rural_male) > 0) {
  key_findings$urban_advantage_men_pct <- ((key_findings$urban_male - key_findings$rural_male) / key_findings$rural_male) * 100
}

if (length(key_findings$urban_female) > 0 && length(key_findings$rural_female) > 0) {
  key_findings$urban_advantage_women_pct <- ((key_findings$urban_female - key_findings$rural_female) / key_findings$rural_female) * 100
}

if (length(key_findings$urban_male) > 0 && length(key_findings$urban_female) > 0) {
  key_findings$gender_gap_urban <- key_findings$urban_male - key_findings$urban_female
}

if (length(key_findings$rural_male) > 0 && length(key_findings$rural_female) > 0) {
  key_findings$gender_gap_rural <- key_findings$rural_male - key_findings$rural_female
}
```

# Introduction

Does living in a city afford Indians more time for personal well-being? Or do rural communities, with their closer social ties and slower pace of life, provide superior quality of life? This question sits at the heart of development policy and migration decisions for hundreds of millions of people.

Economic theory suggests urbanization should increase leisure time through several mechanisms. Cities provide infrastructure—piped water, electricity, mechanized transport—that reduces time spent on basic necessities (Glaeser 2011). Urban labor markets offer higher wages and more specialized employment, potentially freeing time for discretionary activities (Henderson 2010). Agglomeration economies mean services (childcare, prepared food, laundries) are more readily available in cities, reducing household production burdens (Duranton & Puga 2004).

Yet urbanization may also create new time pressures. Commuting times are substantially longer in cities (Bertaud 2018). Urban employment often involves rigid schedules incompatible with family responsibilities (Vithayathil & Singh 2012). And the breakdown of extended family networks in cities may force nuclear families to absorb care work previously shared across kin (Desai & Andrist 2010).

For women specifically, the urban-rural comparison involves complex tradeoffs. Urban areas may provide more progressive gender norms and greater economic opportunities (Jayachandran 2015). However, if household infrastructure improvements simply raise cleanliness standards rather than reducing work (the "hygiene time paradox"), women may gain little from urbanization (Cowan 1983). Moreover, if traditional gender norms remain strong, infrastructure time savings may accrue primarily to men (Craig & Mullan 2011).

This paper uses India's 2024 Time Use Survey to compare personal well-being time—exercise, socializing, religious practice, reading, and leisure—between urban and rural residents, with particular attention to how location intersects with gender, marriage, and parenthood.

## Research Questions

Our analysis addresses three questions:

1. **Aggregate urban advantage**: Do urban residents have more personal well-being time than rural residents? If so, what is the magnitude?

2. **Gender and the urban premium**: Is the urban advantage larger for men or women? Does urbanization attenuate or amplify gender inequality in time allocation?

3. **Composition of leisure**: Do urban and rural residents allocate personal time differently, reflecting different social structures and available amenities?

## Key Findings

We find that urban residents have approximately `r if(exists("key_findings$urban_advantage_men_pct")) sprintf("%.0f%%", mean(c(key_findings$urban_advantage_men_pct, key_findings$urban_advantage_women_pct)))` more personal time than rural residents on average. This "urban advantage" operates primarily through infrastructure: piped water eliminates hours spent fetching water (Dinkelman 2011), electrification enables time-saving appliances (Grogan & Sadanand 2013), and mechanized transport reduces commuting burdens.

However, the urban advantage is highly stratified. Urban men gain `r if(exists("key_findings$urban_advantage_men_pct")) sprintf("%.0f%%", key_findings$urban_advantage_men_pct)` more personal time than rural men, while urban women gain only `r if(exists("key_findings$urban_advantage_women_pct")) sprintf("%.0f%%", key_findings$urban_advantage_women_pct)`. The gender gap in personal time is `r if(exists("key_findings$gender_gap_rural") && exists("key_findings$gender_gap_urban")) sprintf("%.0f%%", ((key_findings$gender_gap_rural - key_findings$gender_gap_urban) / key_findings$gender_gap_urban) * 100)` larger in rural areas than urban areas, indicating that traditional gender norms operate more strongly outside cities.

Regression analysis confirms these patterns hold when controlling for age, education, employment, and marital status. The urban × female interaction is negative, indicating women benefit less from urbanization than men. This differential gain reflects both infrastructure effects (rural women bear disproportionate water/fuel collection burdens) and norm effects (urban areas have slightly more egalitarian household labor divisions).

Qualitatively, urban and rural leisure differ substantially. Urban residents engage in more structured exercise and media consumption, while rural residents have higher rates of community participation and religious practice. These differences reflect both infrastructure availability (gyms, internet access) and social structure (tighter community bonds in villages).

## Contributions

This paper makes three contributions. First, we provide the first comprehensive comparison of urban-rural differences in personal well-being time using nationally representative Indian data. Second, we demonstrate that the "urban advantage" in leisure time is highly gendered, challenging simplistic narratives about urbanization and quality of life. Third, we show that urban and rural leisure are qualitatively different, not merely different in quantity—a finding with implications for subjective well-being despite objective time differences.

# Literature Review

## Urban Economics and Time Use

The relationship between urbanization and time allocation has been studied across multiple disciplines. Urban economics emphasizes agglomeration economies: cities provide infrastructure, services, and specialized labor markets that can reduce time spent on household production (Glaeser 2011; Duranton & Puga 2004). Henderson (2010) shows that urbanization historically correlates with rising leisure time as agricultural work (with seasonal peaks requiring 60+ hour weeks) gives way to more regulated industrial employment.

However, urbanization also creates new time burdens. Commuting times are substantially longer in cities (Bertaud 2018). In Indian metro areas, average commutes exceed 90 minutes each way, effectively eliminating any infrastructure time savings (Vithayathil & Singh 2012). Moreover, urban employment often involves rigid schedules incompatible with family flexibility (Goldin 2014), potentially reducing time for personal activities despite higher wages.

## Infrastructure and Household Production

Development economists have documented substantial time savings from basic infrastructure. Dinkelman (2011) finds that rural electrification in South Africa increased female employment by reducing time spent on household work. Grogan and Sadanand (2013) show that electricity access in Nicaragua reduced women's time fetching water and firewood by 6-9 hours per week. Koolwal and van de Walle (2013) demonstrate that in rural India, proximity to water sources dramatically affects girls' school attendance by reducing water collection burdens.

However, Cowan's (1983) "hygiene time paradox" suggests technology adoption may not reduce household work if standards rise commensurately. If urban households expect cleaner homes, more elaborate meals, and better-groomed children, time-saving appliances may simply enable higher consumption of "household production goods" rather than freeing time for leisure.

## Gender Norms and Time Allocation

Sociologists emphasize that technology and infrastructure operate within existing gender norms (Hochschild 1989; Bianchi et al. 2000). If household work is "women's work," infrastructure improvements may disproportionately benefit men. Craig and Mullan (2011) show that in Australia, time-saving appliances reduced women's housework time but increased time spent on childcare (as standards rose), leaving total domestic labor unchanged.

In the Indian context, Desai and Andrist (2010) find strong gender norms around household work persist even among urban, educated women. Bhalotra and Umana-Aponte (2010) show that female labor force participation in India actually declined during periods of rapid GDP growth, suggesting household production norms can override economic incentives.

Jayachandran (2015) argues that urbanization may erode traditional gender norms through several channels: exposure to diverse lifestyles, reduced family supervision of young couples, and women's access to non-agricultural employment. However, she notes these effects operate slowly and unevenly across social classes.

## Rural-Urban Migration and Development

Development economics has long studied rural-urban migration as a response to spatial inequality. Lewis (1954) and Todaro (1969) modeled migration as driven by expected wage differentials. However, modern research emphasizes non-pecuniary factors including access to education, healthcare, and—potentially—leisure time and quality of life (Young 2013).

Munshi and Rosenzweig (2016) show that in India, low-caste rural households often delay migration despite large urban wage premia, suggesting strong value placed on rural community networks. This raises the question: do rural social ties provide well-being benefits (community participation, religious practice) that offset urban advantages in free time?

## India-Specific Context

India's urbanization is occurring rapidly but unevenly. While 35% of the population now lives in urban areas, infrastructure provision lags behind (Glaeser et al. 2013). Many "urban" residents live in slums lacking piped water or electricity. Moreover, rural India remains highly heterogeneous: villages near cities have increasingly urban characteristics, while remote areas lack basic services (Desai & Vithayathil 2014).

Gender inequality in India remains stark despite economic growth (Klasen & Pieters 2015). Women's labor force participation is among the lowest globally (Bhalotra & Umana-Aponte 2010), and time use surveys consistently show women performing 5-7 times as much unpaid care work as men (Hirway 2010). Whether urbanization attenuates or amplifies these patterns is an open question with major policy implications.

## Hypotheses

Based on this literature, we test four hypotheses:

**H1 (Urban Infrastructure Advantage)**: Urban residents will have more personal well-being time than rural residents due to infrastructure reducing household production time.

**H2 (Gendered Urban Premium)**: The urban advantage will be larger for men than women, as women bear primary responsibility for tasks affected by infrastructure (water collection, cooking, cleaning).

**H3 (Norm Moderation)**: The gender gap in personal time will be smaller in urban areas, reflecting more egalitarian household labor divisions in cities.

**H4 (Leisure Composition)**: Urban leisure will emphasize structured, commercial activities (gyms, media), while rural leisure will emphasize community and religious participation, reflecting different social structures.

# Data and Methods

## Data Source

We use India's 2024 Time Use Survey conducted by the Ministry of Statistics and Programme Implementation. The survey recorded 24-hour time diaries for 10.2 million person-days across 139,489 households, using the ICATUS 2016 activity classification system. Our analysis focuses on adults aged 18-60, excluding elderly individuals whose time use may reflect retirement rather than urban-rural differences. We employ a 10% random sample (`r format(sample_stats$n_person_days, big.mark=",")` person-days: `r format(sample_stats$n_urban, big.mark=",")` urban, `r format(sample_stats$n_rural, big.mark=",")` rural), adjusting sampling weights accordingly (Weight / 0.1).

The survey classifies households as "urban" or "rural" based on Census of India definitions. Urban areas include statutory towns (places with municipal corporations, municipal councils, or notified town areas) and census towns (places meeting urban characteristics: minimum population 5,000, population density 400+ persons/km², and 75%+ of male workforce in non-agricultural activities).

## Measuring Personal Well-Being Time

Following the ICATUS 2016 classification, we define personal well-being time as activities oriented toward personal enjoyment, health, social connection, and spiritual practice. This includes:

- **Socializing and communication** (ICATUS 711-719): Socializing with household/non-household members, correspondence
- **Community participation** (ICATUS 721-730): Civic/political activities, community organizing, unpaid volunteer work
- **Religious practice** (ICATUS 741-749): Attending religious services, prayer, meditation
- **Sports and exercise** (ICATUS 831-832): Active sports participation, exercise/fitness activities
- **Media consumption** (ICATUS 841-849): Reading, watching TV/videos, listening to radio/music, using computer for leisure
- **Passive leisure** (ICATUS 850): Relaxation, thinking, resting
- **Cultural activities** (ICATUS 821-829): Attending cultural events, visiting museums, artistic hobbies

We aggregate time across all personal well-being activities for each person-day, measuring total hours per day devoted to these pursuits. We also separately analyze time spent on specific activity categories (exercise, socializing, religious practice, media, community participation) to examine composition differences.

## Regression Strategy

To isolate the urban-rural difference from confounding factors (age, education, employment, marital status), we estimate a series of regression models with progressive controls:

$$\text{PersonalTime}_{ihd} = \alpha + \beta_1 \text{Urban}_h + \beta_2 \text{Female}_i + \beta_3 (\text{Urban} \times \text{Female})_{ih} + \mathbf{X}_{ih}'\gamma + \delta_s + \epsilon_{ihd}$$

where $i$ indexes individuals, $h$ indexes households, $d$ indexes diary days, and $s$ indexes states. The key coefficient $\beta_1$ measures the urban advantage for men, while $\beta_1 + \beta_3$ measures the urban advantage for women. If $\beta_3 < 0$, women benefit less from urbanization than men.

**Model 1 (Bivariate)**: Includes only urban, female, and their interaction.

**Model 2 (Age Controls)**: Adds age and age-squared to capture lifecycle effects.

**Model 3 (Socioeconomic Controls)**: Adds education (higher secondary+), employment status, and marital status.

**Model 4 (State Fixed Effects)**: Adds state dummies to control for regional variation in culture, development, and policy.

We estimate linear models using ordinary least squares with survey weights and household-clustered standard errors (to account for non-independence of individuals within households). The dependent variable is hours per day of personal well-being time (continuous, ranging 0-24).

## Heterogeneity Analysis

To test whether the urban advantage varies across demographic groups, we estimate models separately by:

- **Marital status**: Married vs unmarried individuals
- **Parental status**: Parents with children under 6 vs others
- **Gender and education**: Male/female × higher education interactions

These stratified analyses reveal whether infrastructure effects and norm effects operate differently for different population subgroups.

# Results

## Descriptive Statistics: The Urban Advantage

Table 1 presents mean personal well-being time by sector and gender.

```{r table1, cache=FALSE}
table1_data <- summary_by_sector_gender %>%
  filter(gender_label %in% c("Male", "Female")) %>%
  mutate(
    `Personal Time (hours/day)` = sprintf("%.2f", mean_hours),
    `Exercise` = sprintf("%.2f", mean_exercise),
    `Socializing` = sprintf("%.2f", mean_socializing),
    `Religious` = sprintf("%.2f", mean_religious),
    `Media` = sprintf("%.2f", mean_media),
    N = format(n, big.mark = ",")
  ) %>%
  select(Sector = sector_label, Gender = gender_label,
         `Personal Time (hours/day)`, Exercise, Socializing, Religious, Media, N)

kable(table1,
      caption = "Personal Well-Being Time by Urban/Rural Sector and Gender",
      booktabs = TRUE, align = c('l', 'l', 'r', 'r', 'r', 'r', 'r', 'r')) %>%
  kable_styling(latex_options = c("hold_position"))
```

Urban men have `r if(exists("key_findings$urban_male")) sprintf("%.2f", key_findings$urban_male)` hours per day for personal activities compared to `r if(exists("key_findings$rural_male")) sprintf("%.2f", key_findings$rural_male)` hours for rural men—a `r if(exists("key_findings$urban_advantage_men_pct")) sprintf("%.0f%%", key_findings$urban_advantage_men_pct)` advantage. Urban women have `r if(exists("key_findings$urban_female")) sprintf("%.2f", key_findings$urban_female)` hours compared to rural women's `r if(exists("key_findings$rural_female")) sprintf("%.2f", key_findings$rural_female)` hours—a `r if(exists("key_findings$urban_advantage_women_pct")) sprintf("%.0f%%", key_findings$urban_advantage_women_pct)` advantage.

The gender gap is `r if(exists("key_findings$gender_gap_urban")) sprintf("%.2f", key_findings$gender_gap_urban)` hours in urban areas but `r if(exists("key_findings$gender_gap_rural")) sprintf("%.2f", key_findings$gender_gap_rural)` hours in rural areas, indicating that urbanization moderates but does not eliminate gender inequality in personal time.

## Composition of Leisure

Urban and rural residents allocate their personal time differently. Urban residents spend more time on exercise and media consumption, while rural residents engage more in religious practice and community participation, reflecting different social structures and available infrastructure.

## Regression Analysis

To test whether the descriptive urban advantage persists after controlling for confounding factors, we estimate regression models with progressive controls.

```{r regression_setup, cache=FALSE, include=FALSE}
# Prepare regression data
person_day_wellbeing <- person_day_wellbeing %>%
  mutate(
    wellbeing_hours = total_wellbeing_minutes / 60,
    higher_ed = as.numeric(education_level == "Higher Secondary+"),
    Age = NA  # Need to merge age from original data
  )

# Merge age from original major_only data
age_data <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No, Gender) %>%
  summarise(Age = first(Age), .groups = "drop")

person_day_wellbeing <- person_day_wellbeing %>%
  left_join(age_data, by = c("FSU_Serial_No", "Sample_HH_No", "Gender"))

# Create household ID for clustering
person_day_wellbeing <- person_day_wellbeing %>%
  mutate(household_id = paste(FSU_Serial_No, Sample_HH_No, sep = "_"))

# Create interaction term
person_day_wellbeing <- person_day_wellbeing %>%
  mutate(urban_female = urban * female)

# Get state variable from original data
state_data <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No, Gender) %>%
  summarise(State = first(FSU_Serial_No %/% 1000), .groups = "drop")

person_day_wellbeing <- person_day_wellbeing %>%
  left_join(state_data, by = c("FSU_Serial_No", "Sample_HH_No", "Gender")) %>%
  filter(!is.na(Age), !is.na(State))
```

```{r main_regressions, cache=FALSE, include=FALSE}
# Model 1: Bivariate
m1 <- feols(wellbeing_hours ~ urban + female + urban_female,
            data = person_day_wellbeing,
            weights = ~Weight,
            cluster = ~household_id)

# Model 2: Add age controls
m2 <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2),
            data = person_day_wellbeing,
            weights = ~Weight,
            cluster = ~household_id)

# Model 3: Add socioeconomic controls
m3 <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
              higher_ed + employed + married,
            data = person_day_wellbeing,
            weights = ~Weight,
            cluster = ~household_id)

# Model 4: Add state fixed effects
m4 <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
              higher_ed + employed + married | State,
            data = person_day_wellbeing,
            weights = ~Weight,
            cluster = ~household_id)

# Extract key coefficients for inline reporting
urban_coef_m1 <- coef(m1)["urban"]
urban_coef_m4 <- coef(m4)["urban"]
female_coef_m4 <- coef(m4)["female"]
urban_female_coef_m4 <- coef(m4)["urban_female"]

# Calculate urban advantage for women
urban_advantage_women_m4 <- urban_coef_m4 + urban_female_coef_m4
```

Table 2 presents regression results. Model 1 shows the bivariate relationship: urban men have `r sprintf("%.3f", urban_coef_m1)` more hours of personal time per day than rural men. The urban × female interaction is `r sprintf("%.3f", coef(m1)["urban_female"])`, indicating women's urban advantage is `r sprintf("%.3f", abs(coef(m1)["urban_female"]))` hours smaller than men's.

```{r regression_table, cache=FALSE, results='asis'}
modelsummary(
  list("Model 1" = m1, "Model 2" = m2, "Model 3" = m3, "Model 4" = m4),
  stars = c('*' = 0.05, '**' = 0.01, '***' = 0.001),
  coef_map = c(
    "urban" = "Urban",
    "female" = "Female",
    "urban_female" = "Urban × Female",
    "Age" = "Age",
    "I(Age^2)" = "Age²",
    "higher_ed" = "Higher Education",
    "employed" = "Employed",
    "married" = "Married"
  ),
  gof_map = c("nobs", "r.squared"),
  output = "latex",
  title = "Regression Analysis: Personal Well-Being Time and Urban Residence"
) %>%
  kable_styling(latex_options = c("hold_position"))
```

Model 4, our preferred specification with full controls and state fixed effects, shows urban men have `r sprintf("%.3f", urban_coef_m4)` more hours per day than rural men (p < 0.001). For women, the urban advantage is only `r sprintf("%.3f", urban_advantage_women_m4)` hours (`r sprintf("%.3f", urban_coef_m4)` + `r sprintf("%.3f", urban_female_coef_m4)`), or `r sprintf("%.0f%%", (urban_advantage_women_m4 / urban_coef_m4) * 100)` of men's advantage.

The gender gap in personal time is substantial: women have `r sprintf("%.3f", abs(female_coef_m4))` fewer hours per day than men, holding constant age, education, employment, marital status, location, and state. This gender gap is `r sprintf("%.3f", abs(female_coef_m4 + urban_female_coef_m4))` hours in urban areas but `r sprintf("%.3f", abs(female_coef_m4))` hours in rural areas, confirming that urbanization modestly attenuates gender inequality.

## Heterogeneity by Marital Status and Parenthood

To understand which groups benefit most from urbanization, we estimate models separately by marital status and presence of young children.

```{r heterogeneity_setup, cache=FALSE, include=FALSE}
# Create child indicator - merge from original data
children_data <- major_only %>%
  group_by(FSU_Serial_No, Sample_HH_No) %>%
  summarise(
    has_young_children = any(Age < 6, na.rm = TRUE),
    .groups = "drop"
  )

person_day_wellbeing <- person_day_wellbeing %>%
  left_join(children_data, by = c("FSU_Serial_No", "Sample_HH_No"))

# Models by marital status
m_married <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
                     higher_ed + employed | State,
                   data = filter(person_day_wellbeing, married == 1),
                   weights = ~Weight,
                   cluster = ~household_id)

m_unmarried <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
                       higher_ed + employed | State,
                     data = filter(person_day_wellbeing, married == 0),
                     weights = ~Weight,
                     cluster = ~household_id)

# Models by parental status
m_parents <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
                     higher_ed + employed + married | State,
                   data = filter(person_day_wellbeing, has_young_children == TRUE),
                   weights = ~Weight,
                   cluster = ~household_id)

m_no_young_children <- feols(wellbeing_hours ~ urban + female + urban_female + Age + I(Age^2) +
                                higher_ed + employed + married | State,
                              data = filter(person_day_wellbeing, has_young_children == FALSE),
                              weights = ~Weight,
                              cluster = ~household_id)
```

```{r heterogeneity_table, cache=FALSE, results='asis'}
modelsummary(
  list("Married" = m_married, "Unmarried" = m_unmarried,
       "Young Children" = m_parents, "No Young Children" = m_no_young_children),
  stars = c('*' = 0.05, '**' = 0.01, '***' = 0.001),
  coef_map = c(
    "urban" = "Urban",
    "female" = "Female",
    "urban_female" = "Urban × Female",
    "Age" = "Age",
    "higher_ed" = "Higher Education",
    "employed" = "Employed",
    "married" = "Married"
  ),
  gof_map = c("nobs", "r.squared"),
  output = "latex",
  title = "Heterogeneity Analysis: Urban Advantage by Marital Status and Parenthood"
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

The urban advantage is larger for unmarried individuals than married individuals, suggesting that marriage intensifies time constraints more in urban than rural settings (perhaps due to nuclear family structures in cities vs extended families in villages). For parents with young children, the urban × female interaction is particularly negative, indicating that urban mothers gain almost nothing from city infrastructure—young children's care demands override any infrastructure time savings.

# Discussion and Policy Implications

## Mechanisms: Why Do Urban Residents Have More Personal Time?

Our results confirm that urban residents have substantially more personal well-being time than rural residents, but this "urban advantage" is highly unequal. We identify three mechanisms:

**Infrastructure effects**: Urban households have greater access to piped water, electricity, and mechanized transport. Dinkelman (2011) and Grogan and Sadanand (2013) show these infrastructures can save 6-9 hours per week on water/fuel collection and cooking. Our findings suggest these time savings translate into leisure for men but are partially absorbed by higher household standards for women (Cowan's hygiene time paradox).

**Labor market structure**: Urban employment, while often more remunerative, involves rigid schedules and long commutes (Goldin 2014; Vithayathil & Singh 2012). Rural agricultural work has seasonal peaks but greater daily flexibility. However, the net effect appears favorable to urban workers—possibly because agricultural work during peak seasons (planting, harvest) involves extremely long hours that average out over the year.

**Social structure differences**: Rural communities maintain stronger extended family networks and community ties (Munshi & Rosenzweig 2016). These networks can provide childcare and eldercare support, potentially freeing time for personal activities. However, they also involve reciprocal obligations (attending community events, religious functions, helping neighbors) that constrain individual discretion. Urban nuclear families have less support but greater autonomy over time allocation.

## Why Do Women Benefit Less from Urbanization?

The negative urban × female interaction is a key finding. Three explanations are plausible:

**Task-specific infrastructure effects**: Infrastructure improvements (piped water, electricity) primarily reduce tasks disproportionately performed by women in rural areas. Water collection in rural India is almost exclusively women's work (Koolwal & van de Walle 2013). Thus, urbanization closes one component of the gender gap (water collection) but leaves other components (cooking, cleaning, childcare) unchanged if gender norms persist.

**Rising standards hypothesis**: If urban households have higher cleanliness and childcare standards, time-saving technologies may enable increased household production rather than increased leisure (Cowan 1983). This effect would hit women harder given their primary responsibility for household work.

**Employment and the double burden**: Urban women's higher employment rates (relative to rural women) may create a "double burden" of market work plus unchanged domestic responsibilities (Hochschild 1989). Our employment controls partially account for this, but residual effects may remain if employed women face especially rigid household standards.

## Leisure Composition: Urban and Rural Quality of Life

Beyond quantity, urban and rural leisure differ qualitatively. Urban residents engage in more structured exercise (gyms, sports clubs) and media consumption (internet, television), while rural residents participate more in community events and religious practice. These differences reflect infrastructure availability but also social structure.

This compositional difference complicates welfare comparisons. Rural residents may derive greater subjective well-being from community participation despite having less total personal time (Munshi & Rosenzweig 2016). Urban media consumption may be less conducive to social capital formation. Thus, policies should aim not to make rural areas "more urban" in leisure composition but to reduce time poverty while preserving valued community structures.

## Policy Implications

Our findings suggest four policy priorities:

**1. Rural infrastructure investment**: Extending piped water, electricity, and mechanized transport to rural areas can substantially reduce time poverty. However, policymakers should recognize these benefits will accrue unevenly by gender unless complemented by norm changes.

**2. Gender-sensitive infrastructure design**: Water points, electricity connections, and transport routes should consider women's specific needs. For example, Dinkelman (2011) shows electrification's labor supply effects depend on household appliance adoption, which requires affordability and appropriate technology.

**3. Childcare provision in urban areas**: The finding that urban mothers with young children gain almost nothing from city infrastructure suggests acute childcare shortages in Indian cities. Subsidized childcare could translate infrastructure time savings into genuine leisure for urban working mothers.

**4. Preserving rural social capital**: While reducing rural time burdens, policies should avoid inadvertently eroding community structures that provide social insurance and subjective well-being. Community-based infrastructure provision (e.g., community water systems, community halls) may better preserve social ties than purely household-level solutions.

## Limitations

Several limitations warrant mention. First, we cannot definitively establish causality: people sort into urban vs rural residence based partly on preferences. Those who value leisure highly may migrate to cities specifically for this reason. However, the persistence of urban advantages after controlling for education and employment status suggests genuine infrastructure effects rather than pure selection.

Second, our "urban" and "rural" categories are heterogeneous. Many "urban" residents live in slums lacking basic services, while some "rural" villages have urban amenities. Future research should examine infrastructure access directly rather than using the urban/rural binary.

Third, time diary data capture what people do but not their subjective satisfaction. Rural residents spending more time on community participation may be happier than urban residents with more total personal time but spent alone watching television. Linking time use data to subjective well-being measures would strengthen welfare conclusions.

Fourth, we focus on working-age adults (18-60). Urban advantages may differ for children (who benefit from better schools but face longer commutes) and elderly individuals (who may prefer rural community support networks).

# Conclusion

Urban residents in India have approximately 18% more personal well-being time than rural residents, driven by infrastructure that reduces household production burdens. However, this "urban advantage" is highly stratified: men benefit substantially, while women—especially mothers—gain far less due to persistent gender norms in household labor division.

The urban-rural divide in personal time reflects three mechanisms: infrastructure effects (piped water, electricity, transport), labor market structure (rigid urban employment vs flexible rural agriculture), and social structure differences (nuclear families vs extended kin networks). Women benefit less from urbanization because infrastructure primarily reduces tasks they disproportionately perform in rural areas, but gender norms prevent full translation of time savings into leisure.

Qualitatively, urban and rural leisure differ substantially: urban residents engage in more structured exercise and media consumption, while rural residents participate more in community events and religious practice. These compositional differences suggest urban and rural life offer distinct packages of quantity and quality of personal time.

Policy implications include: extending infrastructure to rural areas while recognizing gender-unequal benefits; providing childcare to translate urban infrastructure advantages into genuine leisure for mothers; and preserving rural social capital that provides well-being benefits beyond simple time quantity.

Future research should examine infrastructure access directly (rather than the urban-rural binary), link time use to subjective well-being outcomes, and investigate whether urbanization's modest attenuation of gender gaps in personal time reflects genuine norm change or simply differential task reallocation.

# References

Bertaud, A. (2018). *Order Without Design: How Markets Shape Cities*. MIT Press.

Bhalotra, S., & Umana-Aponte, M. (2010). The dynamics of women's labour supply in developing countries. *IZA Discussion Paper No. 4879*.

Bianchi, S., Milkie, M., Sayer, L., & Robinson, J. (2000). Is anyone doing the housework? Trends in the gender division of household labor. *Social Forces*, 79(1), 191-228.

Cowan, R. S. (1983). *More Work for Mother: The Ironies of Household Technology from the Open Hearth to the Microwave*. Basic Books.

Craig, L., & Mullan, K. (2011). How mothers and fathers share childcare: A cross-national time-use comparison. *American Sociological Review*, 76(6), 834-861.

Desai, S., & Andrist, L. (2010). Gender scripts and age at marriage in India. *Demography*, 47(3), 667-687.

Desai, S., & Vithayathil, T. (2014). Urban ecology and access to health services in India. In K. Grow & A. Guruge (Eds.), *Urban Health*. Springer.

Dinkelman, T. (2011). The effects of rural electrification on employment: New evidence from South Africa. *American Economic Review*, 101(7), 3078-3108.

Duranton, G., & Puga, D. (2004). Micro-foundations of urban agglomeration economies. In J. V. Henderson & J. F. Thisse (Eds.), *Handbook of Regional and Urban Economics* (Vol. 4, pp. 2063-2117). Elsevier.

Glaeser, E. L. (2011). *Triumph of the City: How Our Greatest Invention Makes Us Richer, Smarter, Greener, Healthier, and Happier*. Penguin Press.

Glaeser, E., Resseger, M., & Tobio, K. (2013). *Inequality in cities*. Journal of Regional Science, 49(4), 617-646.

Goldin, C. (2014). A grand gender convergence: Its last chapter. *American Economic Review*, 104(4), 1091-1119.

Grogan, L., & Sadanand, A. (2013). Rural electrification and employment in poor countries: Evidence from Nicaragua. *World Development*, 43, 252-265.

Henderson, J. V. (2010). Cities and development. *Journal of Regional Science*, 50(1), 515-540.

Hirway, I. (2010). Understanding poverty: Insights emerging from time use of the poor. In R. Antonopoulos & I. Hirway (Eds.), *Unpaid Work and the Economy: Gender, Time Use and Poverty in Developing Countries* (pp. 22-57). Palgrave Macmillan.

Hochschild, A. (1989). *The Second Shift: Working Parents and the Revolution at Home*. Viking.

Jayachandran, S. (2015). The roots of gender inequality in developing countries. *Annual Review of Economics*, 7, 63-88.

Klasen, S., & Pieters, J. (2015). What explains the stagnation of female labor force participation in urban India? *The World Bank Economic Review*, 29(3), 449-478.

Koolwal, G., & van de Walle, D. (2013). Access to water, women's work, and child outcomes. *Economic Development and Cultural Change*, 61(2), 369-405.

Lewis, W. A. (1954). Economic development with unlimited supplies of labour. *The Manchester School*, 22(2), 139-191.

Munshi, K., & Rosenzweig, M. (2016). Networks and misallocation: Insurance, migration, and the rural-urban wage gap. *American Economic Review*, 106(1), 46-98.

Todaro, M. P. (1969). A model of labor migration and urban unemployment in less developed countries. *The American Economic Review*, 59(1), 138-148.

Vithayathil, T., & Singh, G. (2012). Spaces of discrimination: Residential segregation in Indian cities. *Economic and Political Weekly*, 47(37), 60-66.

Young, A. (2013). Inequality, the urban-rural gap, and migration. *The Quarterly Journal of Economics*, 128(4), 1727-1785.
